---
title: "Steamboat Environmental Data System"
output: html_document
---

# 🌎 Steamboat Environmental Monitoring System

A reproducible pipeline integrating **geochemistry, field observations, and temperature logger data** into a unified relational and GIS-ready system.

---

## 🔁 System Overview

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(DiagrammeR)

DiagrammeR::grViz("
digraph {
  rankdir=LR;
  node [shape=box style=filled fillcolor=lightblue]

  NDEP -> Database
  Field -> Database
  Loggers -> Database
  Database -> GIS
}
")
```

---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)

# Load data quietly (no visible section)
data_path <- "docs/data/ingest_log.csv"

if (file.exists(data_path)) {

  ingest_log <- read_csv(data_path, show_col_types = FALSE) |>
    mutate(timestamp = ymd_hms(timestamp)) |>
    arrange(timestamp)

} else {

  ingest_log <- NULL
}
```

---

## 🧱 Database Composition

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(readr)
library(dplyr)

# optional: export this from pipeline later for GitHub Pages
# for now we assume you will create a CSV like this in run_pipeline

table_path <- "docs/data/table_counts.csv"

if (file.exists(table_path)) {

  table_counts <- read_csv(table_path, show_col_types = FALSE)

  print(
    ggplot(table_counts,
           aes(x = reorder(table, n_rows),
               y = n_rows)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      labs(
        title = "Database Composition by Table",
        x = "Table",
        y = "Number of Records"
      ) +
      theme_minimal()
  )

} else {

  plot.new()
  text(0.5, 0.5, "Database breakdown will appear after export")

}
```

---

## 📈 Database Growth

```{r echo=FALSE, message=FALSE, warning=FALSE}
if (!is.null(ingest_log) && nrow(ingest_log) > 0) {

  plot_data <- ingest_log |>
    mutate(cumulative = cumsum(measurements_inserted))

  print(
    ggplot(plot_data,
           aes(x = timestamp,
               y = cumulative,
               color = data_source)) +

      geom_line(linewidth = 1.2) +
      geom_point(size = 2.5) +

      labs(
        title = "Database Growth Over Time",
        subtitle = "Cumulative records inserted through ingestion pipeline",
        x = "",
        y = "Total Records",
        color = "Source"
      ) +

      theme_minimal() +

      theme(
        plot.title = element_text(face = "bold"),
        legend.position = "bottom"
      )
  )

} else {

  plot.new()
  text(0.5, 0.5, "No data available")

}
```

---

## 🧭 Explore

- 📘 **Project** → scientific context and motivation  
- ⚙️ **Pipeline** → ingestion and processing design  
- 🧱 **Data** → schema and relationships  
- 📊 **Results** → analysis and outputs  
- 👤 **About** → author and tools  

---

## ✅ Status

System is operational and updated via controlled ingestion workflows.
