---
title: "Using Specific Conductance as a Proxy for Chloride Discharge in the Steamboat Hills Geothermal System, Nevada"
subtitle: "Implications for Monitoring Hydrothermal Outflow Following the 2025 Lower Sinter Terrace Eruption"
author: "Tyler Irvin"
date: today
lang: en
format:
  pdf:
    toc: false
    toc-depth: 2
    number-sections: true
    fig-pos: "H"
    keep-tex: false
    include-in-header:
      text: |
        % Quarto calls \maketitle unconditionally right after
        % \begin{document} even with title-block-style: none; blank it
        % out so only the hand-built title page below is used.
        \renewcommand{\maketitle}{}
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
  docx:
    toc: true
    toc-depth: 2
    number-sections: true
title-block-style: none
bibliography: references.bib
csl: gsa.csl
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

::: {.content-visible when-format="pdf"}
\begin{titlepage}
\centering
\vspace*{2.5cm}

{\LARGE \textbf{Using Specific Conductance as a Proxy for Chloride Discharge}}\\[0.3cm]
{\LARGE \textbf{in the Steamboat Hills Geothermal System, Nevada}}\\[0.8cm]

{\large Implications for Monitoring Hydrothermal Outflow Following the 2025 Lower Sinter Terrace Eruption}\\[2cm]

Thesis proposal for the degree of\\
Master of Science in Geology\\[1cm]

University of Nevada -- Reno\\
Nevada Bureau of Mines and Geology\\
Great Basin Center for Geothermal Energy

\vspace{2cm}

Tyler Irvin

\vspace{1cm}

Advisor: Wendy Calvin\\
Committee Member: Alex Lutz\\
Committee Member: Jim Faulds

\vfill

\today
\end{titlepage}

\newpage
\renewcommand*\contentsname{Table of contents}
\setcounter{tocdepth}{2}
\tableofcontents
\newpage
:::

::: {.content-visible unless-format="pdf"}
# Using Specific Conductance as a Proxy for Chloride Discharge in the Steamboat Hills Geothermal System, Nevada {.unnumbered}

**Implications for Monitoring Hydrothermal Outflow Following the 2025 Lower Sinter Terrace Eruption**

Thesis proposal for the degree of Master of Science in Geology, University of Nevada -- Reno, Nevada Bureau of Mines and Geology / Great Basin Center for Geothermal Energy

Tyler Irvin\
Advisor: Wendy Calvin\
Committee Member: Alex Lutz\
Committee Member: Jim Faulds

\newpage
:::

```{r setup}
#| include: false
if (basename(getwd()) == "manuscript") {
  root <- normalizePath(file.path(getwd(), ".."))
} else {
  root <- normalizePath(getwd())
}
knitr::opts_knit$set(root.dir = root)

show_fig <- function(path, caption, alt, id) {
  # `caption` is the full, visible figure caption (becomes the figure's
  # caption text via pandoc's implicit-figure handling). `alt` is a
  # short, separate description for screen readers, passed through via
  # pandoc's image `{alt="..."}` attribute -- per UNR's own accessibility
  # guidance, alt text should be succinct (roughly under 150 characters),
  # while the full descriptive caption stays as real, visible text.
  # `id` sets the figure's Quarto crossref id (e.g. "fig-timeline"), so
  # it renders as a numbered "Figure N" and can be cited by number via
  # @id in the prose.
  abs_path <- file.path(normalizePath(getwd()), path)
  abs_path <- gsub("\\\\", "/", abs_path)
  alt_escaped <- gsub('"', "'", alt)
  knitr::asis_output(sprintf('![%s](%s){#%s fig-alt="%s"}', caption, abs_path, id, alt_escaped))
}
```

# Introduction

## Study area and setting

Steamboat Hills sits a few kilometers south of Reno, Nevada, at the
northern end of a fault-bounded belt of Great Basin geothermal
systems. Unlike most Great Basin systems, which are amagmatic and
derive their heat from regionally elevated mantle heat flow alone,
Steamboat is one of a small number of magmatic-influenced systems in
the province, and it is one of the longest-studied hydrothermal areas
in the western United States, characterized geologically and
hydrologically since the 1950s [@thompsonwhite1964; @white1968;
@white1968c; @silberman1979]. The field has produced electricity
commercially since the mid-1980s and is currently operated by Ormat
Technologies as a binary, fully reinjected system generating on the
order of 79 MW net [@dhakal2025]. Production and injection are
conventionally divided into three thermal zones -- Upper, Middle, and
Lower Steamboat -- reflecting a natural-state upflow of roughly
220-240 degrees Celsius beneath Upper Steamboat that cools
northeastward along strike toward the Middle and Lower fields
[@dhakal2025; @johnsonhulen2006].

At the surface, this deep system expresses itself through the Lower
Sinter Terrace, an area of active and historically active fumaroles,
seeps, and hot springs within the Steamboat Springs Area of Critical
Environmental Concern. That surface expression has not been static.
Sorey and Colvard [@sorey1992] documented a near-total cessation of
natural spring flow at the main and low silica terraces between 1986
and 1989, attributing most -- though not all -- of the decline to
regional shallow-groundwater drawdown rather than to geothermal
production. Sorey [@sorey2000] revisited the same terrace roughly a
decade later and found that although regional groundwater levels had
by then recovered, natural spring flow at the main terrace had not
resumed. That non-recovery is the single most important piece of
historical context for this proposal: it establishes that surface
discharge at Steamboat is not simply a mirror of groundwater level,
and that a system which goes quiet does not necessarily reawaken on
its own.

It was against that backdrop that, beginning around 2022, steaming
ground at the Lower Sinter Terrace began to expand again, and on
June 3, 2025, an uncapped, shallow abandoned well erupted without
warning, sending boiling water roughly 30 meters into the air -- the
first geyser-like activity documented at Steamboat in decades
[@lindsey2026]. The well was capped in the following days to weeks,
but discharge did not stop; it redistributed laterally, opening new
seeps and spitting vents along nearby fractures. This event, and the
still-open question of what it means for the system's deeper
plumbing, motivates this proposal: a direct re-examination of
Sorey's own chloride-discharge method against the modern system, and
a test of whether continuous conductivity monitoring can substitute
for discrete chloride sampling as a practical way to track that
outflow going forward.

## A brief chronology

Three broad periods structure this history, shown in @fig-timeline. The
**historical/characterization period** (roughly 1950-1968) established
the regional geology, structure, and hydrology that every later study,
including this one, still relies on [@thompsonwhite1964; @white1968;
@white1968c]. The **development period** (1986-2007) covers the
transition from a natural system to a commercially produced one, the
1987-89 spring-flow decline documented by Sorey and Colvard
[@sorey1992], its non-recovery by 2000 [@sorey2000], and the buildout
of the groundwater/geothermal monitor-well network described by
Klein, Johnson, and Spielman [@klein2007]. The **reawakening period**
(2022-present) covers the renewed steaming, the June 2025 eruption,
and this project's own response.

```{r fig-timeline}
#| results: asis
show_fig(
  "output/figures/manuscript/steamboat_timeline.png",
  "Steamboat Hills events, 1950-2026, spanning the historical USGS characterization era, the development/production era (including the 1987-89 spring-flow decline and its documented non-recovery by 2000), and the current reawakening. Already produced from real, dated sources cited throughout this proposal.",
  alt = "Horizontal timeline of Steamboat Hills events from 1950 to 2026, from early USGS characterization through the 2025 eruption and current reawakening.",
  id = "fig-timeline"
)
```

| Year | Event |
|---|---|
| 1950 | USGS thermal-gradient and chemistry test holes drilled [@white1968]. |
| 1964 | Regional geology and structure of the Steamboat Springs area described [@thompsonwhite1964; @white1968c]. |
| 1968 | Hydrology, activity, and heat flow of the thermal system characterized in detail [@white1968]. |
| 1986 | Commercial production begins. |
| 1987-1989 | Natural hot-spring flow at the main and low silica terraces declines and ceases [@sorey1992]. |
| 1992 | Sorey and Colvard attribute most of the spring-flow decline to regional groundwater drawdown, not geothermal production [@sorey1992]. |
| 2000 | Groundwater levels recover, but natural spring flow does not resume at the main terrace [@sorey2000]. |
| 2007 | The groundwater/geothermal monitor-well network is documented in detail [@klein2007]. |
| 2022 | Renewed steaming ground is first noted at the Lower Sinter Terrace [@lindsey2026]. |
| 2025 | The Lower Sinter Terrace erupts (June 3); the wellhead is capped and discharge redistributes laterally [@lindsey2026]. Continuous temperature and conductivity monitoring begins. |
| 2026 | The eruption is formally documented [@lindsey2026]; spring and well remapping and renewed chemistry sampling continue under this thesis. By late September 2026, a new seep-fed hot spring and a growing, fresh sulfur deposit are documented at the terrace, independently corroborated by a nearby resident. |

: A brief chronology of Steamboat Hills, from first characterization to the current reawakening. {#tbl-chronology}

## Chloride discharge as a tracer of geothermal outflow

Michael Sorey's own framing of this system -- developed across
[@sorey1992], [@sorey2008], and [@sorey2017] -- is the conceptual
backbone of this proposal. Because chloride behaves conservatively in
this system (it is neither added nor removed by mineral reactions or
degassing on the way to discharge), the difference in chloride mass
flux between two points on Steamboat Creek can be used to estimate how
much thermal water is entering the creek between them, without needing
to locate or measure every individual spring or seep along the way.
Sorey and Spielman applied this method at two gauging sites, Rhodes
Road (SBRR) and Bella Vista (SBBV), reporting 730-1,214 gallons per
minute for 2007-08 [@sorey2008] and an average of 60 liters per second
(range 45-90 L/s) across 24 sampling events from 2008 to 2016
[@sorey2017]. A 2026 poster using the same method and sites,
co-authored with Cary Lindsey, asked a single leading question: did
the 2025 Lower Sinter Terrace geysering event represent activation of
a new geothermal source, or redistribution of fluids within an
existing hydrothermal conduit network [@irvinlindsey2026]? That
question is the direct ancestor of this proposal's own central
question below. The poster's own evidence favored redistribution over
a new source on two independent grounds: major-ion chemistry and
stable isotopes showed the reactivated fissure discharging the same
hydrothermal fluid already expressed in nearby springs and wells, and
paired temperature loggers at the terrace showed a strong inverse
relationship to each other (r = -0.68) -- as one logger warmed, the
other cooled, consistent with discharge shifting between vents within
a single connected, shallow, fracture-controlled flow system rather
than a uniform, system-wide change or a chemically distinct new
source. The poster's own chloride-flux calculation put post-eruption
thermal inflow at roughly 27 L/s, below the 2008-2016 average of 60
L/s [@sorey2017], meaning overall thermal discharge may have declined
even as flow paths reorganized. This proposal's own preliminary
results (below) extend that one-time calculation into a repeatable,
database-grounded estimate, and its own field campaign (also below)
finds that the system has kept changing since the poster was
presented.

The practical problem with the chloride-flux method has always been
that it requires discrete water samples, collected and analyzed in a
laboratory, at a temporal resolution most field campaigns cannot
sustain for long. This thesis's central contribution is testing
whether continuous, field-deployable specific-conductance monitoring
-- already demonstrated as a workable chloride proxy for geothermal
solute flux in Yellowstone's rivers [@mccleskey2012; @mccleskey2016]
-- can do the same job at Steamboat, letting a much sparser chloride-
sampling campaign be calibrated against a continuous conductance
record rather than requiring chloride itself to be sampled at high
frequency indefinitely.

## Purpose of study

**Central question.** Can continuous conductivity measurements be
used to accurately estimate chloride discharge and geothermal outflow
at Steamboat Hills, and what do those measurements reveal about
hydrothermal-system evolution following the 2025 eruption?

This proposal is organized around four research questions, each with
its own concrete deliverable.

**RQ1 -- How stable is the chloride-conductivity relationship through
time and across site types?** Does a single calibration apply
everywhere, or do outlet, creek, monitoring-well, and spring waters
each require their own relationship, and how much uncertainty results
from seasonal and temporal variation? *Deliverable:* one or more real
chloride-to-conductance calibration equations, with stated
uncertainty.

**RQ2 -- What is the minimum chloride-sampling frequency required to
maintain a scientifically defensible, conductivity-based estimate of
chloride flux?** *Deliverable:* a quantitative monitoring-design
recommendation, developed through the Monte Carlo analysis described
in the Methods section below.

**RQ3 -- Are conductivity variations driven primarily by conservative
mixing, or by geochemical reactions that would decouple conductance
from chloride?** Is chloride itself behaving conservatively here?
Is conductivity? Are mineral reactions altering conductance
independently of chloride concentration? *Deliverable:* a
geochemical, PHREEQC-based physical justification -- or an explicit,
documented limitation -- for the chloride-conductivity proxy.

**RQ4 -- Has geothermal outflow at Steamboat changed following the
2025 eruption?** *Deliverable:* a post-eruption interpretation of
chloride flux, conservative-tracer ratios, and outlet chemistry,
read against the 70-year discharge record Sorey and others
established.

# Background: chloride, conductivity, and the role of geochemical modeling

## Chloride and specific conductance as conservative-tracer proxies

Steamboat's thermal water is a sodium-chloride-dominated fluid;
meteoric (background) water in the same area is dilute and
comparatively richer in calcium and magnesium. Chloride's
conservative behavior in this system is what makes it usable both as
a tracer of thermal-water fraction in an individual sample, and, by
mass balance against stream discharge, as an estimate of total
thermal-water discharge [@sorey1992; @sorey2008; @sorey2017]. Specific
conductance is proposed as a proxy for chloride, not because
conductance is itself the quantity of scientific interest, but
because it is inexpensive, continuous, and field-deployable in a way
discrete chloride sampling is not. Whether that substitution actually
holds at Steamboat -- across sites, seasons, and the period before and
after the 2025 eruption -- is this proposal's central empirical
question. This proposal tests that substitution directly; it does not
assume the answer in advance.

## Major-ion chemistry: what each analyte tells us

| Analyte | Role in this system |
|---|---|
| Chloride | The primary conservative tracer. |
| Sodium, potassium | Track with chloride in the real thermal end-member; also the basis of sodium/potassium geothermometry (below). |
| Sulfate | Also strongly correlated with chloride, sodium, and potassium here -- a secondary conservative-ish tracer, though sulfate can be affected by anhydrite/gypsum equilibria at high enough concentration. |
| Calcium, magnesium | Behave oppositely to the chloride/sodium/potassium group -- track a more dilute, rock-reacted water-rock interaction signature. |
| Alkalinity | Drives carbonate mineral saturation (calcite, dolomite, aragonite). |
| Boron, lithium, fluoride | Minor conservative-ish tracers used for independent cross-checks, such as the chloride-to-boron ratio discussed below. |
| Silica | The basis of quartz/chalcedony geothermometry; behaves less conservatively than sodium/potassium on the ascent-to-discharge path at Steamboat. |

## Charge balance and mineral saturation

Every water analysis is first screened for internal consistency via
the standard charge-balance-error statistic,

$$
\text{CBE\%} = \frac{\sum z_i m_i^{\text{cation}} - \sum |z_i| m_i^{\text{anion}}}
{\sum z_i m_i^{\text{cation}} + \sum |z_i| m_i^{\text{anion}}} \times 100,
$$

where $z_i$ and $m_i$ are the charge and molar concentration of each
ion; a sample passing within $\pm 5\%$ is analytically reliable
enough to model further. The mineral saturation index then
quantifies how far a water is from equilibrium with a given mineral,

$$
SI = \log_{10}\!\left(\frac{IAP}{K_{sp}}\right),
$$

where $IAP$ is the ion activity product computed from the sample's
speciated chemistry and $K_{sp}$ is that mineral's temperature-
dependent solubility product; $SI = 0$ is equilibrium, $SI > 0$
implies a tendency to precipitate (scale), and $SI < 0$ implies a
tendency to dissolve.

## Geothermometry: two independent, and divergent, temperature estimates

Activity-based sodium/potassium geothermometers estimate the deep
reservoir temperature a water last equilibrated with, from the ratio
of sodium to potassium activity. Two widely used forms are
Giggenbach's [@giggenbach1988]

$$
T_{\text{Gigg}}(^\circ C) = \frac{1390}{1.75 + \log_{10}(a_{Na}/a_K)} - 273.15
$$

and Fournier's [@fournier1979]

$$
T_{\text{Four}}(^\circ C) = \frac{1217}{1.483 + \log_{10}(a_{Na}/a_K)} - 273.15 .
$$

Silica (quartz/chalcedony) geothermometry instead uses dissolved
silica's own temperature-dependent solubility, read from the real
equilibrium temperature at which a mineral's saturation index crosses
zero. At Steamboat, sodium/potassium estimates (154-256 degrees
Celsius) systematically and substantially exceed quartz/chalcedony
estimates (52-127 degrees Celsius). This divergence is real and
informative, not a contradiction: silica re-
equilibrates toward ambient conditions faster than the sodium/
potassium ratio does as water cools and dilutes on its way to
discharge, so the two geothermometers are expected to diverge exactly
this way along a single cooling, diluting flow path. This
divergence-as-evidence approach follows Shevenell and De Rocher's
[@shevenell2005] own evaluation of multiple geothermometers across
several Nevada geothermal power plants, which found that no single
geothermometer is reliable in isolation and that comparing several
against each other is itself diagnostic of a system's mixing and
cooling history, not just a single reservoir-temperature estimate.

## How PHREEQC will be used: a supporting mechanism, not a modeling thesis

PHREEQC [@phreeqc] is a widely used, open-source geochemical modeling
program that solves aqueous speciation, mineral saturation, mixing,
and inverse mass-balance problems from a water analysis and a chosen
thermodynamic database. This proposal deliberately does **not** treat
PHREEQC as a standalone geochemical-modeling exercise producing
saturation-index tables for their own sake. Instead, PHREEQC answers
four specific questions that directly serve the chloride-conductivity
proxy question above.

**Objective 1 -- Define water types.** Combine major-ion chemistry,
principal component analysis, and hydrochemical facies clustering with
PHREEQC's own fluid classification to separate real samples into a
thermal end-member (outlet and thermal-well samples), mixed waters
(for example, monitoring wells with evidence of partial dilution), and
dilute background waters (domestic wells with no thermal signature).

**Objective 2 -- Evaluate whether conductivity is behaving
conservatively.** The core question is why conductance should track
chloride at all. PHREEQC is used to test whether carbonate
precipitation, silica precipitation, ion exchange, or evaporation
could be altering conductance independently of chloride -- summarized
as a real table of which processes can alter chloride, which can
alter conductance, and whether the two are expected to move together
or diverge. This table is the direct physical justification for, or
documented limitation of, the proxy relationship RQ1 and RQ3 depend
on.

**Objective 3 -- Test thermal end-member stability across decades.**
Real chemistry exists from the 1950s, the 1990s, and the present. Each
era's thermal samples are run through PHREEQC's own equilibrium
calculations (not just raw concentration ratios) to ask whether the
end-member's equilibrium behavior, not only its composition, has
remained stable -- a question directly relevant to whether a
calibration fit today should be expected to hold if extrapolated
backward or forward in time.

**Objective 4 -- Investigate calibration outliers.** Any real
chloride-conductance calibration will have outliers. Rather than
discard them, PHREEQC is used to ask why a given sample deviates --
dilution, mineral precipitation, mixing with a different water, or an
unusual pH -- which is a more informative use of the modeling
capability than reporting saturation indices in isolation.

## Why reactive-transport modeling is not proposed

Full reactive-transport modeling of the Steamboat reservoir --
whether in a dedicated flow-and-transport code or within a
three-dimensional geologic modeling platform's own reactive-transport
module -- is deliberately not part of this proposal. Running such a
model responsibly requires a calibrated permeability field, effective
porosity, defined reservoir boundary conditions, and a transient
pressure history, none of which currently exist for Steamboat at the
resolution a reactive-transport model would need. Attempting it now
would produce a model dominated by untested assumptions rather than
real constraints. A three-dimensional geologic model is still a
genuine and valuable part of this work, but strictly as **conceptual
geologic context** -- a real, well-grounded picture of where
production, injection, and monitoring wells sit relative to known
structure and lithology -- not as a platform for simulating fluid
movement. That distinction is maintained consistently throughout the
Methods section below.

# Previous work

Steamboat's geology, hydrology, and geochemistry have been
characterized across seven decades of prior work this proposal builds
on directly rather than starting over. White, Thompson, and Sandberg
[@white1968; @thompsonwhite1964] established the regional geologic
framework, structure, and thermal history still used today, and
Silberman and others [@silberman1979] constrained the duration of
hydrothermal activity from spatially associated volcanic rocks. Sorey
and Colvard's 1992 administrative report [@sorey1992] is the single
most detailed historical baseline for spring flow, chemistry, and the
1986-89 decline, and Sorey's own 2000 follow-up [@sorey2000] is the
critical observation that spring flow did not resume even once the
groundwater decline initially blamed for it had reversed. Sorey and
Spielman [@sorey2008; @sorey2017] refined the chloride-mass-balance
discharge method this proposal extends, reporting discharge estimates
across 2007-2016 that anchor the long-run record this work compares
against. Mariner and Janik [@mariner1995] provide an independent,
contemporaneous geochemical dataset and conceptual model, including
real gas chemistry this proposal's geothermometer cross-validation
already draws on. Klein, Johnson, and Spielman [@klein2007] documented
the groundwater and geothermal monitor-well network in detail,
providing the modern well-identity framework much of this proposal's
own site resolution work depends on. Skalbeck's dissertation
[@skalbeck2001] supplies geophysical and geochemical subsurface
characterization, including a multi-decade well-monitoring record
used in this proposal's own historical cross-checks. Collar and
Huntley [@collar1990] mapped faults and air-photo lineaments alongside
real 1988-89 discharge measurements using the same conservative-
chloride mass-balance logic, 35 years before this proposal's own
version of that method; their fault map is also the leading real
candidate for the geologic-context digitizing task described below.
Nehring [@nehring1979] produced an early reservoir-temperature
estimate from enthalpy-chloride and deuterium-chloride diagrams on
1977 spring samples that is directly comparable to this proposal's own
geothermometer results, and Vaughan and others [@vaughan2005] mapped
the surface mineral expression of the system with airborne thermal-
infrared imagery, distinguishing actively-forming opaline sinter from
older chalcedony sinter and an acid-sulfate alteration halo around the
fumaroles -- exactly the kind of surface evidence this proposal's own
spring-and-well remapping fieldwork is designed to update for the
present, post-eruption terrace. More recent reservoir-scale work -- a
numerical model update spanning forty years of production
[@dhakal2025], a three-dimensional reservoir model calibrated against
more than 130 wells [@bjornsson2014], a directional-drilling account
for well 83C-6ST1 that supplies this proposal's only real, non-vertical
well survey [@akerley2021], and trace-element and gas evidence for a
magmatic heat source [@arehart2003; @johnsonhulen2006] -- situates this
proposal's own surface and shallow-subsurface focus within the
operator's much larger reservoir picture. Regionally, Cohen and Loeltz
[@cohenloeltz1964] characterized Truckee Meadows groundwater outside
the Steamboat field itself, providing background context for several
wells this proposal draws on for its own barometric-efficiency
calculation. State and federal agency records (the Nevada Divisions of
Environmental Protection, Minerals, and Water Resources, and the U.S.
Geological Survey) supply the production-well chemistry, permit,
well-construction, and streamflow records that extend this historical
baseline into the present monitoring period. Lindsey and others
[@lindsey2026] published the first full account of the 2022-2025
reawakening and the June 2025 eruption itself, the event that directly
motivates this proposal. This project's own preliminary
chloride-discharge estimate following that eruption was presented as a
poster at the 2026 Geothermal Rising Conference in Houston, Texas
[@irvinlindsey2026], and this proposal's Chapter 4 extends that
one-time estimate into a real, repeatable database query.

# Methods and plan of action

## Monitoring network and supporting database

Every method described below draws on field sampling at the Lower
Sinter Terrace; state-agency chemistry, permit, well-construction, and
water-rights records; federal streamflow, water-level, weather, and
historical specific-conductance records; and historical literature
chemistry that has been transcribed directly from Sorey and Colvard
[@sorey1992], Mariner and Janik [@mariner1995], and Skalbeck
[@skalbeck2001] rather than merely cited. These sources are integrated
into a single, queryable database -- 1,600 samples across 269 wells
and additional spring, seep, and creek locations, as of this writing
-- rather than kept as separate spreadsheets, specifically so that a
question spanning multiple decades and multiple agencies' worth of
records can be answered consistently rather than re-assembled by hand
each time it is asked. A standing data-quality process checks for
missing values, implausible readings, and internal consistency on an
ongoing basis. One concrete example materially affects the
preliminary results below: a missing alkalinity parameter in one
major chemistry feed initially made charge-balance error look far
worse than it really was (roughly 60%). Once corrected, the median
charge-balance error for the same samples dropped to about 0.7%, with
most samples now passing the standard plus-or-minus-5% quality
threshold.

**Spring and well remapping.** Several springs at the Lower Sinter
Terrace are listed in the operator's own records as inactive. A
central field task of this thesis is walking that terrace and
confirming, for each one, whether it is in fact dry, seeping, or
actively discharging. This fieldwork has so far produced 51 real,
GPS-tagged survey points at the terrace (28 seeps, 13 steaming-ground
observations, 7 springs, and 3 creek points).

**Chemistry and isotope sampling.** Water samples are collected at
springs, wells, and creek stations, with field parameters (pH,
temperature, specific conductance) measured at the time of collection
and laboratory analyses -- major ions, alkalinity, trace elements, and
delta-18-O/delta-D isotopes -- run afterward. This project's own field
campaign has collected 71 real water samples to date.

**Temperature logger network.** Eleven continuous temperature
loggers are deployed at springs, wells, and along the eruption
fissure itself, giving a direct, continuous view of whether the 2025
eruption's thermal signature has been stable, cooling, or shifting
laterally since the wellhead was capped.

**Conductivity loggers as a chloride proxy.** Two continuous specific-
conductance loggers were deployed in Steamboat Creek in mid-2025: one
co-located with the existing U.S. Geological Survey gauge near Rhodes
Road (the upstream/control site) and one further downstream near
Geiger Grade, matching the same two site roles Sorey and Spielman used
for their own chloride-flux calculations [@sorey2008; @sorey2017].

## PHREEQC workflow: sample eligibility, databases, and interpretation

Aqueous speciation and saturation-index modeling is run through the
real PHREEQC 3 executable [@phreeqc] -- the standalone program itself,
not an R reimplementation -- against every sample with complete enough
chemistry to support it (312 real samples as of this writing), with a
thermodynamic database auto-selected per sample by temperature and
total dissolved solids and a documented fallback when the first choice
fails to converge, as described in the Background section above. What
follows is the actual workflow behind that number: which samples get
modeled, how their chemistry is assembled and cleaned up before
PHREEQC ever sees it, and what the output is actually used to decide.

**Sample eligibility.** Not every sample in the database gets run
through PHREEQC. A sample qualifies once it has pH, temperature, and
major-ion concentrations for chloride, sodium, potassium, calcium,
magnesium, and alkalinity -- the minimum set a real speciation
calculation needs, not just whatever happens to be on hand. Field-
measured pH and temperature are preferred over laboratory-reported
values when both exist, since a lab reading taken days after
collection has already drifted from in-situ conditions; lab values are
used only when no field measurement exists. Additional analytes --
silica, sulfate, lithium, boron, and a handful of trace metals -- are
pulled in when available but aren't themselves gatekeeping. Samples
that don't clear this bar aren't silently dropped, either: each is
logged with the specific field it's missing, so the real, and
currently still substantial, gap between how many samples exist and
how many are PHREEQC-ready stays visible instead of disappearing into
an unexplained smaller number.

**Assembling and cleaning the chemistry first.** Samples come from
several sources spanning more than seventy years -- field sampling,
NDEP's routine and public-records-request chemistry, and historical
literature (Sorey and Colvard, Mariner and Janik, and Skalbeck)
transcribed directly rather than merely cited -- and those sources
don't agree on units or on how alkalinity is reported. Every
concentration is standardized to mg/L before it reaches a SOLUTION
block, and any alkalinity value reported on a CaCO3 basis is converted
to this project's standard HCO3-mass-equivalent basis using the real
50.05/61.02 equivalent-weight ratio (a conversion factor of 1.2189).
Getting this right turned out to matter more than it sounds like it
should: an earlier version of this pipeline simply left alkalinity out
for one major chemistry source, which alone pushed median charge-
balance error from a real ~0.7% up to a misleadingly bad ~60% for the
exact same samples -- a data-handling artifact, not an actual
analytical problem, and a good reminder to check a "bad" charge balance
before assuming the chemistry itself is at fault.

**End-member grouping.** Samples with complete chemistry are assigned
to one of three groups -- a thermal end-member (outlet and thermal-well
samples), mixed waters (chemistry carrying a partial but real thermal
signature), and dilute background waters -- using the hydrochemical-
facies clustering described in the next section, cross-checked against
PHREEQC's own fluid-type classification rather than just asserted from
the site-type label on file. This is what makes the conservative-
tracer check below, and the Cl/B comparison in the preliminary results
section, real chemically-defined groups rather than an arbitrary
site-based split.

**Conservative-tracer evaluation.** Four ratios -- Cl/B, Na/Cl, Li/Cl,
and SiO2/Cl -- are tracked across every era of chemistry on file (the
1950s, the 1990s, and the present) specifically to ask whether the real
thermal end-member's composition has stayed stable over seven decades,
independent of whatever a chloride-conductance calibration eventually
finds. A ratio that drifts meaningfully through time, or differs
systematically between eras constitutes real evidence about
source-fluid stability -- see the preliminary Cl/B result in @fig-clb
below --
rather than folded quietly into one pooled average that would hide
exactly that question.

**Mixing and inverse modeling.** A two-end-member conservative-mixing
fraction is computed directly from chloride, then used to build a real
PHREEQC `MIX` block combining the two end-member solutions at that
fraction and comparing PHREEQC's own predicted speciation against the
sample's observed chemistry -- a genuine test of the mixing hypothesis,
not just an arithmetic ratio. `INVERSE_MODELING` goes further, solving
for mixing fractions and mineral mass transfer simultaneously, within
stated elemental uncertainty bounds. Only one real mixing run and one
real inverse-modeling run exist as of this writing, both using a
representative rather than a time-paired thermal/meteoric end-member
pair, because no two chloride samples have yet been collected close
enough together in time to serve as a genuinely paired end-member set
-- exactly the gap the calibration sampling plan above is meant to
close. The automated parser for PHREEQC's own inverse-modeling text
output also has a known, real gap: it correctly identifies that a
feasible solution was found, but hasn't yet correctly pulled that
solution's mineral-mass-transfer numbers out of PHREEQC's raw text
report. This gap is stated directly rather than replaced with an
unverified number.

**Interpretation framework.** Every PHREEQC run in this project is
ultimately answering one of three questions, not run just because the
chemistry happened to be complete enough to try: are the real water
types here chemically distinct from each other, is conductance
behaving conservatively enough to stand in for chloride, and has the
thermal end-member's own composition held up over the full
seventy-year record. A modeling result that doesn't actually bear on
one of those three questions isn't reported here just because it's
available.

## Hydrochemical endmember characterization

Following Objectives 1 and 3 above, every real water sample with
complete major-ion chemistry -- 157 samples as of this writing -- is
assigned to a hydrochemical facies using unsupervised clustering, cross-checked by principal component
analysis and an independent statistical classification method, and
compared against PHREEQC's own fluid-type classification. The real
thermal end-member's chemistry -- specifically its chloride-to-boron,
sodium-to-chloride, lithium-to-chloride, and silica-to-chloride ratios
-- is then tracked across the 1950s, 1990s, and present eras to test
whether the source fluid's composition, and separately its modeled
equilibrium state, have remained stable over more than seventy years.

## Evaluating conductance as a conservative proxy for chloride

Following Objective 2, every process capable of altering either
chloride or conductance independently is evaluated explicitly rather
than assumed away:

| Process | Alters chloride? | Alters conductance? |
|---|---|---|
| Dilution with meteoric water | Yes | Yes |
| Calcite/dolomite precipitation | No | Yes |
| Thermal/meteoric mixing | Yes | Yes |
| Silica precipitation | No | Slight |
| Ion exchange (sodium-calcium) | No | Slight to moderate |
| Evaporative concentration | Yes (concentrates) | Yes (concentrates), but in a different proportion than dilution/mixing |

Where a process alters conductance without a matching change in
chloride, that is a direct, falsifiable reason the proxy relationship
could fail at a given site or season, and this failure mode is
stated directly rather than smoothed over. The same framework is
then applied to any real
outlier the eventual chloride-conductance calibration produces
(Objective 4): a sample sitting off the fitted relationship is checked
against this same process list before being treated as measurement
error.

## Chloride-conductivity calibration

This is the thesis's core analytical task, and it is explicitly not
yet executable with the real data on hand: no chloride sample
currently overlaps the period the conductivity loggers have been
running closely enough to serve as a real calibration pair. Once
paired samples exist (see the sampling plan below), the calibration
itself will be developed as a statistical regression of chloride
concentration against specific conductance, fit separately by site
type (outlet, creek, monitoring well, spring) and compared against a
single global model to test whether one relationship can serve the
whole system or whether site-specific calibrations are required.
Reported uncertainty will include both the regression's own
confidence interval and, where repeat sampling allows, an estimate of
how much the relationship drifts across seasons.

## Minimum sampling-frequency design via Monte Carlo subsampling

This analysis directly serves RQ2 and has already been carried out,
using the real, densely sampled specific-conductance record as a
stand-in for an equally dense chloride record (since no such dense
chloride record exists yet, this tests how much information is lost
by sampling a continuous signal sparsely, not the calibration itself).
The reconstructed mass flux for a given record is the time integral
of the product of chloride concentration and discharge,

$$
F = \int C(t)\,Q(t)\,dt ,
$$

where $C(t)$ is chloride concentration and $Q(t)$ is discharge,
evaluated numerically by trapezoidal integration.

For each candidate chemistry-sampling interval -- 1, 2, 3, 7, and 14
days -- a random start offset is drawn within the first interval, the
real dense conductance record is subsampled at that fixed interval,
the subsampled series is interpolated back to the original time
resolution, and the reconstruction error (root-mean-square error, as a
percent of the mean) is computed against the real, full-resolution
record. This is repeated 200 times per candidate interval, each with a
new random start day, so that both the typical and the worst-case
error a real field campaign could see -- purely as a function of which
day sampling happened to begin -- are both characterized, not just the
average case.

The real result: daily sampling keeps reconstruction error under
roughly 5% of the mean conductance at both logger sites; weekly
sampling roughly doubles that error (6-8%); biweekly sampling reaches
8-13%. The degradation is smooth, not a sharp cliff, which means the
choice of interval is a genuine precision-versus-cost trade-off rather
than a hard threshold. A specific tiered comparison makes this
concrete: at the upstream site, mean reconstruction error rises from
4.74% under daily sampling to 5.38% under every-other-day sampling --
a modest increase in the typical case -- while the worst-case error
(the 95th percentile across the 200 simulated campaigns) rises more
sharply, from 5.05% to 6.46%, a roughly 28% relative increase in how
bad an unlucky campaign could be. At the downstream site the same
pattern holds: mean error rises from 3.07% to 3.46%, while the worst
case rises from 3.21% to 4.34%, a roughly 35% relative increase. In
plain terms, relaxing from daily to every-other-day chemistry sampling
costs relatively little in the typical case but meaningfully widens
how bad a single unlucky field campaign's result could be -- a direct,
quantitative basis for the sampling-design recommendation below.

**Proposed sampling design.** Because no real chloride sample yet
overlaps the conductivity-logger deployment, the highest-priority
field task is a short, two-to-four-week, daily-to-sub-daily intensive
sampling campaign at both logger sites -- and, given that the loggers
have already been running since mid-2025, there is no real reason to
wait for a conventional spring/summer field season to start it. This
proposal now plans to begin that intensive window as soon as possible
(targeting this fall/winter), specifically so that monthly follow-
through sampling can begin right away rather than months from now.
Following the intensive window, chloride sampling should continue at
roughly monthly intervals at both SBRR and SBGG for a full annual
cycle -- not just "the remainder of the year," but a complete
winter-low-flow-to-spring-runoff-to-summer-baseflow-to-fall-transition
loop -- with additional samples deliberately timed around large storms
and spring snowmelt, when the relationship is most likely to break
down. Starting earlier and running a full year of monthly pairs turns
RQ1's seasonal-stability question from a stated hypothesis into
something this proposal can actually test with real data, rather than
extrapolating from the roughly five-week, summer-only conductance
record the Monte Carlo analysis above currently has to work with.
Every calibration sample should also carry major-ion chemistry, and a
subset should carry delta-18-O/delta-D isotopes, since chloride alone
cannot distinguish simple mixing from evaporative concentration. A
conventional practical minimum of 15-20 paired samples would support a
stable regression across the sampled range; a full year of monthly
sampling plus the intensive daily-to-sub-daily window comfortably
clears the more defensible 30-40-pair target without relying on a
single season's data.

## Post-eruption chloride-flux monitoring

Following RQ4, the chloride mass-balance method is applied to every
real date on which chloride and discharge are both available at the
two gauging stations, and the result is compared against the full
historical record -- White's 1950s/1960s estimates, Sorey and
Spielman's 2008-2016 values, and the 2026 poster estimate -- to ask
whether outflow has measurably changed since the eruption, while
explicitly accounting for the real seasonal variability Sorey and
Spielman themselves documented at these same sites. The same
conservative-ratio tracking used for endmember stability (above) is
applied here as a second, independent line of evidence: a stable
ratio across the eruption itself would support continuity of the same
source fluid, while a shift would support a change in fluid
provenance or mixing pathway.

## Geologic context: a three-dimensional framework, not a flow model

A three-dimensional geologic model -- built from well collars,
completion intervals, and, once available, digitized fault geometry
-- will provide real conceptual context for where the monitored wells
and springs sit relative to known structure, consistent with the
scope decision above. The leading candidate for real fault geometry is
the fault-and-lineament map published by Collar and Huntley
[@collar1990], chosen because it is a real, published map that also
locates several wells already identified in this project's own
records, giving natural control points for georeferencing. Once real
fault geometry exists, it will be used to test -- not assume -- which
chemically plausible end-members identified above are plausibly
connected, following the general principle that a defensible mixing
or reaction-path model should be built from which samples are
genuinely distinct, chemically credible end-members, not from which
samples happen to be close together on a map.

# Preliminary results

Real preliminary work already completed supports the feasibility of
this plan.

**Chloride-to-boron ratio across seven decades.** A two-sample
comparison of the curated thermal end-member (seven historical samples
from Sorey and Colvard [@sorey1992]; six modern samples from this
project's own field campaign) finds the ratio is similar in magnitude
but not statistically identical: 19.6 for 1950-1991 versus 21.9 for
2024-2026 (t = -2.92, 9.9 degrees of freedom, p = 0.015), shown in
@fig-clb. This is real evidence against a wholesale change in source
water, alongside a
small, real shift whose cause -- an analytical-method difference
between eras, versus a genuine geochemical change of roughly 12% --
remains an open question this proposal's work is designed to resolve
further.

```{r fig-clb}
#| results: asis
show_fig(
  "output/figures/manuscript/cl_b_ratio_by_era.png",
  "Chloride-to-boron ratio in real thermal end-member samples, 1950-2026. Already produced.",
  alt = "Scatter plot of chloride-to-boron ratio by well name, comparing 1950-1991 samples to 2024-2026 samples.",
  id = "fig-clb"
)
```

**Geothermometer divergence.** Sodium/potassium and quartz/chalcedony
geothermometers, applied to the same real thermal samples, diverge
systematically: quartz sits below measured discharge temperature,
sodium/potassium sits above it, exactly as expected for a single
cooling, diluting flow path from a genuinely hot (roughly
150-250 degree Celsius) reservoir rather than two disconnected water
types, as shown in @fig-geo.

```{r fig-geo}
#| results: asis
show_fig(
  "output/figures/manuscript/geothermometer_comparison.png",
  "Geothermometer estimates versus measured discharge temperature, per-well mean across all real thermal samples with complete chemistry. Already produced.",
  alt = "Paired-point plot comparing sodium/potassium and quartz/chalcedony geothermometer temperature estimates to measured discharge temperature, per well.",
  id = "fig-geo"
)
```

**Hydrochemical facies recover known geochemistry.** Clustering every
real sample with complete major-ion chemistry, independently
cross-checked by principal component analysis and a classification
method, recovers a pure sodium-chloride thermal cluster, a dilute
background-water cluster, and an intermediate mixed cluster, without
being told each sample's site type in advance -- three independent
statistical approaches converging on the same chemical structure, as
shown in @fig-facies.

```{r fig-facies}
#| results: asis
show_fig(
  "output/figures/facies/facies_pca_biplot.png",
  "Principal-component biplot of major-ion chemistry, colored by persisted hydrochemical-facies cluster assignment. Already produced.",
  alt = "Principal component biplot of major-ion chemistry, points colored by hydrochemical facies cluster, showing thermal, mixed, and background groups.",
  id = "fig-facies"
)
```

**A real, database-grounded discharge estimate.** Applying the
chloride mass-balance method to the one real date with paired
chloride and discharge measurements at both gauging stations gives
58.7 liters per second, a single-day value that sits within, not
clearly above or below, the broader 1955-2026 record (roughly
30-90 L/s depending on year and method) once Sorey and Spielman's own
documented seasonal variability is taken into account, shown in
@fig-discharge.

```{r fig-discharge}
#| results: asis
show_fig(
  "output/figures/manuscript/discharge_through_time.png",
  "Thermal-water discharge to Steamboat Creek, 1955-2026. Already produced.",
  alt = "Line and point plot of estimated thermal-water discharge to Steamboat Creek from 1955 to 2026.",
  id = "fig-discharge"
)
```

**Sampling-frequency feasibility.** The Monte Carlo result described
in the Methods section above and shown in @fig-mc is itself a
completed preliminary analysis, not a planned one.

```{r fig-mc}
#| results: asis
show_fig(
  "output/figures/manuscript/sampling_frequency_error_curve.png",
  "Monte Carlo reconstruction-error curve for candidate chemistry-sampling intervals. Already produced.",
  alt = "Line plot of Monte Carlo reconstruction error versus chemistry-sampling interval, from daily to biweekly sampling.",
  id = "fig-mc"
)
```

**A real three-month field-parameter time series.** A roughly monthly
record of pH, temperature, and specific conductance across fourteen
sites -- six injection-side ports and eight shallow monitoring and
domestic wells -- spanning late September through December 2025 shows
a stable, consistent gap between the injection-side ports (pH 5.8-6.3)
and the shallow well network (pH 7.0-8.1), present in every round
rather than as a single-date artifact. Two individual wells also show
real, opposite-direction multi-month trends not explained by any
change in the other monitored parameter at the same well, a pattern
that cautions against treating any single field parameter as a
universal proxy for thermal fraction across the whole system.

**A still-changing system: September 2026 field observations.** Field
photographs taken September 27, 2026 (GPS- and time-stamped by camera
EXIF data, all within about 150 meters of each other at the terrace)
document that the system has continued to change since the 2026
poster was presented. First, a new seep-fed hot spring, shown in
@fig-hotspring, is now discharging among bedrock boulders at a
location that was not an active feature previously; a nearby resident
independently confirmed that this is a new development, consistent
with the thesis's own field observation rather than resting on a
single observer's judgment. Second, a native-sulfur deposit at the
terrace, shown in @fig-sulfur, has visibly grown in size and remains
texturally fresh, again by both the resident's account and direct
observation -- a real, physical sign of continuing near-surface
fumarolic input, not a static relic of the 2025 event. Third, a
separate point at the terrace, shown in @fig-seep, shows a seep
discharging from the alluvial cover directly onto the underlying hard
sinter bedrock outcrop, a concrete example of the alluvium-to-bedrock
contact this thesis's own spring and well remapping is designed to
document systematically rather than opportunistically. Together,
these three observations are read as further, independent support for
the poster's own conclusion -- fluids continuing to redistribute
within an existing shallow conduit system, rather than the system
having settled into a new, static state since the 2025 eruption -- and
they are a direct, concrete motivation for why the remapping and
monitoring work proposed below needs to happen now rather than later.

```{r fig-hotspring}
#| results: asis
show_fig(
  "output/figures/manuscript/field_photo_new_hot_spring_20260927.jpg",
  "A new seep-fed hot spring discharging among bedrock boulders near Steamboat Creek, photographed September 27, 2026 (39.3804, -119.7398). Not previously an active feature; independently confirmed as new by a nearby resident.",
  alt = "Photograph of a new seep discharging warm water among bedrock boulders toward Steamboat Creek.",
  id = "fig-hotspring"
)
```

```{r fig-sulfur}
#| results: asis
show_fig(
  "output/figures/manuscript/field_photo_sulfur_deposit_20260927.jpg",
  "A fresh, visibly growing native-sulfur deposit (yellow) along the Steamboat Creek bank, photographed September 27, 2026 (39.3805, -119.7396).",
  alt = "Photograph of a bright yellow native-sulfur deposit along a creek bank, with white sinter crust surrounding it.",
  id = "fig-sulfur"
)
```

```{r fig-seep}
#| results: asis
show_fig(
  "output/figures/manuscript/field_photo_alluvium_sinter_seep_20260927.jpg",
  "A seep discharging from alluvial cover directly onto the underlying hard sinter bedrock outcrop, photographed September 27, 2026 (39.3829, -119.7402).",
  alt = "Photograph of a bare alluvial flat transitioning to a hardened white sinter outcrop at a seep location.",
  id = "fig-seep"
)
```

## Planned figures, not yet produced

Three figures central to the finished thesis cannot be produced yet,
and are described here rather than filled with a placeholder or an
unrepresentative result.

**A real chloride-conductivity calibration plot** (chloride
concentration versus specific conductance, with a fitted regression
line and confidence band, by site type) will be the thesis's central
figure. It requires the paired chloride-and-conductance samples the
sampling plan above is designed to collect, and cannot be produced from
the current, unpaired records.

**A monitoring-network map** showing production, injection, and
monitoring wells together with mapped structure and lithology will
provide the geologic context described above. It awaits the
digitized fault geometry discussed in the geologic-context section.

**A final conceptual model** of the Lower Sinter Terrace system,
synthesizing the chemical, discharge, and geologic evidence into a
single interpretive diagram, will be developed once the calibration,
tracer-ratio, and geologic-context results above are complete -- it
is deliberately not sketched in advance of that evidence.

**The three September 27, 2026 field photographs** above (@fig-hotspring,
@fig-sulfur, @fig-seep) carry real GPS coordinates and timestamps in
their own camera EXIF data, confirmed directly rather than assumed,
but have not yet been run through this project's formal photo-location
ingestion pipeline to become real, dated `Locations` entries in the
underlying database -- that promotion step, not the figures
themselves, is the remaining task.

# Preliminary schedule

The schedule below moves the intensive calibration window up to this
fall/winter (rather than a conventional spring field-season start) and
stretches monthly chloride sampling across a full annual cycle, so the
calibration can be tested against all four seasons before it is
treated as final, consistent with the sampling-design reasoning above.
Target dates are a planning structure, not fixed commitments, and will
be adjusted against the real academic calendar and committee
availability as the work proceeds. The most direct risk to this
schedule is weather- or access-limited fieldwork during the
fall/winter calibration window; the contingency is the same full-year
monthly cadence shifted later into spring/summer 2027, at the cost of
testing fewer distinct seasons before the calibration is finalized.
A second risk is that the real chloride-conductance relationship
turns out to be too noisy or site-specific to support a single
practical calibration; the contingency there is reported directly in
RQ1's own deliverable above (site-specific equations with stated
uncertainty, not a forced single model) rather than treated as a
schedule failure.

| Target date | Milestone |
|---|---|
| Oct-Nov 2026 | Submit this proposal and program of study; begin the intensive daily-to-sub-daily Cl-conductance calibration window at SBRR/SBGG (moved up from a spring 2027 start, since the conductivity loggers are already running); continue spring and well remapping at the Lower Sinter Terrace. |
| Dec 2026 | First monthly chloride sample of the full-year cadence (early winter baseflow conditions). |
| Jan 2027 | Monthly chloride sample; winter-season logger maintenance and remapping. |
| Feb 2027 | Monthly chloride sample. |
| Mar 2027 | Monthly chloride sample; watch for early spring runoff onset. |
| Apr 2027 | Monthly chloride sample, with additional unscheduled samples around spring snowmelt/runoff, when the Cl-conductance relationship is most likely to shift. |
| May 2027 | Monthly chloride sample; isotope sampling on the prioritized subset. |
| Jun 2027 | Monthly chloride sample; roughly one full year of continuous conductance logging completed; mid-candidacy committee check-in and preliminary-results presentation. |
| Jul 2027 | Monthly chloride sample; begin fitting a preliminary seasonal calibration now that a full annual cycle of paired samples exists. |
| Aug 2027 | Monthly chloride sample; evaluate calibration residuals by season (winter vs. runoff vs. summer vs. fall). |
| Sep 2027 | Final monthly chloride sample of the full-year campaign (the twelfth month); finalize the chloride-conductance calibration, both site-specific and global models. |
| Oct 2027 | Complete the PHREEQC conservativeness and endmember-stability analyses; apply the post-eruption discharge and tracer-ratio interpretation. |
| Nov 2027 | Finalize the geologic-context framework; begin the full thesis-writing push. |
| Dec 2027 | Complete thesis writing, defend, and submit. |

# Expected products

- A real chloride-to-specific-conductance calibration equation, or set
  of site-specific equations, with stated uncertainty.
- A quantitative minimum-sampling-frequency recommendation for
  ongoing chloride monitoring at Steamboat.
- A PHREEQC-based physical justification -- or documented limitation
  -- for using specific conductance as a chloride proxy at this site.
- An updated interpretation of thermal-water discharge and conservative-
  tracer behavior before and after the 2025 eruption.
- Two candidate manuscripts: *Conductivity-Based Monitoring of
  Chloride Discharge at Steamboat Hills* (targeting *Geothermics* or
  *Geothermal Energy*), and *Hydrochemical Evolution of the Lower
  Sinter Terrace System Following the 2025 Eruption* (drawing on the
  PHREEQC, tracer-ratio, and historical-comparison results above).
- A three-dimensional conceptual geologic framework for the monitored
  wells and springs, explicitly scoped as context rather than a flow
  model.

# Conclusion

This proposal answers a question Sorey originally posed and this
project's own 2026 poster [@irvinlindsey2026] re-asked after the 2025
eruption: is Steamboat's renewed activity a new thermal source, or an
old one finding a new way out? The poster's own evidence -- matching
chemistry, a strong inverse relationship between paired temperature
loggers, and continuing physical change at the terrace through at
least September 2026 -- points toward the latter, but that evidence
was gathered at poster speed, not thesis speed: a handful of samples,
one representative mixing pair, and no real chloride-conductance
calibration yet. What changes once this work is done is that the
same question gets a real, stated-uncertainty answer instead of a
preliminary one. A working chloride-to-conductance calibration turns
a labor-intensive, infrequent sampling program into a continuous,
field-deployable monitoring tool -- not just for this thesis, but for
whoever manages this system's monitoring network after it. A
PHREEQC-based accounting of why conductance should or should not track
chloride turns that tool from a correlation into something with a
stated physical basis and known limits. And a real, if still partial,
three-dimensional geologic framework starts to connect the chemistry
to the structure it is moving through, rather than leaving the two
literatures to talk past each other. None of this closes the book on
Steamboat -- the system is still visibly changing, as the September
2026 field observations above show -- but it leaves whoever studies it
next with a real, working method for tracking that change, rather
than another one-time estimate.

# References {.unnumbered}
