---
title: "Steamboat Hills: Talking Points for Groundwater Hydrology Advisor Meeting"
subtitle: "Where the data stands, and how to think about modeling the system"
date: today
format:
  pdf:
    toc: false
    number-sections: false
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
if (basename(getwd()) == "outreach") {
  root <- normalizePath(file.path(getwd(), "..", ".."))
} else {
  root <- normalizePath(getwd())
}
knitr::opts_knit$set(root.dir = root)
```

## Why this conversation, right now

Over the last few weeks a set of scanned NDEP field-sampling forms got
transcribed into the project database, turning a handful of one-off
grab samples into a real, roughly **monthly time series (late
September through December 2025) across 14 sites** -- six named
injection-side ports (Galena 1/2/3, SB2, SB3, SBHR) and eight shallow
monitoring/domestic wells (Soccer Field, Herz Domestic, Herz Deep,
Eich, NDOT, Boyd, Jeppson, Rogers). This is the densest *repeated*
field-parameter dataset the project has had to date, and it raises a
concrete question worth a faculty opinion: **what modeling approach
does data like this actually support?**

## What the data shows

```{r fig-main}
#| echo: false
#| results: asis
#| message: false
#| warning: false
img_path <- file.path(normalizePath(getwd()), "output", "figures", "manuscript", "u230_timeseries.png")
cap <- "Left: pH across the six injection-side ports vs. the eight shallow monitoring/domestic wells, pooled across three real monthly rounds. Right: two real multi-month trends at individual wells, Sept-Dec 2025."
knitr::asis_output(sprintf("![%s](%s)", cap, img_path))
```

**1. A stable pH gap between the injection side and the shallow
network.** Every one of the six injection-side ports reads pH 5.8-6.3
at every round; every one of the eight shallow wells reads pH
7.0-8.1. That ~1.5-2 unit gap holds across all three rounds, not just
one snapshot -- a real, repeatable signature, not noise.

**2. Two individual wells show real, opposite-direction trends over
the same three months.** Herz Deep's specific conductance rises from
1,877 to 3,024 uS/cm (roughly +60%) while its temperature stays
essentially flat (~30-36 C). NDOT's temperature falls from about 76 C
to 61 C while its conductance stays roughly flat (~3,000-3,500 uS/cm).
Neither well's two parameters move together.

**3. A broader context for these 14 sites**: this project's own
poster finding puts current thermal-water discharge to Steamboat Creek
at roughly 27-59 L/s (a single real paired Cl-discharge measurement
gives 58.7 L/s; the poster's own multi-date average is ~27 L/s),
compared to a historical range of roughly 60-90 L/s reported for
2008-2016 and over 1,000 L/s reported by White (1968) in the 1950s-60s
-- the long-run context this new monthly dataset sits inside.

## Questions for the conversation

- **Does the pH gap argue for reactive transport over conservative
  mixing?** A stable, multi-round pH offset between an injected fluid
  and the shallow network looks like water-rock buffering along a flow
  path -- is that the right read, or could something simpler (e.g. a
  measurement-method difference between the two sample types) explain
  it just as well?
- **How should two wells moving in opposite, uncorrelated directions
  be interpreted?** Herz Deep's conductance rise and NDOT's temperature
  fall don't look like the same mechanism. Is this two independent,
  local processes, or a sign that neither temperature nor conductance
  alone is a safe stand-in for "thermal influence" at this site?
- **What's the right first modeling step given the real data density
  we have?** A full 3-D reactive-transport model is the long-term goal,
  but real continuous water-level coverage *inside* the Steamboat
  field itself is currently limited to 2-3 wells (everything else with
  continuous water level sits in the wider South Truckee Meadows
  basin). Is a simpler lumped-parameter/box model (two- or
  three-end-member mixing with a time-varying fraction) the
  appropriate first step, or is it worth waiting on better spatial
  coverage before committing to any specific model structure?
- **Does this dataset make the planned Cl-conductivity calibration
  campaign more or less urgent?** The project's sampling-frequency
  statistical framework (Monte Carlo subsampling, rolling-origin
  cross-validation) is fully built but still synthetic/proxy-only --
  no real chloride sample yet overlaps the conductivity-logger
  deployment window. Given the pH/trend patterns above, would a
  professor prioritize that calibration field season over, say,
  pursuing the fault-digitization/potentiometric-surface work in
  parallel?

## One flagged data-quality note, for transparency

One value in this new batch (Herz Domestic Well, 10/24/2025 specific
conductance, 9,933 uS/cm) is roughly 7-9x every other reading on file
for that well and is flagged in the database as a likely OCR
transcription error pending a check against the original scanned form
-- not used in the trends described above.

## One-paragraph summary, if time is short

*"We just turned scattered grab samples into a real three-month,
14-site monthly time series. Two patterns stand out: a stable pH gap
between the injection fluid and the shallow groundwater network
(suggesting real water-rock interaction along the flow path), and two
wells trending in opposite directions on temperature vs. conductivity
over the same window (suggesting neither parameter alone is a safe
universal proxy for thermal influence here). I'd like your take on
whether this points toward a reactive-transport framing over
conservative mixing, and what the right first modeling step is given
that our real water-level coverage inside the field is still thin."*
