---
title: "Conductivity Temporal Structure"
subtitle: "Diurnal cycles, autocorrelation, and dominant frequencies at the two Steamboat Creek conductivity loggers"
date: today
format:
  html:
    code-fold: show
---

## Purpose

This notebook runs **today**, on conductivity alone — it does not require
paired chloride samples. It characterizes the temporal structure of the two
HOBO conductivity loggers (`SBRR` upstream/control, `SBGG` downstream) so
that later steps (chloride prediction, Monte Carlo sampling-frequency
simulation) are built on an evidence-based understanding of the signal
rather than assumptions.

::: {.callout-warning}
As of this notebook's first run, each logger has only ~2.5 weeks of record
(mid-July–early August 2026). Diurnal-cycle and short-lag autocorrelation
results are meaningful now; **weekly/monthly-frequency conclusions are not**
— revisit this notebook as more months of data accumulate.
:::

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)

# Notebooks live directly under the project root (see notebooks/_quarto.yml),
# so the database is one level up regardless of the caller's working directory.
db_path <- file.path("..", "database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_demo.sqlite")

con <- dbConnect(SQLite(), db_path, flags = SQLITE_RO)
```

## Load conductivity observations

Excludes rows flagged `spike` or `field_visit_disturbance` by
`scripts/qc/qc_conductivity_checks.R`; `gap_boundary`/`dropout` flags (if
present) are kept but noted.

```{r load-data}
obs <- dbGetQuery(con, "
  SELECT
    co.logger_id,
    cl.role,
    l.external_station_code,
    co.timestamp,
    co.ec_raw,
    co.temperature_c,
    co.sc_25c,
    co.qc_flag
  FROM Conductivity_Observations co
  JOIN Conductivity_Loggers cl ON co.logger_id = cl.logger_id
  LEFT JOIN Locations l ON cl.location_id = l.location_id
  ORDER BY co.logger_id, co.timestamp
") |>
  mutate(timestamp = as.POSIXct(timestamp, tz = "UTC"))

dbDisconnect(con)

n_flagged <- sum(!is.na(obs$qc_flag))
message(nrow(obs), " total observations; ", n_flagged, " QC-flagged (spike/disturbance/gap).")

obs_clean <- obs |> filter(is.na(qc_flag) | !(qc_flag %in% c("spike", "field_visit_disturbance")))

obs_clean |>
  group_by(external_station_code) |>
  summarise(
    n = n(),
    start = min(timestamp),
    end = max(timestamp),
    span_days = round(as.numeric(difftime(max(timestamp), min(timestamp), units = "days")), 1)
  )
```

## Diurnal cycle

```{r diurnal}
#| fig-cap: "Hour-of-day distribution of specific conductance, by logger"
obs_clean |>
  mutate(hour = hour(timestamp) + minute(timestamp) / 60) |>
  ggplot(aes(x = hour, y = sc_25c, group = hour)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~external_station_code, ncol = 1, scales = "free_y") +
  labs(x = "Hour of day (UTC)", y = "Specific conductance @ 25C (uS/cm)") +
  theme_minimal()
```

```{r diurnal-harmonic}
#| fig-cap: "24-hour harmonic fit to specific conductance, by logger"
fit_harmonic <- function(df) {
  t_hours <- as.numeric(difftime(df$timestamp, min(df$timestamp), units = "hours"))
  fit <- lm(sc_25c ~ sin(2 * pi * t_hours / 24) + cos(2 * pi * t_hours / 24), data = df)
  df$fitted <- predict(fit, newdata = df)
  df
}

obs_clean |>
  filter(!is.na(sc_25c)) |>
  group_by(external_station_code) |>
  group_modify(~fit_harmonic(.x)) |>
  ungroup() |>
  ggplot(aes(x = timestamp)) +
  geom_line(aes(y = sc_25c), alpha = 0.3) +
  geom_line(aes(y = fitted), color = "firebrick") +
  facet_wrap(~external_station_code, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Specific conductance @ 25C (uS/cm)",
       subtitle = "Black = observed, red = fitted 24-hr harmonic") +
  theme_minimal()
```

## Autocorrelation structure

```{r acf-native}
#| fig-cap: "ACF at native (~5-min) cadence, by logger"
for (site in unique(obs_clean$external_station_code)) {
  x <- obs_clean |> filter(external_station_code == site) |> pull(sc_25c)
  x <- x[!is.na(x)]
  acf(x, lag.max = 24 * 12, main = paste("ACF —", site, "(lags in 5-min steps, 12/hr)"))
}
```

```{r acf-resampled}
#| fig-cap: "ACF resampled to candidate chemistry-sampling intervals"
resample_to <- function(df, interval_hours) {
  df |>
    mutate(bin = floor((as.numeric(timestamp) - as.numeric(min(timestamp))) / (interval_hours * 3600))) |>
    group_by(bin) |>
    summarise(sc_25c = mean(sc_25c, na.rm = TRUE), .groups = "drop")
}

candidate_intervals_hr <- c(24, 48, 72, 168)  # daily, 2-day, 3-day, weekly

for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |> filter(external_station_code == site)
  for (h in candidate_intervals_hr) {
    rs <- resample_to(site_df, h)
    if (nrow(rs) > 10) {
      acf(rs$sc_25c, na.action = na.pass,
          main = paste0(site, " — resampled every ", h, "h (n=", nrow(rs), ")"))
    } else {
      message(site, " @ ", h, "h: too few points yet (n=", nrow(rs), ") for a meaningful ACF")
    }
  }
}
```

## Spectral analysis

```{r spectral}
#| fig-cap: "Periodogram (dominant frequencies), by logger"
for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |> filter(external_station_code == site, !is.na(sc_25c)) |> arrange(timestamp)

  # Native cadence is nominally 5 min; a plain periodogram assumes even
  # spacing. Any residual gaps after QC are small relative to record
  # length for now — revisit with Lomb-Scargle if gaps grow.
  sp <- spec.pgram(site_df$sc_25c, taper = 0.1, log = "no", plot = FALSE)

  period_hours <- (1 / sp$freq) * (5 / 60)  # 5-min steps -> hours

  plot(period_hours, sp$spec, type = "l", log = "x",
       xlab = "Period (hours)", ylab = "Spectral density",
       main = paste("Periodogram —", site))
  abline(v = 24, col = "firebrick", lty = 2)
  legend("topright", legend = "24-hr", col = "firebrick", lty = 2, bty = "n")
}
```

## Time-series decomposition

```{r decomposition}
#| fig-cap: "STL decomposition (trend / diurnal / remainder), by logger"
for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |>
    filter(external_station_code == site) |>
    arrange(timestamp)

  # Regularize to 5-minute grid via linear interpolation across small gaps
  full_grid <- seq(min(site_df$timestamp), max(site_df$timestamp), by = "5 min")
  interp <- approx(site_df$timestamp, site_df$sc_25c, xout = full_grid)$y

  ts_obj <- ts(interp, frequency = 288)  # 288 = 5-min steps per day

  if (sum(!is.na(interp)) > 2 * 288) {
    decomp <- stl(zoo::na.approx(ts_obj, na.rm = FALSE), s.window = "periodic", robust = TRUE)
    plot(decomp, main = paste("STL decomposition —", site))
  } else {
    message(site, ": record too short yet for a reliable STL decomposition (need > 2 full days).")
  }
}
```

## Summary / next steps

- Both loggers show a clear diurnal signal consistent with evapotranspiration
  / thermal-loading control on discharge and specific conductance.
- Autocorrelation at native cadence is strong out to several hours, as
  expected; resampled ACFs at daily+ intervals will become informative once
  more weeks/months of record accumulate.
- Next: `scripts/analysis/sampling_frequency/02_specific_conductance.R`
  (recalibrate the temperature-compensation coefficient once field
  conductivity checks exist), then
  `notebooks/02_chloride_prediction_models.qmd` once paired Cl samples are
  ingested for `SBRR`/`SBGG`.
