---
title: "Conductivity as a Chloride-Discharge Proxy: Temporal Structure, Calibration, and Sampling-Frequency Design"
subtitle: "Overarching reference for the Steamboat Creek continuous-conductivity / chloride sampling-frequency workflow"
date: today
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

This notebook is the **living scientific reference** for the continuous-conductivity
and chloride-sampling-frequency workstream at Steamboat Hills — it documents *why*
each analysis is done, *what* the math is, and *where* the code that implements it
lives, and it is intended to stay the entry point for this line of work as more
data and methods (new logger deployments, real paired Cl samples, refined models)
accumulate. It complements, but does not replace, the individual R scripts in
`scripts/analysis/sampling_frequency/`, which remain the executable source of
truth; this document narrates and runs representative pieces of them so the
reasoning and the code stay in sync.

Sections below run in dependency order:

1. Scientific motivation (why continuous conductivity, why chloride, why sampling
   frequency matters at all).
2. Instrumentation and site roles, with the provenance of a metadata correction
   that affects everything downstream.
3. Raw EC $\rightarrow$ temperature-compensated specific conductance, and the
   current state of field-calibration.
4. Temporal structure of specific conductance alone (this section requires no
   chloride data and is fully populated today).
5. The conductivity $\rightarrow$ chloride predictive-model framework (currently
   demonstrated on synthetic data pending temporal overlap with real samples).
6. Monte Carlo sampling-frequency design (same caveat).
7. Turning an error curve into an actionable sampling recommendation.
8. Current status, known gaps, and a running changelog.

::: {.callout-warning}
As of this writing, each conductivity logger has only a few weeks of continuous
record (deployed 2026-07-15), and the only chloride/isotope field samples
collected so far (2026-05-01) **predate that deployment by roughly 2.5 months**.
Sections 5–7 therefore run against synthetic self-test data, clearly labeled as
such. Section 4 (temporal structure of conductivity alone) is real and current.
Re-run this notebook once chloride samples exist that overlap the logger record.
:::

## Scientific motivation

Steamboat Hills is a geothermal outflow system in which chloride behaves
approximately conservatively (non-reactive, non-volatile) in the shallow outflow
plume, so the classic approach — used by Michael Sorey in prior studies of this
system — is to treat **chloride discharge**, i.e. the mass flux

$$
\Phi_{Cl} = \int_{t_0}^{t_1} C_{Cl}(t)\, Q(t)\, dt
$$

(concentration $C_{Cl}$ in mg/L $\equiv$ g/m$^3$, discharge $Q$ in m$^3$/s, so
$\Phi_{Cl}$ has units of grams over the interval $[t_0, t_1]$), as a tracer for
the total geothermal outflow reaching the surface stream network. A change in
$\Phi_{Cl}$ following the Lower Sinter Terrace eruption is direct evidence of a
change in the underlying hydrothermal discharge, independent of dilution effects
that would otherwise obscure a pure-concentration or pure-discharge signal.

Two practical problems follow directly from that framing:

- **Chloride analyses are expensive and infrequent** (lab turnaround, sampling
  effort), so $C_{Cl}(t)$ is only ever known at a sparse set of times.
- **Specific conductance is cheap, continuous, and correlates with dissolved
  ion content** (conductivity loggers record every 5 minutes), so it is a
  natural *proxy* for $C_{Cl}(t)$ between chemistry samples — provided the
  conductance-to-chloride relationship is characterized and its stability
  over time is checked.

This notebook's methodological goal is therefore not just "does conductivity
track chloride" but the sharper, decision-relevant question the thesis poses:
**what is the coarsest chemistry (chloride) sampling interval that still lets us
reconstruct $\Phi_{Cl}$, or $C_{Cl}(t)$, within an error the PI considers
acceptable?** Sections 5–7 build the machinery to answer that quantitatively
rather than by convention (e.g. "monthly because that's typical").

## Instrumentation, site roles, and data provenance

Two HOBO/Onset "Full Range" EC loggers were deployed 2026-07-15 in Steamboat
Creek to support this question:

| Location code | Role | Serial | Rationale |
|---|---|---|---|
| `SBRR` | `upstream_control` | 22575724 | Co-located with the USGS gauge near Rhodes Road; represents background/upstream water quality, upstream of the dominant geothermal inflow. |
| `SBGG` | `downstream` | 22575725 | Downstream of the outflow zone; expected to carry the geothermal Cl/EC signature. |

::: {.callout-note collapse="true"}
### Provenance note: role/serial correction (2026-09-05)

The `SBRR`/`SBGG` logger-to-role assignment was originally guessed from
coordinate proximity alone and had the two serials **swapped**: the logger
originally labeled `SBRR` (upstream control) was reading 850–1300 $\mu$S/cm,
which is far too high for background water and matches the expected
downstream/near-outflow signature instead. The assignment was corrected in
`data/raw/conductivity/conductivity_logger_deployments.csv` and re-ingested.
This was a **metadata-only** fix — the already-ingested
`Conductivity_Observations` rows are keyed by `logger_id`, not by location, so
no observation rows needed to change, only which location/role each
`logger_id` points to. After the fix, `SBRR` reads consistently lower than
`SBGG` (means roughly 333 vs. 1069 $\mu$S/cm), which is now physically
sensible for an upstream-control vs. downstream pairing.

This correction matters for every downstream step in this document: any
model, calibration, or figure that joins on `location_id`/`role` inherits the
fix automatically (nothing in the modeling code hard-codes which physical
serial is upstream — see the calibration check in the next section), but any
*prior* interpretation of "SBRR" vs. "SBGG" behavior made before this date
should be re-read with the roles reversed.
:::

The `Conductivity_Loggers` table (below) is the single source of truth for
this mapping; everything else in this document resolves location and role
through it rather than hard-coding a serial number or coordinate.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)

# All chunks in this notebook execute with the working directory pinned to
# the project root, not to notebooks/. This matters because several sourced
# scripts (e.g. 02_specific_conductance.R) themselves source() other helper
# scripts using paths written relative to the project root -- the same
# assumption run_pipeline.R makes -- so pinning root.dir here keeps every
# path below identical to how these scripts are invoked from the pipeline.
knitr::opts_knit$set(root.dir = normalizePath(file.path(getwd(), "..")))
```

```{r connect-db}
#| message: false
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("database", "geochem_demo.sqlite")

con <- dbConnect(SQLite(), db_path, flags = SQLITE_RO)

dbGetQuery(con, "
  SELECT cl.logger_id, cl.location_id, cl.role, cl.serial_number, l.external_station_code
  FROM Conductivity_Loggers cl
  LEFT JOIN Locations l ON cl.location_id = l.location_id
")
```

## From raw EC to specific conductance: temperature compensation and calibration

Raw electrical conductivity ($EC_{raw}$) is temperature-dependent, so it is
converted to **specific conductance at 25$^\circ$C** ($SC_{25}$) before any
comparison across time or between the two loggers. The linear
temperature-compensation model (USGS Techniques and Methods 1-D3) is:

$$
SC_{25} = \frac{EC_{raw}}{1 + \beta\,(T - 25)}
$$

with reference temperature $\theta = 25^\circ C$ and compensation coefficient
$\beta$ (USGS default $\beta = 0.0191\ ^\circ C^{-1}$). This is implemented at
ingest time in `scripts/ingest/helpers/compute_specific_conductance.R`:

```{r sc-formula-demo}
source(file.path("scripts", "ingest", "helpers", "compute_specific_conductance.R"))

# Reproduce the ingest-time formula directly, so the math and the code
# shown here are provably the same function used to populate sc_25c.
compute_specific_conductance
```

Because $\beta$ is an empirical property of the local water chemistry (ionic
composition affects how conductivity responds to temperature), it should be
periodically re-estimated against paired field conductivity checks rather than
assumed to equal the USGS default indefinitely. `02_specific_conductance.R`
does this by fitting

$$
EC_{field,i} = \frac{EC_{raw,i}}{1 + b\,(T_i - 25)} + \varepsilon_i
$$

by nonlinear least squares, where $(EC_{field,i}, EC_{raw,i}, T_i)$ is the
$i$-th field check matched to the nearest logger observation within a time
tolerance (default 15 min), and $b$ is estimated in place of the fixed
default. The fit is only attempted with $\geq 5$ pairs; otherwise the USGS
default is kept and reported as such (no silent extrapolation from too few
points).

**This step is location/role-agnostic by construction**: pairing is done
purely by matching a field check's `location_id` to a logger's `location_id`
(`scripts/analysis/sampling_frequency/02_specific_conductance.R`, `get_calibration_pairs()`),
so the `SBRR`/`SBGG` role correction above required **no code change here** —
it is picked up automatically through the corrected `Conductivity_Loggers`
table. The check below runs the real function against the live database:

```{r calibration-check}
source(file.path("scripts", "analysis", "sampling_frequency", "02_specific_conductance.R"))

pairs_now <- get_calibration_pairs(con)
cat("Field-vs-logger calibration pairs currently available:", nrow(pairs_now), "\n")

calib_result <- recalibrate_sc_coefficient(con)
calib_result[c("coefficient", "n_pairs")]
```

There are currently 0 matched pairs, and the reason is the same temporal gap
noted in the callout above: field conductivity was recorded at `SBRR` (241
$\mu$S/cm, sample date 2026-05-01) and other locations, but no
`Conductivity_Observations` exist yet for that date — logging only began
2026-07-15. As a rough sanity check (not a calibration, since it is not a
time-matched pair), 241 $\mu$S/cm at `SBRR` is the right order of magnitude
relative to `SBRR`'s post-correction logger mean (~333 $\mu$S/cm), which is
mildly reassuring about the corrected role assignment but is not a substitute
for an actual paired check. **Action item:** collect at least 5 field
conductivity checks at `SBRR`/`SBGG` while the loggers are running to enable a
real recalibration.

## Temporal structure of specific conductance

This section requires no chloride data and is fully populated with real
observations. The goal is to characterize the signal conductivity is
expected to reconstruct chloride *from*, before any predictive model is built
on top of it — diurnal cycling, autocorrelation persistence, and dominant
frequencies all directly inform what chemistry-sampling intervals could
plausibly be sufficient in Sections 5–7.

```{r load-data}
obs <- dbGetQuery(con, "
  SELECT
    co.logger_id,
    cl.role,
    l.external_station_code,
    co.timestamp,
    co.ec_raw,
    co.temperature_c,
    co.sc_25c,
    co.qc_flag
  FROM Conductivity_Observations co
  JOIN Conductivity_Loggers cl ON co.logger_id = cl.logger_id
  LEFT JOIN Locations l ON cl.location_id = l.location_id
  ORDER BY co.logger_id, co.timestamp
") |>
  mutate(timestamp = as.POSIXct(timestamp, tz = "UTC"))

dbDisconnect(con)

n_flagged <- sum(!is.na(obs$qc_flag))
message(nrow(obs), " total observations; ", n_flagged, " QC-flagged (spike/disturbance/gap).")

# Excludes rows flagged 'spike' or 'field_visit_disturbance' by
# scripts/qc/qc_conductivity_checks.R; 'gap_boundary'/'dropout' flags (if
# present) are kept but noted, since they mark real data gaps rather than
# untrustworthy values.
obs_clean <- obs |> filter(is.na(qc_flag) | !(qc_flag %in% c("spike", "field_visit_disturbance")))

obs_clean |>
  group_by(external_station_code) |>
  summarise(
    n = n(),
    start = min(timestamp),
    end = max(timestamp),
    span_days = round(as.numeric(difftime(max(timestamp), min(timestamp), units = "days")), 1)
  )
```

### Diurnal cycle

Shallow stream conductivity commonly shows a 24-hour cycle driven by
evapotranspiration and thermal loading modulating the mix of dilute surface
water and higher-EC baseflow/geothermal inflow. We fit a first-harmonic
(single sinusoid) regression at the 24-hour period to each logger:

$$
SC_{25}(t) = \alpha + \beta_1 \sin\!\left(\frac{2\pi t}{24}\right) + \beta_2 \cos\!\left(\frac{2\pi t}{24}\right) + \varepsilon(t)
$$

with $t$ in hours since the start of each logger's record. The
amplitude of the cycle is $\sqrt{\beta_1^2 + \beta_2^2}$.

```{r diurnal}
#| fig-cap: "Hour-of-day distribution of specific conductance, by logger"
obs_clean |>
  mutate(hour = hour(timestamp) + minute(timestamp) / 60) |>
  ggplot(aes(x = hour, y = sc_25c, group = hour)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~external_station_code, ncol = 1, scales = "free_y") +
  labs(x = "Hour of day (UTC)", y = "Specific conductance @ 25C (uS/cm)") +
  theme_minimal()
```

```{r diurnal-harmonic}
#| fig-cap: "24-hour harmonic fit to specific conductance, by logger"
fit_harmonic <- function(df) {
  t_hours <- as.numeric(difftime(df$timestamp, min(df$timestamp), units = "hours"))
  # sc_25c(t) = alpha + b1*sin(2*pi*t/24) + b2*cos(2*pi*t/24) + e
  fit <- lm(sc_25c ~ sin(2 * pi * t_hours / 24) + cos(2 * pi * t_hours / 24), data = df)
  df$fitted <- predict(fit, newdata = df)
  attr(df, "amplitude") <- sqrt(sum(coef(fit)[-1]^2))
  df
}

harmonic_fits <- obs_clean |>
  filter(!is.na(sc_25c)) |>
  group_by(external_station_code) |>
  group_modify(~fit_harmonic(.x)) |>
  ungroup()

harmonic_fits |>
  ggplot(aes(x = timestamp)) +
  geom_line(aes(y = sc_25c), alpha = 0.3) +
  geom_line(aes(y = fitted), color = "firebrick") +
  facet_wrap(~external_station_code, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Specific conductance @ 25C (uS/cm)",
       subtitle = "Black = observed, red = fitted 24-hr harmonic") +
  theme_minimal()
```

### Autocorrelation structure

The autocorrelation function (ACF) at lag $k$,

$$
\rho_k = \frac{\operatorname{Cov}\!\left(SC_{25}(t),\, SC_{25}(t-k)\right)}{\operatorname{Var}\!\left(SC_{25}(t)\right)},
$$

quantifies how far into the future a current reading remains informative.
We look at it twice: at the logger's native ~5-minute cadence (short-lag
persistence), and after resampling to each candidate chemistry-sampling
interval (does a value from $k$ intervals ago still predict the current one
at that coarser resolution?).

```{r acf-native}
#| fig-cap: "ACF at native (~5-min) cadence, by logger"
for (site in unique(obs_clean$external_station_code)) {
  x <- obs_clean |> filter(external_station_code == site) |> pull(sc_25c)
  x <- x[!is.na(x)]
  acf(x, lag.max = 24 * 12, main = paste("ACF —", site, "(lags in 5-min steps, 12/hr)"))
}
```

```{r acf-resampled}
#| fig-cap: "ACF resampled to candidate chemistry-sampling intervals"
resample_to <- function(df, interval_hours) {
  df |>
    mutate(bin = floor((as.numeric(timestamp) - as.numeric(min(timestamp))) / (interval_hours * 3600))) |>
    group_by(bin) |>
    summarise(sc_25c = mean(sc_25c, na.rm = TRUE), .groups = "drop")
}

candidate_intervals_hr <- c(24, 48, 72, 168)  # daily, 2-day, 3-day, weekly

for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |> filter(external_station_code == site)
  for (h in candidate_intervals_hr) {
    rs <- resample_to(site_df, h)
    if (nrow(rs) > 10) {
      acf(rs$sc_25c, na.action = na.pass,
          main = paste0(site, " — resampled every ", h, "h (n=", nrow(rs), ")"))
    } else {
      message(site, " @ ", h, "h: too few points yet (n=", nrow(rs), ") for a meaningful ACF")
    }
  }
}
```

### Spectral analysis

The periodogram estimates the spectral density $\hat{S}(f)$ across
frequencies $f$, decomposing the series into the strength of cyclical
components at each period $1/f$; a peak at 24 hours would confirm the diurnal
signal seen above and reveal any other dominant periods (e.g. a semi-diurnal
component, or longer synoptic-scale cycles as record length grows).

```{r spectral}
#| fig-cap: "Periodogram (dominant frequencies), by logger"
for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |> filter(external_station_code == site, !is.na(sc_25c)) |> arrange(timestamp)

  # Native cadence is nominally 5 min; a plain periodogram assumes even
  # spacing. Any residual gaps after QC are small relative to record
  # length for now — revisit with Lomb-Scargle if gaps grow.
  sp <- spec.pgram(site_df$sc_25c, taper = 0.1, log = "no", plot = FALSE)

  period_hours <- (1 / sp$freq) * (5 / 60)  # 5-min steps -> hours

  plot(period_hours, sp$spec, type = "l", log = "x",
       xlab = "Period (hours)", ylab = "Spectral density",
       main = paste("Periodogram —", site))
  abline(v = 24, col = "firebrick", lty = 2)
  legend("topright", legend = "24-hr", col = "firebrick", lty = 2, bty = "n")
}
```

### Time-series decomposition

STL (Seasonal-Trend decomposition using Loess) additively decomposes the
series into trend, (diurnal) seasonal, and remainder components,

$$
SC_{25}(t) = T(t) + S(t) + R(t),
$$

which separates a slow drift (e.g. seasonal baseflow change) from the fast
diurnal cycle and from residual variability not explained by either — the
remainder $R(t)$ is a useful working estimate of the "noise floor" that any
chloride-prediction model built on $SC_{25}$ will ultimately be limited by.

```{r decomposition}
#| fig-cap: "STL decomposition (trend / diurnal / remainder), by logger"
for (site in unique(obs_clean$external_station_code)) {
  site_df <- obs_clean |>
    filter(external_station_code == site) |>
    arrange(timestamp)

  # Regularize to 5-minute grid via linear interpolation across small gaps
  full_grid <- seq(min(site_df$timestamp), max(site_df$timestamp), by = "5 min")
  interp <- approx(site_df$timestamp, site_df$sc_25c, xout = full_grid)$y

  ts_obj <- ts(interp, frequency = 288)  # 288 = 5-min steps per day

  if (sum(!is.na(interp)) > 2 * 288) {
    decomp <- stl(zoo::na.approx(ts_obj, na.rm = FALSE), s.window = "periodic", robust = TRUE)
    plot(decomp, main = paste("STL decomposition —", site))
  } else {
    message(site, ": record too short yet for a reliable STL decomposition (need > 2 full days).")
  }
}
```

**Interpretation so far:** both loggers show a diurnal signal consistent with
evapotranspiration/thermal-loading control on discharge and specific
conductance, and native-cadence autocorrelation is strong out to several
hours, as expected for a physically smooth process. Resampled (daily+) ACFs
and the STL trend component will only become informative claims — rather than
illustrations of the method — once several weeks to months of record have
accumulated; treat any daily/weekly-scale conclusions drawn today as
provisional.

## From conductivity to chloride: predictive-model framework

Once paired chloride/conductance samples exist, the goal is a model
$\hat{C}_{Cl} = f(SC_{25}, T, \text{season})$ that can be evaluated
continuously from the logger record between chemistry samples. Three
candidate model families are fit and compared
(`scripts/analysis/sampling_frequency/03_chloride_models.R`), all using
$SC_{25}$, stream temperature, and a first-harmonic seasonal term
$\left(\sin\frac{2\pi \cdot doy}{365}, \cos\frac{2\pi \cdot doy}{365}\right)$
as predictors:

$$
\begin{aligned}
\text{Linear:} \quad & C_{Cl} = \gamma_0 + \gamma_1 SC_{25} + \gamma_2 T + \gamma_3 \sin\theta_{doy} + \gamma_4 \cos\theta_{doy} + \varepsilon \\
\text{GAM:} \quad & C_{Cl} = \gamma_0 + g_1(SC_{25}) + g_2(T) + \gamma_3 \sin\theta_{doy} + \gamma_4 \cos\theta_{doy} + \varepsilon,\quad g_1, g_2 \text{ penalized splines} \\
\text{Random Forest:} \quad & C_{Cl} = h(SC_{25}, T, \sin\theta_{doy}, \cos\theta_{doy}) + \varepsilon,\quad h \text{ an ensemble of regression trees}
\end{aligned}
$$

where $\theta_{doy} = 2\pi \cdot \text{day-of-year}/365$. The linear model is
the interpretable baseline; the GAM allows nonlinear (but smooth,
interpretable) response curves; the random forest allows arbitrary
interaction structure at the cost of interpretability. All three are
evaluated identically with **rolling-origin (blocked) cross-validation** —
each fold trains only on data prior to the test window, which respects the
temporal autocorrelation established in the ACF analysis above and avoids the
optimistic bias that ordinary random k-fold CV would produce on an
autocorrelated series.

`get_chloride_conductance_pairs()` currently returns 0 rows for the reason
noted throughout this document (no temporal overlap between the 2026-05-01
chemistry samples and the 2026-07-15+ logger record). The block below
therefore demonstrates the exact same functions against
`simulate_synthetic_pairs()` — a synthetic dataset with a known, injected
linear relationship — purely to confirm the modeling code path is correct
end to end ahead of real data; **the numbers below are not a real
conductivity-chloride relationship for Steamboat Creek.**

```{r chloride-models-demo}
#| message: false
source(file.path("scripts", "analysis", "sampling_frequency", "03_chloride_models.R"))

synthetic_pairs <- simulate_synthetic_pairs()
cv_results <- evaluate_chloride_models(synthetic_pairs, initial = 20, assess = 5, skip = 5)

cv_results |>
  group_by(model) |>
  summarise(mean_rmse = mean(rmse), mean_mae = mean(mae), n_folds = n())
```

## Sampling-frequency design via Monte Carlo subsampling

Given a continuous reference chloride series (in practice, one of the fitted
models above applied to the full-resolution conductivity record), the
sampling-frequency question is answered by simulation
(`scripts/analysis/sampling_frequency/04_monte_carlo_subsampling.R`):
repeatedly subsample the reference at each candidate chemistry interval
$\Delta \in \{1, 2, 3, 7, 14, 30\}$ days with a randomized phase offset,
reconstruct the full series by linear interpolation between the subsampled
points, and compare the reconstruction to the (known) reference. Concentration
error uses

$$
\text{RMSE}(\Delta) = \sqrt{\frac{1}{n}\sum_{i=1}^n \left(C_{Cl,i} - \hat{C}_{Cl,i}^{(\Delta)}\right)^2}, \qquad
\text{MAE}(\Delta) = \frac{1}{n}\sum_{i=1}^n \left|C_{Cl,i} - \hat{C}_{Cl,i}^{(\Delta)}\right|
$$

and, when a paired discharge series $Q(t)$ is available, mass-flux error
compares the trapezoidal-rule mass flux (Section 2's $\Phi_{Cl}$, discretized)
computed from the reconstruction against the true flux:

$$
\Phi_{Cl}^{(\Delta)} \approx \sum_{i} \frac{\Delta t_i}{2}\Big(C_{Cl,i}Q_i + C_{Cl,i+1}Q_{i+1}\Big), \qquad
\text{FluxError\%}(\Delta) = 100 \times \frac{\Phi_{Cl}^{(\Delta)} - \Phi_{Cl}}{\Phi_{Cl}}
$$

Repeating this many times per $\Delta$ (random draw of the sampling phase)
and summarizing across draws (mean, 95th percentile) gives a distribution of
error at each candidate frequency rather than a single, phase-dependent
estimate. As with the previous section, this is demonstrated here on a
synthetic reference series (`simulate_synthetic_reference()`) with known
seasonal + diurnal structure, pending a real fitted reference series:

```{r monte-carlo-demo}
source(file.path("scripts", "analysis", "sampling_frequency", "04_monte_carlo_subsampling.R"))

synthetic_reference <- simulate_synthetic_reference()
mc_draws <- monte_carlo_sampling_frequency(synthetic_reference, n_draws = 100)
mc_summary <- summarize_sampling_frequency(mc_draws)
mc_summary
```

```{r monte-carlo-plot}
#| fig-cap: "Synthetic-data illustration: chloride RMSE vs. candidate sampling interval"
ggplot(mc_summary, aes(x = frequency_days, y = rmse_mean)) +
  geom_ribbon(aes(ymin = rmse_mean, ymax = rmse_p95), alpha = 0.15) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = mc_summary$frequency_days) +
  labs(
    x = "Candidate chemistry-sampling interval (days)",
    y = "Chloride RMSE (mg/L)\n(point = mean across MC draws, band = 95th pct)",
    title = "Synthetic self-test — NOT a real Steamboat Creek result"
  ) +
  theme_minimal()
```

As expected of any subsampling/interpolation scheme, error increases
monotonically (in expectation) with the sampling interval; the question this
framework is built to answer is *how quickly*, and whether an
operationally-realistic interval (e.g. weekly or biweekly) stays within an
error the PI considers acceptable for detecting a real change in chloride
discharge.

## Turning an error curve into an actionable recommendation

`05_recommendation_report.R` intentionally requires an **explicit,
user-supplied error threshold** — `rmse_threshold_mgL` and/or
`flux_pct_threshold` — rather than defaulting to a convention. There is no
statistically "correct" acceptable error here; it is a scientific decision
about what change in chloride discharge would be meaningful to detect, which
belongs with the PI, not with a default argument. Given a threshold, the
recommended interval is the **coarsest** (largest $\Delta$) candidate whose
mean error still satisfies it:

$$
\Delta^{*} = \max\left\{\Delta : \overline{\text{RMSE}}(\Delta) \le \tau_{RMSE} \ \text{and} \ \overline{|\text{FluxError\%}|}(\Delta) \le \tau_{flux}\right\}
$$

The block below applies this logic to the *synthetic* Monte Carlo summary
above purely to illustrate the mechanics — the threshold value (1 mg/L) is
an arbitrary illustrative choice, not a recommended one:

```{r recommendation-demo}
source(file.path("scripts", "analysis", "sampling_frequency", "05_recommendation_report.R"))

# Illustrative threshold only — agree the real threshold with the PI
# before using this on real chloride/conductance data.
illustrative_threshold_mgL <- 1
recommend_sampling_frequency(mc_summary, rmse_threshold_mgL = illustrative_threshold_mgL)
```

When run for real, `build_recommendation_report()` additionally writes the
summary table to `data/derived/sampling_frequency/` and the error-vs-frequency
plot to `output/figures/sampling_frequency/`; it is not called here to avoid
writing synthetic-data artifacts into those directories from this reference
document.

## Current status, gaps, and changelog

**Status of each step against real data:**

| Step | Real data available? | Blocking gap |
|---|---|---|
| Conductivity temporal structure (§4) | Yes | None — record is short (~weeks) but real |
| SC temperature-compensation recalibration (§3) | No | 0 field-vs-logger pairs; need field conductivity checks collected *during* the logger deployment |
| Chloride prediction models (§5) | No | 0 chloride-vs-conductance pairs; existing Cl samples (2026-05-01) predate logger deployment (2026-07-15) by ~2.5 months |
| Monte Carlo sampling-frequency design (§6) | No | Depends on §5 producing a real fitted reference series |
| Final recommendation (§7) | No | Depends on §6, plus an explicit threshold agreed with the PI |

**Immediate next steps:**

1. Collect $\geq 5$ field conductivity checks at `SBRR`/`SBGG` while loggers
   are active, to unblock §3.
2. Collect chloride (and ideally paired discharge) samples that temporally
   overlap the active logger deployment window, to unblock §5–7.
3. Run `scripts/analysis/sampling_frequency/run_sampling_frequency.R`'s
   `run_sampling_frequency_pipeline(con)` for a one-call status check across
   all four steps above (idempotent -- read-only unless `refit_models = TRUE`
   and real data support it). Once (2) is satisfied, also re-run this notebook — Sections 5–7 will
   automatically pull real pairs instead of falling through to synthetic
   self-tests, since `get_chloride_conductance_pairs()` and
   `get_calibration_pairs()` are the same functions used here and in the
   underlying scripts.
4. Agree an explicit $\tau_{RMSE}$ and/or $\tau_{flux}$ threshold with the PI
   before treating any output of §7 as an actual recommendation.

**Changelog:**

- 2026-09-05 — Added `scripts/analysis/sampling_frequency/run_sampling_frequency.R`
  as a single, idempotent entry point sequencing calibration -> chloride
  models -> Monte Carlo -> recommendation, mirroring `run_pipeline.R`'s style;
  each stage reports its status rather than requiring separate manual
  `source()`/function calls.
- 2026-09-05 — `SBRR`/`SBGG` logger role/serial assignment corrected (see
  provenance note in §2); confirmed the calibration join in §3 requires no
  code change as a result. Rewrote this notebook as the overarching
  reference document for the conductivity/chloride workstream, adding
  scientific motivation, explicit formulas alongside their implementing
  code, and worked (synthetic, clearly labeled) demonstrations of the
  chloride-model and Monte Carlo sections.
