---
title: "Specific Conductance vs. Discharge and Weather"
subtitle: "A growing, re-runnable comparison of Steamboat Creek SC against USGS discharge and regional precipitation"
date: today
format:
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose

This notebook is deliberately separate from
`01_conductivity_temporal_structure.qmd` (which characterizes SC's
internal temporal structure) and is meant to **grow** as more discharge
and weather data accumulate: re-running it against the live database is
how the comparison grows, rather than this being a one-off static figure.
It brings together three sources that, until now, had never been joined
in this project:

- `vw_sc_discharge_daily` — daily-mean specific conductance per logger
  (real, from the active `SBRR`/`SBGG` deployment).
- `vw_usgs_discharge` — continuous USGS discharge (currently **not yet
  populated** with live data — see status note below).
- `vw_weather_metric` — NOAA precipitation/temperature, unit-normalized
  (real, from the NOAA station exports in `data/raw/noaa/`).

::: {.callout-warning}
The NOAA stations ingested so far (Reno Airport, Galena, a Rhodes-Road-area
COOP gauge) are **regional**, not on-site at Steamboat Hills — treat any
precipitation comparison here as regional context, not a local
water-balance input, until an on-site rain gauge exists.
:::

```{r setup}
#| message: false
library(knitr)
opts_knit$set(root.dir = normalizePath(file.path(getwd(), "..")))
```

```{r connect}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
library(lubridate)

db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("database", "geochem_demo.sqlite")
con <- dbConnect(SQLite(), db_path, flags = SQLITE_RO)
```

## Load the three sources

```{r load}
sc_daily <- dbGetQuery(con, "SELECT * FROM vw_sc_discharge_daily") |>
  mutate(date = as.Date(date))

discharge_daily <- dbGetQuery(con, "
  SELECT date(datetime) AS date, station_id, AVG(discharge_m3_s) AS discharge_m3_s
  FROM vw_usgs_discharge
  GROUP BY date(datetime), station_id
") |>
  mutate(date = as.Date(date))

precip_daily <- dbGetQuery(con, "
  SELECT date, station_id, value_metric AS precip_mm
  FROM vw_weather_metric
  WHERE parameter = 'PRCP'
") |>
  mutate(date = as.Date(date))

dbDisconnect(con)

message(nrow(sc_daily), " SC-daily rows, ", nrow(discharge_daily),
        " discharge-daily rows, ", nrow(precip_daily), " precip-daily rows loaded.")
```

::: {.callout-note}
As of this run, `discharge_daily` has **`r nrow(discharge_daily)`** rows —
`ingest_usgs.R` has not yet been pointed at a real downloaded discharge
series for this database. The plot/correlation below runs the same either
way and will populate automatically once that ingest happens; nothing
here is hard-coded to "no discharge exists".
:::

## SC and precipitation over time

```{r sc-precip-plot}
#| fig-cap: "Daily-mean specific conductance (by logger) and regional daily precipitation"
precip_total <- precip_daily |>
  group_by(date) |>
  summarise(precip_mm = sum(precip_mm, na.rm = TRUE), .groups = "drop")

ggplot() +
  geom_col(data = precip_total, aes(x = date, y = precip_mm * 3), fill = "steelblue", alpha = 0.4) +
  geom_line(data = sc_daily, aes(x = date, y = sc_25c_mean, color = external_station_code)) +
  labs(
    x = NULL, y = "Specific conductance @ 25C (uS/cm)\n(blue bars = regional precip mm, scaled x3 for visibility)",
    color = "Logger"
  ) +
  theme_minimal()
```

## Overlap check and lagged correlation

A same-day correlation is the wrong question for a dilution/response
signal — precipitation response in a creek's SC is not necessarily
instantaneous. This computes the cross-correlation between total daily
regional precipitation and each logger's daily-mean SC across a range of
lags, but only if there's enough temporally-overlapping record to make
that a meaningful exercise rather than noise on a handful of days.

```{r overlap-check}
overlap_days <- sc_daily |>
  distinct(date) |>
  inner_join(precip_total, by = "date") |>
  nrow()

message("Days with both SC and regional precipitation data: ", overlap_days)

min_overlap_for_ccf <- 14
```

```{r ccf, eval=overlap_days >= min_overlap_for_ccf}
#| fig-cap: "Cross-correlation: daily precipitation vs. daily-mean SC, by logger"
joined <- sc_daily |>
  select(date, external_station_code, sc_25c_mean) |>
  inner_join(precip_total, by = "date") |>
  arrange(date)

for (site in unique(joined$external_station_code)) {
  site_df <- joined |> filter(external_station_code == site)
  ccf(site_df$precip_mm, site_df$sc_25c_mean, na.action = na.pass,
      main = paste("Precip -> SC cross-correlation —", site))
}
```

```{r ccf-skip, eval=overlap_days < min_overlap_for_ccf}
message(
  "Only ", overlap_days, " overlapping day(s) between SC and regional ",
  "precipitation records (need >= ", min_overlap_for_ccf, ") -- skipping ",
  "the cross-correlation as not yet meaningful. This will resolve itself ",
  "as more concurrent record accumulates; no code change needed."
)
```

## Status and next steps

- SC-vs-discharge is architecturally ready (`vw_sc_discharge_daily` joins
  cleanly to `vw_usgs_discharge` on date) but has **no live discharge
  data yet** to actually compare against — run `ingest_usgs.R` against a
  real downloaded discharge series to populate this.
- SC-vs-precipitation has real data on both sides but limited temporal
  overlap so far; the cross-correlation above will strengthen as the
  logger deployment continues and/or an on-site rain gauge is added.
- Temperature-logger-vs-SC correlation (springs/wells vs. creek
  conductivity, at physically distinct sites) is intentionally **not**
  in this notebook — see `notebooks/03_temperature_sc_correlation.qmd`.
