---
title: "Temperature (springs/wells) vs. Specific Conductance (creek) Correlation"
subtitle: "Exploratory, pairwise lagged cross-correlation across spatially distinct monitoring sites"
date: today
format:
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

This notebook correlates two **independently-collected** time series —
temperature at spring/well loggers and specific conductance at the two
creek conductivity loggers — to look for a relationship between thermal
and conductivity signals at spatially distinct points. This is
deliberately **not** a structural join into one table (unlike
`02_sc_discharge_weather.qmd`'s SC/discharge/precipitation view) — it's a
correlation analysis between two datasets that were never designed to be
merged.

::: {.callout-warning}
This is exploratory by construction: every temperature-logger site is
compared against every conductivity-logger site, which means several
pairwise comparisons are being run at once. No individual correlation
below should be read as "significant" without correcting for that — the
purpose here is to see which pairs look worth a closer, targeted look,
not to draw a conclusion from this notebook alone. One pair (`SBRR`) has
**both** a temperature logger and a conductivity logger at the same
physical location, which is the most directly interpretable comparison
in this set — treat cross-site pairs more cautiously than that one.

Also note: both temperature and specific conductance show a diurnal
cycle (see `01_conductivity_temporal_structure.qmd`). A raw
cross-correlation can pick up shared diurnal timing rather than a real
hydrologic link between the two signals — this is not addressed by
detrending below, so a strong correlation at short lags should be
interpreted with that in mind.
:::

```{r setup}
#| message: false
library(knitr)
opts_knit$set(root.dir = normalizePath(file.path(getwd(), "..")))
```

```{r connect}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
library(tidyr)

db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("database", "geochem_demo.sqlite")
con <- dbConnect(SQLite(), db_path, flags = SQLITE_RO)
```

## Load daily-mean series

```{r load}
temp_daily <- dbGetQuery(con, "
  SELECT
    l.external_station_code AS site,
    date(datetime(t.timestamp, 'unixepoch')) AS date,
    AVG(t.temperature) AS temp_c_mean
  FROM Temperature_Observations t
  JOIN Temperature_Loggers tl ON t.logger_id = tl.logger_id
  LEFT JOIN Locations l ON tl.location_id = l.location_id
  WHERE l.external_station_code IS NOT NULL
  GROUP BY l.external_station_code, date(datetime(t.timestamp, 'unixepoch'))
")

sc_daily <- dbGetQuery(con, "SELECT date, external_station_code AS site, sc_25c_mean FROM vw_sc_discharge_daily")

dbDisconnect(con)

temp_daily$date <- as.Date(temp_daily$date)
sc_daily$date <- as.Date(sc_daily$date)

temp_sites <- sort(unique(temp_daily$site))
sc_sites <- sort(unique(sc_daily$site))

message(length(temp_sites), " temperature site(s): ", paste(temp_sites, collapse = ", "))
message(length(sc_sites), " conductivity site(s): ", paste(sc_sites, collapse = ", "))
```

## Pairwise lagged cross-correlation

```{r pairwise-ccf}
#| fig-cap: "Cross-correlation, temperature site vs. conductivity site (only pairs with >= 14 overlapping days shown)"
min_overlap_days <- 14
pair_summary <- list()

for (t_site in temp_sites) {
  for (s_site in sc_sites) {

    joined <- temp_daily |>
      filter(site == t_site) |>
      select(date, temp_c_mean) |>
      inner_join(sc_daily |> filter(site == s_site) |> select(date, sc_25c_mean), by = "date") |>
      arrange(date)

    if (nrow(joined) < min_overlap_days) {
      pair_summary[[length(pair_summary) + 1]] <- data.frame(
        temp_site = t_site, sc_site = s_site, n_overlap_days = nrow(joined),
        max_abs_ccf = NA_real_, lag_at_max = NA_integer_
      )
      next
    }

    cc <- ccf(joined$temp_c_mean, joined$sc_25c_mean, plot = FALSE, na.action = na.pass)
    i_max <- which.max(abs(cc$acf))

    plot(cc, main = paste0(t_site, " (temp) vs ", s_site, " (SC), n=", nrow(joined), " days",
                            if (t_site == s_site) " [co-located]" else ""))

    pair_summary[[length(pair_summary) + 1]] <- data.frame(
      temp_site = t_site, sc_site = s_site, n_overlap_days = nrow(joined),
      max_abs_ccf = round(cc$acf[i_max], 3), lag_at_max = cc$lag[i_max]
    )
  }
}

pair_summary <- do.call(rbind, pair_summary)
```

## Summary table

```{r summary-table}
pair_summary |>
  arrange(desc(abs(max_abs_ccf)))
```

Read `lag_at_max` as days: a positive lag means temperature leads
(predicts) specific conductance by that many days; negative means SC
leads temperature. `max_abs_ccf` close to 0 or `NA` (insufficient
overlap) means no usable relationship was found for that pair yet.

## Status and next steps

- This ran on `r nrow(temp_daily)` temperature-daily rows across
  `r length(temp_sites)` site(s) and `r nrow(sc_daily)` SC-daily rows
  across `r length(sc_sites)` site(s) — re-run as the logger deployment
  continues to accumulate more overlapping record.
- The `SBRR` pair (temperature logger + conductivity logger co-located)
  is the one result here worth following up on directly if it shows a
  notable correlation — everything else is a genuinely exploratory,
  multiple-comparisons screen.
- If a pair does look promising on repeated runs, the sensible next step
  is a proper detrended/deseasonalized correlation (e.g. on STL
  remainder components from `01_conductivity_temporal_structure.qmd`)
  rather than the raw series used here, to rule out shared diurnal timing
  as the explanation.
