---
title: "Photo-Location Workflow"
subtitle: "How field photos become registered Locations and (optionally) field observations, and what stays uncertain along the way"
date: today
format:
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose

`scripts/ingest/ingest_image_locations.R` turns a folder of field photos
into registered `Locations` rows using EXIF GPS metadata, via `exiftool`.
This notebook is the documentation that script's header comments always
pointed to but didn't yet exist: **why** each step works the way it does,
what's assumed, and — per an explicit design goal — where uncertainty is
recorded rather than silently discarded.

As of 2026-09-06, Step 1 also scans video files (mp4/mov/m4v) for
embedded GPS the same way it already handled still images -- tested
against all 7 real 2025-2026 Lower Sinter Terrace event photos and
videos (see `website/timeline.Rmd`), which confirmed exiftool *can*
extract GPS from video containers when a device embeds it, but that
most drone/phone video checked so far simply carries no location
metadata at all. Everything below describing the two-step design
applies identically to video; a video without embedded GPS just means
Step 1 has nothing to extract, not an error, and it still becomes a
mapped point the same way any un-geotagged photo would: a coordinate
or existing station code assigned by hand in `image_location_map.csv`.

## Two deliberately separate steps

```{mermaid}
flowchart LR
  A["Photo dropped in\ndata/raw/images/image_drop/"] --> B["Step 1 (automatic)\nexiftool extracts GPS lat/lon/alt\n+ capture time + GPS accuracy"]
  B --> C["Photo_Location_Candidates\n(raw, append-only, keyed by filename)"]
  C --> D{"Row present in\nimage_location_map.csv?"}
  D -- "no" --> E["Sits in Candidates only --\nno Locations row yet"]
  D -- "yes, new code" --> F["Register new Location\n(coordinate_source = 'photo_gps')"]
  D -- "yes, existing code" --> G["Compare to registered coords --\nlog to Photo_Location_QC,\nNEVER overwrite"]
  D -- "+ observation_note" --> H["Log Field_Observations row\n(qualitative note, optional sample link)"]
```

**Step 1 is automatic and never touches `Locations`.** A photo's GPS tag
is a fact about the photo, not a judgment about which named site it
shows — extracting it requires no human input, so it always runs.

**Step 2 requires a human decision, recorded in
`data/raw/images/image_location_map.csv`, and is the only thing that ever
creates or compares against a `Locations` row.** A photo cannot know
whether it depicts a spring, a seep, or a well, and cannot know that
"this GPS fix" and "SBF_0001" are the same thing — those are always
supplied by whoever fills in the mapping CSV, never inferred.

## Why an existing site's coordinates are never overwritten

If a mapped filename's `external_station_code` already exists in
`Locations`, the photo's GPS coordinate is **compared**, not written —
logged to `Photo_Location_QC` with a haversine distance to the registered
coordinate, and a warning raised if that distance exceeds 50 m.

That threshold is a deliberately loose "something is probably wrong"
trigger, not a precision claim. Typical smartphone GPS horizontal error
is commonly cited around 5-15 m under open sky, and can be considerably
worse under tree canopy or near canyon walls — both common at this site.
50 m is well outside that normal noise floor, so a flag at that distance
is very unlikely to be a false alarm from ordinary GPS jitter.

::: {.callout-note collapse="true"}
### Worked example: the SBF_0001 / IMG_4571 case (2026-09-05)

Before this workflow existed, a photo (`IMG_4571.HEIC`) intended to
register a new site (`SBF_0002`) turned out to have GPS coordinates
identical, to many decimal places, to the *already-registered*
`SBF_0001` — almost certainly the same photo used to originally register
that site, not a distinct location. That was caught by a manual check at
the time. `Photo_Location_QC`'s comparison step (with the 50 m threshold
above) is exactly what would have caught this automatically, the moment
someone tried to map that filename to a *different* code than the one it
actually matches.
:::

## Coordinate uncertainty is recorded, not discarded

A phone-photo GPS fix and a professionally surveyed point are not
equally precise, but until this addition, `Locations` had no way to say
which was which. Two columns now carry that information for **every**
location, not just photo-derived ones, so any future consumer (e.g. the
planned potentiometric-surface interpolation, or a simple nearest-station
join) can check it:

- `coordinate_source` — e.g. `'photo_gps'`, or `NULL` for legacy/unknown-
  provenance rows (not backfilled to `'surveyed'` — silence here means
  "we don't actually know," which is more honest than guessing).
- `coordinate_uncertainty_m` — for `photo_gps` rows, the EXIF
  `GPSHPositioningError` tag when the device recorded one; otherwise a
  documented default of
  **`r knitr::inline_expr("DEFAULT_PHOTO_GPS_UNCERTAINTY_M")`** — 15 m —
  chosen as a conservative mid-point of the typical smartphone accuracy
  range cited above, not a measured value for any specific photo.

```{r constant-check}
#| echo: false
#| message: false
# Confirms the documented default above actually matches the constant
# used in ingest_image_locations.R, so this notebook can't silently drift
# out of sync with the code.
source(file.path("..", "scripts", "ingest", "ingest_image_locations.R"))
cat("DEFAULT_PHOTO_GPS_UNCERTAINTY_M =", DEFAULT_PHOTO_GPS_UNCERTAINTY_M, "meters\n")
```

**This is intentionally descriptive only, for now.** Recording
uncertainty is the first step; actually *using* it — e.g. weighting an
IDW interpolation by `1 / coordinate_uncertainty_m^2`, or excluding
low-confidence points above some threshold from a potentiometric surface
— is real additional design work that belongs with that interpolation
build itself (still design-only, per the plan in the 2026-09-05 project
memory), not invented ad hoc here. Writing the connection down here means
it isn't lost by the time that build happens.

## Sampling/observation linkage

A photo is often taken *because* something was sampled or observed, not
just to document a location. `image_location_map.csv` supports two
optional columns for this:

- `observation_note` — free text; when present and non-blank, a
  `Field_Observations` row is logged (qualitative, timestamped note tied
  to the resolved `location_id`).
- `linked_external_sample_id` — when present, resolves to a `sample_id`
  in `Samples` (via `external_sample_id`), so the note is tied to a
  specific sampling event, not just a place.

`Field_Observations` is deliberately separate from `Field_Measurements`,
which requires a numeric `value` — a note like "spring flowing, no
sample taken" or "new fumarole vent, sulfur smell" has nowhere sensible
to go in a numeric-measurement table.

## Current status

```{r status}
#| message: false
library(DBI); library(RSQLite); library(dplyr)
db_path <- file.path("..", "database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_demo.sqlite")
con <- dbConnect(SQLite(), db_path, flags = SQLITE_RO)

cat("Photos with extracted GPS:", dbGetQuery(con, "SELECT COUNT(*) n FROM Photo_Location_Candidates")$n, "\n")
cat("Locations registered from photo GPS:",
    dbGetQuery(con, "SELECT COUNT(*) n FROM Locations WHERE coordinate_source = 'photo_gps'")$n, "\n")
cat("Photo-vs-registered QC comparisons logged:", dbGetQuery(con, "SELECT COUNT(*) n FROM Photo_Location_QC")$n, "\n")
cat("Field observations logged:", dbGetQuery(con, "SELECT COUNT(*) n FROM Field_Observations")$n, "\n")

dbDisconnect(con)
```

As of this run, no real photos have been dropped into `image_drop/` yet —
all counts above are expected to be zero until that happens. Re-run this
notebook after a real batch of field photos and a filled-in
`image_location_map.csv` to see it populate.
