---
title: "Data Inventory & Well/Facility Flow Network"
format: html
---

## Purpose and Scope

This notebook is the living reference for two threads that, unlike the
conductivity/sampling-frequency workstream (`01_conductivity_temporal_structure.qmd`),
don't have a single home yet: **(1)** what data currently exists in the
database and over what time span, and **(2)** the production well -> port
-> injection well flow network (Dhakal et al. 2025) plus the raw NDWR
well-log records that back up individual well identities/construction.
Update this notebook, not just the underlying scripts, as both threads
evolve -- consistent with this project's convention for `01`.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
source_path <- file.path("scripts", "analysis", "data_availability.R")
if (!file.exists(source_path)) source_path <- file.path("..", "scripts", "analysis", "data_availability.R")
source(source_path)
```

## 1. Data Availability Through Time

`scripts/analysis/data_availability.R` computes, for every major dated
table, the earliest/latest record and count. It handles each table's own
date-storage convention -- and, notably, discovered that
`Sampling_Events.date` mixes **two different numeric epoch conventions**
within the same column (Excel-serial-from-1899-12-30 for recent rows,
Unix-epoch-days-from-1970-01-01 for ~723 older rows) -- a real,
previously-undiscovered data bug, not something introduced by this
notebook. `external_event_id` (e.g. `"SB10_1987-10-21"`) reliably encodes
the true date and is used as ground truth wherever available; only ~10
of 759 rows lack that suffix and still rely on the ambiguous numeric
column.

```{r availability-table}
avail <- compute_data_availability(con)
avail
```

```{r availability-plot}
#| fig-width: 10
#| fig-height: 5.5
plot_data_availability(avail)
```

**Reading this chart:** continuous logger records (temperature,
conductivity, live USGS discharge) all start in 2025-2026 -- this is a
genuinely young, high-frequency instrumentation deployment layered on
top of decades of sparser discrete sampling (chemistry events back to
1974, USGS grab-sample specific conductance 1971-2001, NOAA weather back
to 2001). The gap between the historic USGS SC record ending in 2001 and
today is real, not a data-loading artifact -- worth keeping in mind when
interpreting any long-term trend across that boundary.

**Known follow-up (not fixed here):** the root cause of the
`Sampling_Events.date` dual-epoch bug -- which ingest script wrote the
Unix-epoch-day rows -- has not been identified. `external_event_id`
appears to be a reliable independent record of the true date for nearly
all rows, but the underlying `date` column itself remains wrong for
roughly 95% of rows and should not be trusted directly by any other
script until traced and fixed at the source.

## 2. Well & Facility Flow Network

Production wells feed named commingling ports, which route to injection
wells (Dhakal et al. 2025, Figure 5). `well_role` classifies each well;
`vw_production_to_port` / `vw_port_to_injection` expose each leg of the
network independently since the two legs are confirmed at different
levels of confidence.

```{r well-roles}
dbGetQuery(con, "SELECT well_role, COUNT(*) n FROM Wells GROUP BY well_role ORDER BY n DESC")
```

```{r production-to-port}
dbGetQuery(con, "SELECT * FROM vw_production_to_port")
```

```{r port-to-injection}
dbGetQuery(con, "SELECT * FROM vw_port_to_injection")
```

Both legs are real (transcribed directly from the Dhakal figure), but
**not every well in either table yet has a coordinate** -- see the
GeoPackage's `production_to_port`/`port_to_injection` line layers, which
only draw a line where both endpoints are located. `PW-1/2/3`, `13-5R`,
`21B-5R`, `83B-6R`, and `23-33RD` are the main remaining coordinate gaps
as of this writing.

## 3. Well Logs: OCR Extraction & Identity Matching

12 NDWR "Well Driller's Report" PDFs were OCR'd (via `tesseract`, with a
labeled-field -> unlabeled-scan -> UTM-conversion fallback chain -- see
`scripts/ingest/helpers/parse_well_log_pdf.R`). 5 are confirmed
Gerlach-area wells from an unrelated Ormat project (excluded from
Steamboat analysis); the remaining 7 are genuinely in the Steamboat
field but **none has a confirmed named-well identity** -- matching by
coordinate proximity alone is not reliable in this densely-drilled
field, and matching by completion date is essential to rule out
false-positive proximity matches.

```{r well-log-candidates}
dbGetQuery(con, "
  SELECT log_number, latitude, longitude, completion_date_raw, notes
  FROM Well_Log_Documents
  WHERE well_id IS NOT NULL AND well_name_parsed IS NULL
    AND log_number NOT IN ('146238','146239','146240','146241','146403')
  ORDER BY log_number
")
```

**Worked example -- why proximity alone is misleading:** log `123802`
sits only ~19.5 m from NBMG's "Industrial Production Well 43-33," the
closest lat/lon match found across all 12 logs. But `123802`'s own NDWR
record shows it was completed **9/24/2015**, while 43-33 was drilled in
**2007** -- an 8-year gap that rules out the match despite the tight
spatial coincidence. Applying the same location-*and*-date cross-check
to the other 6 real Steamboat logs found one comparably tight lead --
log `27731` (completed 8/15/1986) sits 133 m from NBMG's "Injection Well
No. 3" (SBGeo, Inc., completion recorded 8/8/1986), a one-week date gap
-- but per-project judgment is that 1986-era Steamboat drilling moved
fast enough that a well finished a week apart is more likely a
genuinely distinct, separate well than the same well misdated. No
identity is written to `well_log_document_map.csv` for any of the 7
without a human confirming it first.

Because none of the 7 real logs cleared the bar for a confident identity
match, each is registered as its own **provisional** `Wells` row (see
`register_provisional_well_logs()` in `scripts/ingest/ingest_well_logs.R`)
so the real coordinate/depth/construction data is visible and mappable
rather than stranded in the staging table -- clearly named and flagged
as unconfirmed, not silently merged into any named well:

```{r provisional-wells}
dbGetQuery(con, "
  SELECT well_id, well_name, latitude, longitude, total_depth,
         top_perforation, bottom_perforation, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name LIKE 'Unidentified Well%'
  ORDER BY well_id
")
```

## 4. NDOM Well-Permit Cross-Validation

NDOM (Nevada Division of Minerals) supplied the official state permit
record for every Ormat Steamboat well -- an independent source from the
NBMG/ArcGIS-digitized coordinates used elsewhere in this project.
`ingest_ndom_wells.R` matches each of the 49 permitted wells to an
existing `Wells` row (exact name, a confirmed spelling variant, or an
existing alias), or registers it as a new provisional well when no match
exists.

```{r ndom-match-summary}
dbGetQuery(con, "
  SELECT match_method, COUNT(*) n
  FROM NDOM_Well_Records
  GROUP BY match_method
  ORDER BY n DESC
")
```

**Coordinate cross-validation:** where a well already had a coordinate
from an earlier source, NDOM's is compared against it -- disagreements
over 100 m are never auto-applied, only logged for manual review:

```{r ndom-discrepancies}
disc_path <- file.path("data", "derived", "ndom_coordinate_discrepancies.csv")
if (!file.exists(disc_path)) disc_path <- file.path("..", "data", "derived", "ndom_coordinate_discrepancies.csv")
if (file.exists(disc_path)) read.csv(disc_path) else cat("No discrepancy report found.")
```

These gaps are large relative to this project's usual ~50-75 m NBMG/
ArcGIS uncertainty and are worth a manual look -- NDOM is the official
permit-of-record coordinate.

**A real, pre-existing unit bug was found via this ingestion**:
`elevation_m` on both `Wells` and `Locations` had been storing raw feet
(inherited from `ingest_ndwr.R`) despite its name -- e.g. "Sky Tavern Ski
Resort" recorded as 7619, when Steamboat's true elevation is
~1450-1750 m. Fixed at the source (`ingest_ndwr.R` now converts on
insert) plus a one-time retroactive migration for already-populated
rows:

```{r elevation-check}
dbGetQuery(con, "
  SELECT 'Wells' AS tbl, MIN(elevation_m) min_m, MAX(elevation_m) max_m FROM Wells WHERE elevation_m IS NOT NULL
  UNION ALL
  SELECT 'Locations', MIN(elevation_m), MAX(elevation_m) FROM Locations WHERE elevation_m IS NOT NULL
")
```

**Literature cross-check:** reading Dhakal et al. (2025) Table 1
("Steamboat Production Well classification") directly confirmed that
`14-33` and `14A-33` are genuinely distinct wells (both listed
separately under Middle Steamboat) -- independent corroboration of the
decision not to merge them, and the basis for adding `14-33` to the
well/port network CSV (Galena 3, matching its sibling).

## 5. NDWR Spring/Stream Flow

Daily manual discharge readings (cfs) at NDWR-monitored spring/stream
sites -- first supplied 2026-09-12 for 4 sites on Whites Creek (Truckee
Meadows basin, part of this project's broader hydrologic context rather
than the Steamboat outflow system itself).

```{r stream-flow-sites}
dbGetQuery(con, "
  SELECT l.name, l.latitude, l.longitude, l.elevation_m,
         COUNT(*) n_obs, MIN(s.date) start_date, MAX(s.date) end_date
  FROM Stream_Flow_Observations s
  JOIN Locations l ON l.location_id = s.location_id
  GROUP BY l.location_id
  ORDER BY l.name
")
```

Two of the four sites share an identical rounded lat/lon (a diversion
structure and its "above"/"below" companion points) -- registration is
keyed on `external_station_code`, not `coord_key`, for exactly this
reason.

## Changelog

| Date | Change |
|---|---|
| 2026-09-05 | Notebook created. Data availability chart built for the first time (and the `Sampling_Events.date` dual-epoch bug discovered while building it). 7 Steamboat well logs cross-checked against NBMG by location *and* completion date together (previously only location had been checked); `123802` conclusively rejected as a match to `43-33`; `27731`/"Injection Well No. 3" flagged as the tightest remaining lead but judged more likely a distinct well. All 7 registered as provisional `Wells` rows. |
| 2026-09-12 | Added NDOM well-permit cross-validation section (49 wells, 6 coordinate discrepancies flagged, elevation_m feet-vs-meters bug found and fixed project-wide) and NDWR spring/stream flow section (4 Whites Creek sites, daily discharge back to 2017). |

```{r cleanup}
#| include: false
dbDisconnect(con)
```