---
title: "Data Inventory & Well/Facility Flow Network"
format: html
---

## Purpose and Scope

This notebook is the living reference for two threads that, unlike the
conductivity/sampling-frequency workstream (`01_conductivity_temporal_structure.qmd`),
don't have a single home yet: **(1)** what data currently exists in the
database and over what time span, and **(2)** the production well -> port
-> injection well flow network (Dhakal et al. 2025) plus the raw NDWR
well-log records that back up individual well identities/construction.
Update this notebook, not just the underlying scripts, as both threads
evolve -- consistent with this project's convention for `01`.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
source_path <- file.path("scripts", "analysis", "data_availability.R")
if (!file.exists(source_path)) source_path <- file.path("..", "scripts", "analysis", "data_availability.R")
source(source_path)
```

## 1. Data Availability Through Time

`scripts/analysis/data_availability.R` computes, for every major dated
table, the earliest/latest record and count. It handles each table's own
date-storage convention -- and, notably, discovered that
`Sampling_Events.date` mixes **two different numeric epoch conventions**
within the same column (Excel-serial-from-1899-12-30 for recent rows,
Unix-epoch-days-from-1970-01-01 for ~723 older rows) -- a real,
previously-undiscovered data bug, not something introduced by this
notebook. `external_event_id` (e.g. `"SB10_1987-10-21"`) reliably encodes
the true date and is used as ground truth wherever available; only ~10
of 759 rows lack that suffix and still rely on the ambiguous numeric
column.

```{r availability-table}
avail <- compute_data_availability(con)
avail
```

```{r availability-plot}
#| fig-width: 10
#| fig-height: 5.5
plot_data_availability(avail)
```

**Reading this chart:** continuous logger records (temperature,
conductivity, live USGS discharge) all start in 2025-2026 -- this is a
genuinely young, high-frequency instrumentation deployment layered on
top of decades of sparser discrete sampling (chemistry events back to
1974, USGS grab-sample specific conductance 1971-2001, NOAA weather back
to 2001). The gap between the historic USGS SC record ending in 2001 and
today is real, not a data-loading artifact -- worth keeping in mind when
interpreting any long-term trend across that boundary.

**Known follow-up (not fixed here):** the root cause of the
`Sampling_Events.date` dual-epoch bug -- which ingest script wrote the
Unix-epoch-day rows -- has not been identified. `external_event_id`
appears to be a reliable independent record of the true date for nearly
all rows, but the underlying `date` column itself remains wrong for
roughly 95% of rows and should not be trusted directly by any other
script until traced and fixed at the source.

## 2. Well & Facility Flow Network

Production wells feed named commingling ports, which route to injection
wells (Dhakal et al. 2025, Figure 5). `well_role` classifies each well;
`vw_production_to_port` / `vw_port_to_injection` expose each leg of the
network independently since the two legs are confirmed at different
levels of confidence.

```{r well-roles}
dbGetQuery(con, "SELECT well_role, COUNT(*) n FROM Wells GROUP BY well_role ORDER BY n DESC")
```

```{r production-to-port}
dbGetQuery(con, "SELECT * FROM vw_production_to_port")
```

```{r port-to-injection}
dbGetQuery(con, "SELECT * FROM vw_port_to_injection")
```

Both legs are real (transcribed directly from the Dhakal figure), but
**not every well in either table yet has a coordinate** -- see the
GeoPackage's `production_to_port`/`port_to_injection` line layers, which
only draw a line where both endpoints are located. `PW-1/2/3`, `13-5R`,
`21B-5R`, `83B-6R`, and `23-33RD` are the main remaining coordinate gaps
as of this writing.

## 3. Well Logs: OCR Extraction, Identity Matching & Status History

The well-log pipeline now spans 76 NDWR "Well Driller's Report" PDFs
across two batches: 12 processed in Sessions 9-11, and 64 more log
numbers (112 files -- most with a second, "(2)"-suffixed reformatted/
streamlined re-scan of the same log) added in Session 24 from NDWR's
public ArcGIS well-log export. Three tables now back this, not just one:

- **`Well_Log_Documents`** -- one row per log number (tracking both the
  primary scan and, when present, its "(2)" alternate on the same row),
  with OCR/cross-reference-derived location, depth, perforation,
  Section/Township/Range, type of work, proposed use, and hole/casing
  diameter.
- **`Well_Work_Events`** -- one row per document once promoted/
  registered, so a well's status history (new -> deepened -> abandoned,
  etc.) is a real time series instead of a single overwritten snapshot.
- **`Well_Lithology`** -- best-effort depth-interval descriptions staged
  from OCR text (`confidence = 'ocr_heuristic_unvalidated'`), joinable
  to perforation depth once a well is identified.

**Location extraction now has four fallback levels**, tried in order:
a labeled Latitude/Longitude field, an unlabeled whole-text scan, UTM
Easting/Northing conversion, and -- new in Session 24 -- a Bureau of
Land Management PLSS (Section/Township/Range) lookup via their public
CadNSDI REST service, used only when the first three all fail. When a
log has both a primary and "(2)" alternate scan, lat/lon specifically
merges by **extraction-confidence rank** (labeled field beats UTM beats
unlabeled scan beats PLSS), not by "primary always wins" -- a real bug
caught during testing (logs `48594`/`126301`: the primary scan's
unlabeled-scan guess landed far outside Nevada, while the "(2)" alt scan
had a properly labeled field).

```{r well-log-table-summary}
dbGetQuery(con, "
  SELECT
    (SELECT COUNT(*) FROM Well_Log_Documents) AS n_documents,
    (SELECT COUNT(*) FROM Well_Log_Documents WHERE match_method = 'duplicate_content_skipped') AS n_duplicate_content,
    (SELECT COUNT(*) FROM Well_Log_Documents WHERE plss_latlon_method = 'plss_section_centroid') AS n_plss_derived,
    (SELECT COUNT(*) FROM Well_Work_Events) AS n_work_events,
    (SELECT COUNT(*) FROM Well_Lithology) AS n_lithology_intervals,
    (SELECT COUNT(DISTINCT well_id) FROM Well_Work_Events WHERE work_type = 'abandonment') AS n_wells_flagged_abandoned
")
```

**Type of work and proposed use**, extracted per document, give a quick
sense of what this batch actually contains:

```{r well-log-work-types}
dbGetQuery(con, "
  SELECT work_type, proposed_use, COUNT(*) n
  FROM Well_Log_Documents
  GROUP BY work_type, proposed_use
  ORDER BY n DESC
")
```

**Well status for mapping (production/injection/monitoring/P&A):**
rather than overloading `Wells.well_role` (a well's *role* doesn't stop
being "injection" just because it's later plugged), `vw_wells_gis` adds a
derived `display_status` column: a well shows as "Plugged & Abandoned" if
it has any `Well_Work_Events` row with `work_type = 'abandonment'`,
otherwise it falls back to `well_role`. This is the field the website's
leaflet maps and the GeoPackage `wells` layer now color by.

```{r display-status-summary}
dbGetQuery(con, "SELECT display_status, COUNT(*) n FROM vw_wells_gis GROUP BY display_status ORDER BY n DESC")
```

**Important, explicitly documented limitation:** this only captures P&A
status where a `Well_Work_Events` row exists (i.e. a driller's report OCR
found "abandonment" as its type of work). Wells known to be P&A only from
earlier sessions' free-text prose (e.g. `64-32`, `IW-3` -- see AGENTS.md)
will NOT show as P&A under this logic unless a `Well_Work_Events` row is
added for them -- deliberately not guess-parsed from notes fields, to
avoid false positives.

**Worked example -- why proximity alone is misleading:** log `123802`
sits only ~19.5 m from NBMG's "Industrial Production Well 43-33," the
closest lat/lon match found across the original 12 logs. But `123802`'s
own NDWR record shows it was completed **9/24/2015**, while 43-33 was
drilled in **2007** -- an 8-year gap that rules out the match despite the
tight spatial coincidence. The same location-*and*-date discipline
applies to every new candidate from the larger batch --
`qc_well_log_matches()` now produces a standing, regenerable candidates
report (`data/derived/well_log_match_candidates.csv`) with a plain-
English **recommendation** per row (e.g. "download the NDWR "(2)" page",
"review as a close match", or "no confident match -- keep unresolved"),
replacing the earlier one-off conversation-driven writeup per batch:

```{r well-log-match-candidates}
cand_path <- file.path("data", "derived", "well_log_match_candidates.csv")
if (!file.exists(cand_path)) cand_path <- file.path("..", "data", "derived", "well_log_match_candidates.csv")
if (file.exists(cand_path)) read.csv(cand_path) else cat("No candidates report found.")
```

No identity is written to `well_log_document_map.csv` for any unmatched
log without a human confirming it first -- both remaining candidates
above are left unresolved deliberately (one, log `80672`, sits 4.5 km
from its nearest named well -- far too distant to treat as a match).

Because none of the newly-processed logs cleared the bar for a confident
identity match, each with a real coordinate is registered as its own
**provisional** `Wells` row (see `register_provisional_well_logs()` in
`scripts/ingest/ingest_well_logs.R`) so the real coordinate/depth/
construction data is visible and mappable rather than stranded in the
staging table -- clearly named and flagged as unconfirmed, not silently
merged into any named well:

```{r provisional-wells}
dbGetQuery(con, "
  SELECT well_id, well_name, latitude, longitude, total_depth,
         top_perforation, bottom_perforation, diameter_in, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name LIKE 'Unidentified Well%'
  ORDER BY well_id
")
```

## 4. NDOM Well-Permit Cross-Validation

NDOM (Nevada Division of Minerals) supplied the official state permit
record for every Ormat Steamboat well -- an independent source from the
NBMG/ArcGIS-digitized coordinates used elsewhere in this project.
`ingest_ndom_wells.R` matches each of the 49 permitted wells to an
existing `Wells` row (exact name, a confirmed spelling variant, or an
existing alias), or registers it as a new provisional well when no match
exists.

```{r ndom-match-summary}
dbGetQuery(con, "
  SELECT match_method, COUNT(*) n
  FROM NDOM_Well_Records
  GROUP BY match_method
  ORDER BY n DESC
")
```

**Coordinate cross-validation:** where a well already had a coordinate
from an earlier source, NDOM's is compared against it -- disagreements
over 100 m are never auto-applied, only logged for manual review:

```{r ndom-discrepancies}
disc_path <- file.path("data", "derived", "ndom_coordinate_discrepancies.csv")
if (!file.exists(disc_path)) disc_path <- file.path("..", "data", "derived", "ndom_coordinate_discrepancies.csv")
if (file.exists(disc_path)) read.csv(disc_path) else cat("No discrepancy report found.")
```

These gaps are large relative to this project's usual ~50-75 m NBMG/
ArcGIS uncertainty and are worth a manual look -- NDOM is the official
permit-of-record coordinate. As of 2026-09-12, the discrepancy report
also includes `existing_coordinate_source`/`existing_coordinate_
uncertainty_m` (all 6 current discrepancies trace back to
`nbmg_geothermal_wells`), so a reviewer can immediately see where the
currently-stored point came from without a separate lookup.
ArcGIS uncertainty and are worth a manual look -- NDOM is the official
permit-of-record coordinate.

**A real, pre-existing unit bug was found via this ingestion**:
`elevation_m` on both `Wells` and `Locations` had been storing raw feet
(inherited from `ingest_ndwr.R`) despite its name -- e.g. "Sky Tavern Ski
Resort" recorded as 7619, when Steamboat's true elevation is
~1450-1750 m. Fixed at the source (`ingest_ndwr.R` now converts on
insert) plus a one-time retroactive migration for already-populated
rows:

```{r elevation-check}
dbGetQuery(con, "
  SELECT 'Wells' AS tbl, MIN(elevation_m) min_m, MAX(elevation_m) max_m FROM Wells WHERE elevation_m IS NOT NULL
  UNION ALL
  SELECT 'Locations', MIN(elevation_m), MAX(elevation_m) FROM Locations WHERE elevation_m IS NOT NULL
")
```

**Literature cross-check:** reading Dhakal et al. (2025) Table 1
("Steamboat Production Well classification") directly confirmed that
`14-33` and `14A-33` are genuinely distinct wells (both listed
separately under Middle Steamboat) -- independent corroboration of the
decision not to merge them, and the basis for adding `14-33` to the
well/port network CSV (Galena 3, matching its sibling).

## 5. NDWR Spring/Stream Flow

Daily manual discharge readings (cfs) at NDWR-monitored spring/stream
sites -- first supplied 2026-09-12 for 4 sites on Whites Creek (Truckee
Meadows basin, part of this project's broader hydrologic context rather
than the Steamboat outflow system itself).

```{r stream-flow-sites}
dbGetQuery(con, "
  SELECT l.name, l.latitude, l.longitude, l.elevation_m,
         COUNT(*) n_obs, MIN(s.date) start_date, MAX(s.date) end_date
  FROM Stream_Flow_Observations s
  JOIN Locations l ON l.location_id = s.location_id
  GROUP BY l.location_id
  ORDER BY l.name
")
```

Two of the four sites share an identical rounded lat/lon (a diversion
structure and its "above"/"below" companion points) -- registration is
keyed on `external_station_code`, not `coord_key`, for exactly this
reason.

## Changelog

| Date | Change |
|---|---|
| 2026-09-05 | Notebook created. Data availability chart built for the first time (and the `Sampling_Events.date` dual-epoch bug discovered while building it). 7 Steamboat well logs cross-checked against NBMG by location *and* completion date together (previously only location had been checked); `123802` conclusively rejected as a match to `43-33`; `27731`/"Injection Well No. 3" flagged as the tightest remaining lead but judged more likely a distinct well. All 7 registered as provisional `Wells` rows. |
| 2026-09-12 | Added NDOM well-permit cross-validation section (49 wells, 6 coordinate discrepancies flagged, elevation_m feet-vs-meters bug found and fixed project-wide) and NDWR spring/stream flow section (4 Whites Creek sites, daily discharge back to 2017). |
| 2026-09-12 (session 24-25) | Session 24: 64 new NDWR log numbers (112 files) ingested -- PLSS lat/lon fallback, work-type/proposed-use/diameter extraction, primary/alt-scan merge-by-confidence, `Well_Work_Events` (well status history), `Well_Lithology` populated for the first time, repeatable match-candidates report with recommendations. Session 25: `146403`'s OCR longitude digit error corrected; derived `display_status` (adds Plugged & Abandoned) added to `vw_wells_gis`/website maps; `ndom_coordinate_discrepancies.csv` now includes the existing coordinate's source. |

```{r cleanup}
#| include: false
dbDisconnect(con)
```
