---
title: "PHREEQC Geochemical Modeling: Speciation, Saturation Indices, Geothermometry, and Mixing"
subtitle: "Overarching reference for the Steamboat Hills PHREEQC workstream"
date: today
format:
  html:
    code-fold: show
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

This notebook is the **living scientific reference** for PHREEQC geochemical
modeling at Steamboat Hills — it documents *why* each analysis is done, *what*
the underlying chemistry/thermodynamics are, and *where* the code that
implements it lives (`scripts/phreeqc/`), and it is intended to stay the entry
point for this workstream as more chemistry (real Cl end-members overlapping
the conductivity logger deployment, alkalinity titrations, well-log
identities) accumulates. It complements, but does not replace, the individual
R scripts; this document narrates and runs representative pieces of them so
the reasoning and the code stay in sync — following the same convention as
`01_conductivity_temporal_structure.qmd` for the Cl-sampling-frequency
workstream.

Sections below run in dependency order:

1. Scientific motivation — why PHREEQC, and what each analysis mode answers.
2. Architecture — SOLUTION-block generation, the thermodynamic-database
   registry, and how the external PHREEQC executable is invoked.
3. Building `PHREEQC_Solutions` from validated chemistry, and why coverage
   went from 8 to 233 samples.
4. Charge-balance QC — and the alkalinity data-quality story behind it.
5. Speciation and saturation indices (real data, runs today).
6. Geothermometry — activity-based Na/K and multicomponent temperature-sweep,
   cross-checked against each other.
7. Mixing, inverse modeling, and gas-phase equilibria (self-tested on
   synthetic end-members; real results pending overlapping Cl chemistry).
8. Known limitations, fixed bugs, and a running changelog.

::: {.callout-warning}
As of this writing, real two-end-member mixing and inverse modeling
(Section 7) cannot yet be run on real Steamboat samples: the only Cl/isotope
field samples collected so far (2026-05-01) predate the conductivity logger
deployment (2026-07-15) by roughly 2.5 months, and the downstream/outflow
proxy site (`SBGG`) has no chemistry at all yet. Section 7 therefore runs
against synthetic self-test data, clearly labeled as such. Sections 3–6 are
real and current as of 2026-09-06.
:::

## Scientific motivation

Three questions this project needs answered, in increasing order of how much
real data they currently have behind them:

1. **What is the current chemical/thermal state of each sampled water, and is
   it internally consistent?** — PHREEQC's aqueous speciation and
   charge-balance calculation answer this directly: given measured
   concentrations, pH, and temperature, what species (free ions, ion pairs,
   complexes) is the dissolved mass actually distributed across, and does the
   analysis close electrically within an acceptable tolerance? A poor charge
   balance is a red flag for the *analysis*, not the water.
2. **Is a thermal water in equilibrium with the minerals it's discharging
   through, and what does that imply about scaling risk and reservoir
   temperature?** — Saturation indices ($SI = \log_{10}(IAP/K_{sp})$, zero at
   equilibrium, positive supersaturated/scaling-prone, negative
   undersaturated) for calcite, quartz, and related phases answer this.
   Silica geothermometry in particular is the classic tool for estimating
   deep reservoir temperature from a *cooled, discharged* fluid, because
   silica dissolution/precipitation kinetics are slow enough that a
   discharge sample can retain a partial memory of higher-temperature
   equilibrium at depth.
3. **Where is the discharge chemistry coming from — how much is deep thermal
   upflow versus meteoric dilution, and does any mineral reaction happen
   along the way?** — This is the mixing/inverse-modeling question this
   thesis's Cl-discharge framing (`01_conductivity_temporal_structure.qmd`)
   is ultimately building toward: PHREEQC's `MIX` and `INVERSE_MODELING`
   keywords test two-end-member mixing hypotheses directly, with or without
   an accompanying mineral reaction. This is the piece still waiting on real
   overlapping chemistry (see the warning above).

## Architecture

```{mermaid}
flowchart LR
    A[Lab_Analyses + Field_Measurements + Isotope_Analyses] --> B["build_phreeqc_solutions(con)<br/>08_build_phreeqc_tables.R"]
    B --> C[(PHREEQC_Solutions)]
    C --> D["run_phreeqc_pipeline(con)<br/>09_run_phreeqc.R"]
    D -->|"system2()"| E[[PHREEQC 3 executable]]
    E --> D
    D --> F[(PHREEQC_Results)]
    D --> G[(PHREEQC_Run_Failures)]
    C --> H["run_phreeqc_temp_sweep(con)<br/>09_run_phreeqc.R"]
    H --> I[(PHREEQC_Temp_Sweep)]
    C --> J["run_mixing_fraction_analysis() /<br/>run_phreeqc_mixing()<br/>10_run_phreeqc_mixing.R"]
    C --> K["run_phreeqc_inverse()<br/>11_run_phreeqc_inverse.R"]
    C --> L["run_phreeqc_gas_phase()<br/>12_run_phreeqc_gas_phase.R"]
    F --> M["interpret_phreeqc_results()<br/>automated human-readable text"]
```

| Script | Role |
|---|---|
| `scripts/phreeqc/utils_phreeqc.R` | SOLUTION-block generation, `system2()` execution of the standalone PHREEQC 3 executable, SELECTED_OUTPUT parsing, the 6-database thermodynamic registry with auto-recommendation, fluid-type classification, automated interpretation text |
| `scripts/phreeqc/08_build_phreeqc_tables.R` | `build_phreeqc_solutions(con)`, `get_phreeqc_eligible(con)` |
| `scripts/phreeqc/09_run_phreeqc.R` | `run_phreeqc_pipeline()`, `run_phreeqc_multi()`, `run_phreeqc_temp_sweep()`, `calculate_activity_geothermometers()` |
| `scripts/phreeqc/10_run_phreeqc_mixing.R` | Two-end-member conservative mixing fractions + PHREEQC `MIX` blocks |
| `scripts/phreeqc/11_run_phreeqc_inverse.R` | PHREEQC `INVERSE_MODELING` (mixing + mineral mass transfer) |
| `scripts/phreeqc/12_run_phreeqc_gas_phase.R` | PHREEQC `GAS_PHASE` (CO2/H2S degassing equilibria) |
| `scripts/phreeqc/run_phreeqc_analysis.R` | Single entry point sequencing all five modes |
| `database/schema/08_phreeqc_schema.R` | `PHREEQC_Solutions`, `PHREEQC_Results`, `PHREEQC_Temp_Sweep`, `PHREEQC_Run_Failures`, and the mixing/inverse/gas-phase table families |

PHREEQC itself is an external executable (USGS PHREEQC 3), not an R package —
this project shells out to it via `system2()` rather than using an R binding,
matching the pattern of the IGNIS project this workstream was ported from.
`PHREEQC_EXE`/`PHREEQC_DB` environment variables override the default search
path (`C:/Program Files/USGS/phreeqc/...`).

Database selection is automatic by default (`phreeqc_db = "auto"`):
temperatures above 100°C route to the LLNL database (validated
high-temperature thermodynamic data to 300°C, standard for geothermal
work), high-TDS waters route to Pitzer or SIT, and everything else uses the
general-purpose default `phreeqc.dat`. If a batch fails to converge, the
pipeline automatically retries with fallback databases, then retries each
sample individually — a single problem sample never silently drops every
other sample in its batch. Convergence failures that persist across every
database tried are logged to `PHREEQC_Run_Failures`, not silently dropped.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)

# Several scripts/phreeqc/*.R functions (build_phreeqc_solutions() in
# particular, via database/schema/lab_analyte_map.R) source() other project
# files using paths relative to the project root, not this notebook's own
# directory -- Quarto's project-level execute-dir default renders .qmd files
# with their own directory (notebooks/) as the working directory, which
# breaks those internal source() calls. A plain setwd() here isn't enough:
# knitr's in_dir() wraps every chunk and restores the working directory
# in between them, so the change would silently revert before the next
# chunk runs -- opts_knit$set(root.dir = ...) is knitr's own supported way
# to persist a working-directory override across all chunks, matching
# scripts/run_pipeline.R's own root-relative-paths convention.
if (basename(getwd()) == "notebooks") {
  knitr::opts_knit$set(root.dir = normalizePath(".."))
  setwd("..")
}

con <- dbConnect(SQLite(), file.path("database", "geochem_operational.sqlite"))
dbExecute(con, "PRAGMA foreign_keys = ON;")

for (f in c("utils_phreeqc.R", "08_build_phreeqc_tables.R", "09_run_phreeqc.R",
            "10_run_phreeqc_mixing.R", "11_run_phreeqc_inverse.R",
            "12_run_phreeqc_gas_phase.R", "13_gas_geothermometry.R",
            "14_gas_mixing.R", "run_phreeqc_analysis.R")) {
  source(file.path("scripts", "phreeqc", f))
}

phreeqc_available <- tryCatch({ check_phreeqc(); TRUE }, error = function(e) FALSE)
```

`phreeqc_available` is `r phreeqc_available` on the machine this notebook was
last rendered on — every live PHREEQC execution below is skipped gracefully
(not silently faked) if the executable isn't found.

## Building `PHREEQC_Solutions`

`build_phreeqc_solutions(con)` derives one row per sample from
`Lab_Analyses`, `Field_Measurements`, and `Isotope_Analyses` — temperature,
pH, and each major/minor ion, plus a real charge balance computed from
`lab_analyte_map.R`'s per-analyte charge and molar mass (not a hardcoded,
partial equivalent-weight table). A sample is *eligible* for a full PHREEQC
run if it has temperature, pH, Na, and Cl at minimum; everything else is
*rejected* with a specific, human-readable reason via `get_phreeqc_eligible()`
— never silently dropped.

```{r build-solutions}
sol <- build_phreeqc_solutions(con)
eligibility <- get_phreeqc_eligible(con)
cat(sprintf("PHREEQC_Solutions: %d rows total, %d eligible for a full run, %d rejected\n",
            nrow(sol), nrow(eligibility$eligible), nrow(eligibility$rejected)))
```

```{r eligibility-by-source}
loc_info <- dbGetQuery(con, "
  SELECT s.sample_id, s.data_source, l.name AS location_name, l.site_type
  FROM Samples s JOIN Locations l ON s.location_id = l.location_id
")
sol |>
  left_join(loc_info, by = "sample_id") |>
  count(data_source, site_type, completeness_flag) |>
  arrange(data_source, site_type)
```

::: {.callout-note collapse="true"}
### Why eligible coverage jumped from 8 to 233 samples (2026-09-06)

The first working version of this pipeline found only 8 eligible samples
database-wide — all from `FIELD` (springs/wells/fumarole/transect). NDEP's
~755 background samples and ~10 NDEP PRR wells all looked like they had zero
pH/temperature data. Investigating why (because the user specifically asked
to run this on existing NDEP data) surfaced three real ingest bugs, not a
genuine data gap:

1. **NDEP's pH and temperature are stored in `Lab_Analyses`, not
   `Field_Measurements`** — `build_phreeqc_solutions()` originally only
   checked the latter. Fixed to pull from both, preferring
   `Field_Measurements` when both exist.
2. **NDEP's minor/trace analytes are stored in mixed units** (the same
   analyte, e.g. `SiO2`/`As`/`B`/`Fe`/`Li`/`Mn`, appears with both `ug/L` and
   `mg/L` rows for different samples) with no ingest-time normalization —
   previously averaged naively, a ~1000x risk for whichever rows were
   "wrong" relative to others in the average. Fixed to convert every row to
   mg/L, per its own `units` column, before any aggregation.
3. **NDEP's authoritative alkalinity parameter, "Total Alkalinity as
   CaCO3", was never ingested at all.** `ndep_analyte_map.R` instead mapped
   two *unit-convention duplicates* of the same measurement ("HCO3 as
   CaCO3"/"HCO3 as HCO3", confirmed via the raw source file to be the exact
   same 647 events, ratio consistently 0.82 — matching the true
   CaCO3/HCO3 equivalent-weight ratio) to a single `HCO3` analyte code,
   with no conversion factor. Added the missing "Total Alkalinity"
   mapping (with the correct CaCO3 -> HCO3 mass conversion,
   $61.017/50.05 = 1.2189$), and remapped the duplicate-convention rows to
   distinct, explicitly-excluded codes so they can never collide again.

A fourth, separate bug in `lab_analyte_map.R` — the `Alkalinity` row had
`charge = 0, molar_mass = NA`, silently excluding it from every
charge-balance calculation even after real data existed — meant the fix
above didn't move the charge-balance numbers at all until this was also
corrected (see the next section).

**A fifth, independent instance of the same class of bug** was found in
`ingest_lab.R` (FIELD's own lab-report parser): `normalize_lab_wide()` ran
raw CSV headers through `make.names()` for R-safe column names, which turns
`"Alkalinity Total"` into `"Alkalinity.Total"` (space $\to$ dot) — and then
used that *mangled* name as the join key against `lab_analyte_map$raw_name`
(`"Alkalinity Total"`, with a space), guaranteed to never match. Every FIELD
sample's Alkalinity (and NO3, the only other multi-word raw name) had been
silently dropped since this pipeline's very first field-lab ingest, not
missing due to a genuine sampling gap as earlier sessions assumed. Fixed to
preserve the original header text as the actual join key.

See `AGENTS.md` (Sessions 15–16) for the full narrative, including the raw
per-row evidence for each fix.
:::

## Charge-balance QC

```{r charge-balance}
complete_sol <- sol |>
  left_join(loc_info, by = "sample_id") |>
  filter(completeness_flag == "complete", !is.na(charge_balance))

complete_sol |>
  mutate(qc_flag = ifelse(abs(charge_balance) > 5, "FLAG (>5%)", "OK (<=5%)")) |>
  count(data_source, qc_flag)
```

```{r charge-balance-hist}
#| fig-width: 7
#| fig-height: 4
ggplot(complete_sol, aes(x = charge_balance)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = c(-5, 5), linetype = "dashed") +
  labs(x = "Charge balance error, 100*(cations-anions)/(cations+anions)  [%]",
       y = "Count", title = "Charge balance across all PHREEQC-complete samples")
```

The dashed lines mark the conventional ±5% acceptability threshold. Before
the alkalinity fixes above, essentially every complete sample failed this
check with a median imbalance around 60% — a pure artifact of excluding
alkalinity (typically the dominant anion in these near-neutral waters) from
the anion sum entirely, not a real geochemical signal. After the fixes, the
median imbalance for NDEP's complete samples is close to zero and most pass
the standard threshold; FIELD's own samples are now fully alkalinity-complete
too (see the callout above).

## Speciation and saturation indices

```{r run-speciation}
#| eval: !expr phreeqc_available
speciation_result <- run_phreeqc_pipeline(con)
```

```{r speciation-summary}
#| eval: !expr phreeqc_available
si_wide <- dbGetQuery(con, "
  SELECT pr.sample_id, l.name AS location_name, l.site_type, pr.parameter, pr.value, pr.db_file
  FROM PHREEQC_Results pr
  JOIN Samples s ON pr.sample_id = s.sample_id
  JOIN Locations l ON s.location_id = l.location_id
  WHERE pr.parameter LIKE 'SI_%'
") |>
  filter(value > -100) |>
  tidyr::pivot_wider(names_from = parameter, values_from = value,
                      id_cols = c(sample_id, location_name, site_type, db_file))

si_wide |> filter(site_type != "background") |> arrange(sample_id)
```

```{r interpretation}
#| eval: !expr phreeqc_available
#| results: asis
cat(interpret_phreeqc_results(con))
```

The automated interpretation above is generated directly from
`PHREEQC_Results` by `interpret_phreeqc_results()` — it is not hand-written
prose, so it stays current every time this notebook (or the pipeline) is
re-run. As of 2026-09-06 it correctly identifies calcite-supersaturated
samples (scaling-relevant) among both the NDEP background set and the real
thermal FIELD samples, and flags the 8 genuinely thermal samples as
approaching quartz equilibrium — see the geothermometry section below for
why that flag needs a caveat.

::: {.callout-important collapse="true"}
### Na/K geothermometry is only physically meaningful for genuinely thermal samples

`interpret_phreeqc_results()`'s SI-based read is database-agnostic and
correct for every sample it covers. But applying **activity-based Na/K
geothermometry** (next section) to NDEP's ~225 dilute, cold (median ~11°C)
background creek/well samples gives geothermometer estimates of 300–390°C —
not a real finding, a violation of the method's core assumption
(water-rock equilibrium, which doesn't hold for shallow, dilute, fast-flowing
surface water). Only the 8 real FIELD thermal samples (measured temperature
59–162°C) are appropriate inputs for Na/K geothermometry; this notebook
restricts that analysis to them explicitly rather than reporting the
background numbers at face value.
:::

## Geothermometry

Two independent geothermometers are computed on the real thermal FIELD
samples: **activity-based Na/K** (Giggenbach 1988; Fournier 1979), computed
directly from PHREEQC's speciated ion activities rather than raw
concentrations — important because these are moderately saline waters where
activity and concentration diverge — and a **multicomponent
mineral-equilibration temperature sweep** (Reed & Spycher 1984 approach):
run each SOLUTION across a temperature ramp (25–300°C by 25°C) and find where
each mineral's saturation index crosses zero.

```{r geothermometry-ids}
thermal_ids <- loc_info |>
  filter(data_source == "FIELD", site_type %in% c("well", "spring", "seep", "fumarole", "transect")) |>
  semi_join(sol |> filter(completeness_flag == "complete"), by = "sample_id") |>
  pull(sample_id)
thermal_ids
```

```{r nak-geothermometer}
#| eval: !expr phreeqc_available
geo_nak <- calculate_activity_geothermometers(con) |>
  filter(sample_id %in% thermal_ids) |>
  select(sample_id, location_name, T_NaK_gigg_activity, T_NaK_four_activity)
geo_nak
```

```{r temp-sweep}
#| eval: !expr phreeqc_available
sweep_res <- run_phreeqc_temp_sweep(con, sample_ids = thermal_ids,
                                     temp_range = c(25, 300), temp_step = 25,
                                     phreeqc_db = "llnl")
```

```{r geothermometer-comparison}
#| eval: !expr phreeqc_available
quartz_chal <- sweep_res$eq_temps |>
  filter(mineral %in% c("Quartz", "Chalcedony")) |>
  select(sample_id, mineral, eq_temperature_C) |>
  tidyr::pivot_wider(names_from = mineral, values_from = eq_temperature_C, names_prefix = "T_")

comparison <- sol |>
  filter(sample_id %in% thermal_ids) |>
  select(sample_id, T_measured = temperature) |>
  left_join(quartz_chal, by = "sample_id") |>
  left_join(geo_nak, by = "sample_id") |>
  mutate(across(where(is.numeric), ~round(., 0)))
comparison
```

```{r geothermometer-plot}
#| eval: !expr phreeqc_available
#| fig-width: 7
#| fig-height: 5
comparison |>
  tidyr::pivot_longer(cols = c(T_measured, T_Quartz, T_Chalcedony, T_NaK_gigg_activity),
                       names_to = "series", values_to = "temperature_C") |>
  ggplot(aes(x = factor(sample_id), y = temperature_C, color = series, group = series)) +
  geom_point(size = 2.5) +
  geom_line(aes(group = series)) +
  labs(x = "sample_id", y = "Temperature (\u00B0C)",
       title = "Measured vs. geothermometer-estimated temperature, real thermal samples",
       color = NULL) +
  theme(legend.position = "bottom")
```

**Reading this comparison**: quartz/chalcedony equilibration temperatures
come out systematically *below* measured discharge temperature for every
sample, while Na/K comes out systematically *above* it. This divergence is
geologically informative, not contradictory: Na/K equilibrates deep and
slowly, retaining a "memory" of a hotter reservoir that persists through
cooling en route to the surface, while silica solubility can re-equilibrate
faster (or the fluid can dilute with a lower-silica meteoric component)
before discharge. A real two-end-member mixing/inverse model (Section 7)
would be the natural next step to formally test this hypothesis once real
overlapping end-member chemistry exists — this divergence is exactly the
kind of independent evidence that motivates doing so, not a discrepancy that
needs to be resolved away.

::: {.callout-note collapse="true"}
### A genuine PHREEQC convergence failure, diagnosed and fixed (2026-09-06)

One real thermal sample (`SBW_0002`) initially failed to converge with
*every* registered thermodynamic database, all failing identically with
`"Alkalinity has not converged... Is non-carbonate alkalinity greater than
total alkalinity?"`. Its pH (9.0) sits close to boric acid's pKa (~9.24), so
a large fraction of its Boron (37.1 mg/L — not even unusually high in
absolute terms) ionizes to alkalinity-contributing borate, pushing
PHREEQC's internally-computed total alkalinity ~26% above the lab-measured
value. `format_solution_block()` already had a boron-exclusion guard for
exactly this failure mode, but it used a crude `B (mg/L) > 0.5 * Alkalinity
(mg/L)` ratio that missed this case (37.1 vs. a threshold of 160.9). Replaced
with a physically-grounded check — estimate borate's charge-equivalent
contribution at the sample's actual pH via Henderson-Hasselbalch
(pKa = 9.24, a 25°C screening approximation) and exclude boron when that
estimate exceeds 15% of specified total Alkalinity — verified against all 81
complete real samples carrying both B and Alkalinity: only this one sample
crosses the new threshold (23.8%; next-highest is 7.0%). `PHREEQC_Run_Failures`
is empty database-wide as of this fix.
:::

## Mixing, inverse modeling, and gas-phase equilibria

These three modes are built and pass synthetic self-tests, but — per the
warning at the top of this notebook — real chemistry doesn't yet overlap the
conductivity logger deployment closely enough to name confident real thermal
and meteoric end-members. They are deliberately **not** run automatically by
`run_pipeline.R`; each requires the caller to name real `sample_id`s
explicitly (never inferred), matching this project's standing rule against
guessing which samples are meaningful end-members.

```{r self-tests}
#| eval: !expr phreeqc_available
demo_results <- demo_phreeqc_analysis()
```

### Two-end-member mixing (`10_run_phreeqc_mixing.R`)

`compute_mixing_fractions()` solves the conservative linear-mixing equation
for chloride (or any other tracer) directly:

$$
f_{\text{thermal}} = \frac{C_{\text{sample}} - C_{\text{meteoric}}}{C_{\text{thermal}} - C_{\text{meteoric}}}
$$

`run_phreeqc_mixing()` then builds a PHREEQC `MIX` block combining the two
end-member `SOLUTION`s at that fraction and compares PHREEQC's predicted full
speciation/SI (which accounts for real, non-conservative effects like a
saturation-index shift on mixing/cooling) against the target sample's
observed chemistry — a genuine fit-quality check on the two-end-member
hypothesis, not just an arithmetic fraction.

```{r mixing-demo-result}
demo_results$mixing[c("true_fraction", "recovered_fraction", "pass")]
```

**Usage once real end-members are available**:

```r
run_phreeqc_analysis_pipeline(con, mode = "mixing",
  thermal_sample_id = <real sample_id>, meteoric_sample_id = <real sample_id>,
  tracer = "Cl")
```

### Inverse modeling (`11_run_phreeqc_inverse.R`)

PHREEQC's `INVERSE_MODELING` is the more rigorous complement to `MIX`
blocks: it solves *simultaneously* for end-member mixing fractions **and**
mineral mass transfer (e.g. calcite dissolution/precipitation along the
flow path) within specified elemental uncertainty limits, so it can test
whether mixing *plus* some reaction fits a target sample better than
conservative mixing alone.

```{r inverse-demo-result}
demo_results$inverse[c("syntax_ok")]
```

::: {.callout-note collapse="true"}
### A documented PHREEQC numerical edge case in the synthetic self-test

The synthetic 2-end-member + phase scenario runs to completion (exit code 0,
after adding an explicit `-balances` sub-block for the conservative tracers
Na/Cl/K/Mg — without it, PHREEQC fails outright with "Not possible to
balance solution N with input uncertainties" even when every input
solution's own charge balance is fine) but finds 0 feasible mixing+phase
solutions, due to a ~1e-9 numerical residual on an unused methane/C(-4)
redox couple ("equality not satisfied for C(-4)"). This is a known PHREEQC
quirk, not a flaw in this project's input generation — reported honestly as
a syntax/execution smoke test, not a numerical-accuracy proof.
:::

### Gas-phase equilibria (`12_run_phreeqc_gas_phase.R`)

Models CO2/H2S degassing along the ascent-to-discharge path via PHREEQC's
`GAS_PHASE` keyword — relevant because boiling/degassing measurably shifts
pH and carbonate speciation between deep upflow and surface discharge, a
real potential confound for the conductivity-as-Cl-proxy workstream (specific
conductance responds to total ionic strength, not just Cl).

```{r gas-phase-demo-result}
demo_results$gas_phase[c("syntax_ok")]
```

A `fixed_pressure` gas phase starting from zero gas moles only exchanges gas
once the solution's own fugacity reaches the target pressure — the self-test
uses an atmospheric CO2 partial pressure ($10^{-3.5}$ atm) rather than 1 atm
pure CO2 for exactly this reason, and visibly raises the synthetic solution's
pH from 6.0 to ~9 on degassing.

## Gas geothermometry and gas mixing (2026-09-06, built ahead of real data)

A UCSB gas-geochemistry sampling campaign (Tobias Fischer) is planned at the
Lower Sinter Terrace geysering fissure and nearby fumaroles: free-gas mass
spectrometry (CO2, H2S, and the rest of the major gas suite) plus He
isotopes and other tracers still to be scoped once real analyses are in
hand. No real gas-sample data exists yet in this database. Per an explicit
"make aware of direction, hold off on schema" instruction, this section
documents two standalone capability modules built and validated against
**real, cited literature data** rather than fabricated numbers -- neither
creates a database table or is wired into `run_pipeline.R`.

### The literature anchor

`docs/literature/"Geochem Data and concept model Mariner & Janik 1995.pdf"`
(GRC Transactions v.19) is this project's only Steamboat-specific
gas-geochemistry reference. Its Table 2 reports real free-gas mole-percent
compositions (CO2, H2S, H2, CH4, NH3, N2, O2, Ar, He) and three
gas-geothermometer temperature columns for the Caithness/Far West wells and
one real fumarole sample (at PW3-4) -- exactly the kind of data Fischer's
campaign will eventually produce.

### D'Amore & Panichi (1980) gas geothermometer

`dap_gas_geothermometer()` (`13_gas_geothermometry.R`) implements the
CO2-H2-H2S-CH4 multicomponent geothermometer:

$$
t(^{\circ}C) = \frac{24775}{\alpha + \beta + 36.05} - 273.15$$

$$\alpha = 2\log_{10}\left(\frac{CH_4}{CO_2}\right) - 6\log_{10}\left(\frac{H_2}{CO_2}\right) - 3\log_{10}\left(\frac{H_2S}{CO_2}\right), \qquad \beta = -7\log_{10}(p_{CO_2})$$

::: {.callout-important collapse="true"}
#### A transcription risk caught by validating against real numbers, not just syntax

Several secondary sources (paper abstracts, review articles) transcribe
alpha's H2 term as a bare `-log10(H2/CO2)` (coefficient 1). Only
coefficient **6** (per Powell 2000, a Stanford GRC short-course review)
reproduces Mariner & Janik's own published T-DP values for real well 23-5
samples -- coefficient 1 overshoots by more than 150°C. This is exactly the
same class of risk as the SO4/`S(6)` PHREEQC element-name bug found earlier
this session: a formula that "looks right" syntactically can still be
numerically wrong. `demo_gas_geothermometry()` below is a literature
cross-check, not a synthetic syntax test.
:::

```{r gas-geothermometer-demo}
gas_geotherm_result <- demo_gas_geothermometry()
gas_geotherm_result
```

Computed T-DP matches Mariner & Janik's own published values to within
0.3°C across all four real 23-5 sampling dates. The other two Table 2
columns (T-HA, attributed to Giggenbach & Goguel 1989; T-CO2, attributed to
Arnórsson & Gunnlaugsson 1985) were deliberately **not** implemented -- only
qualitative descriptions of both could be found, not a citable calibration
verifiable against real numbers the way D'Amore-Panichi's was. Flagged as a
known gap, not silently skipped.

### Forward and inverse gas mixing

`forward_gas_mixing()`/`compute_gas_mixing_fraction()`
(`14_gas_mixing.R`) are the gas-phase analogue of
`compute_mixing_fractions()`'s water-Cl model above, using a conservative
gas ratio (default N2/Ar -- the same axis as Mariner & Janik's Fig. 9
N2-He-Ar plot, used there to assess air dilution of deep thermal gas)
instead of a dissolved-ion concentration.

::: {.callout-important collapse="true"}
#### A ratio does not mix linearly -- caught by the self-test, not assumed away

An earlier version of `compute_gas_mixing_fraction()` reused the water-Cl
model's linear inversion formula directly on the N2/Ar ratio. This is wrong:
a ratio of two linearly-mixing quantities is a *hyperbolic*, not linear,
function of the mixing fraction. The self-test below (forward-mix at a known
fraction, then invert and check the fraction is recovered) caught this
immediately -- it recovered 0.10 instead of the true 0.85. Fixed by solving
the correct hyperbolic relation for the mixing fraction directly from each
end-member's numerator/denominator components, not a pre-collapsed ratio.
:::

```{r gas-mixing-demo}
gas_mixing_result <- demo_gas_mixing()
gas_mixing_result[c("true_fraction", "recovered_fraction", "pass")]
```

The self-test uses well 23-5's real gas composition (11/5/91) as the
"thermal" end-member and standard dry-air composition (NIST/CIPM constants)
as the "meteoric/atmospheric" end-member -- a real sampling artifact
(air contamination) this project will need to screen for once Fischer's
samples arrive.

### Where real Fischer/UCSB data would plug in

Once real gas-sample data (composition, He isotopes, etc.) is in hand, the
next step is a `Gas_Samples`/`Gas_Analyses` schema designed around its
*actual* reporting format -- deliberately not designed now, consistent
with this project's standing rule against building storage for a data shape
that hasn't been seen yet.

## Known limitations and status (2026-09-06)

- **Real mixing/inverse modeling is blocked on field data, not code**: the
  single biggest next step for this workstream is chloride sampling that
  overlaps the conductivity logger deployment window closely enough to name
  confident real thermal and meteoric end-members.
- **`parse_phreeqc_inverse_output()`** is a best-effort, line-anchored regex
  parser of PHREEQC's free-text inverse-modeling output (PHREEQC does not
  punch inverse-model solutions to `SELECTED_OUTPUT`) — always cross-check
  the raw `.pqo` file for anything decision-critical.
- **`parse_phreeqc_gas_output()`**'s moles/pressure columns are gas-phase
  *totals*, exactly correct only for a single-gas-component phase.
- **FIELD's historical Alkalinity values are correctly ingested as of this
  session**, but the underlying convention (`mg/L CaCO3eq`, confirmed
  directly from the raw lab report's units row) has only been spot-checked
  for the one lab report reviewed so far — worth re-confirming if a
  differently-formatted lab report is ever added.
- **`run_phreeqc_pipeline()`'s printed "Result rows stored" summary** is
  approximate (not an exact DB count) specifically for batches that succeed
  via a fallback database rather than the first-choice one — cosmetic only,
  the real per-batch "Stored N rows" messages and the final DB row count are
  both correct.

### Changelog

| Date | Change |
|---|---|
| 2026-09-06 | Notebook created. PHREEQC integration wired into `run_pipeline.R`; SO4/Si element-name bug and Alkalinity charge/molar_mass bug fixed in `lab_analyte_map.R`. |
| 2026-09-06 | NDEP pH/temperature-location bug, mixed-unit-averaging bug, and missing "Total Alkalinity" ingest bug found and fixed; PHREEQC-eligible samples went from 8 to 233. |
| 2026-09-06 | Duplicate HCO3/CO3 rows cleaned up in both NDEP ingest pathways; FIELD's own `make.names()`-mangled-join-key bug found and fixed, backfilling Alkalinity (and NO3) for all 8 real thermal samples for the first time. |
| 2026-09-06 | `SBW_0002` PHREEQC convergence failure diagnosed (boron/alkalinity interaction at pH 9) and the boron-exclusion guard replaced with a physically-grounded, pH-aware check. `PHREEQC_Run_Failures` empty database-wide. |
| 2026-09-06 | Gas geothermometry (D'Amore-Panichi 1980, validated against real Mariner & Janik 1995 numbers) and forward/inverse gas mixing added (`13_gas_geothermometry.R`, `14_gas_mixing.R`) -- standalone, no schema/pipeline wiring, ahead of the planned UCSB (Fischer) gas-sampling campaign. Also fixed a latent PHREEQC SO4/sulfide element-name ambiguity (bare `S` -> explicit `S(6)`) and made the PHREEQC pipeline stage auto-detect new/changed chemistry and expose mixing/inverse/gas-phase runs via toggleable config CSVs. |

```{r cleanup}
#| include: false
dbDisconnect(con)
```
