---
title: "Historical Context: Sorey & Colvard (1992) and the Long-Run Steamboat Record"
subtitle: "Cross-referencing a 1988-1992 USGS study against this project's database"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

`docs/literature/Sorey_StmbtSprgsHSActivity_1992.pdf` -- Sorey, M.L. and
Colvard, E.M. (1992), *Factors Affecting the Decline in Hot-Spring
Activity in the Steamboat Springs Area of Critical Environmental
Concern, Washoe County, Nevada* (USGS Administrative Report for the
Bureau of Land Management, 297 p.) -- was added to this project's
literature folder on 2026-09-12. It is a real, OCR'd text layer
(extracted directly with `pdftotext -layout`, not guessed from a
citation database), and it turns out to be the single most detailed
pre-Ormat-era dataset this project has direct access to: dated 1950-1991
thermal-water chemistry, a 1955-1989 timeline of thermal-water discharge
estimates, a documented monitoring-site list, and a reproducible
barometric-pressure-correction method for spring/well water-level data.

This notebook does four things, mirroring the living-reference-notebook
convention established by `01`/`05`/`06`:

1. Summarizes what the report actually says (Sections 2-3 below), reading
   its own figures/tables/conclusions directly rather than from a
   secondary source.
2. Cross-references every well, spring, and monitoring site it names
   against this project's `Wells`/`Locations`/`Well_Aliases` tables
   (Section 4) -- several names turn out to already be independently
   resolved from entirely different sources (NDEP, Klein 2007, NDOM),
   which is itself a form of validation.
3. Documents the new ingestion of its Table 1 chemistry
   (`scripts/ingest/ingest_historical_sorey1992.R`,
   `data/raw/historical/sorey1992_table1_chemistry.csv`) and uses it to
   test whether Steamboat's thermal-water Cl/B and Cl/Li ratios -- a
   conservative-element signature Sorey found consistent across the CPI
   and SB GEO reservoirs in 1990-91 -- still hold in this project's own
   2024-2026 chemistry (Section 5).
4. Assembles a single discharge-through-time comparison spanning
   White (1968)'s 1955 estimate through this project's own 2026 poster
   result (Section 6), and documents barometric efficiency as a
   recommended future capability (Section 7) -- this project does not
   yet have the paired continuous water-level/barometric-pressure data
   needed to run Sorey's own method for real.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
```

## What the 1992 report says

**Study design.** The study began in 1988 to explain a systematic
decline in natural hot-spring flow at the main and low silica terraces
(the Steamboat Springs Area of Critical Environmental Concern, "ACEC")
-- the same Lower Sinter Terrace this thesis's own spring-remapping work
covers. Declines were first noted in spring 1986/1987; all natural
spring flow had ceased by mid-1989, with water-level declines in spring
vents of 1-17 feet.

**Conclusions (its own numbered list, p. 100-101, paraphrased):**

- ~80-95% of the head decline beneath the ACEC is attributed to decline
  in the regional shallow ground-water system (increased South Truckee
  Meadows General Improvement District pumping, plus below-normal
  precipitation beginning water-year 1986-87).
- ~5-20% is attributed to drawdown from the Caithness Power Incorporated
  (CPI) production/injection well field (full production began Feb
  1988), transmitted through a shallow ~1,000-ft thermal-water flow zone
  and steeply dipping faults.
- SB GEO well-field influence (full production began Jan 1987, 0.5 mi
  NW of the ACEC) on the ACEC springs was judged limited, despite 15-20
  ft of head decline within the SB GEO field itself -- possibly lower
  permeability/anisotropy between the two areas.

**Figures (40 total) and Tables (10 total)** cover hydrographs for
individual springs and wells (Figs. 5, 27, 29-33), precipitation and its
cumulative deviation from the mean (Figs. 15-18), a stable-isotope
altitude/boiling diagram (Fig. 9), electrical-resistivity mapping (Fig.
12), and the barometric-pressure hydrographs used in Section 7 below
(Figs. 39-40). Tables 1, 2, and 10 are used directly in this notebook;
Tables 3-9 (well-completion data, production/injection intervals,
reservoir parameters) were reviewed but are not yet transcribed here.

## Well and monitoring-site cross-reference

Every well/spring name in Table 1 (chemistry), the surrounding
discussion text, and Table 10 (monitoring sites) was checked against
this project's `Wells`, `Well_Aliases`, and `Locations` tables. Several
names were **already independently resolved** from entirely different
sources (NDEP promotions, Klein et al. 2007, NDOM permit data) years
before this 1992 report was read -- a real cross-validation, not a
coincidence engineered for this notebook.

```{r crossref-table}
crossref <- tribble(
  ~sorey_name, ~feature_type, ~status, ~note,
  "21-5 (CPI)", "well", "Resolved", "Well_Aliases '21-5' -> canonical 21-5R (well_id 94)",
  "23-5 (CPI)", "well", "Resolved", "Exact Wells.well_name match (well_id 96)",
  "83A-6 (CPI)", "well", "Resolved (2026-09-12)", "Exact NBMG Geothermal_Wells name match ('Well No. 83A-6', API 27-031-90080)",
  "Cox 1-1 / Cox well", "well (injection)", "Resolved (2026-09-12)", "Exact NBMG match ('Cox I-1', API 27-031-90051); previously flagged unresolved as COX-1/Cox1-1",
  "PW-1, PW-2, PW-3 (SB GEO)", "well", "Resolved", "Exact Wells.well_name matches (well_id 79/80/81)",
  "IW-1, IW-2, IW-3 (SB GEO)", "well (injection)", "Resolved", "IW-1 alias -> 45-28; IW-2/IW-3 exact canonical names",
  "OW-1, OW-2 (SB GEO observation)", "well (monitor)", "Partial", "OW-2 exact match (well_id 110); OW-1 not found",
  "28-32", "well", "Resolved", "Exact Wells.well_name match (well_id 125)",
  "Strat wells 2, 5, 6, 7, 8, 9, 13, 14", "well (stratigraphic test)", "Registered, no coordinate (2026-09-12)", "Real provisional Wells rows with depth/formation/temperature/water-level data (Table 8, Section 4.2); no coordinate -- see proximity notes for strat 13 (near 23-5) and strats 2/5/9 (near Cox I-1)",
  "Well GS-58, Spring GS-59", "well/seep (pre-Ormat)", "Resolved (2026-09-12)", "Both are OCR artifacts of well GS-5 (see Section 4.1) -- exact NBMG match, API 27-031-80055",
  "Hot spring 6 (main terrace ACEC)", "spring", "Unresolved", "Coordinate-less provisional Locations row; no modern counterpart identified",
  "Steamboat Creek at Rhodes Road", "surface water", "Resolved (independently)", "Locations 'SB5'/'SB6' -- same site this project's SBRR conductivity logger occupies",
  "Curti Barn Well, Curti Domestic Well", "domestic/geothermal well", "Resolved (independently)", "Klein et al. (2007)-sourced Locations rows",
  "Boyd Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Boyd Domestic Well'",
  "Rogers Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Rogers Domestic Well'",
  "Jeppson Well", "geothermal well", "Resolved (independently)", "NDEP-promoted Locations row 'Jeppson Geothermal Well'",
  "Steinhardt Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG name match ('Steinhardt Geothermal Well')",
  "Brown School Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG owner-name match ('Brown Elementary School')",
  "Bianco Well", "domestic well", "Tentative (2026-09-12)", "NDWR owner-name match ('BIANCO, U J'); PLSS-section coordinate only, and a flagged depth discrepancy (232 ft vs. Table 3's 400 ft)",
  "Herz-2 (geothermal)", "well", "Inconclusive (2026-09-12)", "Could be either of NBMG's two ambiguous 'Harold Herz Geothermal Well 1/2' candidates (flagged since Session 3); Table 3's own OCR is too garbled to disambiguate -- not guessed",
  "PTR-1, PTR-2 (Pine Tree Ranch)", "domestic well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows with real depth/chloride/water-level data; fresh NDWR/NBMG owner-name search found nothing",
  "STMGID MW-3, MW-4", "monitor well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows; NBMG's only STMGID-named record ('Well 4 Shadowridge') is a production well, not this monitor-well pair -- not used",
  "MacKay Well, Woods Well, Tangen Well, STMGID Went",
    "domestic/monitor well", "Unresolved", "Consistent with prior sessions' independent 'unresolved' flags for these"
)
crossref
```

**Reading this table:** the "Resolved (independently)" rows are the
most interesting -- this project matched Boyd, Rogers, Jeppson (via
NDEP), Curti (via Klein 2007), and Rhodes Road (via its own SBRR
conductivity-logger siting) to real coordinates in entirely separate
sessions, months before anyone here had read this 1992 report. Finding
those same names in Sorey & Colvard's own 1988-92 monitoring-site table
is independent confirmation that these are the *same*, real,
multi-decade monitoring locations -- not a coincidence of common
surnames in a small area.

The remaining genuinely unresolved rows (MacKay, Woods, Tangen, STMGID
Went, PTR-1/2, STMGID MW-3/4, and the inconclusive Herz-2/Bianco cases)
are a concrete follow-up list, consistent with (and in several cases
literally the same names as) gaps already flagged in `AGENTS.md` from
unrelated well-matching sessions. None are guessed here.

### 4.1 Resolving 83A-6, Cox 1-1, and the "GS-58"/"GS-59" OCR artifact (2026-09-12)

A direct search of NBMG's statewide `Geothermal_Wells.csv` and
`GEOTHERM06102019.csv` compilations (`data/raw/nbmg/`) found exact
well-name matches for `83A-6` (API 27-031-90080, a CPI production well
drilled 1987) and `Cox I-1` (API 27-031-90051, a CPI injection well,
previously drilled as observation well "Cox No. 1" for Phillips
Petroleum in 1981) -- both are now real `Wells` rows with NBMG-sourced
coordinates (`data/raw/wells/sorey1992_nbmg_resolved_wells.csv`,
registered via `register_sorey1992_resolved_wells()`), and `Cox I-1`
picked up `Well_Aliases` entries for the `Cox 1-1`/`COX-1`/`Cox1-1`/
`Cox well` spellings used across this report and prior sessions.

More interestingly, **"GS-58" and "GS-59" are not real well names at
all -- they are an OCR artifact.** Re-reading Table 1's own footnote
list resolves this precisely:

> `8From White (1968).`
> `9New seep adjacent to well GS-5, analysis by Nevada Division of Health Laboratory.`

Both footnote markers ("8" and "9") were superscripts attached to a
single real well name, **GS-5**, in the original typeset table, and the
PDF's OCR layer fused each digit directly onto the name it followed --
producing "GS-58" (really "GS-5" + footnote 8) and "GS-59" (really
"GS-5" + footnote 9). This is confirmed two independent ways: (1) NBMG's
statewide compilation has real wells `GS-1` through `GS-8` (the pre-Ormat
USGS 1950-51 thermal-gradient/chemistry test-hole program described in
White 1968, Plate 1) but nothing numbered `GS-58` or `GS-59`; (2) GS-5's
own NBMG record lists a chemistry suite (pH, SC, Cl, Au, Ag, As, Sb, Hg,
Tl, B, Cu, Zn, Pb) drilled 1950-1951 -- consistent with "Well GS-58"'s
1950 sample date -- and footnote 9's own text ("new seep adjacent to
well GS-5") is a direct, literal match for what "Spring GS-59" actually
is. Both rows in `sorey1992_table1_chemistry.csv` now carry
`matched_well_name = "GS-5"` and have been renamed accordingly (the
"GS-58" row to "Well GS-5, 1950 sample (White 1968)"; the "GS-59" row to
"New seep adjacent to well GS-5 (1991)", stored as a `seep`-type
Location rather than a `spring`, since that is literally what the
footnote calls it).

```{r resolved-wells-check}
dbGetQuery(con, "
  SELECT external_station_code, name, latitude, longitude, site_type
  FROM Locations WHERE external_station_code LIKE 'SOREY1992%'
  ORDER BY external_station_code
")
```

**A thematically important find:** "Hot spring 6" on the main terrace,
sampled by USGS in 1977 and again described here as *ceasing to flow
entirely by 1987*, is exactly the class of historically-active,
currently-dormant Lower Sinter Terrace spring this thesis's own
spring-remapping fieldwork is specifically designed to find and
re-evaluate. It has no modern counterpart in this database yet -- worth
keeping in mind during remapping fieldwork.

### 4.2 Table 3: South Truckee Meadows ground-water wells (2026-09-12)

Table 3 ("Data for selected wells completed in the ground-water system
of South Truckee Meadows and Pleasant Valley") ties directly to the
report's own conclusion that 80-95% of the ACEC spring decline traces to
regional shallow ground-water decline. It gives real depth/temperature/
chloride-range/water-level-decline data for PTR-1, PTR-2, STMGID MW-3,
MW-4, Bianco, Boyd, Steinhardt, Brown School, and Herz-2 -- but, like
Table 8, has **no coordinates of its own**. A fresh NDWR/NBMG
owner-name search (this session), cross-checked against the report's
own directional descriptions (e.g. Steinhardt "northeast of the ACEC"),
found:

- **Steinhardt** and **Brown School**: exact NBMG name matches
  ("Steinhardt Geothermal Well", "Brown School Geothermal Well" --
  the latter owned by "Brown Elementary School", an unambiguous
  identity confirmation). Both directions check out against the
  report's own description. Now real, coordinate-having `Wells` rows.
- **Bianco**: only a surname match in NDWR's Truckee Meadows
  WellLogQuery table ("BIANCO, U J", Log 5916, Sec 28 T18N R20E,
  completed 1961) -- resolved to a bare PLSS-section centroid (not a
  survey point, `coordinate_uncertainty_m = 800`), and its drill depth
  (232 ft) does not match Table 3's own Bianco depth (400 ft). Treated
  as **tentative**, not confirmed -- the direction (northeast of the
  ACEC) is consistent, but the depth conflict is real and unresolved.
- **PTR-1, PTR-2, STMGID MW-3, MW-4**: still **not found** in either
  NDWR basin file (TM or PV) or NBMG's statewide compilations, even
  with the new directional clues -- consistent with, not an improvement
  on, prior sessions' flags for these names. All four are still
  registered as real, coordinate-less provisional `Wells` rows so their
  Table 3 depth/temperature/chloride/water-level facts aren't stranded
  in prose only.
- **Herz-2**: the report's own text is explicit that this is a
  *geothermal* well, distinct from "the shallower Herz domestic well"
  in the same paragraph -- but NBMG has two ambiguous "Harold Herz
  Geothermal Well 1/2" candidates about 1.3 km apart (flagged unresolved
  since Session 3), and Table 3's own OCR text for Herz-2 is split
  across multiple garbled rows in a way that could not be confidently
  disentangled this session. Left inconclusive rather than guessed.

```{r table3-check}
dbGetQuery(con, "
  SELECT well_name, well_role, latitude, longitude, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name IN ('Steinhardt Geothermal Well', 'Brown School Geothermal Well', 'Bianco',
                       'Pine Tree Ranch-1', 'Pine Tree Ranch-2', 'STMGID MW-3', 'STMGID MW-4')
  ORDER BY well_name
")
```

### 4.3 Table 8: stratigraphic test wells (2026-09-12)

Table 8 ("Selected data for stratigraphic test wells in the Steamboat
Hills area") gives perforated depth, formation, bottom-hole temperature,
and 1987-1989 water-level change for strat wells 2, 5, 6, 7, 8, 9, 13,
and 14 -- the wells behind the "Strat wells... Unresolved" cross-
reference gap. Still **no coordinates anywhere in the table**, but the
surrounding text gives two real proximity clues used here (as prose
notes, never as a fabricated coordinate): *"Strat 13 is located next to
CPI production well 23-5"* (independently confirmed by a real
hydraulic-connection signal -- strat 13's water level rose during a
shut-in of well 23-5), and *"Strats 2, 5, and 9 are located near the
northern end of the Steamboat Hills, in the general vicinity of the Cox
1-1 injection well."* All eight are now real, coordinate-less
provisional `Wells` rows carrying their real depth/formation/
temperature/water-level facts.

```{r table8-check}
dbGetQuery(con, "
  SELECT well_name, well_role, notes
  FROM Wells WHERE well_name LIKE 'strat %'
  ORDER BY well_name
")
```

Two things worth noting from the table: strat 5's bottom-hole
temperature (44 C) is far cooler than strats 2 and 9 (171-179 C) despite
a similar depth range -- the report attributes this to strat 5 tapping a
distinct, non-thermal zone, not the same reservoir. And strats 8 and 14
both show small water-level *rises* (1-2 ft, 1987-1989) with "no clear
explanation" given in the report -- a genuine open question in the
source material, not something resolved here either.

### 4.4 Tables 6/7: well-completion data fills a currently-empty field (2026-09-12)

`Wells.top_perforation`/`bottom_perforation` were NULL for every one of
the CPI/SB GEO wells already resolved in this database, even though
Tables 6 and 7 ("CPI well-completion information", "SB GEO well
completion information") report real casing-depth and open-hole-
interval data for most of them. The table's own OCR layout has visible
column-alignment slippage, so each well's numbers were individually
cross-checked (casing depth + open-hole thickness = total depth,
confirmed to reconcile exactly for every SB GEO well and for 83A-6) before
being trusted:

```{r table67-check}
dbGetQuery(con, "
  SELECT well_name, total_depth, top_perforation, bottom_perforation
  FROM Wells
  WHERE well_name IN ('83A-6','Cox I-1','PW-1','PW-2','PW-3','21-5R','IW-3','23-5','IW-2')
  ORDER BY well_name
")
```

**Two real, deliberately-unfilled conflicts**: `23-5`'s 1990-era report
depth (2422 ft) is 19% shallower than the 3001 ft already on record
(NDOM-sourced), and `IW-2`'s 1990 report depth (1403 ft) is less than a
third of the 4700 ft on record -- both almost certainly explained by
subsequent deepening of these still-active wells, not a data error.
Rather than overwrite (or awkwardly average) these, the 1990 construction
facts are recorded in `Wells.notes` as explicitly historical, and
`top_perforation`/`bottom_perforation` were left untouched for both.
Every other well's numbers were close enough to (or had no) existing
`total_depth` to fill with confidence.


## Newly ingested chemistry: Table 1 (1950-1991)

`ingest_historical_sorey1992.R` (wired into `run_pipeline.R` as
`RUN_INGEST$historical_sorey1992`) transcribes Table 1's ~10 dated
analyses directly into `Lab_Analyses`, extending this project's own
chemistry record back to 1950 (well GS-58) for the first time.

```{r sorey-chem-table}
sorey_chem <- dbGetQuery(con, "
  SELECT l.name AS feature, s.collection_time AS sample_date,
         la.analyte, la.value, la.units
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE l.external_station_code LIKE 'SOREY1992%'
  ORDER BY s.collection_time, l.name
")
sorey_chem_wide <- sorey_chem %>%
  select(feature, sample_date, analyte, value) %>%
  pivot_wider(names_from = analyte, values_from = value)
sorey_chem_wide
```

### Cl/B and Cl/Li ratio check against modern chemistry

Sorey & Colvard found Cl/B = 19.3 +/- 1.7 and Cl/Li = 122.5 +/- 9.9
across the CPI and SB GEO reservoirs (excluding the more dilute Cox
well), and used this constancy of conservative-element ratios to argue
the two well fields draw on chemically related source water. Recomputing
the same ratios from the newly-ingested 1950-1991 values, and comparing
against this project's own real 2024-2026 thermal-water samples (the
FIELD spring/well/seep samples, and the NDEP-promoted domestic wells),
is a direct test of whether that same geochemical signature has
persisted for over three decades.

```{r ratio-check}
hist_ratios <- sorey_chem %>%
  filter(analyte %in% c("Cl", "B", "Li")) %>%
  pivot_wider(id_cols = c(feature, sample_date), names_from = analyte, values_from = value) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li, era = "1950-1991 (Sorey & Colvard 1992)")

modern_ratios <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, la.analyte, la.value
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte IN ('Cl','B','Li')
    AND s.data_source != 'Sorey & Colvard 1992'
") %>%
  pivot_wider(id_cols = c(sample_id, feature, site_type), names_from = analyte,
              values_from = value, values_fn = mean) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li,
         era = if_else(site_type %in% c("fumarole", "seep", "spring"),
                        "2024-2026 real thermal (FIELD)",
                        "2024-2026 NDEP domestic/background wells"))

ratio_compare <- bind_rows(
  hist_ratios %>% select(feature, era, Cl, B, Li, Cl_B, Cl_Li),
  modern_ratios %>% select(feature, era, Cl, B, Li, Cl_B, Cl_Li)
)
ratio_compare
```

```{r ratio-plot}
#| fig-width: 8
#| fig-height: 5
ggplot(ratio_compare %>% filter(!is.na(Cl_B)), aes(x = Cl_B, y = era)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_vline(xintercept = 19.3, linetype = "dashed") +
  annotate("text", x = 19.3, y = 0.55, label = "Sorey & Colvard (1992)\nmean Cl/B = 19.3", size = 3, hjust = -0.05) +
  labs(x = "Cl / B (mass ratio)", y = NULL,
       title = "Chloride-to-boron ratio: 1950-1991 vs. present-day chemistry",
       subtitle = "Real thermal samples cluster near Sorey's 1990-91 value; dilute domestic/background wells scatter widely near their B detection limit")
```

**Reading this:** this project's own 2024-2026 real thermal FIELD
samples (fumarole/seep/spring) give Cl/B roughly 20-23, closely matching
Sorey & Colvard's 1990-91 CPI/SB GEO value of 19.3 +/- 1.7 -- a genuine,
quantitative "then vs. now" consistency check, not just a qualitative
one. The NDEP domestic/background well samples (Boyd, Jeppson, Rogers)
show much more scattered Cl/B ratios; their B values sit at or near
0.05-0.18 mg/L, close to a plausible analytical detection limit, so
those ratios should be read as noisy near-DL artifacts of dilute,
non-thermal ground water rather than a real geochemical signal -- fully
consistent with this project's established finding that these are
background, not thermal, waters.

## Discharge-through-time: extending Sorey's Table 2

Table 2 of the 1992 report tabulates dated estimates of thermal-water
discharge to Steamboat Creek from 1955 through 1989. These values are
**transcribed literature numbers, not live database queries** -- they
predate any table this project's own pipeline builds -- assembled here
alongside the already-cited Sorey & Spielman (2008, 2017) follow-up
estimates and this project's own 2026 poster result to give a single,
~70-year view.

```{r discharge-timeline}
discharge_hist <- tribble(
  ~period, ~year_mid, ~gpm, ~Ls, ~source,
  "White (1968), Apr 1955",         1955, 1110, 1110 * 0.0630902, "White (1968), chloride-flux method",
  "White (1968), Apr 1964",         1964, 1385, 1385 * 0.0630902, "White (1968), chloride-flux method",
  "Shump (1985), 1981-82 avg",      1981.5, 1300, 1300 * 0.0630902, "Shump (1985), specific-conductance method (less reliable per Sorey & Colvard)",
  "Collar (1990), Jun 1988",        1988.5, 523, 523 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Collar (1990), Aug 1988",        1988.6, NA, NA, "Not counting net CPI production; unseen seepage component only reported",
  "Collar (1990), Mar 1989",        1989.2, 663, 663 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Sorey & Spielman (2008), 2007-08", 2007.5, NA, mean(c(730, 1214)) * 0.0630902, "Chloride-flux, SBRR/SBBV sites",
  "Sorey & Spielman (2017), 2008-2016 avg", 2012, NA, 60, "Chloride-flux, SBRR/SBBV sites, 24 sampling events",
  "Irvin & Lindsey (2026 poster), post-eruption", 2026, NA, 27, "This project's own chloride mass-balance estimate"
)
discharge_hist
```

```{r discharge-plot}
#| fig-width: 8
#| fig-height: 5
ggplot(discharge_hist %>% filter(!is.na(Ls)), aes(x = year_mid, y = Ls)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.4) +
  labs(x = "Year", y = "Thermal-water discharge (L/s)",
       title = "Thermal-water discharge to Steamboat Creek, 1955-2026",
       subtitle = "White (1968) through this project's own 2026 poster estimate")
```

**Reading this:** discharge estimates were broadly 45-90 L/s
(660-1,385 gpm-equivalent) from 1955 through the 2008-2016 Sorey &
Spielman baseline, with the 1988-89 Collar (1990) values sitting at the
low end even before accounting for that they explicitly exclude net CPI
well-field production. This project's own post-eruption ~27 L/s estimate
sits below every prior estimate in this 70-year record -- consistent
with the poster's own conclusion (conduit reactivation/redistribution,
not necessarily a new source) but worth treating as one point in a
long, variable record rather than a single before/after comparison.
None of the pre-2007 methods used the same sites (SBRR/SBBV) as the
modern chloride-flux estimates, so method consistency, not just
magnitude, matters when reading this chart -- White (1968) and Collar
(1990)'s methods and gauge locations are not identical to Sorey &
Spielman's.

## Appendix G and Table 9: what's behind the discharge/reservoir numbers (2026-09-12)

### Appendix G: the raw synoptic Cl-flux surveys behind Table 2

Appendix G ("Stream Discharge and Chloride Flux") documents the real
method behind the Collar (1990) discharge estimates already used in the
discharge-through-time chart above: chloride treated as a conservative
tracer, an 820 mg/L thermal / 6 mg/L non-thermal end-member split, and
gaining/losing-reach mass-balance equations -- methodologically almost
identical to this project's own SBRR/SBBV Cl mass-balance approach, just
predating it by ~35 years. Chloride was measured by mercuric-nitrate or
argentometric titration (SDSU and UURI, respectively); discharge by
pygmy current meter. The underlying raw data (Table G-1) comes from four
real synoptic surveys: **6/26-6/30/1988, 7/1-7/2/1988, 8/9/1988, and
3/4/1989**, at ~40 named stations along Steamboat Creek, Galena Creek,
and the Steamboat/Chandler/Crane irrigation ditches.

**Deliberately not parsed further**: Table G-1's raw per-station
discharge/chloride/boron values are visibly OCR-misaligned in the
extracted text (values shifted relative to station-name rows, in some
blocks by a full row) -- the same class of "real data trapped in a
badly-OCR'd table" problem already flagged elsewhere in this project
(the NDEP TFT Appendix D geochemistry tables, well-log lithology
tables). The three aggregate numbers we actually use (Collar 1990's
1988-89 discharge estimates in the Table 2 chart above) are already the
synthesized output of this same appendix -- Appendix G is the raw
survey behind data already in this notebook, not a source of new,
different data points. If a denser historical Cl-flux time series is
ever wanted, it would need to be read directly from the scanned page
images (not the OCR text) station-by-station -- a real, larger effort
than anything else in this session, not attempted here.

### Table 9: reservoir parameters (transmissivity, storage coefficient)

Table 9 ("Reservoir parameters determined for the geothermal system in
the Steamboat Hills area from well test analyses and spring
hydrographs") gives real aquifer-test results from four dated tests:

```{r table9-reference}
table9 <- tibble::tribble(
  ~well_and_year, ~injection, ~transmissivity_ft2_day, ~storage_coefficient, ~observation_well, ~source,
  "Steamboat No.1, 1980", "no", "3270-9500", "1.2e-3 (calc.)", "strat 2", "Collar (1990)",
  "Steamboat No.1, Mar-May 1986", "no", "nd", "nd", "23-5?", "Berkeley Group (1987)",
  "Steamboat No.1, Mar-May 1987", "no", "nd", "1.2e-3 (calc.)", "strat 5", "Berkeley Group (1987)",
  "23-5/Cox 1-1, May-June 1987", "yes", "7930-9500", "7.8e-4 to 9.0e-4", "strat 2, 21-5", "Faulder (1987); Goranson (1989)",
  "21-5, 23-5, 83A-6, 1988", "yes", "1250-2400", "2.5e-3", "strat 9", "Collar (1990)",
  "21-5, 23-5, 83A-6, 1988 (spring response)", "yes", "nd", "2.8e-3", "spring 12, spring 42w", "Collar (1990)"
)
table9
```

No table in the current schema models a dated aquifer/interference
test -- this is kept as reference documentation only, not a new
database table, since there is no immediate analytical use for it beyond
the qualitative reservoir-drawdown narrative already summarized above
("What the 1992 report says"). Worth revisiting if this project's own
hydraulic-gradient work ever needs a real transmissivity/storage
reference range for the Steamboat Hills reservoir.


## Table 10 monitoring network vs. this project's Locations

```{r table10-check}
dbGetQuery(con, "
  SELECT location_id, external_station_code, name, site_type,
         latitude, longitude, coordinate_source
  FROM Locations
  WHERE name LIKE '%Boyd%' OR name LIKE '%Curti%' OR name LIKE '%Rogers%'
     OR name LIKE '%Jeppson%' OR name LIKE '%Herz%' OR name LIKE '%Zolezzi%'
     OR external_station_code IN ('SB5','SB6')
  ORDER BY name
")
```

Sorey & Colvard's Table 10 ("Current monitoring sites for CPI and SBG
geothermal developments") describes exactly this same set of domestic
wells and the Rhodes Road creek-monitoring point as the active 1988-92
monitoring network -- confirming this project's own SBRR site and its
NDEP/Klein-resolved domestic wells sit at genuinely long-monitored
locations, not newly-chosen ones. None of these have a linked `Wells`
row or `Water_Level_Observations` history yet, which is a real,
documented gap (see Section 7 caveats) rather than an oversight to
silently work around here.

**Aside (found and fixed, 2026-09-12): the malformed `"7-07-10"`
timestamp.** `Water_Level_Observations` has 73 wells and 153,437 rows
spanning 1969-06-23 to 2026-01-01; one row (well-log document 104216, a
"WELL DRILLER'S PLUGGING REPORT" with `completion_date_raw = "7/10/07"`)
had timestamp `"7-07-10"`. Root cause: `ingest_well_logs.R`'s
`.safe_completion_date()` tries `as.Date(x, format = "%m/%d/%Y")` first,
and R's `%Y` silently accepts a 2-digit year as a literal, unpadded year
value (year 7 CE) rather than rejecting the format mismatch the way the
function's design assumed it would -- so the very first format "matched"
with a nonsensical date instead of falling through to a better one.
Fixed by adding an explicit `%m/%d/%y` attempt and a plausible-year-range
sanity check (1900 to next calendar year) that every candidate parse must
pass before being accepted; the one bad row is corrected to `2007-07-10`.
No other row in `Well_Log_Documents`/`Water_Level_Observations` was
affected -- this was the only 2-digit-year-formatted completion date in
the batch.

## Barometric efficiency: a documented, not-yet-implemented capability

Sorey & Colvard (p. 85-88, Figs. 39-40) computed **barometric
efficiency** (BE) -- the ratio of water-level change in a well/spring to
the corresponding barometric-pressure change -- by linear regression of
spring water level against Reno Airport barometric pressure, from 3-4x
daily manual readings over one week in October 1988. Reported values:
BE = 0.42 for spring 6 (r = 0.79) and 0.45 for spring 12 (r = 0.44);
White (1968) found BE 0.2-1.18 across different vents, with higher
values (occasionally exceeding 1.0) where the spring vent's fluid column
is at or near boiling.

**Why this project can't run this for real yet:** it requires
*continuous* barometric pressure paired with *continuous* water level at
the same site. Checking both:

```{r barometric-check}
dbGetQuery(con, "SELECT DISTINCT parameter, unit FROM Weather_Observations")
```

NOAA's ingested weather parameters are precipitation and air temperature
only -- **no barometric/station pressure parameter currently exists
anywhere in this database.** `Water_Level_Observations` itself is
discrete manual/transducer readings, not always continuous, and none of
the wells in Section 4's cross-reference table currently overlap in time
with any pressure record anyway.

This mirrors the same "build the capability, wait for real overlapping
data" pattern already used for PHREEQC's mixing/inverse/gas-phase
modules (`06_phreeqc_geochemical_modeling.qmd`): a stub function with
only a synthetic self-test, not wired into `run_pipeline.R`.

```{r barometric-stub}
#| eval: false
# Not run by default -- synthetic self-test only, requires real paired
# continuous barometric-pressure + water-level data to be meaningful.
compute_barometric_efficiency <- function(water_level, barometric_pressure) {
  fit <- lm(water_level ~ barometric_pressure)
  list(
    BE = unname(coef(fit)[2]),
    r_squared = summary(fit)$r.squared,
    model = fit
  )
}

demo_barometric_efficiency <- function(seed = 4172) {
  set.seed(seed)
  n <- 168 # one week, hourly
  baro <- 30 + cumsum(rnorm(n, 0, 0.02))          # synthetic barometric pressure, inHg
  wl <- 5 - 0.4 * (baro - mean(baro)) + rnorm(n, 0, 0.02)  # synthetic BE ~ 0.4
  compute_barometric_efficiency(wl, baro)
}
demo_barometric_efficiency()
```

**Recommendation, not a request for new instrumentation decisions here:**
if a future logger deployment or manual monitoring effort pairs
continuous water level at a spring/well with continuous (or even just
synchronized 3-4x/day manual) barometric pressure, this function is
ready to compute a real, Sorey-comparable BE immediately.

## Why this matters for the project

Sorey & Colvard (1992) does three things no other single source in this
project's literature folder does at once: it gives this project's
chemistry record a **35-40 year deeper baseline** (Table 1, now
ingested), it gives the chloride-discharge thesis question a **~70-year
timeline** instead of just the 2008-2026 window (Section 6), and it
independently corroborates several of this project's own
well/site-identity resolutions made from entirely unrelated sources in
earlier sessions (Section 4) -- itself a form of validation this
project didn't have before. It also demonstrates, with real 1988-era
field methods, that this project's own instrumentation program (SBRR/SBGG
conductivity loggers, temperature loggers, NDEP-resolved domestic wells)
sits on genuinely decades-old monitoring ground, not a newly invented
network. The one capability it demonstrates that this project cannot yet
reproduce -- barometric efficiency -- is now a documented, ready-to-run
gap rather than an unrecognized one.

## Changelog

| Date | Change |
|---|---|
| 2026-09-12 | Notebook created. Sorey & Colvard (1992) read directly (`pdftotext -layout`); Table 1 chemistry (9 samples, 1950-1991) ingested for the first time via `ingest_historical_sorey1992.R`; well/site cross-reference built against `Wells`/`Well_Aliases`/`Locations`; Cl/B ratio confirmed consistent between 1990-91 and 2024-2026 real thermal samples; discharge-through-time table assembled (White 1968 through this project's 2026 poster); barometric efficiency documented as a recommended, not-yet-runnable capability (no barometric-pressure parameter exists in this database yet). |
| 2026-09-12 (follow-up) | Resolved `83A-6` and `Cox 1-1` via exact NBMG Geothermal_Wells name matches; discovered and resolved that `"GS-58"`/`"GS-59"` are OCR artifacts of well `GS-5` (its own table footnote superscripts fused onto the name); all four now have real coordinates in `Locations`/`Wells`. Also found and fixed a 2-digit-year date-parsing bug in `ingest_well_logs.R`'s `.safe_completion_date()` that had produced one malformed `Water_Level_Observations` timestamp (`"7-07-10"`, corrected to `2007-07-10`). |
| 2026-09-12 (Tables 3/6/7/8/9, Appendix G) | Registered strat wells 2/5/6/7/8/9/13/14 (Table 8) and Table 3's South Truckee Meadows wells as real Wells rows; resolved Steinhardt and Brown School via exact NBMG matches, Bianco tentatively via NDWR owner-name + PLSS centroid; PTR-1/2 and STMGID MW-3/4 remain unresolved despite a fresh search; Herz-2 left inconclusive between two ambiguous NBMG candidates. Filled `Wells.top_perforation`/`bottom_perforation` (previously 100% NULL) for 7 wells from Tables 6/7, deliberately leaving 23-5 and IW-2 unfilled due to real depth conflicts with current records (documented as historical facts instead). Documented Appendix G's real methodology/survey dates (not parsed further -- too OCR-misaligned) and Table 9's reservoir-parameter values as reference only. |

```{r cleanup}
#| include: false
dbDisconnect(con)
```
