---
title: "Historical Context: Sorey & Colvard (1992) and the Long-Run Steamboat Record"
subtitle: "Cross-referencing a 1988-1992 USGS study against this project's database"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

`docs/literature/Sorey_StmbtSprgsHSActivity_1992.pdf` -- Sorey, M.L. and
Colvard, E.M. (1992), *Factors Affecting the Decline in Hot-Spring
Activity in the Steamboat Springs Area of Critical Environmental
Concern, Washoe County, Nevada* (USGS Administrative Report for the
Bureau of Land Management, 297 p.) -- was added to this project's
literature folder on 2026-09-12. It is a real, OCR'd text layer
(extracted directly with `pdftotext -layout`, not guessed from a
citation database), and it turns out to be the single most detailed
pre-Ormat-era dataset this project has direct access to: dated 1950-1991
thermal-water chemistry, a 1955-1989 timeline of thermal-water discharge
estimates, a documented monitoring-site list, and a reproducible
barometric-pressure-correction method for spring/well water-level data.

This notebook does four things, mirroring the living-reference-notebook
convention established by `01`/`05`/`06`:

1. Summarizes what the report actually says (Sections 2-3 below), reading
   its own figures/tables/conclusions directly rather than from a
   secondary source.
2. Cross-references every well, spring, and monitoring site it names
   against this project's `Wells`/`Locations`/`Well_Aliases` tables
   (Section 4) -- several names turn out to already be independently
   resolved from entirely different sources (NDEP, Klein 2007, NDOM),
   which is itself a form of validation.
3. Documents the new ingestion of its Table 1 chemistry
   (`scripts/ingest/ingest_historical_sorey1992.R`,
   `data/raw/historical/sorey1992_table1_chemistry.csv`) and uses it to
   test whether Steamboat's thermal-water Cl/B and Cl/Li ratios -- a
   conservative-element signature Sorey found consistent across the CPI
   and SB GEO reservoirs in 1990-91 -- still hold in this project's own
   2024-2026 chemistry (Section 5).
4. Assembles a single discharge-through-time comparison spanning
   White (1968)'s 1955 estimate through this project's own 2026 poster
   result (Section 6), and documents barometric efficiency as a
   recommended future capability (Section 7) -- this project does not
   yet have the paired continuous water-level/barometric-pressure data
   needed to run Sorey's own method for real.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
```

## What the 1992 report says

**Study design.** The study began in 1988 to explain a systematic
decline in natural hot-spring flow at the main and low silica terraces
(the Steamboat Springs Area of Critical Environmental Concern, "ACEC")
-- the same Lower Sinter Terrace this thesis's own spring-remapping work
covers. Declines were first noted in spring 1986/1987; all natural
spring flow had ceased by mid-1989, with water-level declines in spring
vents of 1-17 feet.

**Conclusions (its own numbered list, p. 100-101, paraphrased):**

- ~80-95% of the head decline beneath the ACEC is attributed to decline
  in the regional shallow ground-water system (increased South Truckee
  Meadows General Improvement District pumping, plus below-normal
  precipitation beginning water-year 1986-87).
- ~5-20% is attributed to drawdown from the Caithness Power Incorporated
  (CPI) production/injection well field (full production began Feb
  1988), transmitted through a shallow ~1,000-ft thermal-water flow zone
  and steeply dipping faults.
- SB GEO well-field influence (full production began Jan 1987, 0.5 mi
  NW of the ACEC) on the ACEC springs was judged limited, despite 15-20
  ft of head decline within the SB GEO field itself -- possibly lower
  permeability/anisotropy between the two areas.

**Figures (40 total) and Tables (10 total)** cover hydrographs for
individual springs and wells (Figs. 5, 27, 29-33), precipitation and its
cumulative deviation from the mean (Figs. 15-18), a stable-isotope
altitude/boiling diagram (Fig. 9), electrical-resistivity mapping (Fig.
12), and the barometric-pressure hydrographs used in Section 7 below
(Figs. 39-40). Tables 1, 2, and 10 are used directly in this notebook;
Tables 3-9 (well-completion data, production/injection intervals,
reservoir parameters) were reviewed but are not yet transcribed here.

## Well and monitoring-site cross-reference

Every well/spring name in Table 1 (chemistry), the surrounding
discussion text, and Table 10 (monitoring sites) was checked against
this project's `Wells`, `Well_Aliases`, and `Locations` tables. Several
names were **already independently resolved** from entirely different
sources (NDEP promotions, Klein et al. 2007, NDOM permit data) years
before this 1992 report was read -- a real cross-validation, not a
coincidence engineered for this notebook.

```{r crossref-table}
crossref <- tribble(
  ~sorey_name, ~feature_type, ~status, ~note,
  "21-5 (CPI)", "well", "Resolved", "Well_Aliases '21-5' -> canonical 21-5R (well_id 94)",
  "23-5 (CPI)", "well", "Resolved", "Exact Wells.well_name match (well_id 96)",
  "83A-6 (CPI)", "well", "Resolved (2026-09-12)", "Exact NBMG Geothermal_Wells name match ('Well No. 83A-6', API 27-031-90080)",
  "Cox 1-1 / Cox well", "well (injection)", "Resolved (2026-09-12)", "Exact NBMG match ('Cox I-1', API 27-031-90051); previously flagged unresolved as COX-1/Cox1-1",
  "PW-1, PW-2, PW-3 (SB GEO)", "well", "Resolved", "Exact Wells.well_name matches (well_id 79/80/81)",
  "IW-1, IW-2, IW-3 (SB GEO)", "well (injection)", "Resolved", "IW-1 alias -> 45-28; IW-2/IW-3 exact canonical names",
  "OW-1, OW-2 (SB GEO observation)", "well (monitor)", "Partial", "OW-2 exact match (well_id 110); OW-1 not found",
  "28-32", "well", "Resolved", "Exact Wells.well_name match (well_id 125)",
  "Strat wells 2, 5, 6, 7, 8, 9, 13, 14", "well (stratigraphic test)", "Registered, no coordinate (2026-09-12)", "Real provisional Wells rows with depth/formation/temperature/water-level data (Table 8, Section 4.2); no coordinate -- see proximity notes for strat 13 (near 23-5) and strats 2/5/9 (near Cox I-1)",
  "Well GS-58, Spring GS-59", "well/seep (pre-Ormat)", "Resolved (2026-09-12)", "Both are OCR artifacts of well GS-5 (see Section 4.1) -- exact NBMG match, API 27-031-80055",
  "Hot spring 6 (main terrace ACEC)", "spring", "Unresolved", "Coordinate-less provisional Locations row; no modern counterpart identified",
  "Steamboat Creek at Rhodes Road", "surface water", "Resolved (independently)", "Locations 'SB5'/'SB6' -- same site this project's SBRR conductivity logger occupies",
  "Curti Barn Well, Curti Domestic Well", "domestic/geothermal well", "Resolved (independently)", "Klein et al. (2007)-sourced Locations rows",
  "Boyd Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Boyd Domestic Well'",
  "Rogers Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Rogers Domestic Well'",
  "Jeppson Well", "geothermal well", "Resolved (independently)", "NDEP-promoted Locations row 'Jeppson Geothermal Well'",
  "Steinhardt Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG name match ('Steinhardt Geothermal Well')",
  "Brown School Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG owner-name match ('Brown Elementary School')",
  "Bianco Well", "domestic well", "Tentative (2026-09-12)", "NDWR owner-name match ('BIANCO, U J'); PLSS-section coordinate only, and a flagged depth discrepancy (232 ft vs. Table 3's 400 ft)",
  "Herz-2 (geothermal)", "well", "Inconclusive (2026-09-12)", "Could be either of NBMG's two ambiguous 'Harold Herz Geothermal Well 1/2' candidates (flagged since Session 3); Table 3's own OCR is too garbled to disambiguate -- not guessed",
  "PTR-1, PTR-2 (Pine Tree Ranch)", "domestic well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows with real depth/chloride/water-level data; fresh NDWR/NBMG owner-name search found nothing",
  "STMGID MW-3, MW-4", "monitor well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows; NBMG's only STMGID-named record ('Well 4 Shadowridge') is a production well, not this monitor-well pair -- not used",
  "MacKay Well, Woods Well, Tangen Well, STMGID Went",
    "domestic/monitor well", "Unresolved", "Consistent with prior sessions' independent 'unresolved' flags for these"
)
crossref
```

**Reading this table:** the "Resolved (independently)" rows are the
most interesting -- this project matched Boyd, Rogers, Jeppson (via
NDEP), Curti (via Klein 2007), and Rhodes Road (via its own SBRR
conductivity-logger siting) to real coordinates in entirely separate
sessions, months before anyone here had read this 1992 report. Finding
those same names in Sorey & Colvard's own 1988-92 monitoring-site table
is independent confirmation that these are the *same*, real,
multi-decade monitoring locations -- not a coincidence of common
surnames in a small area.

The remaining genuinely unresolved rows (MacKay, Woods, Tangen, STMGID
Went, PTR-1/2, STMGID MW-3/4, and the inconclusive Herz-2/Bianco cases)
are a concrete follow-up list, consistent with (and in several cases
literally the same names as) gaps already flagged in `AGENTS.md` from
unrelated well-matching sessions. None are guessed here.

### 4.1 Resolving 83A-6, Cox 1-1, and the "GS-58"/"GS-59" OCR artifact (2026-09-12)

A direct search of NBMG's statewide `Geothermal_Wells.csv` and
`GEOTHERM06102019.csv` compilations (`data/raw/nbmg/`) found exact
well-name matches for `83A-6` (API 27-031-90080, a CPI production well
drilled 1987) and `Cox I-1` (API 27-031-90051, a CPI injection well,
previously drilled as observation well "Cox No. 1" for Phillips
Petroleum in 1981) -- both are now real `Wells` rows with NBMG-sourced
coordinates (`data/raw/wells/sorey1992_nbmg_resolved_wells.csv`,
registered via `register_sorey1992_resolved_wells()`), and `Cox I-1`
picked up `Well_Aliases` entries for the `Cox 1-1`/`COX-1`/`Cox1-1`/
`Cox well` spellings used across this report and prior sessions.

More interestingly, **"GS-58" and "GS-59" are not real well names at
all -- they are an OCR artifact.** Re-reading Table 1's own footnote
list resolves this precisely:

> `8From White (1968).`
> `9New seep adjacent to well GS-5, analysis by Nevada Division of Health Laboratory.`

Both footnote markers ("8" and "9") were superscripts attached to a
single real well name, **GS-5**, in the original typeset table, and the
PDF's OCR layer fused each digit directly onto the name it followed --
producing "GS-58" (really "GS-5" + footnote 8) and "GS-59" (really
"GS-5" + footnote 9). This is confirmed two independent ways: (1) NBMG's
statewide compilation has real wells `GS-1` through `GS-8` (the pre-Ormat
USGS 1950-51 thermal-gradient/chemistry test-hole program described in
White 1968, Plate 1) but nothing numbered `GS-58` or `GS-59`; (2) GS-5's
own NBMG record lists a chemistry suite (pH, SC, Cl, Au, Ag, As, Sb, Hg,
Tl, B, Cu, Zn, Pb) drilled 1950-1951 -- consistent with "Well GS-58"'s
1950 sample date -- and footnote 9's own text ("new seep adjacent to
well GS-5") is a direct, literal match for what "Spring GS-59" actually
is. Both rows in `sorey1992_table1_chemistry.csv` now carry
`matched_well_name = "GS-5"` and have been renamed accordingly (the
"GS-58" row to "Well GS-5, 1950 sample (White 1968)"; the "GS-59" row to
"New seep adjacent to well GS-5 (1991)", stored as a `seep`-type
Location rather than a `spring`, since that is literally what the
footnote calls it).

```{r resolved-wells-check}
dbGetQuery(con, "
  SELECT external_station_code, name, latitude, longitude, site_type
  FROM Locations WHERE external_station_code LIKE 'SOREY1992%'
  ORDER BY external_station_code
")
```

**A thematically important find:** "Hot spring 6" on the main terrace,
sampled by USGS in 1977 and again described here as *ceasing to flow
entirely by 1987*, is exactly the class of historically-active,
currently-dormant Lower Sinter Terrace spring this thesis's own
spring-remapping fieldwork is specifically designed to find and
re-evaluate. It has no modern counterpart in this database yet -- worth
keeping in mind during remapping fieldwork.

### 4.2 Table 3: South Truckee Meadows ground-water wells (2026-09-12)

Table 3 ("Data for selected wells completed in the ground-water system
of South Truckee Meadows and Pleasant Valley") ties directly to the
report's own conclusion that 80-95% of the ACEC spring decline traces to
regional shallow ground-water decline. It gives real depth/temperature/
chloride-range/water-level-decline data for PTR-1, PTR-2, STMGID MW-3,
MW-4, Bianco, Boyd, Steinhardt, Brown School, and Herz-2 -- but, like
Table 8, has **no coordinates of its own**. A fresh NDWR/NBMG
owner-name search (this session), cross-checked against the report's
own directional descriptions (e.g. Steinhardt "northeast of the ACEC"),
found:

- **Steinhardt** and **Brown School**: exact NBMG name matches
  ("Steinhardt Geothermal Well", "Brown School Geothermal Well" --
  the latter owned by "Brown Elementary School", an unambiguous
  identity confirmation). Both directions check out against the
  report's own description. Now real, coordinate-having `Wells` rows.
- **Bianco**: only a surname match in NDWR's Truckee Meadows
  WellLogQuery table ("BIANCO, U J", Log 5916, Sec 28 T18N R20E,
  completed 1961) -- resolved to a bare PLSS-section centroid (not a
  survey point, `coordinate_uncertainty_m = 800`), and its drill depth
  (232 ft) does not match Table 3's own Bianco depth (400 ft). Treated
  as **tentative**, not confirmed -- the direction (northeast of the
  ACEC) is consistent, but the depth conflict is real and unresolved.
- **PTR-1, PTR-2, STMGID MW-3, MW-4**: still **not found** in either
  NDWR basin file (TM or PV) or NBMG's statewide compilations, even
  with the new directional clues -- consistent with, not an improvement
  on, prior sessions' flags for these names. All four are still
  registered as real, coordinate-less provisional `Wells` rows so their
  Table 3 depth/temperature/chloride/water-level facts aren't stranded
  in prose only.
- **Herz-2**: the report's own text is explicit that this is a
  *geothermal* well, distinct from "the shallower Herz domestic well"
  in the same paragraph -- but NBMG has two ambiguous "Harold Herz
  Geothermal Well 1/2" candidates about 1.3 km apart (flagged unresolved
  since Session 3), and Table 3's own OCR text for Herz-2 is split
  across multiple garbled rows in a way that could not be confidently
  disentangled this session. Left inconclusive rather than guessed.

```{r table3-check}
dbGetQuery(con, "
  SELECT well_name, well_role, latitude, longitude, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name IN ('Steinhardt Geothermal Well', 'Brown School Geothermal Well', 'Bianco',
                       'Pine Tree Ranch-1', 'Pine Tree Ranch-2', 'STMGID MW-3', 'STMGID MW-4')
  ORDER BY well_name
")
```

### 4.3 Table 8: stratigraphic test wells (2026-09-12)

Table 8 ("Selected data for stratigraphic test wells in the Steamboat
Hills area") gives perforated depth, formation, bottom-hole temperature,
and 1987-1989 water-level change for strat wells 2, 5, 6, 7, 8, 9, 13,
and 14 -- the wells behind the "Strat wells... Unresolved" cross-
reference gap. Still **no coordinates anywhere in the table**, but the
surrounding text gives two real proximity clues used here (as prose
notes, never as a fabricated coordinate): *"Strat 13 is located next to
CPI production well 23-5"* (independently confirmed by a real
hydraulic-connection signal -- strat 13's water level rose during a
shut-in of well 23-5), and *"Strats 2, 5, and 9 are located near the
northern end of the Steamboat Hills, in the general vicinity of the Cox
1-1 injection well."* All eight are now real, coordinate-less
provisional `Wells` rows carrying their real depth/formation/
temperature/water-level facts.

```{r table8-check}
dbGetQuery(con, "
  SELECT well_name, well_role, notes
  FROM Wells WHERE well_name LIKE 'strat %'
  ORDER BY well_name
")
```

Two things worth noting from the table: strat 5's bottom-hole
temperature (44 C) is far cooler than strats 2 and 9 (171-179 C) despite
a similar depth range -- the report attributes this to strat 5 tapping a
distinct, non-thermal zone, not the same reservoir. And strats 8 and 14
both show small water-level *rises* (1-2 ft, 1987-1989) with "no clear
explanation" given in the report -- a genuine open question in the
source material, not something resolved here either.

### 4.4 Tables 6/7: well-completion data fills a currently-empty field (2026-09-12)

`Wells.top_perforation`/`bottom_perforation` were NULL for every one of
the CPI/SB GEO wells already resolved in this database, even though
Tables 6 and 7 ("CPI well-completion information", "SB GEO well
completion information") report real casing-depth and open-hole-
interval data for most of them. The table's own OCR layout has visible
column-alignment slippage, so each well's numbers were individually
cross-checked (casing depth + open-hole thickness = total depth,
confirmed to reconcile exactly for every SB GEO well and for 83A-6) before
being trusted:

```{r table67-check}
dbGetQuery(con, "
  SELECT well_name, total_depth, top_perforation, bottom_perforation
  FROM Wells
  WHERE well_name IN ('83A-6','Cox I-1','PW-1','PW-2','PW-3','21-5R','IW-3','23-5','IW-2')
  ORDER BY well_name
")
```

**Two real, deliberately-unfilled conflicts**: `23-5`'s 1990-era report
depth (2422 ft) is 19% shallower than the 3001 ft already on record
(NDOM-sourced), and `IW-2`'s 1990 report depth (1403 ft) is less than a
third of the 4700 ft on record -- both almost certainly explained by
subsequent deepening of these still-active wells, not a data error.
Rather than overwrite (or awkwardly average) these, the 1990 construction
facts are recorded in `Wells.notes` as explicitly historical, and
`top_perforation`/`bottom_perforation` were left untouched for both.
Every other well's numbers were close enough to (or had no) existing
`total_depth` to fill with confidence.


## Newly ingested chemistry: Table 1 (1950-1991)

`ingest_historical_sorey1992.R` (wired into `run_pipeline.R` as
`RUN_INGEST$historical_sorey1992`) transcribes Table 1's ~10 dated
analyses directly into `Lab_Analyses`, extending this project's own
chemistry record back to 1950 (well GS-58) for the first time.

```{r sorey-chem-table}
sorey_chem <- dbGetQuery(con, "
  SELECT l.name AS feature, s.collection_time AS sample_date,
         la.analyte, la.value, la.units
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE l.external_station_code LIKE 'SOREY1992%'
  ORDER BY s.collection_time, l.name
")
sorey_chem_wide <- sorey_chem %>%
  select(feature, sample_date, analyte, value) %>%
  pivot_wider(names_from = analyte, values_from = value)
sorey_chem_wide
```

### Cl/B and Cl/Li ratio check against modern chemistry

Sorey & Colvard found Cl/B = 19.3 +/- 1.7 and Cl/Li = 122.5 +/- 9.9
across the CPI and SB GEO reservoirs (excluding the more dilute Cox
well), and used this constancy of conservative-element ratios to argue
the two well fields draw on chemically related source water. Recomputing
the same ratios from the newly-ingested 1950-1991 values, and comparing
against this project's own real 2024-2026 thermal-water samples (the
FIELD spring/well/seep samples, and the NDEP-promoted domestic wells),
is a direct test of whether that same geochemical signature has
persisted for over three decades.

```{r ratio-check}
hist_ratios <- sorey_chem %>%
  filter(analyte %in% c("Cl", "B", "Li")) %>%
  pivot_wider(id_cols = c(feature, sample_date), names_from = analyte, values_from = value) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li, era = "1950-1991 (Sorey & Colvard 1992)",
         year = as.integer(substr(sample_date, 1, 4)))

modern_ratios <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte IN ('Cl','B','Li')
    AND s.data_source != 'Sorey & Colvard 1992'
") %>%
  pivot_wider(id_cols = c(sample_id, feature, site_type, collection_time), names_from = analyte,
              values_from = value, values_fn = mean) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li,
         era = if_else(site_type %in% c("fumarole", "seep", "spring"),
                        "2024-2026 real thermal (FIELD)",
                        "2024-2026 NDEP domestic/background wells"),
         year = lubridate::year(lubridate::mdy_hm(collection_time, quiet = TRUE)))  # real parsed sample year (2024 or 2026)

ratio_compare <- bind_rows(
  hist_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li),
  modern_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li)
)
ratio_compare
```

```{r ratio-plot}
#| fig-width: 8
#| fig-height: 5
# Single unified plot (year on the x-axis, not era-on-y as in an earlier
# draft) -- well-name labels via ggrepel, and the t-test statistics moved
# into the figure caption rather than left as a separate prose paragraph,
# per explicit feedback this session. Excludes the Cox well (per Sorey &
# Colvard's own stated convention) and the SBRR/SBBV creek stations
# (mixing points, not end-member samples -- including them the first time
# this was tried pulled in SBRR's near-zero-B ratio and badly distorted
# the comparison).
cl_b_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "SB GEO production well PW-1" = "PW-1", "SB GEO production well PW-2" = "PW-2",
  "SB GEO production well PW-3" = "PW-3", "Hot spring 6, main terrace ACEC" = "Spring 6",
  "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001", "SBW_0002" = "SBW_0002",
  "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003"
)
ratio_plot_data <- ratio_compare %>%
  filter(!is.na(Cl_B), !grepl("Cox", feature), !feature %in% c("SBRR", "SBBV")) %>%
  mutate(
    era_short = if_else(grepl("^19", era) | era == "1950-1991", "1990-91", "2024-26"),
    label = coalesce(cl_b_labels[feature], feature)
  )
t_res_ratio <- t.test(Cl_B ~ era_short, data = ratio_plot_data)
caption_ratio <- sprintf(
  "Welch two-sample t-test: mean Cl/B = %.1f (1990-91, n=%d) vs. %.1f (2024-26, n=%d); t(%.1f) = %.2f, p = %.3f, 95%% CI [%.2f, %.2f]",
  t_res_ratio$estimate[1], sum(ratio_plot_data$era_short == "1990-91"),
  t_res_ratio$estimate[2], sum(ratio_plot_data$era_short == "2024-26"),
  t_res_ratio$parameter, t_res_ratio$statistic, t_res_ratio$p.value,
  t_res_ratio$conf.int[1], t_res_ratio$conf.int[2]
)
ggplot(ratio_plot_data, aes(x = year, y = Cl_B, color = era_short)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20, seed = 4172, show.legend = FALSE) +
  labs(x = "Year", y = "Cl / B (mass ratio)", color = "Era",
       title = "Chloride-to-boron ratio in real thermal samples, 1950-2026",
       caption = caption_ratio) +
  theme(plot.caption = element_text(hjust = 0, size = 9))
```

**Reading this (revised 2026-09-12 with a real statistical test -- see
the new Statistical Analysis section below):** this project's own 2024-2026 real thermal FIELD
samples (fumarole/seep/spring) give Cl/B roughly 20-23, in the same
general range as Sorey & Colvard's 1990-91 CPI/SB GEO value of 19.3 +/-
1.7. A Welch two-sample t-test on the 7 historical vs. 6 modern real
thermal Cl/B values (excluding the Cox well, per Sorey & Colvard's own
convention, and excluding the SBRR/SBBV creek stations, which are
mixing points, not end-member samples) finds this difference is
**small but statistically significant**: mean 19.6 (1950-1991) vs. 21.9
(2024-2026), Welch t(9.87) = -2.92, p = 0.015, 95% CI on the difference
[-4.03, -0.54]. This is a more precise, and slightly different,
statement than this notebook's original (Session 27) "closely
matching... essentially unchanged" framing -- the conservative-element
signature is genuinely similar in magnitude across 35 years, but the
two eras are not statistically indistinguishable. A plausible, not yet
tested, explanation is a systematic difference in analytical method
(UURI/AMTEC ICP for the 1990-91 samples vs. this project's own modern
lab) rather than a real 12% geochemical shift -- flagged as an open
question, not resolved here. The NDEP domestic/background well samples
(Boyd, Jeppson, Rogers) show much more scattered Cl/B ratios; their B
values sit at or near 0.05-0.18 mg/L, close to a plausible analytical
detection limit, so those ratios should be read as noisy near-DL
artifacts of dilute, non-thermal ground water rather than a real
geochemical signal -- fully consistent with this project's established
finding that these are background, not thermal, waters.

## Discharge-through-time: extending Sorey's Table 2

Table 2 of the 1992 report tabulates dated estimates of thermal-water
discharge to Steamboat Creek from 1955 through 1989. These values are
**transcribed literature numbers, not live database queries** -- they
predate any table this project's own pipeline builds -- assembled here
alongside the already-cited Sorey & Spielman (2008, 2017) follow-up
estimates and this project's own 2026 poster result to give a single,
~70-year view.

```{r discharge-timeline}
# Extended 2026-09-25: added a new database-computed point (this project's
# only real paired Cl+discharge measurement, 2026-05-01) and a second
# Collar & Huntley (1990) point (their own stated "about 80% of White's
# 1955 baseline" 1988-89 natural-only estimate, distinct from the two
# dated Table 2 values already used).
discharge_hist <- tribble(
  ~period, ~year_mid, ~gpm, ~Ls, ~source,
  "White (1968), Apr 1955",         1955, 1110, 1110 * 0.0630902, "White (1968), chloride-flux method",
  "White (1968), Apr 1964",         1964, 1385, 1385 * 0.0630902, "White (1968), chloride-flux method",
  "Shump (1985), 1981-82 avg",      1981.5, 1300, 1300 * 0.0630902, "Shump (1985), specific-conductance method (less reliable per Sorey & Colvard)",
  "Collar (1990), Jun 1988",        1988.5, 523, 523 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Collar (1990), Aug 1988",        1988.6, NA, NA, "Not counting net CPI production; unseen seepage component only reported",
  "Collar (1990), Mar 1989",        1989.2, 663, 663 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Collar & Huntley (1990), 1988-89 natural-only", 1988.8, NA, 0.80 * 1120 * 0.0630902, "Collar & Huntley (1990): 'about 80% of that estimated by White (1968) in 1955' (1120 gpm baseline, natural discharge only, excludes CPI production)",
  "Sorey & Spielman (2008), 2007-08", 2007.5, NA, mean(c(730, 1214)) * 0.0630902, "Chloride-flux, SBRR/SBBV sites",
  "Sorey & Spielman (2017), 2008-2016 avg", 2012, NA, 60, "Chloride-flux, SBRR/SBBV sites, 24 sampling events",
  "This project (database), 5/1/2026", 2026.33, NA, 42.1, "Computed this session directly from real paired Cl (SBRR=17.1, SBBV=126 mg/L, both 2026-05-01) and USGS discharge (station 10349300, 11.2 cfs near the SBRR sample time) -- same formula as the poster estimate below, but a single day, not an averaging period",
  "Irvin & Lindsey (2026 poster), post-eruption avg", 2026, NA, 27, "This project's own chloride mass-balance estimate (averaged over its own sampling period, not a single day)"
)
discharge_hist
```

```{r discharge-plot}
#| fig-width: 9
#| fig-height: 5.5
ggplot(discharge_hist %>% filter(!is.na(Ls)), aes(x = year_mid, y = Ls)) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = grepl("This project", period)), size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(aes(label = period), size = 2.6, max.overlaps = 30, seed = 4172) +
  scale_color_manual(values = c(`TRUE` = "#d95f02", `FALSE` = "black")) +
  labs(x = "Year", y = "Thermal-water discharge (L/s)",
       title = "Thermal-water discharge to Steamboat Creek, 1955-2026",
       subtitle = "No confidence intervals shown -- no source reports a measurement uncertainty. Orange point is a\nnew, reproducible single-day estimate computed directly from this project's own database this session.")
```

**Reading this:** discharge estimates were broadly 45-90 L/s
(660-1,385 gpm-equivalent) from 1955 through the 2008-2016 Sorey &
Spielman baseline, with the 1988-89 Collar (1990) values sitting at the
low end even before accounting for that they explicitly exclude net CPI
well-field production. This project's own post-eruption ~27 L/s estimate
sits below every prior estimate in this 70-year record -- consistent
with the poster's own conclusion (conduit reactivation/redistribution,
not necessarily a new source) but worth treating as one point in a
long, variable record rather than a single before/after comparison.
None of the pre-2007 methods used the same sites (SBRR/SBBV) as the
modern chloride-flux estimates, so method consistency, not just
magnitude, matters when reading this chart -- White (1968) and Collar
(1990)'s methods and gauge locations are not identical to Sorey &
Spielman's.

**Seasonal Cl-flux variability -- a real, previously-published effect
worth naming here (2026-09-25).** Sorey & Spielman (2017) -- already
cited above, not a new document -- explicitly document a real seasonal
pattern in their own SBRR/SBBV Cl-flux computations: *"lower computed
chemical flux at downstream sites in SB Creek each summer and higher
chemical loads in creek water in the wintertime runoff period,"*
attributed in part to seasonal buildup and flushing of chloride in
shallow soils from irrigation/evaporation cycles downstream of Crane
Ditch diversions -- though they state *"the magnitude of such seasonal
changes appears to be only on the order of 25% or less."* This is a
plausible, real contributor to the gap between this session's new
single-day database estimate (42.1 L/s, 2026-05-01, spring runoff
season) and the poster's own averaged ~27 L/s figure -- a ~56% relative
difference, larger than Sorey & Spielman's own documented ~25% seasonal
envelope, so seasonality alone does not fully explain it, but it is a
real, published, and directly relevant effect rather than pure noise.
This also matches this project's own poster, which discusses Cl outflow
variability in the diluted creek -- worth a dedicated seasonal-Cl-flux
analysis once more paired Cl+discharge dates exist in the database (a
single day, as computed here, cannot separate a real seasonal signal
from ordinary day-to-day variability).

## Appendix G and Table 9: what's behind the discharge/reservoir numbers (2026-09-12)

### Appendix G: the raw synoptic Cl-flux surveys behind Table 2

Appendix G ("Stream Discharge and Chloride Flux") documents the real
method behind the Collar (1990) discharge estimates already used in the
discharge-through-time chart above: chloride treated as a conservative
tracer, an 820 mg/L thermal / 6 mg/L non-thermal end-member split, and
gaining/losing-reach mass-balance equations -- methodologically almost
identical to this project's own SBRR/SBBV Cl mass-balance approach, just
predating it by ~35 years. Chloride was measured by mercuric-nitrate or
argentometric titration (SDSU and UURI, respectively); discharge by
pygmy current meter. The underlying raw data (Table G-1) comes from four
real synoptic surveys: **6/26-6/30/1988, 7/1-7/2/1988, 8/9/1988, and
3/4/1989**, at ~40 named stations along Steamboat Creek, Galena Creek,
and the Steamboat/Chandler/Crane irrigation ditches.

**Deliberately not parsed further**: Table G-1's raw per-station
discharge/chloride/boron values are visibly OCR-misaligned in the
extracted text (values shifted relative to station-name rows, in some
blocks by a full row) -- the same class of "real data trapped in a
badly-OCR'd table" problem already flagged elsewhere in this project
(the NDEP TFT Appendix D geochemistry tables, well-log lithology
tables). The three aggregate numbers we actually use (Collar 1990's
1988-89 discharge estimates in the Table 2 chart above) are already the
synthesized output of this same appendix -- Appendix G is the raw
survey behind data already in this notebook, not a source of new,
different data points. If a denser historical Cl-flux time series is
ever wanted, it would need to be read directly from the scanned page
images (not the OCR text) station-by-station -- a real, larger effort
than anything else in this session, not attempted here.

### Table 9: reservoir parameters (transmissivity, storage coefficient)

Table 9 ("Reservoir parameters determined for the geothermal system in
the Steamboat Hills area from well test analyses and spring
hydrographs") gives real aquifer-test results from four dated tests:

```{r table9-reference}
table9 <- tibble::tribble(
  ~well_and_year, ~injection, ~transmissivity_ft2_day, ~storage_coefficient, ~observation_well, ~source,
  "Steamboat No.1, 1980", "no", "3270-9500", "1.2e-3 (calc.)", "strat 2", "Collar (1990)",
  "Steamboat No.1, Mar-May 1986", "no", "nd", "nd", "23-5?", "Berkeley Group (1987)",
  "Steamboat No.1, Mar-May 1987", "no", "nd", "1.2e-3 (calc.)", "strat 5", "Berkeley Group (1987)",
  "23-5/Cox 1-1, May-June 1987", "yes", "7930-9500", "7.8e-4 to 9.0e-4", "strat 2, 21-5", "Faulder (1987); Goranson (1989)",
  "21-5, 23-5, 83A-6, 1988", "yes", "1250-2400", "2.5e-3", "strat 9", "Collar (1990)",
  "21-5, 23-5, 83A-6, 1988 (spring response)", "yes", "nd", "2.8e-3", "spring 12, spring 42w", "Collar (1990)",
  "Spring 42w, Jun-Aug 1988 (Birsoy-Summers method)", "yes", "3000", "2.6e-3", "spring 42w", "Collar & Huntley (1990)"
)
table9
```

No table in the current schema models a dated aquifer/interference
test -- this is kept as reference documentation only, not a new
database table, since there is no immediate analytical use for it beyond
the qualitative reservoir-drawdown narrative already summarized above
("What the 1992 report says"). Worth revisiting if this project's own
hydraulic-gradient work ever needs a real transmissivity/storage
reference range for the Steamboat Hills reservoir.

### 4.6 Collar & Huntley (1990): a fault map and a discharge-deficit statistic (2026-09-12)

A second Sorey-team document, `Collar.pdf` (Collar, R.J. and Huntley, D.
1990, 12th New Zealand Geothermal Workshop), was added to the literature
folder and read directly. It largely overlaps the 1992 report above (same
wells, same hydrograph story), but contributes two things not already
captured:

**A real fault map.** Its Figure 1 ("Fault and air photo lineament map of
the Steamboat Springs geothermal area, including the location of selected
wells") shows mapped faults (dashed = inferred, dotted = concealed) and
air-photo lineaments of probable structural origin, including the named
Pleasant Valley Fault and Steamboat Springs Fault Zone, with `21-5`,
`83A-6`, `23-5`, `Cox I-1`, `PW-1/2/3`, `IW-2/3`, strat wells, `Brown
School`, `Pine Tree Ranch`, `Steinhardt`, and `GS-1`/`GS-8` all located on
it by name. Ten of these already have real coordinates in this database:

```{r collar-fig1-control-points}
dbGetQuery(con, "
  SELECT well_name, latitude, longitude FROM Wells
  WHERE well_name IN ('21-5R','23-5','83A-6','Cox I-1','PW-2','PW-3','IW-2',
                       'Brown School Geothermal Well','Steinhardt Geothermal Well','GS-5')
  ORDER BY well_name
")
```

This is a materially better fault-digitizing candidate than the
Dhakal-diagram source discussed in `README.md`'s "Planned: Leapfrog 3D
geologic model export" section (a production/injection flow diagram, not
a fault map) -- see that section (updated 2026-09-12) for the proposed
digitizing approach using these ten points as control points.

**A discharge-deficit statistic.** The paper states that 1988 spring
discharge, measured at 3-4 gal/min, was far below the 34-40 gal/min
*predicted* from White (1968)'s own spring-discharge/annual-precipitation
relationship -- a specific, citable number distinct from (and more
precise than) the aggregate gpm figures already in the Section 6
discharge-through-time chart.

**Elevation cross-check.** The paper gives well-field elevations (OESI/
SB GEO ~4700 ft; CPI 5350-5660 ft; Cox I-1 5050 ft; springs ~4650 ft),
checked against this project's own `Wells.elevation_m`:

```{r elevation-crosscheck}
dbGetQuery(con, "
  SELECT well_name, elevation_m, ROUND(elevation_m/0.3048,0) AS elevation_ft
  FROM Wells WHERE well_name IN ('PW-2','PW-3','IW-2','21-5R','Cox I-1')
  ORDER BY well_name
")
```

`IW-2`/`PW-2`/`PW-3` (4634-4730 ft) are consistent with the paper's OESI
well-field figure (~4700 ft). `21-5R` (5722 ft) sits just above the
paper's stated CPI range (5350-5660 ft) -- a minor, unresolved
discrepancy, plausibly because the paper gives one representative
figure for the whole field rather than a per-well value. `Cox I-1` had
no recorded elevation at all; the paper's specific value (5050 ft =
1539.24 m) is filled in below since, unlike the CPI range, it is
directly attributable to a single named well.



## Table 10 monitoring network vs. this project's Locations

```{r table10-check}
dbGetQuery(con, "
  SELECT location_id, external_station_code, name, site_type,
         latitude, longitude, coordinate_source
  FROM Locations
  WHERE name LIKE '%Boyd%' OR name LIKE '%Curti%' OR name LIKE '%Rogers%'
     OR name LIKE '%Jeppson%' OR name LIKE '%Herz%' OR name LIKE '%Zolezzi%'
     OR external_station_code IN ('SB5','SB6')
  ORDER BY name
")
```

Sorey & Colvard's Table 10 ("Current monitoring sites for CPI and SBG
geothermal developments") describes exactly this same set of domestic
wells and the Rhodes Road creek-monitoring point as the active 1988-92
monitoring network -- confirming this project's own SBRR site and its
NDEP/Klein-resolved domestic wells sit at genuinely long-monitored
locations, not newly-chosen ones. None of these have a linked `Wells`
row or `Water_Level_Observations` history yet, which is a real,
documented gap (see Section 7 caveats) rather than an oversight to
silently work around here.

**Aside (found and fixed, 2026-09-12): the malformed `"7-07-10"`
timestamp.** `Water_Level_Observations` has 73 wells and 153,437 rows
spanning 1969-06-23 to 2026-01-01; one row (well-log document 104216, a
"WELL DRILLER'S PLUGGING REPORT" with `completion_date_raw = "7/10/07"`)
had timestamp `"7-07-10"`. Root cause: `ingest_well_logs.R`'s
`.safe_completion_date()` tries `as.Date(x, format = "%m/%d/%Y")` first,
and R's `%Y` silently accepts a 2-digit year as a literal, unpadded year
value (year 7 CE) rather than rejecting the format mismatch the way the
function's design assumed it would -- so the very first format "matched"
with a nonsensical date instead of falling through to a better one.
Fixed by adding an explicit `%m/%d/%y` attempt and a plausible-year-range
sanity check (1900 to next calendar year) that every candidate parse must
pass before being accepted; the one bad row is corrected to `2007-07-10`.
No other row in `Well_Log_Documents`/`Water_Level_Observations` was
affected -- this was the only 2-digit-year-formatted completion date in
the batch.

## Statistical Analysis: Is There a Real Chloride Trend? (2026-09-12)

Sections above establish real Cl values across 1950-2026 and a
qualitative sense that thermal-water chemistry looks broadly similar
over time. This section applies actual statistical tests to that claim
-- per an explicit request that this notebook's historical-comparison
work be statistic-driven, not just descriptive.

### A combined 1950-2026 Cl dataset

```{r cl-timeline-data}
library(lubridate)
cl_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, s.data_source, la.value AS Cl
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte = 'Cl'
")

# Sampling_Events/Samples.collection_time mixes Unix-epoch-seconds,
# MM/DD/YYYY H:MM, and ISO YYYY-MM-DD formats across sources -- the
# same class of dual/mixed-format issue already flagged project-wide
# for Sampling_Events.date (notebook 05). Parsed here defensively,
# trying each format in turn rather than assuming one.
.parse_any_date <- function(x) {
  num <- suppressWarnings(as.numeric(x))
  out <- as.Date(rep(NA, length(x)))
  is_num <- !is.na(num) & !grepl('[A-Za-z]', x)
  out[is_num] <- as.Date(as.POSIXct(num[is_num], origin = '1970-01-01', tz = 'UTC'))
  rem <- !is_num
  out[rem] <- suppressWarnings(as.Date(x[rem], format = '%m/%d/%Y %H:%M'))
  rem2 <- rem & is.na(out)
  out[rem2] <- suppressWarnings(as.Date(x[rem2], format = '%Y-%m-%d'))
  out
}
cl_data$date <- .parse_any_date(cl_data$collection_time)
cl_data$year <- year(cl_data$date)

# Added 2026-09-25: Steinhardt's two narrative-based bracket samples
# (1987/1990, see Section 4.2 correction above), now that they exist.
focus_wells <- c(
  'CPI production well 21-5','CPI production well 83A-6','CPI production well 23-5',
  'CPI injection well Cox 1-1 (Cox well)','SB GEO production well PW-1','SB GEO production well PW-2',
  'SB GEO production well PW-3','Well GS-5, 1950 sample (White 1968)','New seep adjacent to well GS-5 (1991)',
  'Hot spring 6, main terrace ACEC',
  'SBF_0001','SBW_0001','SBW_0002','SBS_0001','SBS_0002','SBS_0003',
  'Eich Well','Soccer Field Monitoring Well','Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well',
  'Steinhardt Geothermal Well (narrative Cl trend)'
)
background_names <- c('Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well')
cl_focus <- cl_data[cl_data$feature %in% focus_wells, ]
cl_focus$group2 <- ifelse(cl_focus$feature %in% background_names, 'background/domestic',
                    ifelse(cl_focus$feature %in% c('Eich Well','Soccer Field Monitoring Well'), 'intermediate (NDEP)', 'thermal'))

well_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "CPI injection well Cox 1-1 (Cox well)" = "Cox I-1", "SB GEO production well PW-1" = "PW-1",
  "SB GEO production well PW-2" = "PW-2", "SB GEO production well PW-3" = "PW-3",
  "Well GS-5, 1950 sample (White 1968)" = "GS-5", "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "Hot spring 6, main terrace ACEC" = "Spring 6", "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001",
  "SBW_0002" = "SBW_0002", "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003",
  "Eich Well" = "Eich", "Soccer Field Monitoring Well" = "Soccer Field", "Boyd Domestic Well" = "Boyd",
  "Jeppson Geothermal Well" = "Jeppson", "Rogers Domestic Well" = "Rogers",
  "Steinhardt Geothermal Well (narrative Cl trend)" = "Steinhardt"
)
cl_focus$label <- well_labels[cl_focus$feature]
nrow(cl_focus)
```

```{r cl-timeline-plot}
#| fig-width: 9
#| fig-height: 5.5
# Reworked 2026-09-25 per explicit feedback: well-name labels (not just
# color groups), lines connecting repeat samples of the same well/spring
# through time, and the 6 single-snapshot 2026 FIELD thermal samples
# (SBF_0001, SBW_0001/2, SBS_0001-3) dropped from THIS plot -- they add
# no temporal information (all one date) and were crowding the 2020s
# cluster; they remain visible in the Cl/B ratio plot above.
exclude_2026_single <- c("SBF_0001","SBW_0001","SBW_0002","SBS_0001","SBS_0002","SBS_0003")
cl_focus_timeline <- cl_focus[!cl_focus$feature %in% exclude_2026_single, ]
# 2026-09-26: faceted by well vs. spring/seep per explicit request --
# almost everything in this curated focus set is a well; GS-5 seep and
# Hot spring 6 are the only spring/seep entries, but keeping them in a
# separate facet panel avoids visually implying they belong to the same
# "population" as the production/domestic wells.
cl_focus_timeline$site_category <- ifelse(cl_focus_timeline$site_type %in% c("spring","seep"),
                                            "Springs & Seeps", "Wells")
ggplot(cl_focus_timeline, aes(x = date, y = Cl, color = group2)) +
  geom_line(aes(group = feature), alpha = 0.4, linewidth = 0.5, show.legend = FALSE) +
  geom_point(size = 2.5) +
  ggrepel::geom_text_repel(aes(label = label), size = 3.2, max.overlaps = 20, seed = 4172) +
  scale_y_log10() +
  facet_wrap(~ site_category, scales = "free_x") +
  labs(x = NULL, y = "Chloride (mg/L, log scale)", color = "Sample group",
       title = "Chloride concentration by well/spring through time, 1950-2026",
       subtitle = "Faceted by well vs. spring/seep; wells with real repeat measurements over time (Steinhardt,\nBoyd, Jeppson, Rogers, Eich, Soccer Field) plus single historical thermal anchors for context",
       caption = "The 6 single-snapshot 2026 FIELD thermal samples (SBF_0001, SBW_0001/2, SBS_0001-3) are shown separately in the Cl/B ratio plot above to reduce crowding here.")
```

### Is there a statistically significant Cl trend over time?

A naive linear regression of Cl against year across every
thermal-influenced sample (thermal + intermediate groups, n = 19)
finds a significant negative slope:

```{r cl-trend-naive}
fit_all <- lm(Cl ~ year, data = cl_focus[cl_focus$group2 != 'background/domestic', ])
summary(fit_all)
```

**But this is confounded**, and re-running it after excluding the two
'intermediate' NDEP wells (Eich, Soccer Field -- only ever sampled in
2024, and already established elsewhere in this notebook as
partially-diluted, not pure end-member thermal water) weakens the
trend to marginal significance:

```{r cl-trend-pure}
pure_thermal <- cl_focus[cl_focus$group2 == 'thermal', ]
fit_pure <- lm(Cl ~ year, data = pure_thermal)
summary(fit_pure)

pure_thermal$era <- ifelse(pure_thermal$year < 2000, 'pre-2000', '2024-2026')
aggregate(Cl ~ era, data = pure_thermal, FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
```

**Honest reading:** including the intermediate wells (p = 0.0099,
R2 = 0.33) versus excluding them (p = 0.099, R2 = 0.20) changes the
conclusion from 'significant decline' to 'suggestive but not
significant at the conventional 0.05 threshold' -- purely because of
which samples are grouped as 'thermal.' The pre-2000 real-thermal mean
(811 mg/L, n=9) is numerically higher than the 2024-2026 mean (699
mg/L, n=6), but with only 15 points spanning 76 years and no repeated
measurements of the same individual well across both eras, this
dataset does not have the statistical power to distinguish a real
long-term decline from era-to-era sampling differences (different
wells sampled in 1990-91 than in 2024-2026). This is reported as an
open, unresolved question -- not a confirmed trend -- consistent with
this notebook's standing practice of not overstating what a small
historical dataset can support.

### A cluster-aware re-analysis: hydrochemical facies (2026-09-26)

The naive/pure-thermal regression above pools every "thermal" sample
regardless of how similar their overall water chemistry actually is.
`scripts/analysis/cluster_hydrochemical_facies.R` takes a more
defensible approach: cluster every real sample with complete major-ion
chemistry (Na, K, Ca, Mg, Cl, SO4, Alkalinity -- the seven analytes
every source in this database reports in common) into hydrochemical
facies, and only then test for a temporal trend *within* a facies,
rather than across an ad hoc "thermal" label decided by hand.

**Method**: log1p-transform and z-score each of the 7 major ions,
then hierarchically cluster (Ward's method, Euclidean distance). The
number of clusters is chosen objectively by silhouette width (not
picked by eye), and cross-checked against an independent Gaussian-
mixture fit (`mclust`) -- if the two disagree substantially, that
disagreement is reported rather than silently resolved.

```{r facies-clustering}
#| fig-width: 7
#| fig-height: 5.5
facies_script <- file.path("scripts", "analysis", "cluster_hydrochemical_facies.R")
if (!file.exists(facies_script)) facies_script <- file.path("..", facies_script)
source(facies_script)
fc <- run_facies_clustering(con)
fc$plot_silhouette
fc$plot_cluster
fc$site_type_summary
```

Of 846 chemistry-bearing samples in the database, only 157 have all
seven core majors measured (no imputation is attempted -- this is a
real, stated constraint on statistical power). Silhouette width
selects k = `r fc$k` clusters, with `r fc$agreement_pct`% row-wise
agreement between the hierarchical and Gaussian-mixture fits -- a
moderate, not perfect, agreement, reported as-is.

The four facies are chemically interpretable, not arbitrary: a dilute
background-water cluster (median Cl < 1 mg/L), an intermediate/mixed
cluster, a pure Na-Cl thermal cluster (median Cl ~ 800 mg/L, low
Ca/Mg), and a distinct, elevated-Ca/Mg "diluted-and-water-rock-
reacted" cluster containing Brown School, Curti Domestic Well, Eich
Well, and this project's own 2026 SBW_0001 sample together -- which is
exactly the Ca-Na-Cl albitization signature Mariner & Janik (1995)
described narratively for the Brown's School well (Section 4 above),
now independently recovered by unsupervised clustering on modern *and*
historical samples together. Notably, the 1991-94 "Stuart" domestic
well -- chemically dilute in absolute terms (Cl 64-86 mg/L) -- clusters
with the *concentrated* pure-thermal group, not the background group:
its ion *ratios* (Na-dominant, low Ca/Mg) match the thermal signature
even though its magnitude is heavily diluted, consistent with Mariner
& Janik's own text describing Stuart as "highly diluted thermal
water." This is a real, substantive validation of the clustering
approach, not just a plotting exercise.

**Within-facies temporal trend**: restricting to the pure-thermal
facies and fitting a mixed-effects model (well as a random intercept,
year as the fixed effect) -- rather than pooling naively -- gives a
properly-powered estimate that doesn't let heavily-repeat-sampled
wells dominate:

```{r facies-glmmTMB}
library(glmmTMB)
c3 <- fc$data %>% dplyr::filter(facies_cluster == fc$data$facies_cluster[which.max(fc$data$Cl)]) # pure-thermal = highest-Cl cluster
# (re-derive the pure-thermal cluster id robustly rather than hardcoding "3")
thermal_cluster_id <- fc$data %>%
  group_by(facies_cluster) %>%
  summarise(med_cl = median(Cl)) %>%
  slice_max(med_cl, n = 1) %>%
  pull(facies_cluster)
c3 <- fc$data %>% dplyr::filter(facies_cluster == thermal_cluster_id)

parse_date_flex <- function(x) {
  x <- as.character(x)
  out <- as.Date(rep(NA, length(x)))
  is_num <- grepl("^[0-9]+(\\.0)?$", x)
  out[is_num] <- as.Date(as.numeric(sub("\\.0$", "", x[is_num])) / 86400, origin = "1970-01-01")
  rest <- !is_num & !is.na(x)
  d1 <- suppressWarnings(as.Date(x[rest], format = "%m/%d/%Y %H:%M"))
  d2 <- suppressWarnings(as.Date(x[rest], format = "%m/%d/%Y"))
  d3 <- suppressWarnings(as.Date(x[rest], format = "%Y-%m-%d"))
  out[rest] <- dplyr::coalesce(d1, d2, d3)
  out
}
c3$date <- parse_date_flex(c3$collection_time)
c3$year_frac <- as.numeric(format(c3$date, "%Y")) + (as.numeric(format(c3$date, "%j")) - 1) / 365.25
c3$well_label <- sub(" \\(Mariner.*", "", c3$location_name)

repeat_wells <- c3 %>% dplyr::count(well_label) %>% dplyr::filter(n >= 2) %>% dplyr::pull(well_label)
mod_dat <- c3 %>% dplyr::filter(well_label %in% repeat_wells, !is.na(date))
mod_dat$year_c <- mod_dat$year_frac - mean(mod_dat$year_frac)

m1 <- glmmTMB(log(Cl) ~ year_c + (1 | well_label), data = mod_dat)
summary(m1)
```

**Result**: once between-well baseline chemistry is properly absorbed
by a random intercept (n = `r nrow(mod_dat)` observations across
`r length(unique(mod_dat$well_label))` repeat-sampled thermal-facies
wells spanning 1991-2026), there is **no statistically significant
temporal trend** in Cl within this facies (year coefficient on the log
scale, p > 0.05). This revises the naive/pure-thermal regression's
"suggestive but not significant" reading above to a cleaner null
result once the analysis stops conflating between-well and within-well
variation -- consistent with (not contradicting) that earlier,
honestly-hedged conclusion, just on firmer statistical footing. As
with the naive regression, this remains limited by the fact that few
individual wells were sampled in *both* the 1990s and 2020s, so the
model still can't fully separate a true multi-decade trend from
era-to-era differences in which wells happened to be sampled -- flagged
here again rather than glossed over.

### Spatial overlay: facies clusters mapped onto well positions with approximate depth (2026-09-26)

The clustering above is purely chemical; it says nothing about where
these facies sit in the field or how deep they're being produced from.
`scripts/analysis/facies_depth_map.R` overlays each well/location's
majority facies cluster onto its real map position, sized by an
approximate reservoir/completion depth.

```{r facies-depth-map}
#| fig-width: 8.5
#| fig-height: 6.5
depth_map_script <- file.path("scripts", "analysis", "facies_depth_map.R")
if (!file.exists(depth_map_script)) depth_map_script <- file.path("..", depth_map_script)
source(depth_map_script)
map_result <- plot_facies_depth_map(con, fc)
map_result$plot
```

**Depth is a coarse proxy, not a real geologic model**: it's
`midpoint(top_perforation, bottom_perforation)` where a real screened
interval exists, falling back to `total_depth` where it doesn't, and
left as an explicit hollow point where neither exists. As of this
writing that's only 5 of 39 wells shown -- most PW2-x/PW3-x wells (real
coordinates, no depth record) and every Locations-only entity (Eich,
Soccer Field, Boyd, Jeppson, Rogers, Curti, Steinhardt -- domestic wells
promoted from NDEP chemistry, never registered in the `Wells` table with
a depth) have no depth on record at all. This is a real, stated data
gap, not a plotting artifact -- a genuine 3D subsurface model (real
lithology + fault-bounded compartments) is exactly what the planned
Leapfrog export is meant to eventually support once shallow-layer and
fault data exist (see README.md).

Even with that limitation, the map is informative: the pure-thermal
facies (cluster 3 in the run above) sits tightly clustered around the
CPI/SB GEO production wells near the historic terraces, while the
background/intermediate facies fall at the field's margins -- a spatial
pattern consistent with (not independently discovered beyond) the
chemistry-only clustering result above.


### Cl/B ratio: a real two-sample test (not just eyeballing)

The Cl/B ratio comparison introduced earlier in this notebook is
repeated here as an explicit, reproducible statistical test rather
than a visual read of the plot:

```{r cl-b-test}
cl_b_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la JOIN Samples s ON s.sample_id=la.sample_id JOIN Locations l ON l.location_id=s.location_id
  WHERE la.analyte IN ('Cl','B') AND s.data_source != 'NDEP'
")
cl_b_wide <- reshape(cl_b_data[, c('sample_id','feature','collection_time','analyte','value')],
                      idvar = c('sample_id','feature','collection_time'), timevar = 'analyte', direction = 'wide')
names(cl_b_wide) <- gsub('value\\.', '', names(cl_b_wide))
cl_b_wide <- cl_b_wide[!is.na(cl_b_wide$Cl) & !is.na(cl_b_wide$B), ]
cl_b_wide$Cl_B <- cl_b_wide$Cl / cl_b_wide$B
cl_b_wide$year <- year(.parse_any_date(cl_b_wide$collection_time))
# Exclude Cox (per Sorey & Colvard's own stated convention) and the
# SBRR/SBBV creek stations (mixing points, not end-member samples --
# including them wrongly pulled in a Cl/B ratio near 1500 dominated by
# SBRR's near-zero B).
cl_b_wide <- cl_b_wide[!grepl('Cox', cl_b_wide$feature) & !cl_b_wide$feature %in% c('SBRR','SBBV'), ]
cl_b_wide$era <- ifelse(cl_b_wide$year < 2000, '1950-1991', '2024-2026')
cl_b_wide[, c('feature','year','Cl','B','Cl_B','era')]

t.test(Cl_B ~ era, data = cl_b_wide)
```

This confirms the result already folded into the "Cl/B and Cl/Li ratio
check" section above: a small (19.6 vs. 21.9) but real, statistically
significant (p = 0.015) difference between the two eras -- similar in
magnitude, not statistically identical.

### Discharge-estimate uncertainty: an honest limitation, not a fabricated CI

The discharge-through-time chart earlier in this notebook (White 1968,
Shump 1985, Collar 1990, Sorey & Spielman 2008/2017, this project's
2026 poster) plots point estimates with no confidence interval. This
is deliberate, not an oversight: none of the source documents report a
measurement-uncertainty or replicate-based error bar for their
discharge estimates, and the raw synoptic measurements behind the
pre-2007 estimates are not available in a form this project can
re-derive an uncertainty from (Appendix G's per-station table, as
documented above, is real but too OCR-misaligned to trust for this).
Fabricating a confidence interval from nothing would be worse than
reporting none. A future, real improvement would be to propagate this
project's own analytical Cl-measurement uncertainty (a stated lab
detection-limit/precision figure, not yet compiled here) through the
chloride mass-balance formula for the SBRR/SBBV-based estimates only,
where the raw measurements are directly available -- flagged as a
concrete next step, not attempted in this session.

## Barometric efficiency: now a real (not synthetic) calculation

Sorey & Colvard (p. 85-88, Figs. 39-40) computed **barometric
efficiency** (BE) -- the ratio of water-level change in a well/spring to
the corresponding barometric-pressure change -- by linear regression of
spring water level against Reno Airport barometric pressure, from 3-4x
daily manual readings over one week in October 1988. Reported values:
BE = 0.42 for spring 6 (r = 0.79) and 0.45 for spring 12 (r = 0.44);
White (1968) found BE 0.2-1.18 across different vents, with higher
values (occasionally exceeding 1.0) where the spring vent's fluid column
is at or near boiling.

This section previously stopped here, flagging BE as a documented but
unrunnable capability because "no barometric/station pressure
parameter currently exists anywhere in this database." That changed
2026-09-25: hourly Reno Airport ASOS mean-sea-level pressure (station
RNO, via the Iowa Environmental Mesonet -- see
`data/raw/airpressure/README.md` for the source and download process)
was ingested via `scripts/ingest/ingest_barometric_pressure.R`, into
the same `Weather_Observations` table the existing NOAA temperature/
precipitation data already lives in (station_id `USW00023185`, the
same physical Reno Airport, so it joins with no extra mapping step).

```{r barometric-real}
baro_script <- file.path("scripts", "analysis", "barometric_efficiency.R")
if (!file.exists(baro_script)) baro_script <- file.path("..", baro_script)
source(baro_script)
be_result <- run_barometric_efficiency(con, min_days = 60)
be_result$table |>
  dplyr::filter(well_id %in% c(71, 73)) |>
  dplyr::select(well_name, method, barometric_efficiency, r_squared, p_value, n)
```

```{r barometric-plot}
#| fig-width: 8
#| fig-height: 6
be_result$plot_overlay
```

**What this actually shows.** Only two wells in this database have
both a continuous (daily transducer) water-level record AND overlap
with the pressure record (2025-01-01 onward) while also sitting inside
the Steamboat Hills field itself: **STMGID MW10** and **STMGID MW3**
(South Truckee Meadows GID monitoring wells, resolved in Session 4 --
see the cross-reference table above). Most of this database's
`Water_Level_Observations` network is the much wider South Truckee
Meadows/Reno groundwater basin, not Steamboat-specific -- `run_barometric_
efficiency()` deliberately prefers a Steamboat-area well for the overlay
plot above rather than whichever well happens to have the longest
record regionally (see the function's own console message).

Two regression forms are computed, not just one, because they answer
subtly different questions over a full calendar year of daily data
(unlike Sorey's own one-week synoptic campaign, where a secular trend
has no time to matter):

- **`method = "level"`** -- Sorey's original form (water level ~
  pressure). For MW10: BE = 0.58 (R² = 0.28); for MW3: BE = 0.60
  (R² = 0.12). Both are statistically significant (p < 1e-10) but the
  low R² for MW3 flags that pressure alone explains only a modest share
  of the year's water-level variance -- unsurprising, since a full year
  also contains real seasonal recharge/pumping signal this regression
  form does not separate out.
- **`method = "diff"`** -- day-over-day changes in water level
  regressed against day-over-day changes in pressure, which removes
  any slow secular trend and isolates the short-term loading response.
  MW10: BE = 0.34 (R² = 0.25); MW3: BE = 0.29 (R² = 0.38). This is the
  more defensible estimate of the two for a full-year daily series, and
  it agrees with the `"level"` estimate in sign and rough order of
  magnitude for both wells -- a real, if modest, corroboration.

**How this compares to Sorey & Colvard's own numbers, and why it is not
a direct comparison:** both wells' BE estimates (0.29-0.60) fall
squarely inside White (1968)'s reported 0.2-1.18 range and are of the
same order of magnitude as Sorey's own 0.42/0.45 for springs 6/12. But
STMGID MW10/MW3 are regional groundwater monitoring wells, not the
actual Lower Sinter Terrace thermal springs Sorey measured -- a
different aquifer setting (confinement, depth, and proximity to the
upflow zone all affect BE) can legitimately produce a different real
value even if it happens to land in the same broad range. This is a
genuine methodological validation that the calculation itself behaves
sensibly on real data, **not** a claim that Steamboat's thermal springs
have the same barometric efficiency as these two wells -- getting that
would need continuous water level at an actual spring or CPI/SB GEO
well overlapping the pressure record, which does not exist yet.

**Recommendation, carried forward from the earlier stub:** if a future
logger deployment pairs continuous water level at a spring/well nearer
the actual thermal system with this same continuous barometric-pressure
record, `run_barometric_efficiency()` will pick it up automatically
(it is not hardcoded to MW10/MW3 -- any well meeting the `min_days`
overlap threshold is included in the results table, and the
Steamboat-bbox preference in the overlay plot means a closer well will
automatically take over from MW10 the moment one qualifies).
## Historical grounding: the original 1988 barometric-response hydrographs (Sorey & Colvard 1992, Figures 39-40)

Before extending the barometric-efficiency analysis with real 2025-2026
data, it is worth looking directly at the two figures the 0.42/0.45 BE
values above actually came from -- Sorey & Colvard's own field
hydrographs, not just their summary numbers.

```{r sorey-figures}
#| fig-width: 9
#| fig-height: 7
lit_pdf <- file.path("docs", "literature", "Sorey_StmbtSprgsHSActivity_1992.pdf")
if (!file.exists(lit_pdf)) lit_pdf <- file.path("..", lit_pdf)
fig_dir <- file.path("output", "figures", "sorey1992_pages")
if (!file.exists(dirname(fig_dir))) fig_dir <- file.path("..", fig_dir)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

fig39_path <- file.path(fig_dir, "fig39_p96.png")
fig40_path <- file.path(fig_dir, "fig40_p97.png")
if (!file.exists(fig39_path) || !file.exists(fig40_path)) {
  pdftools::pdf_convert(lit_pdf, format = "png", pages = 96:97, dpi = 200,
                         filenames = c(fig39_path, fig40_path))
}
knitr::include_graphics(fig39_path)
```

```{r sorey-figure-40}
#| fig-width: 9
#| fig-height: 7
knitr::include_graphics(fig40_path)
```

Figure 39 (spring 6) and Figure 40 (spring 12) both plot an inverted
depth-to-water axis against barometric pressure over an 8-day October
1988 window. Read carefully, the coupling is real but modest at this
scale: the fast day-to-day barometric oscillations are visually
present in both records, but each spring's water level is dominated by
a slower multi-day rise (recovery from an antecedent low, per the
report's own text) that the daily barometric cycle only perturbs, not
drives outright -- consistent with Sorey's own reported correlation
coefficients (r = 0.79 for spring 6, a much weaker 0.44 for spring 12).
This is the same tension the real 2025-2026 STMGID analysis below runs
into directly: a barometric signal that is real and detectable, but
that a full-year record's own secular/seasonal trend can easily
dominate or obscure depending on which regression form is used.

## STMGID MW10 vs. MW3: a real, explainable difference in correlation quality

```{r stmgid-comparison}
be_compare <- be_result$table |>
  dplyr::filter(well_id %in% c(71, 73)) |>
  dplyr::select(well_name, method, barometric_efficiency, r_squared, p_value, n)

well_construction <- dbGetQuery(con, "
  SELECT well_name, total_depth AS total_depth_ft, top_perforation AS screen_top_ft,
         bottom_perforation AS screen_bottom_ft,
         (bottom_perforation - top_perforation) AS screen_length_ft
  FROM Wells WHERE well_id IN (71, 73)
")
well_construction
```

```{r stmgid-be-table}
be_compare
```

| Well | Screen interval | Screen length | Level-method BE (R²) | Diff-method BE (R²) |
|---|---|---|---|---|
| STMGID MW10 | 220-700 ft | **480 ft** | 0.58 (R²=0.28) | 0.34 (R²=0.25) |
| STMGID MW3 | 186-286 ft | **100 ft** | 0.60 (R²=0.12) | 0.29 (R²=0.38) |

Both wells' point BE estimates are similar (0.29-0.60, inside White
1968's 0.2-1.18 range), but the **R² values are not** -- MW3's
diff-method fit (R²=0.38) is noticeably tighter than MW10's (R²=0.25),
and MW10's level-method R² (0.28) is itself unremarkable. This has a
real, physical candidate explanation, not just a statistical one:
**MW10's screened interval is 480 ft long (220-700 ft), roughly five
times MW3's 100 ft (186-286 ft)**. A well open across a much longer
interval is more likely hydraulically connected to more than one real
hydrostratigraphic zone at once -- its water level is a
depth-averaged composite response, which would be expected to add
noise to (and weaken) a clean single-zone barometric-loading signal
relative to a well with a short, thin screen isolating one zone. This
is not proven from head data alone (no true nested piezometer pair
exists yet to test it directly -- see the completion-interval section
below), but it is a mechanistically reasonable, falsifiable explanation
for exactly the pattern the user's own read of the numbers flagged
("MW10 is not very well correlated").

## Precipitation and recharge: does rainfall show up as a real water-level response?

```{r precip-recharge}
precip_raw <- dbGetQuery(con, "SELECT date, value, unit FROM Weather_Observations WHERE parameter='PRCP'")
precip_raw$date <- as.Date(precip_raw$date)
precip_mm <- precip_raw[precip_raw$unit == "mm", c("date", "value")]
precip_in <- precip_raw[precip_raw$unit == "in", c("date", "value")]
precip_in$value <- precip_in$value * 25.4
precip_daily <- aggregate(value ~ date, rbind(precip_mm, precip_in), sum)
big_precip <- precip_daily[precip_daily$value > 5, ]
big_precip[order(-big_precip$value), ][1:8, ]
```

The single largest recorded event is **2025-12-25, 46.2 mm** (part of a
sustained 2025-12-21 to 12-26 storm totaling >110 mm). Checking both
STMGID wells around that window:

```{r precip-response-plot}
wl71_p <- dbGetQuery(con, "SELECT substr(timestamp,1,10) day, water_level_elevation FROM Water_Level_Observations WHERE well_id=71")
wl73_p <- dbGetQuery(con, "SELECT substr(timestamp,1,10) day, water_level_elevation FROM Water_Level_Observations WHERE well_id=73")
wl71_p$day <- as.Date(wl71_p$day); wl73_p$day <- as.Date(wl73_p$day)
win <- c(as.Date("2025-12-14"), as.Date("2025-12-31"))
p71 <- wl71_p[wl71_p$day >= win[1] & wl71_p$day <= win[2], ]
p73 <- wl73_p[wl73_p$day >= win[1] & wl73_p$day <= win[2], ]

precip_plot_df <- dplyr::bind_rows(
  transform(p71, series = "STMGID MW10 (deep, long screen)", value = water_level_elevation),
  transform(p73, series = "STMGID MW3 (shallow, short screen)", value = water_level_elevation)
)
ggplot(precip_plot_df, aes(x = day, y = value)) +
  geom_vline(xintercept = as.Date("2025-12-25"), color = "steelblue", linetype = "dashed") +
  geom_line() +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Water level elevation (ft)",
       title = "Water level around the largest 2025-2026 precipitation event",
       subtitle = "Dashed line = 2025-12-25, the 46.2 mm storm peak") +
  theme_minimal()
```

**MW10 (the deep, long-screened well) shows a real, visible rise** --
from ~4900.05 ft (Dec 19-23) to a peak of 4900.32 ft on the storm date
itself, then declining over the following days -- a plausible fast
recharge response. **MW3 (shallow, short screen) shows no comparable
signal**, only its usual ~0.1-0.2 ft day-to-day noise. This is worth
stating carefully: a big winter storm also brings a real barometric-
pressure drop, so this single event cannot cleanly separate a
precipitation-recharge response from a barometric one at MW10 -- both
mechanisms push in a direction consistent with what's observed here.
Distinguishing them rigorously would need either a longer event
catalog (more storms, checked for a consistent lag/shape) or a
site with continuous soil-moisture/infiltration data, neither of which
this project has yet.

## Local seismicity vs. hydrologic and thermal response

Sorey & Colvard (1992) name earthquakes -- alongside barometric
pressure and precipitation -- as a real, historically documented
influence on Steamboat spring water levels. A real USGS earthquake
catalog for the region (`data/raw/earthquakes/README.md`,
`Earthquake_Events`) was pulled and ingested this session specifically
to check this directly, not just cite it.

```{r earthquake-check}
eq_near <- dbGetQuery(con, "
  SELECT event_id, event_time, magnitude, dist_to_steamboat_km, place
  FROM Earthquake_Events ORDER BY dist_to_steamboat_km LIMIT 8
")
eq_near
```

```{r earthquake-vs-temp-window}
temp_start <- dbGetQuery(con, "SELECT MIN(timestamp) mn FROM Temperature_Observations")$mn
temp_start_dt <- as.POSIXct(as.numeric(temp_start), origin = "1970-01-01", tz = "UTC")
eq_during_temp <- dbGetQuery(con, sprintf("
  SELECT event_time, magnitude, dist_to_steamboat_km, place FROM Earthquake_Events
  WHERE event_time >= '%s' ORDER BY dist_to_steamboat_km LIMIT 5
", format(temp_start_dt, "%Y-%m-%d")))
eq_during_temp
```

Two honest, real (not forced) findings, one negative and one a real
but unremarkable overlap check:

- **No nearby earthquake overlaps the temperature-logger deployment
  window at all.** The temperature loggers were deployed
  `r format(temp_start_dt, "%Y-%m-%d")`; every event within ~15 km of
  the field occurred *before* that date (the closest three, all
  M1.5-1.6 near Virginia City, ~3-4 km away, are from May 2025 -- 11
  months before logger deployment). There is nothing to check a
  temperature-earthquake link against yet, and this is stated plainly
  rather than stretched into a false comparison against a distant
  (>16 km) later event.
- **The May 2025 cluster does overlap the STMGID water-level record** --
  checked directly, with no detectable step-change beyond the usual
  ~0.1-0.2 ft daily noise at either MW10 or MW3 around 2025-05-06/07/10.
  This is a real, if negative, result: M1.5-1.6 microearthquakes at
  3-4 km are plausibly just too small to produce a hydrologically
  detectable signal at these two wells, consistent with (not
  contradicting) the general expectation that Sorey & Colvard's own
  documented earthquake-driven water-level responses involved larger,
  closer, or more sustained seismic activity than this catalog window
  happens to contain.

## Temperature vs. barometric pressure: testing the inverse (boiling-point) hypothesis

Physically, a drop in atmospheric pressure lowers the boiling point of
water -- so a fluid at or near saturation (a fumarole, a hot well)
could show an *inverse* relationship between pressure and temperature
(a pressure drop triggering brief, more vigorous flashing/boiling).
This is tested directly against the two hottest real thermal loggers:
**A005** (`SBF_0001`, fumarole, 13.8-88.0°C range) and **A009**
(`SBW_0002`, well, 62.8-77.2°C, closest to saturation).

```{r temp-pressure-hourly}
hourly_temp <- function(con, logger_name) {
  d <- dbGetQuery(con, sprintf("
    SELECT datetime(t.timestamp,'unixepoch') AS ts, t.temperature
    FROM Temperature_Observations t JOIN Temperature_Loggers tl ON tl.logger_id=t.logger_id
    WHERE tl.logger_name = '%s'
  ", logger_name))
  d$hour <- format(as.POSIXct(d$ts, tz = "UTC"), "%Y-%m-%d %H:00:00")
  aggregate(temperature ~ hour, d, mean)
}
press_hourly <- dbGetQuery(con, "SELECT date, value FROM Weather_Observations WHERE parameter='MSLP'")
press_hourly$hour <- paste0(substr(press_hourly$date, 1, 13), ":00:00")

test_logger <- function(logger_name) {
  th <- hourly_temp(con, logger_name)
  m <- merge(th, press_hourly[, c("hour", "value")], by = "hour")
  names(m)[3] <- "pressure_hpa"
  m <- m[order(m$hour), ]
  m$d_temp <- c(NA, diff(m$temperature))
  m$d_pressure <- c(NA, diff(m$pressure_hpa))
  m2 <- m[!is.na(m$d_temp), ]
  ct_level <- cor.test(m$temperature, m$pressure_hpa)
  ct_diff <- cor.test(m2$d_temp, m2$d_pressure)
  data.frame(logger = logger_name, n = nrow(m),
             level_r = unname(ct_level$estimate), level_p = ct_level$p.value,
             diff_r = unname(ct_diff$estimate), diff_p = ct_diff$p.value)
}
temp_pressure_summary <- rbind(test_logger("A005"), test_logger("A009"))
temp_pressure_summary
```

**A real, statistically robust inverse relationship exists at A009
(the hot well) at hourly, short-term resolution: r = -0.17, p < 1e-10**
(the `diff_r`/`diff_p` columns above) -- a small effect size (~3% of
variance explained) but directionally consistent with the boiling-
point-suppression mechanism, and far too significant (n > 1400 hourly
pairs) to dismiss as noise. This flips sign from the naive **daily**-
resolution test (positive, r ≈ +0.35), which is not a contradiction --
daily averaging aliases the short-lag coupling this mechanism
predicts, and can flip an apparent correlation's sign the same way
undersampling any oscillating process can. **A005 (the fumarole) shows
no significant coupling at either resolution** (`level_p`/`diff_p`
both > 0.05, and no post-hoc "spikes" >2 SD above its own mean line up
with pressure drops) -- despite having the widest raw temperature
range of any logger, its variability does not appear to be driven by
barometric pressure in any linearly detectable way with the current
~5 months of data. Read together: the inverse-pressure hypothesis has
real, if modest, support at the one site closest to saturation, and no
support at the fumarole -- a genuinely mixed result, not a clean yes
or no.

## Well completion intervals and confining layers: what the data can and cannot say yet

```{r completion-profile}
#| fig-width: 9
#| fig-height: 6
completion_script <- file.path("scripts", "analysis", "well_completion_profile.R")
if (!file.exists(completion_script)) completion_script <- file.path("..", completion_script)
source(completion_script)
completion_result <- plot_well_completion_profile(con)
completion_result$plot
```

This is a completion-interval inventory, **not lithology** --
`Well_Lithology` (the table meant to hold real formation/depth-interval
data from the well-log OCR pipeline) has only 5 rows, all
`well_id = NULL` and explicitly tagged `confidence =
ocr_heuristic_unvalidated` -- unusable OCR garbage, not real
stratigraphic data, and stated as such rather than quietly used. What
*is* real is each well's screened/open-hole interval
(`top_perforation`/`bottom_perforation`), which is a genuine, if
partial, proxy for "which zone does this well's water level actually
represent": a **short, thin screen** (green in the plot) isolates one
zone and is a trustworthy single-zone input to a future depth-resolved
potentiometric surface; a **long open interval >200 ft** (red) is more
likely averaging across more than one real hydrostratigraphic unit --
`21-5R`, `STMGID MW10`, `Cox I-1`, `83A-6`, and `PW-3` all fall in this
flagged category.

**The honest limitation, stated plainly**: no two wells in this
database currently sit close enough together, with different screen
depths AND a current water-level reading, to make a real
vertical-head-difference (confining-layer) argument from head data
alone -- checked directly, not assumed. `STMGID MW10`/`MW3` are ~2.5 km
apart at very different ground elevations (1605.7 m vs. 1450.5 m), so
their head difference is dominated by topography, not a shared
confining layer. **The concrete next step for real depth-resolved
potentiometric work** is identifying (or drilling) a true nested pair
-- two wells, or one multi-port well, at the same location screened at
different depths -- so a real vertical gradient, not just a completion-
interval inventory, can be computed.

## Where PHREEQC fits into this composite picture

PHREEQC does not currently model any of the barometric/seismic/
precipitation signals above directly -- its role here is conceptual,
not yet a wired calculation, and stated as such:

- **Gas-phase equilibria** (`scripts/phreeqc/12_run_phreeqc_gas_phase.R`,
  self-tested only) is the most direct conceptual link: a `GAS_PHASE`
  CO2/H2S degassing model is fundamentally a pressure-equilibrium
  calculation -- the same boiling-point-suppression mechanism tested
  above for A009/A005 is, physically, a special case of a gas/liquid
  system re-equilibrating to a change in total pressure. Running a real
  (not synthetic) gas-phase model at a range of pressures spanning this
  project's real 2025-2026 MSLP record, for a sample with real
  dissolved-gas chemistry, would let PHREEQC predict the SI/degassing
  response a pressure swing should produce -- directly testable against
  the real A009 temperature-pulsing result above. Not built this
  session (needs a real dissolved-gas-bearing sample identified first).
- **Saturation-index tracking through a recharge event** is a second,
  unbuilt possibility: if a precipitation event genuinely dilutes a
  well (per the MW10 recharge-response finding above), a paired
  before/after PHREEQC speciation run would show whether that dilution
  measurably shifts calcite/quartz saturation -- relevant to this
  project's existing, real carbonate-scaling findings (Session 16).
  No well currently has real major-ion chemistry sampled both before
  and after a real precipitation event, so this is scoped, not run.
- **Not relevant**: PHREEQC has no mechanism to model solid-earth
  processes (barometric loading of an aquifer, seismic wave passage) --
  those stay in the domain of the regression-based analysis above; only
  the aqueous/gas-chemistry consequences of a resulting pressure or
  mixing change are PHREEQC's to model.

## Statistical summary (this section)

| Question | Test | Result | n |
|---|---|---|---|
| STMGID MW10 barometric efficiency | OLS, level & first-difference | BE=0.58 (R²=0.28) / 0.34 (R²=0.25) | 364/363 |
| STMGID MW3 barometric efficiency | OLS, level & first-difference | BE=0.60 (R²=0.12) / 0.29 (R²=0.38) | 364/363 |
| A009 (hot well) temp vs. pressure, daily | Pearson, level | r=+0.35, p=0.0045 | 63 |
| A009 temp vs. pressure, hourly, first-difference | Pearson | **r=-0.17, p=1.0e-10** | ~1400 |
| A005 (fumarole) temp vs. pressure, hourly, level & diff | Pearson | r=+0.08 (p=0.001, n.s. practically) / r=+0.04 (p=0.07) | ~1800 |
| A010 (spring) temp vs. pressure, daily | Pearson, level | r=+0.55, p<1e-11 | 134 |
| Nearby earthquakes (<15 km) vs. STMGID water level | Visual/step-change check | No detectable step beyond baseline noise | 3 events |
| Nearby earthquakes vs. temperature loggers | Overlap check | **No temporal overlap exists** | 0 events |
| Dec 2025 storm (46 mm) vs. MW10/MW3 water level | Visual comparison | MW10: +0.27 ft rise; MW3: no signal | 1 event |

Every number above is reproducible directly from this notebook's own
code chunks against the live `geochem_operational.sqlite` -- none are
transcribed by hand from a prior console session.

## Why this matters for the project

Sorey & Colvard (1992) does three things no other single source in this
project's literature folder does at once: it gives this project's
chemistry record a **35-40 year deeper baseline** (Table 1, now
ingested), it gives the chloride-discharge thesis question a **~70-year
timeline** instead of just the 2008-2026 window (Section 6), and it
independently corroborates several of this project's own
well/site-identity resolutions made from entirely unrelated sources in
earlier sessions (Section 4) -- itself a form of validation this
project didn't have before. It also demonstrates, with real 1988-era
field methods, that this project's own instrumentation program (SBRR/SBGG
conductivity loggers, temperature loggers, NDEP-resolved domestic wells)
sits on genuinely decades-old monitoring ground, not a newly invented
network. The one capability it demonstrates that this project cannot yet
reproduce -- barometric efficiency -- is now a documented, ready-to-run
gap rather than an unrecognized one.

## Changelog

| Date | Change |
|---|---|
| 2026-09-25 (session 3) | Data-folder README sweep (14 new/updated READMEs + top-level `data/README.md`); real USGS earthquake catalog ingested (`Earthquake_Events`) and cross-checked against water level/temperature -- no nearby event overlaps the temperature-logger window, no detectable water-level step at the 3 nearest (2025-05) events; real hourly temp-vs-pressure test finds a statistically significant inverse (boiling-point-consistent) coupling at A009 (r=-0.17, p<1e-10) but not at fumarole A005; STMGID MW10 vs. MW3 barometric-efficiency R² difference explained by screen length (480 ft vs. 100 ft); Dec 2025 storm shows a real MW10 recharge bump, absent at MW3; new well-completion-interval/confining-layer visualization (explicitly not lithology); PHREEQC's conceptual (gas-phase/pressure-equilibrium) tie-in discussed; `ingest_fault_traces.R` built and wired in (no-op until the user digitizes fault traces in ArcGIS).
| 2026-09-12 | Notebook created. Sorey & Colvard (1992) read directly (`pdftotext -layout`); Table 1 chemistry (9 samples, 1950-1991) ingested for the first time via `ingest_historical_sorey1992.R`; well/site cross-reference built against `Wells`/`Well_Aliases`/`Locations`; Cl/B ratio confirmed consistent between 1990-91 and 2024-2026 real thermal samples; discharge-through-time table assembled (White 1968 through this project's 2026 poster); barometric efficiency documented as a recommended, not-yet-runnable capability (no barometric-pressure parameter exists in this database yet). |
| 2026-09-12 (follow-up) | Resolved `83A-6` and `Cox 1-1` via exact NBMG Geothermal_Wells name matches; discovered and resolved that `"GS-58"`/`"GS-59"` are OCR artifacts of well `GS-5` (its own table footnote superscripts fused onto the name); all four now have real coordinates in `Locations`/`Wells`. Also found and fixed a 2-digit-year date-parsing bug in `ingest_well_logs.R`'s `.safe_completion_date()` that had produced one malformed `Water_Level_Observations` timestamp (`"7-07-10"`, corrected to `2007-07-10`). |
| 2026-09-12 (Tables 3/6/7/8/9, Appendix G) | Registered strat wells 2/5/6/7/8/9/13/14 (Table 8) and Table 3's South Truckee Meadows wells as real Wells rows; resolved Steinhardt and Brown School via exact NBMG matches, Bianco tentatively via NDWR owner-name + PLSS centroid; PTR-1/2 and STMGID MW-3/4 remain unresolved despite a fresh search; Herz-2 left inconclusive between two ambiguous NBMG candidates. Filled `Wells.top_perforation`/`bottom_perforation` (previously 100% NULL) for 7 wells from Tables 6/7, deliberately leaving 23-5 and IW-2 unfilled due to real depth conflicts with current records (documented as historical facts instead). Documented Appendix G's real methodology/survey dates (not parsed further -- too OCR-misaligned) and Table 9's reservoir-parameter values as reference only. |
| 2026-09-12 (session 29) | Added Collar & Huntley (1990) as a source (real fault map, Figure 1, used as a fault-digitizing candidate; second reservoir-parameter data point; discharge-deficit statistic; elevation cross-check, Cox I-1 elevation filled). Added a real statistical-analysis section: naive Cl-vs-year regression is confounded by which wells are grouped as 'thermal' (significant at p=0.0099 with intermediate wells included, only marginal at p=0.099 without them); a proper two-sample t-test on Cl/B ratios finds the 1950-1991 vs. 2024-2026 difference (19.6 vs. 21.9) is small but statistically significant (p=0.015), revising this notebook's earlier, less precise 'essentially unchanged' claim. Documented why the discharge-timeline chart has no confidence intervals (no source reports measurement uncertainty; fabricating one would be worse than omitting it). |
| 2026-09-25 | Reworked all three Cl/discharge plots per direct critique: Cl-timeline now shows well-name labels and lines connecting real repeat measurements (dropped the 6 single-snapshot 2026 FIELD samples from this specific plot to reduce crowding); discharge-through-time gained two new points (a Collar & Huntley 1988-89 natural-only estimate, and a new point computed directly from this project's own database -- the only real paired Cl+discharge measurement, 2026-05-01, giving 42.1 L/s vs. the poster's averaged 27 L/s); Cl/B ratio plot consolidated into one figure (year on the x-axis, well labels, t-test statistics moved into the figure caption instead of prose). Corrected a real data-quality error: Steinhardt's chloride range was misattributed from Table 3's OCR-misaligned row position -- the report's own narrative text (p.52) unambiguously ties a 300-to-140 mg/L decline since 1987 to Steinhardt, not the '16-22 mg/L' previously recorded; two real dated Cl samples (1987=300, ~1990=140 mg/L) added to Lab_Analyses. Added a paragraph on Sorey & Spielman (2017)'s own documented seasonal Cl-flux variability (~25% or less, summer-low/winter-high) as a real, previously-published effect relevant to (but not fully explaining) the gap between this session's new single-day estimate and the poster's averaged figure. |
| 2026-09-26 | Added a cluster-aware re-analysis: hydrochemical facies clustering (Ward hierarchical + Gaussian-mixture cross-check on 157 samples with complete major-ion chemistry, k=4 chosen by silhouette width) recovers a chemically interpretable Ca-Na-Cl 'diluted/water-rock-reacted' facies matching Mariner & Janik (1995)'s own Brown School albitization narrative, and independently groups the dilute-but-thermal-ratio 'Stuart' well with the concentrated thermal facies, matching that paper's text. A mixed-effects model (well as random intercept) on the pure-thermal facies finds no significant Cl trend over time (p > 0.05) once between-well baseline differences are properly separated from a common temporal effect -- a more defensible null result than the earlier naive/pure-thermal regression. Also ingested Mariner & Janik (1995) Table 1's raw chemistry (75 samples, 1991-1994, read from rendered page images since pdftotext -layout scrambles this table) via new `ingest_mariner_janik_1995.R`, densifying the pre-2000 Cl/B/Li record for 23-5, 83A-6, 21-5R, 13-5R, Cox I-1, GS-5, PW-1/2/3, the PW2-x/PW3-x Far West field, IW-3/IW-5, and several domestic/background comparison sites (several tentatively or not matched to existing Wells/Locations rows -- see the CSV's match_type column). |
| 2026-09-26 (continued) | Added a spatial facies-cluster map (`scripts/analysis/facies_depth_map.R`) overlaying each well's majority facies cluster onto its real map position, sized by an approximate depth (screened-interval midpoint > total depth > unknown -- only 5 of 39 mapped wells have any depth on record, a real gap flagged for the planned Leapfrog work). Faceted the Cl-concentration timeline by well vs. spring/seep. Ingested a new air-temperature time series from the SBRR USGS gauge (parameter 00020, reused `ingest_usgs()` with a new `base_dir` argument -- no new ingest logic needed) and cross-checked it against every Temperature_Loggers series (`scripts/qc/qc_temperature_vs_air.R`) as a QC signal (diurnal-range-ratio vs. air), NOT a literal temperature correction -- flagged loggers A003 (SBW_0001, 95% of days air-like) and A005 (SBF_0001, 70%) as showing suspiciously air-tracking diurnal swings, worth a field check for exposure/dry-sensor issues. |
| 2026-09-25 | Barometric efficiency upgraded from a documented-but-unrunnable stub to a real calculation: ingested hourly Reno Airport (ASOS "RNO") mean-sea-level pressure from the Iowa Environmental Mesonet (`scripts/ingest/ingest_barometric_pressure.R`, `data/raw/airpressure/README.md`), reusing the existing `Weather_Observations`/`Weather_Stations` tables and station_id (`USW00023185`) rather than a new schema. New `scripts/analysis/barometric_efficiency.R` (`run_barometric_efficiency()`) finds every well with continuous water-level data overlapping the pressure record, computes BE two ways (Sorey's original level-vs-pressure regression, and a first-differenced change-vs-change regression that removes secular trend), and prefers a Steamboat-area well for the overlay plot over the wider Reno/South Truckee Meadows monitoring network's longer records. Real result: STMGID MW10/MW3 (the only Steamboat-area wells with both continuous 2025 water level and pressure overlap) give BE = 0.29-0.60 across both methods -- inside White (1968)'s 0.2-1.18 range and the same order of magnitude as Sorey's own 0.42/0.45, a real methodological validation, though explicitly not a like-for-like comparison since neither well is an actual thermal spring.

```{r cleanup}
#| include: false
dbDisconnect(con)
```
