---
title: "Historical Context: Sorey & Colvard (1992) and the Long-Run Steamboat Record"
subtitle: "Cross-referencing a 1988-1992 USGS study against this project's database"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

`docs/literature/Sorey_StmbtSprgsHSActivity_1992.pdf` -- Sorey, M.L. and
Colvard, E.M. (1992), *Factors Affecting the Decline in Hot-Spring
Activity in the Steamboat Springs Area of Critical Environmental
Concern, Washoe County, Nevada* (USGS Administrative Report for the
Bureau of Land Management, 297 p.) -- was added to this project's
literature folder on 2026-09-12. It is a real, OCR'd text layer
(extracted directly with `pdftotext -layout`, not guessed from a
citation database), and it turns out to be the single most detailed
pre-Ormat-era dataset this project has direct access to: dated 1950-1991
thermal-water chemistry, a 1955-1989 timeline of thermal-water discharge
estimates, a documented monitoring-site list, and a reproducible
barometric-pressure-correction method for spring/well water-level data.

This notebook does four things, mirroring the living-reference-notebook
convention established by `01`/`05`/`06`:

1. Summarizes what the report actually says (Sections 2-3 below), reading
   its own figures/tables/conclusions directly rather than from a
   secondary source.
2. Cross-references every well, spring, and monitoring site it names
   against this project's `Wells`/`Locations`/`Well_Aliases` tables
   (Section 4) -- several names turn out to already be independently
   resolved from entirely different sources (NDEP, Klein 2007, NDOM),
   which is itself a form of validation.
3. Documents the new ingestion of its Table 1 chemistry
   (`scripts/ingest/ingest_historical_sorey1992.R`,
   `data/raw/historical/sorey1992_table1_chemistry.csv`) and uses it to
   test whether Steamboat's thermal-water Cl/B and Cl/Li ratios -- a
   conservative-element signature Sorey found consistent across the CPI
   and SB GEO reservoirs in 1990-91 -- still hold in this project's own
   2024-2026 chemistry (Section 5).
4. Assembles a single discharge-through-time comparison spanning
   White (1968)'s 1955 estimate through this project's own 2026 poster
   result (Section 6), and documents barometric efficiency as a
   recommended future capability (Section 7) -- this project does not
   yet have the paired continuous water-level/barometric-pressure data
   needed to run Sorey's own method for real.

```{r setup}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
```

## What the 1992 report says

**Study design.** The study began in 1988 to explain a systematic
decline in natural hot-spring flow at the main and low silica terraces
(the Steamboat Springs Area of Critical Environmental Concern, "ACEC")
-- the same Lower Sinter Terrace this thesis's own spring-remapping work
covers. Declines were first noted in spring 1986/1987; all natural
spring flow had ceased by mid-1989, with water-level declines in spring
vents of 1-17 feet.

**Conclusions (its own numbered list, p. 100-101, paraphrased):**

- ~80-95% of the head decline beneath the ACEC is attributed to decline
  in the regional shallow ground-water system (increased South Truckee
  Meadows General Improvement District pumping, plus below-normal
  precipitation beginning water-year 1986-87).
- ~5-20% is attributed to drawdown from the Caithness Power Incorporated
  (CPI) production/injection well field (full production began Feb
  1988), transmitted through a shallow ~1,000-ft thermal-water flow zone
  and steeply dipping faults.
- SB GEO well-field influence (full production began Jan 1987, 0.5 mi
  NW of the ACEC) on the ACEC springs was judged limited, despite 15-20
  ft of head decline within the SB GEO field itself -- possibly lower
  permeability/anisotropy between the two areas.

**Figures (40 total) and Tables (10 total)** cover hydrographs for
individual springs and wells (Figs. 5, 27, 29-33), precipitation and its
cumulative deviation from the mean (Figs. 15-18), a stable-isotope
altitude/boiling diagram (Fig. 9), electrical-resistivity mapping (Fig.
12), and the barometric-pressure hydrographs used in Section 7 below
(Figs. 39-40). Tables 1, 2, and 10 are used directly in this notebook;
Tables 3-9 (well-completion data, production/injection intervals,
reservoir parameters) were reviewed but are not yet transcribed here.

## Well and monitoring-site cross-reference

Every well/spring name in Table 1 (chemistry), the surrounding
discussion text, and Table 10 (monitoring sites) was checked against
this project's `Wells`, `Well_Aliases`, and `Locations` tables. Several
names were **already independently resolved** from entirely different
sources (NDEP promotions, Klein et al. 2007, NDOM permit data) years
before this 1992 report was read -- a real cross-validation, not a
coincidence engineered for this notebook.

```{r crossref-table}
crossref <- tribble(
  ~sorey_name, ~feature_type, ~status, ~note,
  "21-5 (CPI)", "well", "Resolved", "Well_Aliases '21-5' -> canonical 21-5R (well_id 94)",
  "23-5 (CPI)", "well", "Resolved", "Exact Wells.well_name match (well_id 96)",
  "83A-6 (CPI)", "well", "Resolved (2026-09-12)", "Exact NBMG Geothermal_Wells name match ('Well No. 83A-6', API 27-031-90080)",
  "Cox 1-1 / Cox well", "well (injection)", "Resolved (2026-09-12)", "Exact NBMG match ('Cox I-1', API 27-031-90051); previously flagged unresolved as COX-1/Cox1-1",
  "PW-1, PW-2, PW-3 (SB GEO)", "well", "Resolved", "Exact Wells.well_name matches (well_id 79/80/81)",
  "IW-1, IW-2, IW-3 (SB GEO)", "well (injection)", "Resolved", "IW-1 alias -> 45-28; IW-2/IW-3 exact canonical names",
  "OW-1, OW-2 (SB GEO observation)", "well (monitor)", "Partial", "OW-2 exact match (well_id 110); OW-1 not found",
  "28-32", "well", "Resolved", "Exact Wells.well_name match (well_id 125)",
  "Strat wells 2, 5, 6, 7, 8, 9, 13, 14", "well (stratigraphic test)", "Registered, no coordinate (2026-09-12)", "Real provisional Wells rows with depth/formation/temperature/water-level data (Table 8, Section 4.2); no coordinate -- see proximity notes for strat 13 (near 23-5) and strats 2/5/9 (near Cox I-1)",
  "Well GS-58, Spring GS-59", "well/seep (pre-Ormat)", "Resolved (2026-09-12)", "Both are OCR artifacts of well GS-5 (see Section 4.1) -- exact NBMG match, API 27-031-80055",
  "Hot spring 6 (main terrace ACEC)", "spring", "Unresolved", "Coordinate-less provisional Locations row; no modern counterpart identified",
  "Steamboat Creek at Rhodes Road", "surface water", "Resolved (independently)", "Locations 'SB5'/'SB6' -- same site this project's SBRR conductivity logger occupies",
  "Curti Barn Well, Curti Domestic Well", "domestic/geothermal well", "Resolved (independently)", "Klein et al. (2007)-sourced Locations rows",
  "Boyd Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Boyd Domestic Well'",
  "Rogers Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Rogers Domestic Well'",
  "Jeppson Well", "geothermal well", "Resolved (independently)", "NDEP-promoted Locations row 'Jeppson Geothermal Well'",
  "Steinhardt Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG name match ('Steinhardt Geothermal Well')",
  "Brown School Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG owner-name match ('Brown Elementary School')",
  "Bianco Well", "domestic well", "Tentative (2026-09-12)", "NDWR owner-name match ('BIANCO, U J'); PLSS-section coordinate only, and a flagged depth discrepancy (232 ft vs. Table 3's 400 ft)",
  "Herz-2 (geothermal)", "well", "Inconclusive (2026-09-12)", "Could be either of NBMG's two ambiguous 'Harold Herz Geothermal Well 1/2' candidates (flagged since Session 3); Table 3's own OCR is too garbled to disambiguate -- not guessed",
  "PTR-1, PTR-2 (Pine Tree Ranch)", "domestic well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows with real depth/chloride/water-level data; fresh NDWR/NBMG owner-name search found nothing",
  "STMGID MW-3, MW-4", "monitor well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows; NBMG's only STMGID-named record ('Well 4 Shadowridge') is a production well, not this monitor-well pair -- not used",
  "MacKay Well, Woods Well, Tangen Well, STMGID Went",
    "domestic/monitor well", "Unresolved", "Consistent with prior sessions' independent 'unresolved' flags for these"
)
crossref
```

**Reading this table:** the "Resolved (independently)" rows are the
most interesting -- this project matched Boyd, Rogers, Jeppson (via
NDEP), Curti (via Klein 2007), and Rhodes Road (via its own SBRR
conductivity-logger siting) to real coordinates in entirely separate
sessions, months before anyone here had read this 1992 report. Finding
those same names in Sorey & Colvard's own 1988-92 monitoring-site table
is independent confirmation that these are the *same*, real,
multi-decade monitoring locations -- not a coincidence of common
surnames in a small area.

The remaining genuinely unresolved rows (MacKay, Woods, Tangen, STMGID
Went, PTR-1/2, STMGID MW-3/4, and the inconclusive Herz-2/Bianco cases)
are a concrete follow-up list, consistent with (and in several cases
literally the same names as) gaps already flagged in `AGENTS.md` from
unrelated well-matching sessions. None are guessed here.

### 4.1 Resolving 83A-6, Cox 1-1, and the "GS-58"/"GS-59" OCR artifact (2026-09-12)

A direct search of NBMG's statewide `Geothermal_Wells.csv` and
`GEOTHERM06102019.csv` compilations (`data/raw/nbmg/`) found exact
well-name matches for `83A-6` (API 27-031-90080, a CPI production well
drilled 1987) and `Cox I-1` (API 27-031-90051, a CPI injection well,
previously drilled as observation well "Cox No. 1" for Phillips
Petroleum in 1981) -- both are now real `Wells` rows with NBMG-sourced
coordinates (`data/raw/wells/sorey1992_nbmg_resolved_wells.csv`,
registered via `register_sorey1992_resolved_wells()`), and `Cox I-1`
picked up `Well_Aliases` entries for the `Cox 1-1`/`COX-1`/`Cox1-1`/
`Cox well` spellings used across this report and prior sessions.

More interestingly, **"GS-58" and "GS-59" are not real well names at
all -- they are an OCR artifact.** Re-reading Table 1's own footnote
list resolves this precisely:

> `8From White (1968).`
> `9New seep adjacent to well GS-5, analysis by Nevada Division of Health Laboratory.`

Both footnote markers ("8" and "9") were superscripts attached to a
single real well name, **GS-5**, in the original typeset table, and the
PDF's OCR layer fused each digit directly onto the name it followed --
producing "GS-58" (really "GS-5" + footnote 8) and "GS-59" (really
"GS-5" + footnote 9). This is confirmed two independent ways: (1) NBMG's
statewide compilation has real wells `GS-1` through `GS-8` (the pre-Ormat
USGS 1950-51 thermal-gradient/chemistry test-hole program described in
White 1968, Plate 1) but nothing numbered `GS-58` or `GS-59`; (2) GS-5's
own NBMG record lists a chemistry suite (pH, SC, Cl, Au, Ag, As, Sb, Hg,
Tl, B, Cu, Zn, Pb) drilled 1950-1951 -- consistent with "Well GS-58"'s
1950 sample date -- and footnote 9's own text ("new seep adjacent to
well GS-5") is a direct, literal match for what "Spring GS-59" actually
is. Both rows in `sorey1992_table1_chemistry.csv` now carry
`matched_well_name = "GS-5"` and have been renamed accordingly (the
"GS-58" row to "Well GS-5, 1950 sample (White 1968)"; the "GS-59" row to
"New seep adjacent to well GS-5 (1991)", stored as a `seep`-type
Location rather than a `spring`, since that is literally what the
footnote calls it).

```{r resolved-wells-check}
dbGetQuery(con, "
  SELECT external_station_code, name, latitude, longitude, site_type
  FROM Locations WHERE external_station_code LIKE 'SOREY1992%'
  ORDER BY external_station_code
")
```

**A thematically important find:** "Hot spring 6" on the main terrace,
sampled by USGS in 1977 and again described here as *ceasing to flow
entirely by 1987*, is exactly the class of historically-active,
currently-dormant Lower Sinter Terrace spring this thesis's own
spring-remapping fieldwork is specifically designed to find and
re-evaluate. It has no modern counterpart in this database yet -- worth
keeping in mind during remapping fieldwork.

### 4.2 Table 3: South Truckee Meadows ground-water wells (2026-09-12)

Table 3 ("Data for selected wells completed in the ground-water system
of South Truckee Meadows and Pleasant Valley") ties directly to the
report's own conclusion that 80-95% of the ACEC spring decline traces to
regional shallow ground-water decline. It gives real depth/temperature/
chloride-range/water-level-decline data for PTR-1, PTR-2, STMGID MW-3,
MW-4, Bianco, Boyd, Steinhardt, Brown School, and Herz-2 -- but, like
Table 8, has **no coordinates of its own**. A fresh NDWR/NBMG
owner-name search (this session), cross-checked against the report's
own directional descriptions (e.g. Steinhardt "northeast of the ACEC"),
found:

- **Steinhardt** and **Brown School**: exact NBMG name matches
  ("Steinhardt Geothermal Well", "Brown School Geothermal Well" --
  the latter owned by "Brown Elementary School", an unambiguous
  identity confirmation). Both directions check out against the
  report's own description. Now real, coordinate-having `Wells` rows.
- **Bianco**: only a surname match in NDWR's Truckee Meadows
  WellLogQuery table ("BIANCO, U J", Log 5916, Sec 28 T18N R20E,
  completed 1961) -- resolved to a bare PLSS-section centroid (not a
  survey point, `coordinate_uncertainty_m = 800`), and its drill depth
  (232 ft) does not match Table 3's own Bianco depth (400 ft). Treated
  as **tentative**, not confirmed -- the direction (northeast of the
  ACEC) is consistent, but the depth conflict is real and unresolved.
- **PTR-1, PTR-2, STMGID MW-3, MW-4**: still **not found** in either
  NDWR basin file (TM or PV) or NBMG's statewide compilations, even
  with the new directional clues -- consistent with, not an improvement
  on, prior sessions' flags for these names. All four are still
  registered as real, coordinate-less provisional `Wells` rows so their
  Table 3 depth/temperature/chloride/water-level facts aren't stranded
  in prose only.
- **Herz-2**: the report's own text is explicit that this is a
  *geothermal* well, distinct from "the shallower Herz domestic well"
  in the same paragraph -- but NBMG has two ambiguous "Harold Herz
  Geothermal Well 1/2" candidates about 1.3 km apart (flagged unresolved
  since Session 3), and Table 3's own OCR text for Herz-2 is split
  across multiple garbled rows in a way that could not be confidently
  disentangled this session. Left inconclusive rather than guessed.

```{r table3-check}
dbGetQuery(con, "
  SELECT well_name, well_role, latitude, longitude, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name IN ('Steinhardt Geothermal Well', 'Brown School Geothermal Well', 'Bianco',
                       'Pine Tree Ranch-1', 'Pine Tree Ranch-2', 'STMGID MW-3', 'STMGID MW-4')
  ORDER BY well_name
")
```

### 4.3 Table 8: stratigraphic test wells (2026-09-12)

Table 8 ("Selected data for stratigraphic test wells in the Steamboat
Hills area") gives perforated depth, formation, bottom-hole temperature,
and 1987-1989 water-level change for strat wells 2, 5, 6, 7, 8, 9, 13,
and 14 -- the wells behind the "Strat wells... Unresolved" cross-
reference gap. Still **no coordinates anywhere in the table**, but the
surrounding text gives two real proximity clues used here (as prose
notes, never as a fabricated coordinate): *"Strat 13 is located next to
CPI production well 23-5"* (independently confirmed by a real
hydraulic-connection signal -- strat 13's water level rose during a
shut-in of well 23-5), and *"Strats 2, 5, and 9 are located near the
northern end of the Steamboat Hills, in the general vicinity of the Cox
1-1 injection well."* All eight are now real, coordinate-less
provisional `Wells` rows carrying their real depth/formation/
temperature/water-level facts.

```{r table8-check}
dbGetQuery(con, "
  SELECT well_name, well_role, notes
  FROM Wells WHERE well_name LIKE 'strat %'
  ORDER BY well_name
")
```

Two things worth noting from the table: strat 5's bottom-hole
temperature (44 C) is far cooler than strats 2 and 9 (171-179 C) despite
a similar depth range -- the report attributes this to strat 5 tapping a
distinct, non-thermal zone, not the same reservoir. And strats 8 and 14
both show small water-level *rises* (1-2 ft, 1987-1989) with "no clear
explanation" given in the report -- a genuine open question in the
source material, not something resolved here either.

### 4.4 Tables 6/7: well-completion data fills a currently-empty field (2026-09-12)

`Wells.top_perforation`/`bottom_perforation` were NULL for every one of
the CPI/SB GEO wells already resolved in this database, even though
Tables 6 and 7 ("CPI well-completion information", "SB GEO well
completion information") report real casing-depth and open-hole-
interval data for most of them. The table's own OCR layout has visible
column-alignment slippage, so each well's numbers were individually
cross-checked (casing depth + open-hole thickness = total depth,
confirmed to reconcile exactly for every SB GEO well and for 83A-6) before
being trusted:

```{r table67-check}
dbGetQuery(con, "
  SELECT well_name, total_depth, top_perforation, bottom_perforation
  FROM Wells
  WHERE well_name IN ('83A-6','Cox I-1','PW-1','PW-2','PW-3','21-5R','IW-3','23-5','IW-2')
  ORDER BY well_name
")
```

**Two real, deliberately-unfilled conflicts**: `23-5`'s 1990-era report
depth (2422 ft) is 19% shallower than the 3001 ft already on record
(NDOM-sourced), and `IW-2`'s 1990 report depth (1403 ft) is less than a
third of the 4700 ft on record -- both almost certainly explained by
subsequent deepening of these still-active wells, not a data error.
Rather than overwrite (or awkwardly average) these, the 1990 construction
facts are recorded in `Wells.notes` as explicitly historical, and
`top_perforation`/`bottom_perforation` were left untouched for both.
Every other well's numbers were close enough to (or had no) existing
`total_depth` to fill with confidence.


## Newly ingested chemistry: Table 1 (1950-1991)

`ingest_historical_sorey1992.R` (wired into `run_pipeline.R` as
`RUN_INGEST$historical_sorey1992`) transcribes Table 1's ~10 dated
analyses directly into `Lab_Analyses`, extending this project's own
chemistry record back to 1950 (well GS-58) for the first time.

```{r sorey-chem-table}
sorey_chem <- dbGetQuery(con, "
  SELECT l.name AS feature, s.collection_time AS sample_date,
         la.analyte, la.value, la.units
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE l.external_station_code LIKE 'SOREY1992%'
  ORDER BY s.collection_time, l.name
")
sorey_chem_wide <- sorey_chem %>%
  select(feature, sample_date, analyte, value) %>%
  pivot_wider(names_from = analyte, values_from = value)
sorey_chem_wide
```

### Cl/B and Cl/Li ratio check against modern chemistry

Sorey & Colvard found Cl/B = 19.3 +/- 1.7 and Cl/Li = 122.5 +/- 9.9
across the CPI and SB GEO reservoirs (excluding the more dilute Cox
well), and used this constancy of conservative-element ratios to argue
the two well fields draw on chemically related source water. Recomputing
the same ratios from the newly-ingested 1950-1991 values, and comparing
against this project's own real 2024-2026 thermal-water samples (the
FIELD spring/well/seep samples, and the NDEP-promoted domestic wells),
is a direct test of whether that same geochemical signature has
persisted for over three decades.

```{r ratio-check}
hist_ratios <- sorey_chem %>%
  filter(analyte %in% c("Cl", "B", "Li")) %>%
  pivot_wider(id_cols = c(feature, sample_date), names_from = analyte, values_from = value) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li, era = "1950-1991 (Sorey & Colvard 1992)",
         year = as.integer(substr(sample_date, 1, 4)))

modern_ratios <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte IN ('Cl','B','Li')
    AND s.data_source != 'Sorey & Colvard 1992'
") %>%
  pivot_wider(id_cols = c(sample_id, feature, site_type, collection_time), names_from = analyte,
              values_from = value, values_fn = mean) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li,
         era = if_else(site_type %in% c("fumarole", "seep", "spring"),
                        "2024-2026 real thermal (FIELD)",
                        "2024-2026 NDEP domestic/background wells"),
         year = lubridate::year(lubridate::mdy_hm(collection_time, quiet = TRUE)))  # real parsed sample year (2024 or 2026)

ratio_compare <- bind_rows(
  hist_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li),
  modern_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li)
)
ratio_compare
```

```{r ratio-plot}
#| fig-width: 8
#| fig-height: 5
# Single unified plot (year on the x-axis, not era-on-y as in an earlier
# draft) -- well-name labels via ggrepel, and the t-test statistics moved
# into the figure caption rather than left as a separate prose paragraph,
# per explicit feedback this session. Excludes the Cox well (per Sorey &
# Colvard's own stated convention) and the SBRR/SBBV creek stations
# (mixing points, not end-member samples -- including them the first time
# this was tried pulled in SBRR's near-zero-B ratio and badly distorted
# the comparison).
cl_b_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "SB GEO production well PW-1" = "PW-1", "SB GEO production well PW-2" = "PW-2",
  "SB GEO production well PW-3" = "PW-3", "Hot spring 6, main terrace ACEC" = "Spring 6",
  "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001", "SBW_0002" = "SBW_0002",
  "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003"
)
ratio_plot_data <- ratio_compare %>%
  filter(!is.na(Cl_B), !grepl("Cox", feature), !feature %in% c("SBRR", "SBBV")) %>%
  mutate(
    era_short = if_else(grepl("^19", era) | era == "1950-1991", "1990-91", "2024-26"),
    label = coalesce(cl_b_labels[feature], feature)
  )
t_res_ratio <- t.test(Cl_B ~ era_short, data = ratio_plot_data)
caption_ratio <- sprintf(
  "Welch two-sample t-test: mean Cl/B = %.1f (1990-91, n=%d) vs. %.1f (2024-26, n=%d); t(%.1f) = %.2f, p = %.3f, 95%% CI [%.2f, %.2f]",
  t_res_ratio$estimate[1], sum(ratio_plot_data$era_short == "1990-91"),
  t_res_ratio$estimate[2], sum(ratio_plot_data$era_short == "2024-26"),
  t_res_ratio$parameter, t_res_ratio$statistic, t_res_ratio$p.value,
  t_res_ratio$conf.int[1], t_res_ratio$conf.int[2]
)
ggplot(ratio_plot_data, aes(x = year, y = Cl_B, color = era_short)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20, seed = 4172, show.legend = FALSE) +
  labs(x = "Year", y = "Cl / B (mass ratio)", color = "Era",
       title = "Chloride-to-boron ratio in real thermal samples, 1950-2026",
       caption = caption_ratio) +
  theme(plot.caption = element_text(hjust = 0, size = 9))
```

**Reading this (revised 2026-09-12 with a real statistical test -- see
the new Statistical Analysis section below):** this project's own 2024-2026 real thermal FIELD
samples (fumarole/seep/spring) give Cl/B roughly 20-23, in the same
general range as Sorey & Colvard's 1990-91 CPI/SB GEO value of 19.3 +/-
1.7. A Welch two-sample t-test on the 7 historical vs. 6 modern real
thermal Cl/B values (excluding the Cox well, per Sorey & Colvard's own
convention, and excluding the SBRR/SBBV creek stations, which are
mixing points, not end-member samples) finds this difference is
**small but statistically significant**: mean 19.6 (1950-1991) vs. 21.9
(2024-2026), Welch t(9.87) = -2.92, p = 0.015, 95% CI on the difference
[-4.03, -0.54]. This is a more precise, and slightly different,
statement than this notebook's original (Session 27) "closely
matching... essentially unchanged" framing -- the conservative-element
signature is genuinely similar in magnitude across 35 years, but the
two eras are not statistically indistinguishable. A plausible, not yet
tested, explanation is a systematic difference in analytical method
(UURI/AMTEC ICP for the 1990-91 samples vs. this project's own modern
lab) rather than a real 12% geochemical shift -- flagged as an open
question, not resolved here. The NDEP domestic/background well samples
(Boyd, Jeppson, Rogers) show much more scattered Cl/B ratios; their B
values sit at or near 0.05-0.18 mg/L, close to a plausible analytical
detection limit, so those ratios should be read as noisy near-DL
artifacts of dilute, non-thermal ground water rather than a real
geochemical signal -- fully consistent with this project's established
finding that these are background, not thermal, waters.

## Discharge-through-time: extending Sorey's Table 2

Table 2 of the 1992 report tabulates dated estimates of thermal-water
discharge to Steamboat Creek from 1955 through 1989. These values are
**transcribed literature numbers, not live database queries** -- they
predate any table this project's own pipeline builds -- assembled here
alongside the already-cited Sorey & Spielman (2008, 2017) follow-up
estimates and this project's own 2026 poster result to give a single,
~70-year view.

```{r discharge-timeline}
# Extended 2026-09-25: added a new database-computed point (this project's
# only real paired Cl+discharge measurement, 2026-05-01) and a second
# Collar & Huntley (1990) point (their own stated "about 80% of White's
# 1955 baseline" 1988-89 natural-only estimate, distinct from the two
# dated Table 2 values already used).
discharge_hist <- tribble(
  ~period, ~year_mid, ~gpm, ~Ls, ~source,
  "White (1968), Apr 1955",         1955, 1110, 1110 * 0.0630902, "White (1968), chloride-flux method",
  "White (1968), Apr 1964",         1964, 1385, 1385 * 0.0630902, "White (1968), chloride-flux method",
  "Shump (1985), 1981-82 avg",      1981.5, 1300, 1300 * 0.0630902, "Shump (1985), specific-conductance method (less reliable per Sorey & Colvard)",
  "Collar (1990), Jun 1988",        1988.5, 523, 523 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Collar (1990), Aug 1988",        1988.6, NA, NA, "Not counting net CPI production; unseen seepage component only reported",
  "Collar (1990), Mar 1989",        1989.2, 663, 663 * 0.0630902, "Collar (1990) via Sorey & Colvard (1992) Table 2",
  "Collar & Huntley (1990), 1988-89 natural-only", 1988.8, NA, 0.80 * 1120 * 0.0630902, "Collar & Huntley (1990): 'about 80% of that estimated by White (1968) in 1955' (1120 gpm baseline, natural discharge only, excludes CPI production)",
  "Sorey & Spielman (2008), 2007-08", 2007.5, NA, mean(c(730, 1214)) * 0.0630902, "Chloride-flux, SBRR/SBBV sites",
  "Sorey & Spielman (2017), 2008-2016 avg", 2012, NA, 60, "Chloride-flux, SBRR/SBBV sites, 24 sampling events",
  "This project (database), 5/1/2026", 2026.33, NA, 42.1, "Computed this session directly from real paired Cl (SBRR=17.1, SBBV=126 mg/L, both 2026-05-01) and USGS discharge (station 10349300, 11.2 cfs near the SBRR sample time) -- same formula as the poster estimate below, but a single day, not an averaging period",
  "Irvin & Lindsey (2026 poster), post-eruption avg", 2026, NA, 27, "This project's own chloride mass-balance estimate (averaged over its own sampling period, not a single day)"
)
discharge_hist
```

```{r discharge-plot}
#| fig-width: 9
#| fig-height: 5.5
ggplot(discharge_hist %>% filter(!is.na(Ls)), aes(x = year_mid, y = Ls)) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = grepl("This project", period)), size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(aes(label = period), size = 2.6, max.overlaps = 30, seed = 4172) +
  scale_color_manual(values = c(`TRUE` = "#d95f02", `FALSE` = "black")) +
  labs(x = "Year", y = "Thermal-water discharge (L/s)",
       title = "Thermal-water discharge to Steamboat Creek, 1955-2026",
       subtitle = "No confidence intervals shown -- no source reports a measurement uncertainty. Orange point is a\nnew, reproducible single-day estimate computed directly from this project's own database this session.")
```

**Reading this:** discharge estimates were broadly 45-90 L/s
(660-1,385 gpm-equivalent) from 1955 through the 2008-2016 Sorey &
Spielman baseline, with the 1988-89 Collar (1990) values sitting at the
low end even before accounting for that they explicitly exclude net CPI
well-field production. This project's own post-eruption ~27 L/s estimate
sits below every prior estimate in this 70-year record -- consistent
with the poster's own conclusion (conduit reactivation/redistribution,
not necessarily a new source) but worth treating as one point in a
long, variable record rather than a single before/after comparison.
None of the pre-2007 methods used the same sites (SBRR/SBBV) as the
modern chloride-flux estimates, so method consistency, not just
magnitude, matters when reading this chart -- White (1968) and Collar
(1990)'s methods and gauge locations are not identical to Sorey &
Spielman's.

**Seasonal Cl-flux variability -- a real, previously-published effect
worth naming here (2026-09-25).** Sorey & Spielman (2017) -- already
cited above, not a new document -- explicitly document a real seasonal
pattern in their own SBRR/SBBV Cl-flux computations: *"lower computed
chemical flux at downstream sites in SB Creek each summer and higher
chemical loads in creek water in the wintertime runoff period,"*
attributed in part to seasonal buildup and flushing of chloride in
shallow soils from irrigation/evaporation cycles downstream of Crane
Ditch diversions -- though they state *"the magnitude of such seasonal
changes appears to be only on the order of 25% or less."* This is a
plausible, real contributor to the gap between this session's new
single-day database estimate (42.1 L/s, 2026-05-01, spring runoff
season) and the poster's own averaged ~27 L/s figure -- a ~56% relative
difference, larger than Sorey & Spielman's own documented ~25% seasonal
envelope, so seasonality alone does not fully explain it, but it is a
real, published, and directly relevant effect rather than pure noise.
This also matches this project's own poster, which discusses Cl outflow
variability in the diluted creek -- worth a dedicated seasonal-Cl-flux
analysis once more paired Cl+discharge dates exist in the database (a
single day, as computed here, cannot separate a real seasonal signal
from ordinary day-to-day variability).

## Appendix G and Table 9: what's behind the discharge/reservoir numbers (2026-09-12)

### Appendix G: the raw synoptic Cl-flux surveys behind Table 2

Appendix G ("Stream Discharge and Chloride Flux") documents the real
method behind the Collar (1990) discharge estimates already used in the
discharge-through-time chart above: chloride treated as a conservative
tracer, an 820 mg/L thermal / 6 mg/L non-thermal end-member split, and
gaining/losing-reach mass-balance equations -- methodologically almost
identical to this project's own SBRR/SBBV Cl mass-balance approach, just
predating it by ~35 years. Chloride was measured by mercuric-nitrate or
argentometric titration (SDSU and UURI, respectively); discharge by
pygmy current meter. The underlying raw data (Table G-1) comes from four
real synoptic surveys: **6/26-6/30/1988, 7/1-7/2/1988, 8/9/1988, and
3/4/1989**, at ~40 named stations along Steamboat Creek, Galena Creek,
and the Steamboat/Chandler/Crane irrigation ditches.

**Deliberately not parsed further**: Table G-1's raw per-station
discharge/chloride/boron values are visibly OCR-misaligned in the
extracted text (values shifted relative to station-name rows, in some
blocks by a full row) -- the same class of "real data trapped in a
badly-OCR'd table" problem already flagged elsewhere in this project
(the NDEP TFT Appendix D geochemistry tables, well-log lithology
tables). The three aggregate numbers we actually use (Collar 1990's
1988-89 discharge estimates in the Table 2 chart above) are already the
synthesized output of this same appendix -- Appendix G is the raw
survey behind data already in this notebook, not a source of new,
different data points. If a denser historical Cl-flux time series is
ever wanted, it would need to be read directly from the scanned page
images (not the OCR text) station-by-station -- a real, larger effort
than anything else in this session, not attempted here.

### Table 9: reservoir parameters (transmissivity, storage coefficient)

Table 9 ("Reservoir parameters determined for the geothermal system in
the Steamboat Hills area from well test analyses and spring
hydrographs") gives real aquifer-test results from four dated tests:

```{r table9-reference}
table9 <- tibble::tribble(
  ~well_and_year, ~injection, ~transmissivity_ft2_day, ~storage_coefficient, ~observation_well, ~source,
  "Steamboat No.1, 1980", "no", "3270-9500", "1.2e-3 (calc.)", "strat 2", "Collar (1990)",
  "Steamboat No.1, Mar-May 1986", "no", "nd", "nd", "23-5?", "Berkeley Group (1987)",
  "Steamboat No.1, Mar-May 1987", "no", "nd", "1.2e-3 (calc.)", "strat 5", "Berkeley Group (1987)",
  "23-5/Cox 1-1, May-June 1987", "yes", "7930-9500", "7.8e-4 to 9.0e-4", "strat 2, 21-5", "Faulder (1987); Goranson (1989)",
  "21-5, 23-5, 83A-6, 1988", "yes", "1250-2400", "2.5e-3", "strat 9", "Collar (1990)",
  "21-5, 23-5, 83A-6, 1988 (spring response)", "yes", "nd", "2.8e-3", "spring 12, spring 42w", "Collar (1990)",
  "Spring 42w, Jun-Aug 1988 (Birsoy-Summers method)", "yes", "3000", "2.6e-3", "spring 42w", "Collar & Huntley (1990)"
)
table9
```

No table in the current schema models a dated aquifer/interference
test -- this is kept as reference documentation only, not a new
database table, since there is no immediate analytical use for it beyond
the qualitative reservoir-drawdown narrative already summarized above
("What the 1992 report says"). Worth revisiting if this project's own
hydraulic-gradient work ever needs a real transmissivity/storage
reference range for the Steamboat Hills reservoir.

### 4.6 Collar & Huntley (1990): a fault map and a discharge-deficit statistic (2026-09-12)

A second Sorey-team document, `Collar.pdf` (Collar, R.J. and Huntley, D.
1990, 12th New Zealand Geothermal Workshop), was added to the literature
folder and read directly. It largely overlaps the 1992 report above (same
wells, same hydrograph story), but contributes two things not already
captured:

**A real fault map.** Its Figure 1 ("Fault and air photo lineament map of
the Steamboat Springs geothermal area, including the location of selected
wells") shows mapped faults (dashed = inferred, dotted = concealed) and
air-photo lineaments of probable structural origin, including the named
Pleasant Valley Fault and Steamboat Springs Fault Zone, with `21-5`,
`83A-6`, `23-5`, `Cox I-1`, `PW-1/2/3`, `IW-2/3`, strat wells, `Brown
School`, `Pine Tree Ranch`, `Steinhardt`, and `GS-1`/`GS-8` all located on
it by name. Ten of these already have real coordinates in this database:

```{r collar-fig1-control-points}
dbGetQuery(con, "
  SELECT well_name, latitude, longitude FROM Wells
  WHERE well_name IN ('21-5R','23-5','83A-6','Cox I-1','PW-2','PW-3','IW-2',
                       'Brown School Geothermal Well','Steinhardt Geothermal Well','GS-5')
  ORDER BY well_name
")
```

This is a materially better fault-digitizing candidate than the
Dhakal-diagram source discussed in `README.md`'s "Planned: Leapfrog 3D
geologic model export" section (a production/injection flow diagram, not
a fault map) -- see that section (updated 2026-09-12) for the proposed
digitizing approach using these ten points as control points.

**A discharge-deficit statistic.** The paper states that 1988 spring
discharge, measured at 3-4 gal/min, was far below the 34-40 gal/min
*predicted* from White (1968)'s own spring-discharge/annual-precipitation
relationship -- a specific, citable number distinct from (and more
precise than) the aggregate gpm figures already in the Section 6
discharge-through-time chart.

**Elevation cross-check.** The paper gives well-field elevations (OESI/
SB GEO ~4700 ft; CPI 5350-5660 ft; Cox I-1 5050 ft; springs ~4650 ft),
checked against this project's own `Wells.elevation_m`:

```{r elevation-crosscheck}
dbGetQuery(con, "
  SELECT well_name, elevation_m, ROUND(elevation_m/0.3048,0) AS elevation_ft
  FROM Wells WHERE well_name IN ('PW-2','PW-3','IW-2','21-5R','Cox I-1')
  ORDER BY well_name
")
```

`IW-2`/`PW-2`/`PW-3` (4634-4730 ft) are consistent with the paper's OESI
well-field figure (~4700 ft). `21-5R` (5722 ft) sits just above the
paper's stated CPI range (5350-5660 ft) -- a minor, unresolved
discrepancy, plausibly because the paper gives one representative
figure for the whole field rather than a per-well value. `Cox I-1` had
no recorded elevation at all; the paper's specific value (5050 ft =
1539.24 m) is filled in below since, unlike the CPI range, it is
directly attributable to a single named well.



## Table 10 monitoring network vs. this project's Locations

```{r table10-check}
dbGetQuery(con, "
  SELECT location_id, external_station_code, name, site_type,
         latitude, longitude, coordinate_source
  FROM Locations
  WHERE name LIKE '%Boyd%' OR name LIKE '%Curti%' OR name LIKE '%Rogers%'
     OR name LIKE '%Jeppson%' OR name LIKE '%Herz%' OR name LIKE '%Zolezzi%'
     OR external_station_code IN ('SB5','SB6')
  ORDER BY name
")
```

Sorey & Colvard's Table 10 ("Current monitoring sites for CPI and SBG
geothermal developments") describes exactly this same set of domestic
wells and the Rhodes Road creek-monitoring point as the active 1988-92
monitoring network -- confirming this project's own SBRR site and its
NDEP/Klein-resolved domestic wells sit at genuinely long-monitored
locations, not newly-chosen ones. None of these have a linked `Wells`
row or `Water_Level_Observations` history yet, which is a real,
documented gap (see Section 7 caveats) rather than an oversight to
silently work around here.

**Aside (found and fixed, 2026-09-12): the malformed `"7-07-10"`
timestamp.** `Water_Level_Observations` has 73 wells and 153,437 rows
spanning 1969-06-23 to 2026-01-01; one row (well-log document 104216, a
"WELL DRILLER'S PLUGGING REPORT" with `completion_date_raw = "7/10/07"`)
had timestamp `"7-07-10"`. Root cause: `ingest_well_logs.R`'s
`.safe_completion_date()` tries `as.Date(x, format = "%m/%d/%Y")` first,
and R's `%Y` silently accepts a 2-digit year as a literal, unpadded year
value (year 7 CE) rather than rejecting the format mismatch the way the
function's design assumed it would -- so the very first format "matched"
with a nonsensical date instead of falling through to a better one.
Fixed by adding an explicit `%m/%d/%y` attempt and a plausible-year-range
sanity check (1900 to next calendar year) that every candidate parse must
pass before being accepted; the one bad row is corrected to `2007-07-10`.
No other row in `Well_Log_Documents`/`Water_Level_Observations` was
affected -- this was the only 2-digit-year-formatted completion date in
the batch.

## Statistical Analysis: Is There a Real Chloride Trend? (2026-09-12)

Sections above establish real Cl values across 1950-2026 and a
qualitative sense that thermal-water chemistry looks broadly similar
over time. This section applies actual statistical tests to that claim
-- per an explicit request that this notebook's historical-comparison
work be statistic-driven, not just descriptive.

### A combined 1950-2026 Cl dataset

```{r cl-timeline-data}
library(lubridate)
cl_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, s.data_source, la.value AS Cl
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte = 'Cl'
")

# Sampling_Events/Samples.collection_time mixes Unix-epoch-seconds,
# MM/DD/YYYY H:MM, and ISO YYYY-MM-DD formats across sources -- the
# same class of dual/mixed-format issue already flagged project-wide
# for Sampling_Events.date (notebook 05). Parsed here defensively,
# trying each format in turn rather than assuming one.
.parse_any_date <- function(x) {
  num <- suppressWarnings(as.numeric(x))
  out <- as.Date(rep(NA, length(x)))
  is_num <- !is.na(num) & !grepl('[A-Za-z]', x)
  out[is_num] <- as.Date(as.POSIXct(num[is_num], origin = '1970-01-01', tz = 'UTC'))
  rem <- !is_num
  out[rem] <- suppressWarnings(as.Date(x[rem], format = '%m/%d/%Y %H:%M'))
  rem2 <- rem & is.na(out)
  out[rem2] <- suppressWarnings(as.Date(x[rem2], format = '%Y-%m-%d'))
  out
}
cl_data$date <- .parse_any_date(cl_data$collection_time)
cl_data$year <- year(cl_data$date)

# Added 2026-09-25: Steinhardt's two narrative-based bracket samples
# (1987/1990, see Section 4.2 correction above), now that they exist.
focus_wells <- c(
  'CPI production well 21-5','CPI production well 83A-6','CPI production well 23-5',
  'CPI injection well Cox 1-1 (Cox well)','SB GEO production well PW-1','SB GEO production well PW-2',
  'SB GEO production well PW-3','Well GS-5, 1950 sample (White 1968)','New seep adjacent to well GS-5 (1991)',
  'Hot spring 6, main terrace ACEC',
  'SBF_0001','SBW_0001','SBW_0002','SBS_0001','SBS_0002','SBS_0003',
  'Eich Well','Soccer Field Monitoring Well','Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well',
  'Steinhardt Geothermal Well (narrative Cl trend)'
)
background_names <- c('Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well')
cl_focus <- cl_data[cl_data$feature %in% focus_wells, ]
cl_focus$group2 <- ifelse(cl_focus$feature %in% background_names, 'background/domestic',
                    ifelse(cl_focus$feature %in% c('Eich Well','Soccer Field Monitoring Well'), 'intermediate (NDEP)', 'thermal'))

well_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "CPI injection well Cox 1-1 (Cox well)" = "Cox I-1", "SB GEO production well PW-1" = "PW-1",
  "SB GEO production well PW-2" = "PW-2", "SB GEO production well PW-3" = "PW-3",
  "Well GS-5, 1950 sample (White 1968)" = "GS-5", "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "Hot spring 6, main terrace ACEC" = "Spring 6", "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001",
  "SBW_0002" = "SBW_0002", "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003",
  "Eich Well" = "Eich", "Soccer Field Monitoring Well" = "Soccer Field", "Boyd Domestic Well" = "Boyd",
  "Jeppson Geothermal Well" = "Jeppson", "Rogers Domestic Well" = "Rogers",
  "Steinhardt Geothermal Well (narrative Cl trend)" = "Steinhardt"
)
cl_focus$label <- well_labels[cl_focus$feature]
nrow(cl_focus)
```

```{r cl-timeline-plot}
#| fig-width: 9
#| fig-height: 5.5
# Reworked 2026-09-25 per explicit feedback: well-name labels (not just
# color groups), lines connecting repeat samples of the same well/spring
# through time, and the 6 single-snapshot 2026 FIELD thermal samples
# (SBF_0001, SBW_0001/2, SBS_0001-3) dropped from THIS plot -- they add
# no temporal information (all one date) and were crowding the 2020s
# cluster; they remain visible in the Cl/B ratio plot above.
exclude_2026_single <- c("SBF_0001","SBW_0001","SBW_0002","SBS_0001","SBS_0002","SBS_0003")
cl_focus_timeline <- cl_focus[!cl_focus$feature %in% exclude_2026_single, ]
ggplot(cl_focus_timeline, aes(x = date, y = Cl, color = group2)) +
  geom_line(aes(group = feature), alpha = 0.4, linewidth = 0.5, show.legend = FALSE) +
  geom_point(size = 2.5) +
  ggrepel::geom_text_repel(aes(label = label), size = 3.2, max.overlaps = 20, seed = 4172) +
  scale_y_log10() +
  labs(x = NULL, y = "Chloride (mg/L, log scale)", color = "Sample group",
       title = "Chloride concentration by well/spring through time, 1950-2026",
       subtitle = "Wells with real repeat measurements over time (Steinhardt, Boyd, Jeppson, Rogers, Eich,\nSoccer Field) plus single historical thermal anchors for context",
       caption = "The 6 single-snapshot 2026 FIELD thermal samples (SBF_0001, SBW_0001/2, SBS_0001-3) are shown separately in the Cl/B ratio plot above to reduce crowding here.")
```

### Is there a statistically significant Cl trend over time?

A naive linear regression of Cl against year across every
thermal-influenced sample (thermal + intermediate groups, n = 19)
finds a significant negative slope:

```{r cl-trend-naive}
fit_all <- lm(Cl ~ year, data = cl_focus[cl_focus$group2 != 'background/domestic', ])
summary(fit_all)
```

**But this is confounded**, and re-running it after excluding the two
'intermediate' NDEP wells (Eich, Soccer Field -- only ever sampled in
2024, and already established elsewhere in this notebook as
partially-diluted, not pure end-member thermal water) weakens the
trend to marginal significance:

```{r cl-trend-pure}
pure_thermal <- cl_focus[cl_focus$group2 == 'thermal', ]
fit_pure <- lm(Cl ~ year, data = pure_thermal)
summary(fit_pure)

pure_thermal$era <- ifelse(pure_thermal$year < 2000, 'pre-2000', '2024-2026')
aggregate(Cl ~ era, data = pure_thermal, FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
```

**Honest reading:** including the intermediate wells (p = 0.0099,
R2 = 0.33) versus excluding them (p = 0.099, R2 = 0.20) changes the
conclusion from 'significant decline' to 'suggestive but not
significant at the conventional 0.05 threshold' -- purely because of
which samples are grouped as 'thermal.' The pre-2000 real-thermal mean
(811 mg/L, n=9) is numerically higher than the 2024-2026 mean (699
mg/L, n=6), but with only 15 points spanning 76 years and no repeated
measurements of the same individual well across both eras, this
dataset does not have the statistical power to distinguish a real
long-term decline from era-to-era sampling differences (different
wells sampled in 1990-91 than in 2024-2026). This is reported as an
open, unresolved question -- not a confirmed trend -- consistent with
this notebook's standing practice of not overstating what a small
historical dataset can support.

### Cl/B ratio: a real two-sample test (not just eyeballing)

The Cl/B ratio comparison introduced earlier in this notebook is
repeated here as an explicit, reproducible statistical test rather
than a visual read of the plot:

```{r cl-b-test}
cl_b_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la JOIN Samples s ON s.sample_id=la.sample_id JOIN Locations l ON l.location_id=s.location_id
  WHERE la.analyte IN ('Cl','B') AND s.data_source != 'NDEP'
")
cl_b_wide <- reshape(cl_b_data[, c('sample_id','feature','collection_time','analyte','value')],
                      idvar = c('sample_id','feature','collection_time'), timevar = 'analyte', direction = 'wide')
names(cl_b_wide) <- gsub('value\\.', '', names(cl_b_wide))
cl_b_wide <- cl_b_wide[!is.na(cl_b_wide$Cl) & !is.na(cl_b_wide$B), ]
cl_b_wide$Cl_B <- cl_b_wide$Cl / cl_b_wide$B
cl_b_wide$year <- year(.parse_any_date(cl_b_wide$collection_time))
# Exclude Cox (per Sorey & Colvard's own stated convention) and the
# SBRR/SBBV creek stations (mixing points, not end-member samples --
# including them wrongly pulled in a Cl/B ratio near 1500 dominated by
# SBRR's near-zero B).
cl_b_wide <- cl_b_wide[!grepl('Cox', cl_b_wide$feature) & !cl_b_wide$feature %in% c('SBRR','SBBV'), ]
cl_b_wide$era <- ifelse(cl_b_wide$year < 2000, '1950-1991', '2024-2026')
cl_b_wide[, c('feature','year','Cl','B','Cl_B','era')]

t.test(Cl_B ~ era, data = cl_b_wide)
```

This confirms the result already folded into the "Cl/B and Cl/Li ratio
check" section above: a small (19.6 vs. 21.9) but real, statistically
significant (p = 0.015) difference between the two eras -- similar in
magnitude, not statistically identical.

### Discharge-estimate uncertainty: an honest limitation, not a fabricated CI

The discharge-through-time chart earlier in this notebook (White 1968,
Shump 1985, Collar 1990, Sorey & Spielman 2008/2017, this project's
2026 poster) plots point estimates with no confidence interval. This
is deliberate, not an oversight: none of the source documents report a
measurement-uncertainty or replicate-based error bar for their
discharge estimates, and the raw synoptic measurements behind the
pre-2007 estimates are not available in a form this project can
re-derive an uncertainty from (Appendix G's per-station table, as
documented above, is real but too OCR-misaligned to trust for this).
Fabricating a confidence interval from nothing would be worse than
reporting none. A future, real improvement would be to propagate this
project's own analytical Cl-measurement uncertainty (a stated lab
detection-limit/precision figure, not yet compiled here) through the
chloride mass-balance formula for the SBRR/SBBV-based estimates only,
where the raw measurements are directly available -- flagged as a
concrete next step, not attempted in this session.

## Barometric efficiency: a documented, not-yet-implemented capability

Sorey & Colvard (p. 85-88, Figs. 39-40) computed **barometric
efficiency** (BE) -- the ratio of water-level change in a well/spring to
the corresponding barometric-pressure change -- by linear regression of
spring water level against Reno Airport barometric pressure, from 3-4x
daily manual readings over one week in October 1988. Reported values:
BE = 0.42 for spring 6 (r = 0.79) and 0.45 for spring 12 (r = 0.44);
White (1968) found BE 0.2-1.18 across different vents, with higher
values (occasionally exceeding 1.0) where the spring vent's fluid column
is at or near boiling.

**Why this project can't run this for real yet:** it requires
*continuous* barometric pressure paired with *continuous* water level at
the same site. Checking both:

```{r barometric-check}
dbGetQuery(con, "SELECT DISTINCT parameter, unit FROM Weather_Observations")
```

NOAA's ingested weather parameters are precipitation and air temperature
only -- **no barometric/station pressure parameter currently exists
anywhere in this database.** `Water_Level_Observations` itself is
discrete manual/transducer readings, not always continuous, and none of
the wells in Section 4's cross-reference table currently overlap in time
with any pressure record anyway.

This mirrors the same "build the capability, wait for real overlapping
data" pattern already used for PHREEQC's mixing/inverse/gas-phase
modules (`06_phreeqc_geochemical_modeling.qmd`): a stub function with
only a synthetic self-test, not wired into `run_pipeline.R`.

```{r barometric-stub}
#| eval: false
# Not run by default -- synthetic self-test only, requires real paired
# continuous barometric-pressure + water-level data to be meaningful.
compute_barometric_efficiency <- function(water_level, barometric_pressure) {
  fit <- lm(water_level ~ barometric_pressure)
  list(
    BE = unname(coef(fit)[2]),
    r_squared = summary(fit)$r.squared,
    model = fit
  )
}

demo_barometric_efficiency <- function(seed = 4172) {
  set.seed(seed)
  n <- 168 # one week, hourly
  baro <- 30 + cumsum(rnorm(n, 0, 0.02))          # synthetic barometric pressure, inHg
  wl <- 5 - 0.4 * (baro - mean(baro)) + rnorm(n, 0, 0.02)  # synthetic BE ~ 0.4
  compute_barometric_efficiency(wl, baro)
}
demo_barometric_efficiency()
```

**Recommendation, not a request for new instrumentation decisions here:**
if a future logger deployment or manual monitoring effort pairs
continuous water level at a spring/well with continuous (or even just
synchronized 3-4x/day manual) barometric pressure, this function is
ready to compute a real, Sorey-comparable BE immediately.

## Why this matters for the project

Sorey & Colvard (1992) does three things no other single source in this
project's literature folder does at once: it gives this project's
chemistry record a **35-40 year deeper baseline** (Table 1, now
ingested), it gives the chloride-discharge thesis question a **~70-year
timeline** instead of just the 2008-2026 window (Section 6), and it
independently corroborates several of this project's own
well/site-identity resolutions made from entirely unrelated sources in
earlier sessions (Section 4) -- itself a form of validation this
project didn't have before. It also demonstrates, with real 1988-era
field methods, that this project's own instrumentation program (SBRR/SBGG
conductivity loggers, temperature loggers, NDEP-resolved domestic wells)
sits on genuinely decades-old monitoring ground, not a newly invented
network. The one capability it demonstrates that this project cannot yet
reproduce -- barometric efficiency -- is now a documented, ready-to-run
gap rather than an unrecognized one.

## Changelog

| Date | Change |
|---|---|
| 2026-09-12 | Notebook created. Sorey & Colvard (1992) read directly (`pdftotext -layout`); Table 1 chemistry (9 samples, 1950-1991) ingested for the first time via `ingest_historical_sorey1992.R`; well/site cross-reference built against `Wells`/`Well_Aliases`/`Locations`; Cl/B ratio confirmed consistent between 1990-91 and 2024-2026 real thermal samples; discharge-through-time table assembled (White 1968 through this project's 2026 poster); barometric efficiency documented as a recommended, not-yet-runnable capability (no barometric-pressure parameter exists in this database yet). |
| 2026-09-12 (follow-up) | Resolved `83A-6` and `Cox 1-1` via exact NBMG Geothermal_Wells name matches; discovered and resolved that `"GS-58"`/`"GS-59"` are OCR artifacts of well `GS-5` (its own table footnote superscripts fused onto the name); all four now have real coordinates in `Locations`/`Wells`. Also found and fixed a 2-digit-year date-parsing bug in `ingest_well_logs.R`'s `.safe_completion_date()` that had produced one malformed `Water_Level_Observations` timestamp (`"7-07-10"`, corrected to `2007-07-10`). |
| 2026-09-12 (Tables 3/6/7/8/9, Appendix G) | Registered strat wells 2/5/6/7/8/9/13/14 (Table 8) and Table 3's South Truckee Meadows wells as real Wells rows; resolved Steinhardt and Brown School via exact NBMG matches, Bianco tentatively via NDWR owner-name + PLSS centroid; PTR-1/2 and STMGID MW-3/4 remain unresolved despite a fresh search; Herz-2 left inconclusive between two ambiguous NBMG candidates. Filled `Wells.top_perforation`/`bottom_perforation` (previously 100% NULL) for 7 wells from Tables 6/7, deliberately leaving 23-5 and IW-2 unfilled due to real depth conflicts with current records (documented as historical facts instead). Documented Appendix G's real methodology/survey dates (not parsed further -- too OCR-misaligned) and Table 9's reservoir-parameter values as reference only. |
| 2026-09-12 (session 29) | Added Collar & Huntley (1990) as a source (real fault map, Figure 1, used as a fault-digitizing candidate; second reservoir-parameter data point; discharge-deficit statistic; elevation cross-check, Cox I-1 elevation filled). Added a real statistical-analysis section: naive Cl-vs-year regression is confounded by which wells are grouped as 'thermal' (significant at p=0.0099 with intermediate wells included, only marginal at p=0.099 without them); a proper two-sample t-test on Cl/B ratios finds the 1950-1991 vs. 2024-2026 difference (19.6 vs. 21.9) is small but statistically significant (p=0.015), revising this notebook's earlier, less precise 'essentially unchanged' claim. Documented why the discharge-timeline chart has no confidence intervals (no source reports measurement uncertainty; fabricating one would be worse than omitting it). |
| 2026-09-25 | Reworked all three Cl/discharge plots per direct critique: Cl-timeline now shows well-name labels and lines connecting real repeat measurements (dropped the 6 single-snapshot 2026 FIELD samples from this specific plot to reduce crowding); discharge-through-time gained two new points (a Collar & Huntley 1988-89 natural-only estimate, and a new point computed directly from this project's own database -- the only real paired Cl+discharge measurement, 2026-05-01, giving 42.1 L/s vs. the poster's averaged 27 L/s); Cl/B ratio plot consolidated into one figure (year on the x-axis, well labels, t-test statistics moved into the figure caption instead of prose). Corrected a real data-quality error: Steinhardt's chloride range was misattributed from Table 3's OCR-misaligned row position -- the report's own narrative text (p.52) unambiguously ties a 300-to-140 mg/L decline since 1987 to Steinhardt, not the '16-22 mg/L' previously recorded; two real dated Cl samples (1987=300, ~1990=140 mg/L) added to Lab_Analyses. Added a paragraph on Sorey & Spielman (2017)'s own documented seasonal Cl-flux variability (~25% or less, summer-low/winter-high) as a real, previously-published effect relevant to (but not fully explaining) the gap between this session's new single-day estimate and the poster's averaged figure. |

```{r cleanup}
#| include: false
dbDisconnect(con)
```
