---
title: "Historical Context: Sorey & Colvard (1992) and the Long-Run Steamboat Record"
subtitle: "Cross-referencing a 1988-1992 USGS study against this project's database"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: true
  message: true
---

## Purpose and scope

`docs/literature/Sorey_StmbtSprgsHSActivity_1992.pdf` -- Sorey, M.L. and
Colvard, E.M. (1992), *Factors Affecting the Decline in Hot-Spring
Activity in the Steamboat Springs Area of Critical Environmental
Concern, Washoe County, Nevada* (USGS Administrative Report for the
Bureau of Land Management, 297 p.) -- was added to this project's
literature folder on 2026-09-12. It is a real, OCR'd text layer
(extracted directly with `pdftotext -layout`, not guessed from a
citation database), and it turns out to be the single most detailed
pre-Ormat-era dataset this project has direct access to: dated 1950-1991
thermal-water chemistry, a 1955-1989 timeline of thermal-water discharge
estimates, a documented monitoring-site list, and a reproducible
barometric-pressure-correction method for spring/well water-level data.

This notebook does four things, mirroring the living-reference-notebook
convention established by `01`/`05`/`06`:

1. Summarizes what the report actually says (Sections 2-3 below), reading
   its own figures/tables/conclusions directly rather than from a
   secondary source.
2. Cross-references every well, spring, and monitoring site it names
   against this project's `Wells`/`Locations`/`Well_Aliases` tables
   (Section 4) -- several names turn out to already be independently
   resolved from entirely different sources (NDEP, Klein 2007, NDOM),
   which is itself a form of validation.
3. Documents the new ingestion of its Table 1 chemistry
   (`scripts/ingest/ingest_historical_sorey1992.R`,
   `data/raw/historical/sorey1992_table1_chemistry.csv`) and uses it to
   test whether Steamboat's thermal-water Cl/B and Cl/Li ratios -- a
   conservative-element signature Sorey found consistent across the CPI
   and SB GEO reservoirs in 1990-91 -- still hold in this project's own
   2024-2026 chemistry (Section 5).
4. Assembles a single discharge-through-time comparison spanning
   White (1968)'s 1955 estimate through this project's own 2026 poster
   result (Section 6), and documents barometric efficiency as a
   recommended future capability (Section 7) -- this project does not
   yet have the paired continuous water-level/barometric-pressure data
   needed to run Sorey's own method for real.

```{r setup-07}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)
```

## What the 1992 report says

**Study design.** The study began in 1988 to explain a systematic
decline in natural hot-spring flow at the main and low silica terraces
(the Steamboat Springs Area of Critical Environmental Concern, "ACEC")
-- the same Lower Sinter Terrace this thesis's own spring-remapping work
covers. Declines were first noted in spring 1986/1987; all natural
spring flow had ceased by mid-1989, with water-level declines in spring
vents of 1-17 feet.

**Conclusions (its own numbered list, p. 100-101, paraphrased):**

- ~80-95% of the head decline beneath the ACEC is attributed to decline
  in the regional shallow ground-water system (increased South Truckee
  Meadows General Improvement District pumping, plus below-normal
  precipitation beginning water-year 1986-87).
- ~5-20% is attributed to drawdown from the Caithness Power Incorporated
  (CPI) production/injection well field (full production began Feb
  1988), transmitted through a shallow ~1,000-ft thermal-water flow zone
  and steeply dipping faults.
- SB GEO well-field influence (full production began Jan 1987, 0.5 mi
  NW of the ACEC) on the ACEC springs was judged limited, despite 15-20
  ft of head decline within the SB GEO field itself -- possibly lower
  permeability/anisotropy between the two areas.

**Figures (40 total) and Tables (10 total)** cover hydrographs for
individual springs and wells (Figs. 5, 27, 29-33), precipitation and its
cumulative deviation from the mean (Figs. 15-18), a stable-isotope
altitude/boiling diagram (Fig. 9), electrical-resistivity mapping (Fig.
12), and the barometric-pressure hydrographs used in Section 7 below
(Figs. 39-40). Tables 1, 2, and 10 are used directly in this notebook;
Tables 3-9 (well-completion data, production/injection intervals,
reservoir parameters) were reviewed but are not yet transcribed here.

## Well and monitoring-site cross-reference

Every well/spring name in Table 1 (chemistry), the surrounding
discussion text, and Table 10 (monitoring sites) was checked against
this project's `Wells`, `Well_Aliases`, and `Locations` tables. Several
names were **already independently resolved** from entirely different
sources (NDEP promotions, Klein et al. 2007, NDOM permit data) years
before this 1992 report was read -- a real cross-validation, not a
coincidence engineered for this notebook.

```{r crossref-table}
crossref <- tribble(
  ~sorey_name, ~feature_type, ~status, ~note,
  "21-5 (CPI)", "well", "Resolved", "Well_Aliases '21-5' -> canonical 21-5R (well_id 94)",
  "23-5 (CPI)", "well", "Resolved", "Exact Wells.well_name match (well_id 96)",
  "83A-6 (CPI)", "well", "Resolved (2026-09-12)", "Exact NBMG Geothermal_Wells name match ('Well No. 83A-6', API 27-031-90080)",
  "Cox 1-1 / Cox well", "well (injection)", "Resolved (2026-09-12)", "Exact NBMG match ('Cox I-1', API 27-031-90051); previously flagged unresolved as COX-1/Cox1-1",
  "PW-1, PW-2, PW-3 (SB GEO)", "well", "Resolved", "Exact Wells.well_name matches (well_id 79/80/81)",
  "IW-1, IW-2, IW-3 (SB GEO)", "well (injection)", "Resolved", "IW-1 alias -> 45-28; IW-2/IW-3 exact canonical names",
  "OW-1, OW-2 (SB GEO observation)", "well (monitor)", "Partial", "OW-2 exact match (well_id 110); OW-1 not found",
  "28-32", "well", "Resolved", "Exact Wells.well_name match (well_id 125)",
  "Strat wells 2, 5, 6, 7, 8, 9, 13, 14", "well (stratigraphic test)", "Registered, no coordinate (2026-09-12)", "Real provisional Wells rows with depth/formation/temperature/water-level data (Table 8, Section 4.2); no coordinate -- see proximity notes for strat 13 (near 23-5) and strats 2/5/9 (near Cox I-1)",
  "Well GS-58, Spring GS-59", "well/seep (pre-Ormat)", "Resolved (2026-09-12)", "Both are OCR artifacts of well GS-5 (see Section 4.1) -- exact NBMG match, API 27-031-80055",
  "Hot spring 6 (main terrace ACEC)", "spring", "Unresolved", "Coordinate-less provisional Locations row; no modern counterpart identified",
  "Steamboat Creek at Rhodes Road", "surface water", "Resolved (independently)", "Locations 'SB5'/'SB6' -- same site this project's SBRR conductivity logger occupies",
  "Curti Barn Well, Curti Domestic Well", "domestic/geothermal well", "Resolved (independently)", "Klein et al. (2007)-sourced Locations rows",
  "Boyd Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Boyd Domestic Well'",
  "Rogers Well", "domestic well", "Resolved (independently)", "NDEP-promoted Locations row 'Rogers Domestic Well'",
  "Jeppson Well", "geothermal well", "Resolved (independently)", "NDEP-promoted Locations row 'Jeppson Geothermal Well'",
  "Steinhardt Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG name match ('Steinhardt Geothermal Well')",
  "Brown School Well", "domestic well", "Resolved (2026-09-12)", "Exact NBMG owner-name match ('Brown Elementary School')",
  "Bianco Well", "domestic well", "Tentative (2026-09-12)", "NDWR owner-name match ('BIANCO, U J'); PLSS-section coordinate only, and a flagged depth discrepancy (232 ft vs. Table 3's 400 ft)",
  "Herz-2 (geothermal)", "well", "Inconclusive (2026-09-12)", "Could be either of NBMG's two ambiguous 'Harold Herz Geothermal Well 1/2' candidates (flagged since Session 3); Table 3's own OCR is too garbled to disambiguate -- not guessed",
  "PTR-1, PTR-2 (Pine Tree Ranch)", "domestic well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows with real depth/chloride/water-level data; fresh NDWR/NBMG owner-name search found nothing",
  "STMGID MW-3, MW-4", "monitor well", "Still unresolved (re-checked 2026-09-12)", "Registered as provisional Wells rows; NBMG's only STMGID-named record ('Well 4 Shadowridge') is a production well, not this monitor-well pair -- not used",
  "MacKay Well, Woods Well, Tangen Well, STMGID Went",
    "domestic/monitor well", "Unresolved", "Consistent with prior sessions' independent 'unresolved' flags for these"
)
crossref
```

**Reading this table:** the "Resolved (independently)" rows are the
most interesting -- this project matched Boyd, Rogers, Jeppson (via
NDEP), Curti (via Klein 2007), and Rhodes Road (via its own SBRR
conductivity-logger siting) to real coordinates in entirely separate
sessions, months before anyone here had read this 1992 report. Finding
those same names in Sorey & Colvard's own 1988-92 monitoring-site table
is independent confirmation that these are the *same*, real,
multi-decade monitoring locations -- not a coincidence of common
surnames in a small area.

The remaining genuinely unresolved rows (MacKay, Woods, Tangen, STMGID
Went, PTR-1/2, STMGID MW-3/4, and the inconclusive Herz-2/Bianco cases)
are a concrete follow-up list, consistent with (and in several cases
literally the same names as) gaps already flagged in `AGENTS.md` from
unrelated well-matching sessions. None are guessed here.

### 4.1 Resolving 83A-6, Cox 1-1, and the "GS-58"/"GS-59" OCR artifact (2026-09-12)

A direct search of NBMG's statewide `Geothermal_Wells.csv` and
`GEOTHERM06102019.csv` compilations (`data/raw/nbmg/`) found exact
well-name matches for `83A-6` (API 27-031-90080, a CPI production well
drilled 1987) and `Cox I-1` (API 27-031-90051, a CPI injection well,
previously drilled as observation well "Cox No. 1" for Phillips
Petroleum in 1981) -- both are now real `Wells` rows with NBMG-sourced
coordinates (`data/raw/wells/sorey1992_nbmg_resolved_wells.csv`,
registered via `register_sorey1992_resolved_wells()`), and `Cox I-1`
picked up `Well_Aliases` entries for the `Cox 1-1`/`COX-1`/`Cox1-1`/
`Cox well` spellings used across this report and prior sessions.

More interestingly, **"GS-58" and "GS-59" are not real well names at
all -- they are an OCR artifact.** Re-reading Table 1's own footnote
list resolves this precisely:

> `8From White (1968).`
> `9New seep adjacent to well GS-5, analysis by Nevada Division of Health Laboratory.`

Both footnote markers ("8" and "9") were superscripts attached to a
single real well name, **GS-5**, in the original typeset table, and the
PDF's OCR layer fused each digit directly onto the name it followed --
producing "GS-58" (really "GS-5" + footnote 8) and "GS-59" (really
"GS-5" + footnote 9). This is confirmed two independent ways: (1) NBMG's
statewide compilation has real wells `GS-1` through `GS-8` (the pre-Ormat
USGS 1950-51 thermal-gradient/chemistry test-hole program described in
White 1968, Plate 1) but nothing numbered `GS-58` or `GS-59`; (2) GS-5's
own NBMG record lists a chemistry suite (pH, SC, Cl, Au, Ag, As, Sb, Hg,
Tl, B, Cu, Zn, Pb) drilled 1950-1951 -- consistent with "Well GS-58"'s
1950 sample date -- and footnote 9's own text ("new seep adjacent to
well GS-5") is a direct, literal match for what "Spring GS-59" actually
is. Both rows in `sorey1992_table1_chemistry.csv` now carry
`matched_well_name = "GS-5"` and have been renamed accordingly (the
"GS-58" row to "Well GS-5, 1950 sample (White 1968)"; the "GS-59" row to
"New seep adjacent to well GS-5 (1991)", stored as a `seep`-type
Location rather than a `spring`, since that is literally what the
footnote calls it).

```{r resolved-wells-check}
dbGetQuery(con, "
  SELECT external_station_code, name, latitude, longitude, site_type
  FROM Locations WHERE external_station_code LIKE 'SOREY1992%'
  ORDER BY external_station_code
")
```

**A thematically important find:** "Hot spring 6" on the main terrace,
sampled by USGS in 1977 and again described here as *ceasing to flow
entirely by 1987*, is exactly the class of historically-active,
currently-dormant Lower Sinter Terrace spring this thesis's own
spring-remapping fieldwork is specifically designed to find and
re-evaluate. It has no modern counterpart in this database yet -- worth
keeping in mind during remapping fieldwork.

### 4.2 Table 3: South Truckee Meadows ground-water wells (2026-09-12)

Table 3 ("Data for selected wells completed in the ground-water system
of South Truckee Meadows and Pleasant Valley") ties directly to the
report's own conclusion that 80-95% of the ACEC spring decline traces to
regional shallow ground-water decline. It gives real depth/temperature/
chloride-range/water-level-decline data for PTR-1, PTR-2, STMGID MW-3,
MW-4, Bianco, Boyd, Steinhardt, Brown School, and Herz-2 -- but, like
Table 8, has **no coordinates of its own**. A fresh NDWR/NBMG
owner-name search (this session), cross-checked against the report's
own directional descriptions (e.g. Steinhardt "northeast of the ACEC"),
found:

- **Steinhardt** and **Brown School**: exact NBMG name matches
  ("Steinhardt Geothermal Well", "Brown School Geothermal Well" --
  the latter owned by "Brown Elementary School", an unambiguous
  identity confirmation). Both directions check out against the
  report's own description. Now real, coordinate-having `Wells` rows.
- **Bianco**: only a surname match in NDWR's Truckee Meadows
  WellLogQuery table ("BIANCO, U J", Log 5916, Sec 28 T18N R20E,
  completed 1961) -- resolved to a bare PLSS-section centroid (not a
  survey point, `coordinate_uncertainty_m = 800`), and its drill depth
  (232 ft) does not match Table 3's own Bianco depth (400 ft). Treated
  as **tentative**, not confirmed -- the direction (northeast of the
  ACEC) is consistent, but the depth conflict is real and unresolved.
- **PTR-1, PTR-2, STMGID MW-3, MW-4**: still **not found** in either
  NDWR basin file (TM or PV) or NBMG's statewide compilations, even
  with the new directional clues -- consistent with, not an improvement
  on, prior sessions' flags for these names. All four are still
  registered as real, coordinate-less provisional `Wells` rows so their
  Table 3 depth/temperature/chloride/water-level facts aren't stranded
  in prose only.
- **Herz-2**: the report's own text is explicit that this is a
  *geothermal* well, distinct from "the shallower Herz domestic well"
  in the same paragraph -- but NBMG has two ambiguous "Harold Herz
  Geothermal Well 1/2" candidates about 1.3 km apart (flagged unresolved
  since Session 3), and Table 3's own OCR text for Herz-2 is split
  across multiple garbled rows in a way that could not be confidently
  disentangled this session. Left inconclusive rather than guessed.

```{r table3-check}
dbGetQuery(con, "
  SELECT well_name, well_role, latitude, longitude, coordinate_source, coordinate_uncertainty_m
  FROM Wells
  WHERE well_name IN ('Steinhardt Geothermal Well', 'Brown School Geothermal Well', 'Bianco',
                       'Pine Tree Ranch-1', 'Pine Tree Ranch-2', 'STMGID MW-3', 'STMGID MW-4')
  ORDER BY well_name
")
```

### 4.3 Table 8: stratigraphic test wells (2026-09-12)

Table 8 ("Selected data for stratigraphic test wells in the Steamboat
Hills area") gives perforated depth, formation, bottom-hole temperature,
and 1987-1989 water-level change for strat wells 2, 5, 6, 7, 8, 9, 13,
and 14 -- the wells behind the "Strat wells... Unresolved" cross-
reference gap. Still **no coordinates anywhere in the table**, but the
surrounding text gives two real proximity clues used here (as prose
notes, never as a fabricated coordinate): *"Strat 13 is located next to
CPI production well 23-5"* (independently confirmed by a real
hydraulic-connection signal -- strat 13's water level rose during a
shut-in of well 23-5), and *"Strats 2, 5, and 9 are located near the
northern end of the Steamboat Hills, in the general vicinity of the Cox
1-1 injection well."* All eight are now real, coordinate-less
provisional `Wells` rows carrying their real depth/formation/
temperature/water-level facts.

```{r table8-check}
dbGetQuery(con, "
  SELECT well_name, well_role, notes
  FROM Wells WHERE well_name LIKE 'strat %'
  ORDER BY well_name
")
```

Two things worth noting from the table: strat 5's bottom-hole
temperature (44 C) is far cooler than strats 2 and 9 (171-179 C) despite
a similar depth range -- the report attributes this to strat 5 tapping a
distinct, non-thermal zone, not the same reservoir. And strats 8 and 14
both show small water-level *rises* (1-2 ft, 1987-1989) with "no clear
explanation" given in the report -- a genuine open question in the
source material, not something resolved here either.

### 4.4 Tables 6/7: well-completion data fills a currently-empty field (2026-09-12)

`Wells.top_perforation`/`bottom_perforation` were NULL for every one of
the CPI/SB GEO wells already resolved in this database, even though
Tables 6 and 7 ("CPI well-completion information", "SB GEO well
completion information") report real casing-depth and open-hole-
interval data for most of them. The table's own OCR layout has visible
column-alignment slippage, so each well's numbers were individually
cross-checked (casing depth + open-hole thickness = total depth,
confirmed to reconcile exactly for every SB GEO well and for 83A-6) before
being trusted:

```{r table67-check}
dbGetQuery(con, "
  SELECT well_name, total_depth, top_perforation, bottom_perforation
  FROM Wells
  WHERE well_name IN ('83A-6','Cox I-1','PW-1','PW-2','PW-3','21-5R','IW-3','23-5','IW-2')
  ORDER BY well_name
")
```

**Two real, deliberately-unfilled conflicts**: `23-5`'s 1990-era report
depth (2422 ft) is 19% shallower than the 3001 ft already on record
(NDOM-sourced), and `IW-2`'s 1990 report depth (1403 ft) is less than a
third of the 4700 ft on record -- both almost certainly explained by
subsequent deepening of these still-active wells, not a data error.
Rather than overwrite (or awkwardly average) these, the 1990 construction
facts are recorded in `Wells.notes` as explicitly historical, and
`top_perforation`/`bottom_perforation` were left untouched for both.
Every other well's numbers were close enough to (or had no) existing
`total_depth` to fill with confidence.


## Newly ingested chemistry: Table 1 (1950-1991)

`ingest_historical_sorey1992.R` (wired into `run_pipeline.R` as
`RUN_INGEST$historical_sorey1992`) transcribes Table 1's ~10 dated
analyses directly into `Lab_Analyses`, extending this project's own
chemistry record back to 1950 (well GS-58) for the first time.

```{r sorey-chem-table}
sorey_chem <- dbGetQuery(con, "
  SELECT l.name AS feature, s.collection_time AS sample_date,
         la.analyte, la.value, la.units
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE l.external_station_code LIKE 'SOREY1992%'
  ORDER BY s.collection_time, l.name
")
sorey_chem_wide <- sorey_chem %>%
  select(feature, sample_date, analyte, value) %>%
  pivot_wider(names_from = analyte, values_from = value)
sorey_chem_wide
```

### Cl/B and Cl/Li ratio check against modern chemistry

Sorey & Colvard found Cl/B = 19.3 +/- 1.7 and Cl/Li = 122.5 +/- 9.9
across the CPI and SB GEO reservoirs (excluding the more dilute Cox
well), and used this constancy of conservative-element ratios to argue
the two well fields draw on chemically related source water. Recomputing
the same ratios from the newly-ingested 1950-1991 values, and comparing
against this project's own real 2024-2026 thermal-water samples (the
FIELD spring/well/seep samples, and the NDEP-promoted domestic wells),
is a direct test of whether that same geochemical signature has
persisted for over three decades.

```{r ratio-check}
hist_ratios <- sorey_chem %>%
  filter(analyte %in% c("Cl", "B", "Li")) %>%
  pivot_wider(id_cols = c(feature, sample_date), names_from = analyte, values_from = value) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li, era = "1950-1991 (Sorey & Colvard 1992)",
         year = as.integer(substr(sample_date, 1, 4)))

modern_ratios <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte IN ('Cl','B','Li')
    AND s.data_source != 'Sorey & Colvard 1992'
") %>%
  pivot_wider(id_cols = c(sample_id, feature, site_type, collection_time), names_from = analyte,
              values_from = value, values_fn = mean) %>%
  filter(!is.na(Cl), !is.na(B)) %>%
  mutate(Cl_B = Cl / B, Cl_Li = Cl / Li,
         era = if_else(site_type %in% c("fumarole", "seep", "spring"),
                        "2024-2026 real thermal (FIELD)",
                        "2024-2026 NDEP domestic/background wells"),
         year = lubridate::year(lubridate::mdy_hm(collection_time, quiet = TRUE)))  # real parsed sample year (2024 or 2026)

ratio_compare <- bind_rows(
  hist_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li),
  modern_ratios %>% select(feature, era, year, Cl, B, Li, Cl_B, Cl_Li)
)
ratio_compare
```

```{r ratio-plot}
#| fig-width: 8
#| fig-height: 5
# Single unified plot (year on the x-axis, not era-on-y as in an earlier
# draft) -- well-name labels via ggrepel, and the t-test statistics moved
# into the figure caption rather than left as a separate prose paragraph,
# per explicit feedback this session. Excludes the Cox well (per Sorey &
# Colvard's own stated convention) and the SBRR/SBBV creek stations
# (mixing points, not end-member samples -- including them the first time
# this was tried pulled in SBRR's near-zero-B ratio and badly distorted
# the comparison).
cl_b_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "SB GEO production well PW-1" = "PW-1", "SB GEO production well PW-2" = "PW-2",
  "SB GEO production well PW-3" = "PW-3", "Hot spring 6, main terrace ACEC" = "Spring 6",
  "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001", "SBW_0002" = "SBW_0002",
  "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003"
)
ratio_plot_data <- ratio_compare %>%
  filter(!is.na(Cl_B), !grepl("Cox", feature), !feature %in% c("SBRR", "SBBV")) %>%
  mutate(
    era_short = if_else(grepl("^19", era) | era == "1950-1991", "1990-91", "2024-26"),
    label = coalesce(cl_b_labels[feature], feature)
  )
t_res_ratio <- t.test(Cl_B ~ era_short, data = ratio_plot_data)
caption_ratio <- sprintf(
  "Welch two-sample t-test: mean Cl/B = %.1f (1990-91, n=%d) vs. %.1f (2024-26, n=%d); t(%.1f) = %.2f, p = %.3f, 95%% CI [%.2f, %.2f]",
  t_res_ratio$estimate[1], sum(ratio_plot_data$era_short == "1990-91"),
  t_res_ratio$estimate[2], sum(ratio_plot_data$era_short == "2024-26"),
  t_res_ratio$parameter, t_res_ratio$statistic, t_res_ratio$p.value,
  t_res_ratio$conf.int[1], t_res_ratio$conf.int[2]
)
ggplot(ratio_plot_data, aes(x = year, y = Cl_B, color = era_short)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = label), size = 3, max.overlaps = 20, seed = 4172, show.legend = FALSE) +
  labs(x = "Year", y = "Cl / B (mass ratio)", color = "Era",
       title = "Chloride-to-boron ratio in real thermal samples, 1950-2026",
       caption = caption_ratio) +
  theme(plot.caption = element_text(hjust = 0, size = 9))
```

**Reading this (revised 2026-09-12 with a real statistical test -- see
the new Statistical Analysis section below):** this project's own 2024-2026 real thermal FIELD
samples (fumarole/seep/spring) give Cl/B roughly 20-23, in the same
general range as Sorey & Colvard's 1990-91 CPI/SB GEO value of 19.3 +/-
1.7. A Welch two-sample t-test on the 7 historical vs. 6 modern real
thermal Cl/B values (excluding the Cox well, per Sorey & Colvard's own
convention, and excluding the SBRR/SBBV creek stations, which are
mixing points, not end-member samples) finds this difference is
**small but statistically significant**: mean 19.6 (1950-1991) vs. 21.9
(2024-2026), Welch t(9.87) = -2.92, p = 0.015, 95% CI on the difference
[-4.03, -0.54]. This is a more precise, and slightly different,
statement than this notebook's original (Session 27) "closely
matching... essentially unchanged" framing -- the conservative-element
signature is genuinely similar in magnitude across 35 years, but the
two eras are not statistically indistinguishable. A plausible, not yet
tested, explanation is a systematic difference in analytical method
(UURI/AMTEC ICP for the 1990-91 samples vs. this project's own modern
lab) rather than a real 12% geochemical shift -- flagged as an open
question, not resolved here. The NDEP domestic/background well samples
(Boyd, Jeppson, Rogers) show much more scattered Cl/B ratios; their B
values sit at or near 0.05-0.18 mg/L, close to a plausible analytical
detection limit, so those ratios should be read as noisy near-DL
artifacts of dilute, non-thermal ground water rather than a real
geochemical signal -- fully consistent with this project's established
finding that these are background, not thermal, waters.

## Skalbeck (2001) Table B-2: a real, densified pre-2000 monitoring record

Skalbeck (2001) (see `docs/literature/annotated_bibliography.qmd` for the
full annotation) contributes a second, independent historical baseline
alongside Sorey & Colvard (1992) above -- its Table B-2 is a genuine
**monthly** Cl/B/temperature(/depth-to-water) monitoring record, not the
sparser dated snapshots Table 1 (1950-1991) provides. As of this
session, real monthly series have been transcribed and ingested for 8
wells: Brown School, Curti Geothermal, Curti Domestic, Herz Geothermal,
Herz Domestic, Peigh Domestic, Pine Tree Ranch #1, **and, added this
session, Flame and Steinhardt** -- the two wells `ROADMAP.md` had
previously flagged as unresolvable from OCR text alone (their real
Cl/temperature ranges overlap enough that an ordered, range-based
parser produced inconsistent results); both were instead transcribed
directly from a clean user-supplied image of the source table.

```{r skalbeck-summary-07}
skalbeck_summary <- dbGetQuery(con, "
  SELECT l.name AS well_or_spring,
         COUNT(DISTINCT la.sample_id) AS n_samples,
         MIN(se.date) AS earliest, MAX(se.date) AS latest
  FROM Lab_Analyses la
  JOIN Samples s ON la.sample_id = s.sample_id
  JOIN Sampling_Events se ON s.event_id = se.event_id
  JOIN Locations l ON s.location_id = l.location_id
  WHERE la.source_id IN (
    SELECT source_id FROM Data_Sources WHERE name LIKE '%Skalbeck%'
  )
  GROUP BY l.name
  ORDER BY earliest
")
tbl_out <- knitr::kable(skalbeck_summary, caption = "Real Skalbeck (2001)-sourced chemistry, by well/spring")
if (knitr::is_latex_output()) {
  tbl_out <- kableExtra::kable_styling(tbl_out,
    latex_options = c("scale_down", "hold_position"))
}
tbl_out
```

Two things make this real addition more useful than "more chemistry
points": (1) it extends the **facies-clustering and Cl-trend-over-time**
analysis below (Section "Statistical Analysis") with real,
densely-repeated pre-2000 observations for wells that otherwise have
only 1-2 dated samples on file, meaningfully improving the within-well
baseline for any future mixed-effects trend model; and (2) Skalbeck
(2001)'s own Table A-2 alluvium/bedrock-depth point grid -- a separate,
already-ingested dataset from the same dissertation -- feeds
`Geophysical_Depth_Model_Points` and, through it, the real horizon-point
layer `export_leapfrog_geophysical_horizons()` writes for Leapfrog (see
the "Planned: Leapfrog 3D geologic model export" section of
`README.md`, and the new Dhakal-permeability-zonation discussion added
there this session). Flame and Steinhardt's own real depth-to-water
readings from Table B-2 are stored in `Water_Level_Observations` with
`method='skalbeck2001_monitoring'` and are candidate future inputs to
the barometric-efficiency workflow above, once/if their date ranges are
found to overlap a pressure record -- not yet checked in this session.


## Skalbeck (2001), continued: Table A-2 completed, Table 3 well control, TH-1/2/3 water levels, and a shared formation-unit vocabulary

This session substantially extended the Skalbeck (2001) work above:

- **Table A-2 (the alluvium/Tv/AltKgdpKm/Kgd depth-point grid)
  re-checked against the raw OCR text directly**: it has 353 total
  UTM-coordinate rows across all 8 page-repeats, not 240. Of the
  remaining 113, exactly 1 was a real, safely-recoverable OCR
  artifact (a stray apostrophe, `"'124"` instead of `"124"`,
  confirmed by exact arithmetic: 19+124+0=143 matches the row's own
  Kgd depth exactly) -- now added, for 241 total points. The other
  112 have genuine column-position ambiguity that could NOT be
  safely resolved: testing one plausible column assignment produced
  physically impossible results (an Alt-Kgd-pKm thickness exceeding
  its own row's Kgd depth), and the already-ingested data's own
  documented caveat (Kgd depth does not always equal the sum of the
  three thicknesses) means arithmetic cross-checking alone cannot
  disambiguate the rest either way. These 112 rows are logged, not
  guessed, in `data/derived/skalbeck2001_table_a2_ambiguous_rows.csv`
  for a future pass against the actual scanned page images.
- **Table 3 ("Well data used in 2.75-D forward models and 3-D model
  depth to bedrock", p.74-75)** -- 41 real named well-control
  points, each with a real depth-to-Tv/Kgd/pKm/total-depth log --
  ingested for the first time. 10 resolve to already-known wells
  (`Cox I-1`, `GS-5`, `IW-2`, `IW-3`, `PW-1`, `21-5R`, `28-32`,
  `Steinhardt Geothermal Well`, `Brown School Geothermal Well`,
  `Herz Domestic Well`); 31 register as new, coordinate-less
  provisional wells (Table 3 has no coordinates of its own, only the
  raster Figure 5 map). One well name, `32-5`, is flagged rather than
  merged: its profile (20270) groups it with wells immediately
  adjacent to the real `23-5`, making a `2`/`3` digit-transposition
  OCR error plausible -- but its surface geology (`pKm`, i.e. bare
  metasediment at ground surface) doesn't match what would be
  expected for that well, so it stays a distinct, unconfirmed entry.
- **TH-1/TH-2/TH-3's real monthly water-depth series (p.210-213,
  Jan-1985 through 1998, 160 real readings)** -- these are SBG-
  TH1/2/3 (cross-confirmed via Table 1's own 1991-completion
  entries), and this table section has no chemistry at all, water
  depth only. A real, unresolved discrepancy: the series' own
  earliest real readings (1989) predate that 1991 completion date by
  two years -- flagged, not resolved, since the column-position
  parsing itself was independently verified self-consistent (a
  smooth, continuous, physically plausible record with no
  discontinuity at 1991).
- **A shared formation-unit vocabulary.** `Well_Lithology` gained a
  `formation_unit` column (Qal/Tv/AltKgdpKm/Kgd/unknown), alongside
  its original free-text `description` (never overwritten). Table
  3's own well-control depths map directly onto this scheme (no
  classification needed -- the source table IS the unit); the
  existing free-text lithology (83C-6ST1's Akerley et al. 2021
  interval, and the still-candidate log 61248 / 23-5 interval) was
  run through a documented keyword classifier
  (`classify_formation_unit()`), which correctly resolved
  "basaltic andesite"/"metamorphosed volcanic rock" to Tv and
  "metamorphosed sedimentary rock"/"schist, conglomerate, and
  quartzite" to AltKgdpKm, while leaving genuinely ambiguous text
  (the swelling-clay interval, and several generic driller terms
  like "hard fractured rock" that don't specify a real geologic
  unit) as `unknown` rather than guessing.

```{r skalbeck-table3-summary}
table3_summary <- dbGetQuery(con, "
  SELECT w.well_name, w.well_role, w.elevation_m,
         COUNT(wl.lithology_id) AS n_intervals,
         GROUP_CONCAT(DISTINCT wl.formation_unit) AS units_present
  FROM Wells w
  JOIN Well_Lithology wl ON wl.well_id = w.well_id
  WHERE wl.formation_unit_basis = 'source_table_column'
  GROUP BY w.well_id
  ORDER BY w.well_name
")
tbl_t3 <- knitr::kable(table3_summary, caption = "Skalbeck (2001) Table 3 well-control points, real formation-unit intervals")
if (knitr::is_latex_output()) {
  tbl_t3 <- kableExtra::kable_styling(tbl_t3, latex_options = c("scale_down", "hold_position"))
}
tbl_t3
```

### A mini-well-log view: Table 3 wells vs. Table A-2 grid points, same units

```{r skalbeck-minilog-plot}
#| fig-width: 10
#| fig-height: 6
minilog_wells <- dbGetQuery(con, "
  SELECT w.well_name AS HoleID, wl.depth_from_ft AS From_m, wl.depth_to_ft AS To_m, wl.formation_unit
  FROM Well_Lithology wl JOIN Wells w ON w.well_id = wl.well_id
  WHERE wl.formation_unit_basis = 'source_table_column' AND wl.units = 'm'
")
minilog_a2 <- dbGetQuery(con, "
  SELECT point_id, qal_thickness_m, tv_thickness_m, alt_kgd_km_thickness_m, depth_to_bedrock_m
  FROM Geophysical_Depth_Model_Points
  WHERE qal_thickness_m IS NOT NULL
  ORDER BY point_id LIMIT 15
")
# Kgd drawn as the full bedrock column (base of Tv -> a per-point
# visual floor, real_bedrock_depth * 1.3 -- a plotting choice, not a
# measured thickness -- Kgd is genuinely open-ended at depth here),
# with AltKgdpKm cut out of its shallow part afterward. Matches
# scripts/analysis/skalbeck_cross_sections.R's corrected model:
# AltKgdpKm is the named altered CAP of the Kgd body, not a separate
# pre-Kgd layer -- it is cut out of Kgd, not out of Tv.
a2_intervals <- purrr::pmap_dfr(minilog_a2, function(point_id, qal_thickness_m, tv_thickness_m,
                                                       alt_kgd_km_thickness_m, depth_to_bedrock_m) {
  cursor <- 0
  rows <- list()
  if (!is.na(qal_thickness_m) && qal_thickness_m > 0) {
    rows[[length(rows)+1]] <- data.frame(HoleID = paste0('A2PT_', point_id), From_m = cursor,
                                          To_m = cursor + qal_thickness_m, formation_unit = 'Qal')
    cursor <- cursor + qal_thickness_m
  }
  if (!is.na(tv_thickness_m) && tv_thickness_m > 0) {
    rows[[length(rows)+1]] <- data.frame(HoleID = paste0('A2PT_', point_id), From_m = cursor,
                                          To_m = cursor + tv_thickness_m, formation_unit = 'Tv')
    cursor <- cursor + tv_thickness_m
  }
  bedrock_top <- cursor
  bedrock_floor <- if (!is.na(depth_to_bedrock_m)) max(depth_to_bedrock_m, bedrock_top + 1) * 1.3 else NA
  if (!is.na(bedrock_floor)) {
    rows[[length(rows)+1]] <- data.frame(HoleID = paste0('A2PT_', point_id), From_m = bedrock_top,
                                          To_m = bedrock_floor, formation_unit = 'Kgd')
  }
  if (!is.na(alt_kgd_km_thickness_m) && alt_kgd_km_thickness_m > 0) {
    rows[[length(rows)+1]] <- data.frame(HoleID = paste0('A2PT_', point_id), From_m = bedrock_top,
                                          To_m = bedrock_top + alt_kgd_km_thickness_m, formation_unit = 'AltKgdpKm')
  }
  dplyr::bind_rows(rows)
})
combined_minilog <- dplyr::bind_rows(
  dplyr::mutate(minilog_wells, source = 'Table 3 well control'),
  dplyr::mutate(a2_intervals, source = 'Table A-2 grid point')
)
unit_colors <- c(Qal = "#e6ab02", Tv = "#7570b3", AltKgdpKm = "#1b9e77", Kgd = "#f781bf")
ggplot(combined_minilog) +
  geom_rect(aes(xmin = as.numeric(factor(HoleID)) - 0.4, xmax = as.numeric(factor(HoleID)) + 0.4,
                ymin = From_m, ymax = To_m, fill = formation_unit)) +
  scale_x_continuous(breaks = seq_along(unique(combined_minilog$HoleID)),
                      labels = levels(factor(combined_minilog$HoleID))) +
  scale_y_reverse() +
  facet_grid(. ~ source, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = unit_colors, name = "Formation unit",
                     breaks = c("Qal", "Tv", "AltKgdpKm", "Kgd")) +
  labs(x = NULL, y = "Depth (m)",
       title = 'Real formation-unit "mini well logs": Table 3 wells vs. Table A-2 grid points',
       subtitle = "Kgd (pink) is the full bedrock column; AltKgdpKm (green) is cut out of its shallow, altered part") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
        legend.position = "bottom")
```

This is a direct visual check that the shared vocabulary actually
lines up: real well-log-derived intervals (left facet) and the
model-grid "mini well logs" (right facet, first 15 of 241 real
points, for readability) use the identical color scheme. **On the
AltKgdpKm/Kgd relationship, corrected 2026-09-26**: AltKgdpKm is the
named altered CAP overlying/within the Kgd bedrock body, not a
separate unit stacked between Tv and a disconnected deeper Kgd --
the fix is to draw Kgd first as the full bedrock column starting at
the base of Tv, then cut AltKgdpKm out of its shallow part, matching
how `scripts/analysis/skalbeck_cross_sections.R`'s cross-sections now
render the same real data. Table 3's own well-control intervals were
separately corrected in the same pass: its "Kgd" column is the
driller's own contact name (not an alteration claim), so those
intervals are now labeled Kgd, not AltKgdpKm, in `Well_Lithology`.

### From scattered points to real layers: a gridded Steamboat surface for Leapfrog

Table A-2's 241 confidently-parsed points are real but irregularly
spaced (they follow Skalbeck's own aeromagnetic/gravity profile
lines, not a uniform survey grid). `export_leapfrog_geophysical_grid_surfaces()`
interpolates them onto a real, regular 100 m grid (matching Dhakal
et al. (2025)'s own numerical-model cell size) via inverse-distance
weighting (`gstat::idw()`) -- turning the point cloud into three real
layered surfaces (base of Qal, base of Tv, top of fresh Kgd) that
Leapfrog (or ArcGIS) can import directly as GM surfaces, rather than
leaving every downstream consumer to interpolate the raw points
themselves. This is a real geostatistical interpolation (IDW, not
kriging -- no variogram has been fitted for these specific surfaces
yet), and grid cells far from any real point are a genuine
extrapolation, tracked via each cell's own `n_control_points` count.

```{r skalbeck-grid-plot}
#| fig-width: 9
#| fig-height: 4
grid_script <- file.path("scripts", "leapfrog", "export_leapfrog.R")
if (!file.exists(grid_script)) grid_script <- file.path("..", "scripts", "leapfrog", "export_leapfrog.R")
source(grid_script)
grid_tmp_dir <- tempfile()
invisible(export_leapfrog_geophysical_grid_surfaces(con, out_dir = grid_tmp_dir))
grid_surfaces <- readr::read_csv(file.path(grid_tmp_dir, "geophysical_grid_surfaces.csv"), show_col_types = FALSE)
ggplot(grid_surfaces, aes(x = UTM_E, y = UTM_N, fill = Depth_m)) +
  geom_raster() +
  facet_wrap(~ Layer, ncol = 3) +
  scale_fill_viridis_c(direction = -1, name = "Depth (m)") +
  coord_fixed() +
  labs(title = "Real IDW-interpolated depth surfaces, 100 m grid",
       subtitle = "From the 241 confidently-parsed Table A-2 points -- three real layers, not a single flat model") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### A quasi-3D view of the same three layers

A real, interactive 3D scatter (rotate/zoom in the HTML version of
this notebook) of the same grid cells, colored by layer and
positioned at their real UTM X/Y and (negative) depth Z -- the
closest this project currently comes to an actual Leapfrog-style 3D
view without opening Leapfrog itself.

```{r skalbeck-3d-plot}
#| results: asis
grid_thin <- grid_surfaces[seq(1, nrow(grid_surfaces), by = 8), ]
if (knitr::is_latex_output()) {
  # A real interactive 3D scatter only renders in HTML output -- PDF/LaTeX
  # gets a static 2D projection (Northing vs. depth) as a fallback so the
  # same real data is still shown, just not rotatably.
  print(ggplot(grid_thin, aes(x = UTM_N, y = -Depth_m, color = Layer)) +
    geom_point(size = 0.6, alpha = 0.5) +
    scale_color_manual(values = c("#e6ab02", "#7570b3", "#d95f02")) +
    labs(title = "Static fallback for PDF: Northing vs. depth (interactive 3D view is in the HTML notebook)",
         x = "UTM Northing (m)", y = "Depth below surface (m, negative down)"))
} else {
  plotly::plot_ly(grid_thin, x = ~UTM_E, y = ~UTM_N, z = ~ -Depth_m, color = ~Layer,
                   colors = c("#e6ab02", "#7570b3", "#d95f02"),
                   type = "scatter3d", mode = "markers", marker = list(size = 2)) %>%
    plotly::layout(scene = list(
      xaxis = list(title = "UTM Easting (m)"),
      yaxis = list(title = "UTM Northing (m)"),
      zaxis = list(title = "Depth below surface (m, negative down)")
    ))
}
```

**On hydraulic conductivity**: no literature source currently on
file (Skalbeck 2001, Dhakal et al. 2025, Collar & Huntley 1990, Sorey
& Colvard 1992) gives a real, numeric hydraulic-conductivity or
permeability value *specific to* the Qal/Tv/AltKgdpKm/Kgd units
themselves. The only real numeric permeability value on file anywhere
is Dhakal et al. (2025)'s single 0.001 mD top-boundary condition
(not a property of any of these four units); the only real
aquifer-scale hydraulic values on file are Sorey & Colvard (1992)'s
Table 9 transmissivity/storage-coefficient results and Collar &
Huntley (1990)'s own transmissivity point (both already in this
notebook's earlier sections) -- basin-scale pumping-test results, not
per-formation-unit properties. This is a real, honestly-reported
gap, not a placeholder value.


### Real logger calibration checks and measurement uncertainty

Per your request to track real field calibration/reference checks
during data pulls: `build_logger_calibration_checks()` scans
`Field_Measurements` for any temperature/conductivity reading taken
at a site that also hosts one of this project's own loggers, and
matches it to the nearest real logger observation (within 4 hours).
This needs no new manual data-entry form -- it is derived entirely
from data this project already collects.

```{r logger-calibration-checks}
calchecks <- dbGetQuery(con, "
  SELECT logger_type, logger_id, check_timestamp, reference_instrument,
         round(reference_value, 2) AS reference_value,
         round(matched_observation_value, 2) AS logger_value,
         round(time_offset_seconds / 60, 1) AS offset_min,
         round(difference, 2) AS difference
  FROM Logger_Calibration_Checks ORDER BY logger_type, check_timestamp
")
cc_caption <- if (knitr::is_latex_output()) {
  "Real field calibration/reference checks, auto-detected from Field\\_Measurements"
} else {
  "Real field calibration/reference checks, auto-detected from Field_Measurements"
}
tbl_cc <- knitr::kable(calchecks, caption = cc_caption)
if (knitr::is_latex_output()) {
  tbl_cc <- kableExtra::kable_styling(tbl_cc, latex_options = c("scale_down", "hold_position"), font_size = 8)
}
tbl_cc
```

10 real temperature checks exist so far (all using a FLUKE reference
thermometer during field visits), with differences ranging from
near-zero up to ~14 C at a couple of sites -- worth a field follow-up
on those specific loggers, not just reported as an abstract number.
Only 1 real conductivity check candidate exists project-wide, and it
falls outside the 4-hour matching tolerance (see `ROADMAP.md`).
Combining each logger's own manufacturer-published accuracy
(`Logger_Specifications`, real HOBO/Onset U24-002-C and Elitech
LogEt 8 PTE/Tlog 10E datasheet values, not fabricated) with these
real field checks via root-sum-square
(`get_combined_uncertainty()`/`summarize_logger_uncertainty()`) gives
a combined uncertainty per logger that is *larger* than the
manufacturer spec alone wherever a real field check shows drift --
exactly the behavior intended.

```{r logger-uncertainty-plot}
#| fig-width: 8
#| fig-height: 4.5
unc_script <- file.path("scripts", "analysis", "measurement_uncertainty.R")
if (!file.exists(unc_script)) unc_script <- file.path("..", "scripts", "analysis", "measurement_uncertainty.R")
source(unc_script)
unc_summary <- summarize_logger_uncertainty(con)
if (nrow(unc_summary) > 0) {
  ggplot(unc_summary, aes(x = factor(logger_id), y = combined_uncertainty)) +
    geom_col(aes(fill = n_checks > 0)) +
    geom_point(aes(y = manufacturer_uncertainty), shape = 4, size = 3, color = 'black') +
    scale_fill_manual(values = c(`TRUE` = '#d95f02', `FALSE` = 'grey70'), guide = 'none') +
    labs(x = 'Logger ID', y = 'Uncertainty (deg C)',
         title = 'Combined logger temperature uncertainty: manufacturer spec (X) vs. real field-check-informed value (bar)',
         subtitle = 'Every logger shown has at least one real calibration check')
} else {
  message('No loggers with real calibration checks yet.')
}
```


## Discharge-through-time: extending Sorey's Table 2

Table 2 of the 1992 report tabulates dated estimates of thermal-water
discharge to Steamboat Creek from 1955 through 1989. These values are
**transcribed literature numbers, not live database queries** -- they
predate any table this project's own pipeline builds -- assembled here
alongside the already-cited Sorey & Spielman (2008, 2017) follow-up
estimates and this project's own 2026 poster result to give a single,
~70-year view.

```{r discharge-timeline}
# Extended 2026-09-25: added a new database-computed point (this project's
# only real paired Cl+discharge measurement, 2026-05-01) and a second
# Collar & Huntley (1990) point (their own stated "about 80% of White's
# 1955 baseline" 1988-89 natural-only estimate, distinct from the two
# dated Table 2 values already used).
discharge_hist <- tribble(
  ~period, ~year_mid, ~gpm, ~Ls, ~source,

  "White (1968), Apr 1955", 1955, 1110, 1110 * 0.0630902,
  "White (1968), chloride-flux method",

  "White (1968), Apr 1964", 1964, 1385, 1385 * 0.0630902,
  "White (1968), chloride-flux method",

  "Shump (1985), 1981-82 avg", 1981.5, 1300, 1300 * 0.0630902,
  "Shump (1985), specific-conductance method",

  "Collar (1990), Jun 1988", 1988.5, 523, 523 * 0.0630902,
  "Collar (1990) via Sorey & Colvard (1992) Table 2",

  "Collar (1990), Aug 1988", 1988.6, NA, NA,
  "Not counting net CPI production; unseen seepage component only",

  "Collar (1990), Mar 1989", 1989.2, 663, 663 * 0.0630902,
  "Collar (1990) via Sorey & Colvard (1992) Table 2",

  "Collar & Huntley (1990), 1988-89 natural-only", 1988.8, NA,
  0.80 * 1120 * 0.0630902,
  "Collar & Huntley (1990): ~80% of White's 1955 estimate (natural only)",

  "Sorey & Spielman (2008), 2007-08", 2007.5, NA,
  mean(c(730, 1214)) * 0.0630902,
  "Chloride-flux, SBRR/SBBV sites",

  "Sorey & Spielman (2017), 2008-2016 avg", 2012, NA, 60,
  "Chloride-flux, SBRR/SBBV sites, 24 sampling events",

  "This project (database), 5/1/2026", 2026.33, NA, 42.1,
  "Real paired Cl (SBRR=17.1, SBBV=126 mg/L, 2026-05-01) + USGS Q",

  "Irvin & Lindsey (2026 poster), post-eruption avg", 2026, NA, 27,
  "Own Cl mass-balance estimate, averaged over the sampling period"
)
# Full source-attribution text (trimmed above for print width) is in
# this notebook's own prose immediately below the table -- see
# "Reading this:" further down, not just the `source` column.
tbl_disch <- knitr::kable(discharge_hist, digits = 1, booktabs = TRUE,
  caption = "Thermal-water discharge to Steamboat Creek, 1955-2026")
if (knitr::is_latex_output()) {
  # scale_down doesn't apply to a longtable (kableExtra warns and
  # skips it) -- constraining the wide 'source' column's width directly
  # (with word-wrap inside the cell) is the fix that actually holds.
  tbl_disch <- kableExtra::column_spec(tbl_disch, 5, width = "2in")
  tbl_disch <- kableExtra::kable_styling(tbl_disch,
    latex_options = c("hold_position"), font_size = 9)
}
tbl_disch
```

```{r discharge-plot}
#| fig-width: 9
#| fig-height: 5.5
ggplot(discharge_hist %>% filter(!is.na(Ls)),
       aes(x = year_mid, y = Ls)) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = grepl("This project", period)),
             size = 3, show.legend = FALSE) +
  ggrepel::geom_text_repel(aes(label = period), size = 2.6,
                           max.overlaps = 30, seed = 4172) +
  scale_color_manual(values = c(`TRUE` = "#d95f02", `FALSE` = "black")) +
  labs(
    x = "Year", y = "Thermal-water discharge (L/s)",
    title = "Thermal-water discharge to Steamboat Creek, 1955-2026",
    subtitle = paste(
      "No confidence intervals shown -- no source reports a measurement",
      "uncertainty. Orange point is a new, reproducible single-day",
      "estimate computed directly from this project's own database."
    )
  )
```

**Reading this:** discharge estimates were broadly 45-90 L/s
(660-1,385 gpm-equivalent) from 1955 through the 2008-2016 Sorey &
Spielman baseline, with the 1988-89 Collar (1990) values sitting at the
low end even before accounting for that they explicitly exclude net CPI
well-field production. This project's own post-eruption ~27 L/s estimate
sits below every prior estimate in this 70-year record -- consistent
with the poster's own conclusion (conduit reactivation/redistribution,
not necessarily a new source) but worth treating as one point in a
long, variable record rather than a single before/after comparison.
None of the pre-2007 methods used the same sites (SBRR/SBBV) as the
modern chloride-flux estimates, so method consistency, not just
magnitude, matters when reading this chart -- White (1968) and Collar
(1990)'s methods and gauge locations are not identical to Sorey &
Spielman's.

**Seasonal Cl-flux variability -- a real, previously-published effect
worth naming here (2026-09-25).** Sorey & Spielman (2017) -- already
cited above, not a new document -- explicitly document a real seasonal
pattern in their own SBRR/SBBV Cl-flux computations: *"lower computed
chemical flux at downstream sites in SB Creek each summer and higher
chemical loads in creek water in the wintertime runoff period,"*
attributed in part to seasonal buildup and flushing of chloride in
shallow soils from irrigation/evaporation cycles downstream of Crane
Ditch diversions -- though they state *"the magnitude of such seasonal
changes appears to be only on the order of 25% or less."* This is a
plausible, real contributor to the gap between this session's new
single-day database estimate (42.1 L/s, 2026-05-01, spring runoff
season) and the poster's own averaged ~27 L/s figure -- a ~56% relative
difference, larger than Sorey & Spielman's own documented ~25% seasonal
envelope, so seasonality alone does not fully explain it, but it is a
real, published, and directly relevant effect rather than pure noise.
This also matches this project's own poster, which discusses Cl outflow
variability in the diluted creek -- worth a dedicated seasonal-Cl-flux
analysis once more paired Cl+discharge dates exist in the database (a
single day, as computed here, cannot separate a real seasonal signal
from ordinary day-to-day variability).

**Sorey (2000)'s reawakening baseline -- a critical reference, cited
directly, not just added to the bibliography (2026-09-26).** Sorey
(2000, World Geothermal Congress) is the direct, real historical
follow-up to Sorey & Colvard (1992) that this discharge-through-time
chart itself does not otherwise carry: by 2000, precipitation had
returned to normal-or-above levels and South Truckee Meadows groundwater
levels had risen back to nearly their late-1980s levels, **but Sorey
(2000) reports no evidence of any renewed natural spring flow on the
main terrace despite that recovery.** This is the direct historical
baseline against which the 2025 Lower Sinter Terrace eruption should be
read: if the pre-2025 hydrologic system had genuinely not "reawakened"
even once one of its own documented drivers (regional groundwater
decline) reversed, then the 2025 event raises a real, specific question
this thesis's own spring-remapping fieldwork can help answer -- is 2025
a delayed reawakening of the *same* system Sorey tracked, or evidence of
a mechanistically different process (e.g. a discrete conduit
reactivation, consistent with this project's own 2026 poster
conclusion, rather than a broad system-wide recovery)? Framed this way,
the discharge-through-time chart above and the 27-42 L/s post-2025
estimates are one data point in that specific, testable question, not
just a continuation of the same 70-year record.

## Appendix G and Table 9: what's behind the discharge/reservoir numbers (2026-09-12)

### Appendix G: the raw synoptic Cl-flux surveys behind Table 2

Appendix G ("Stream Discharge and Chloride Flux") documents the real
method behind the Collar (1990) discharge estimates already used in the
discharge-through-time chart above: chloride treated as a conservative
tracer, an 820 mg/L thermal / 6 mg/L non-thermal end-member split, and
gaining/losing-reach mass-balance equations -- methodologically almost
identical to this project's own SBRR/SBBV Cl mass-balance approach, just
predating it by ~35 years. Chloride was measured by mercuric-nitrate or
argentometric titration (SDSU and UURI, respectively); discharge by
pygmy current meter. The underlying raw data (Table G-1) comes from four
real synoptic surveys: **6/26-6/30/1988, 7/1-7/2/1988, 8/9/1988, and
3/4/1989**, at ~40 named stations along Steamboat Creek, Galena Creek,
and the Steamboat/Chandler/Crane irrigation ditches.

**Deliberately not parsed further**: Table G-1's raw per-station
discharge/chloride/boron values are visibly OCR-misaligned in the
extracted text (values shifted relative to station-name rows, in some
blocks by a full row) -- the same class of "real data trapped in a
badly-OCR'd table" problem already flagged elsewhere in this project
(the NDEP TFT Appendix D geochemistry tables, well-log lithology
tables). The three aggregate numbers we actually use (Collar 1990's
1988-89 discharge estimates in the Table 2 chart above) are already the
synthesized output of this same appendix -- Appendix G is the raw
survey behind data already in this notebook, not a source of new,
different data points. If a denser historical Cl-flux time series is
ever wanted, it would need to be read directly from the scanned page
images (not the OCR text) station-by-station -- a real, larger effort
than anything else in this session, not attempted here.

### Table 9: reservoir parameters (transmissivity, storage coefficient)

Table 9 ("Reservoir parameters determined for the geothermal system in
the Steamboat Hills area from well test analyses and spring
hydrographs") gives real aquifer-test results from four dated tests:

```{r table9-reference}
table9 <- tibble::tribble(
  ~well_and_year, ~injection, ~transmissivity_ft2_day, ~storage_coefficient, ~observation_well, ~source,
  "Steamboat No.1, 1980", "no", "3270-9500", "1.2e-3 (calc.)", "strat 2", "Collar (1990)",
  "Steamboat No.1, Mar-May 1986", "no", "nd", "nd", "23-5?", "Berkeley Group (1987)",
  "Steamboat No.1, Mar-May 1987", "no", "nd", "1.2e-3 (calc.)", "strat 5", "Berkeley Group (1987)",
  "23-5/Cox 1-1, May-June 1987", "yes", "7930-9500", "7.8e-4 to 9.0e-4", "strat 2, 21-5", "Faulder (1987); Goranson (1989)",
  "21-5, 23-5, 83A-6, 1988", "yes", "1250-2400", "2.5e-3", "strat 9", "Collar (1990)",
  "21-5, 23-5, 83A-6, 1988 (spring response)", "yes", "nd", "2.8e-3", "spring 12, spring 42w", "Collar (1990)",
  "Spring 42w, Jun-Aug 1988 (Birsoy-Summers method)", "yes", "3000", "2.6e-3", "spring 42w", "Collar & Huntley (1990)"
)
table9
```

No table in the current schema models a dated aquifer/interference
test -- this is kept as reference documentation only, not a new
database table, since there is no immediate analytical use for it beyond
the qualitative reservoir-drawdown narrative already summarized above
("What the 1992 report says"). Worth revisiting if this project's own
hydraulic-gradient work ever needs a real transmissivity/storage
reference range for the Steamboat Hills reservoir.

### 4.6 Collar & Huntley (1990): a fault map and a discharge-deficit statistic (2026-09-12)

A second Sorey-team document, `Collar.pdf` (Collar, R.J. and Huntley, D.
1990, 12th New Zealand Geothermal Workshop), was added to the literature
folder and read directly. It largely overlaps the 1992 report above (same
wells, same hydrograph story), but contributes two things not already
captured:

**A real fault map.** Its Figure 1 ("Fault and air photo lineament map of
the Steamboat Springs geothermal area, including the location of selected
wells") shows mapped faults (dashed = inferred, dotted = concealed) and
air-photo lineaments of probable structural origin, including the named
Pleasant Valley Fault and Steamboat Springs Fault Zone, with `21-5`,
`83A-6`, `23-5`, `Cox I-1`, `PW-1/2/3`, `IW-2/3`, strat wells, `Brown
School`, `Pine Tree Ranch`, `Steinhardt`, and `GS-1`/`GS-8` all located on
it by name. Ten of these already have real coordinates in this database:

```{r collar-fig1-control-points}
dbGetQuery(con, "
  SELECT well_name, latitude, longitude FROM Wells
  WHERE well_name IN ('21-5R','23-5','83A-6','Cox I-1','PW-2','PW-3','IW-2',
                       'Brown School Geothermal Well','Steinhardt Geothermal Well','GS-5')
  ORDER BY well_name
")
```

This is a materially better fault-digitizing candidate than the
Dhakal-diagram source discussed in `README.md`'s "Planned: Leapfrog 3D
geologic model export" section (a production/injection flow diagram, not
a fault map) -- see that section (updated 2026-09-12) for the proposed
digitizing approach using these ten points as control points.

**A discharge-deficit statistic.** The paper states that 1988 spring
discharge, measured at 3-4 gal/min, was far below the 34-40 gal/min
*predicted* from White (1968)'s own spring-discharge/annual-precipitation
relationship -- a specific, citable number distinct from (and more
precise than) the aggregate gpm figures already in the Section 6
discharge-through-time chart.

**Elevation cross-check.** The paper gives well-field elevations (OESI/
SB GEO ~4700 ft; CPI 5350-5660 ft; Cox I-1 5050 ft; springs ~4650 ft),
checked against this project's own `Wells.elevation_m`:

```{r elevation-crosscheck}
dbGetQuery(con, "
  SELECT well_name, elevation_m, ROUND(elevation_m/0.3048,0) AS elevation_ft
  FROM Wells WHERE well_name IN ('PW-2','PW-3','IW-2','21-5R','Cox I-1')
  ORDER BY well_name
")
```

`IW-2`/`PW-2`/`PW-3` (4634-4730 ft) are consistent with the paper's OESI
well-field figure (~4700 ft). `21-5R` (5722 ft) sits just above the
paper's stated CPI range (5350-5660 ft) -- a minor, unresolved
discrepancy, plausibly because the paper gives one representative
figure for the whole field rather than a per-well value. `Cox I-1` had
no recorded elevation at all; the paper's specific value (5050 ft =
1539.24 m) is filled in below since, unlike the CPI range, it is
directly attributable to a single named well.



## Table 10 monitoring network vs. this project's Locations

```{r table10-check}
dbGetQuery(con, "
  SELECT location_id, external_station_code, name, site_type,
         latitude, longitude, coordinate_source
  FROM Locations
  WHERE name LIKE '%Boyd%' OR name LIKE '%Curti%' OR name LIKE '%Rogers%'
     OR name LIKE '%Jeppson%' OR name LIKE '%Herz%' OR name LIKE '%Zolezzi%'
     OR external_station_code IN ('SB5','SB6')
  ORDER BY name
")
```

Sorey & Colvard's Table 10 ("Current monitoring sites for CPI and SBG
geothermal developments") describes exactly this same set of domestic
wells and the Rhodes Road creek-monitoring point as the active 1988-92
monitoring network -- confirming this project's own SBRR site and its
NDEP/Klein-resolved domestic wells sit at genuinely long-monitored
locations, not newly-chosen ones. None of these have a linked `Wells`
row or `Water_Level_Observations` history yet, which is a real,
documented gap (see Section 7 caveats) rather than an oversight to
silently work around here.

**Aside (found and fixed, 2026-09-12): the malformed `"7-07-10"`
timestamp.** `Water_Level_Observations` has 73 wells and 153,437 rows
spanning 1969-06-23 to 2026-01-01; one row (well-log document 104216, a
"WELL DRILLER'S PLUGGING REPORT" with `completion_date_raw = "7/10/07"`)
had timestamp `"7-07-10"`. Root cause: `ingest_well_logs.R`'s
`.safe_completion_date()` tries `as.Date(x, format = "%m/%d/%Y")` first,
and R's `%Y` silently accepts a 2-digit year as a literal, unpadded year
value (year 7 CE) rather than rejecting the format mismatch the way the
function's design assumed it would -- so the very first format "matched"
with a nonsensical date instead of falling through to a better one.
Fixed by adding an explicit `%m/%d/%y` attempt and a plausible-year-range
sanity check (1900 to next calendar year) that every candidate parse must
pass before being accepted; the one bad row is corrected to `2007-07-10`.
No other row in `Well_Log_Documents`/`Water_Level_Observations` was
affected -- this was the only 2-digit-year-formatted completion date in
the batch.

## Statistical Analysis: Is There a Real Chloride Trend? (2026-09-12)

Sections above establish real Cl values across 1950-2026 and a
qualitative sense that thermal-water chemistry looks broadly similar
over time. This section applies actual statistical tests to that claim
-- per an explicit request that this notebook's historical-comparison
work be statistic-driven, not just descriptive.

### A combined 1950-2026 Cl dataset

```{r cl-timeline-data}
library(lubridate)
cl_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, l.site_type, s.collection_time, s.data_source, la.value AS Cl
  FROM Lab_Analyses la
  JOIN Samples s ON s.sample_id = la.sample_id
  JOIN Locations l ON l.location_id = s.location_id
  WHERE la.analyte = 'Cl'
")

# Sampling_Events/Samples.collection_time mixes Unix-epoch-seconds,
# MM/DD/YYYY H:MM, and ISO YYYY-MM-DD formats across sources -- the
# same class of dual/mixed-format issue already flagged project-wide
# for Sampling_Events.date (notebook 05). Parsed here defensively,
# trying each format in turn rather than assuming one.
.parse_any_date <- function(x) {
  num <- suppressWarnings(as.numeric(x))
  out <- as.Date(rep(NA, length(x)))
  is_num <- !is.na(num) & !grepl('[A-Za-z]', x)
  out[is_num] <- as.Date(as.POSIXct(num[is_num], origin = '1970-01-01', tz = 'UTC'))
  rem <- !is_num
  out[rem] <- suppressWarnings(as.Date(x[rem], format = '%m/%d/%Y %H:%M'))
  rem2 <- rem & is.na(out)
  out[rem2] <- suppressWarnings(as.Date(x[rem2], format = '%Y-%m-%d'))
  out
}
cl_data$date <- .parse_any_date(cl_data$collection_time)
cl_data$year <- year(cl_data$date)

# Added 2026-09-25: Steinhardt's two narrative-based bracket samples
# (1987/1990, see Section 4.2 correction above), now that they exist.
focus_wells <- c(
  'CPI production well 21-5','CPI production well 83A-6','CPI production well 23-5',
  'CPI injection well Cox 1-1 (Cox well)','SB GEO production well PW-1','SB GEO production well PW-2',
  'SB GEO production well PW-3','Well GS-5, 1950 sample (White 1968)','New seep adjacent to well GS-5 (1991)',
  'Hot spring 6, main terrace ACEC',
  'SBF_0001','SBW_0001','SBW_0002','SBS_0001','SBS_0002','SBS_0003',
  'Eich Well','Soccer Field Monitoring Well','Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well',
  'Steinhardt Geothermal Well (narrative Cl trend)'
)
background_names <- c('Boyd Domestic Well','Jeppson Geothermal Well','Rogers Domestic Well')
cl_focus <- cl_data[cl_data$feature %in% focus_wells, ]
cl_focus$group2 <- ifelse(cl_focus$feature %in% background_names, 'background/domestic',
                    ifelse(cl_focus$feature %in% c('Eich Well','Soccer Field Monitoring Well'), 'intermediate (NDEP)', 'thermal'))

well_labels <- c(
  "CPI production well 21-5" = "21-5", "CPI production well 83A-6" = "83A-6",
  "CPI injection well Cox 1-1 (Cox well)" = "Cox I-1", "SB GEO production well PW-1" = "PW-1",
  "SB GEO production well PW-2" = "PW-2", "SB GEO production well PW-3" = "PW-3",
  "Well GS-5, 1950 sample (White 1968)" = "GS-5", "New seep adjacent to well GS-5 (1991)" = "GS-5 seep",
  "Hot spring 6, main terrace ACEC" = "Spring 6", "SBF_0001" = "SBF_0001", "SBW_0001" = "SBW_0001",
  "SBW_0002" = "SBW_0002", "SBS_0001" = "SBS_0001", "SBS_0002" = "SBS_0002", "SBS_0003" = "SBS_0003",
  "Eich Well" = "Eich", "Soccer Field Monitoring Well" = "Soccer Field", "Boyd Domestic Well" = "Boyd",
  "Jeppson Geothermal Well" = "Jeppson", "Rogers Domestic Well" = "Rogers",
  "Steinhardt Geothermal Well (narrative Cl trend)" = "Steinhardt"
)
cl_focus$label <- well_labels[cl_focus$feature]
nrow(cl_focus)
```

```{r cl-timeline-plot}
#| fig-width: 9
#| fig-height: 5.5
# Reworked 2026-09-25 per explicit feedback: well-name labels (not just
# color groups), lines connecting repeat samples of the same well/spring
# through time, and the 6 single-snapshot 2026 FIELD thermal samples
# (SBF_0001, SBW_0001/2, SBS_0001-3) dropped from THIS plot -- they add
# no temporal information (all one date) and were crowding the 2020s
# cluster; they remain visible in the Cl/B ratio plot above.
exclude_2026_single <- c("SBF_0001","SBW_0001","SBW_0002","SBS_0001","SBS_0002","SBS_0003")
cl_focus_timeline <- cl_focus[!cl_focus$feature %in% exclude_2026_single, ]
# 2026-09-26: faceted by well vs. spring/seep per explicit request --
# almost everything in this curated focus set is a well; GS-5 seep and
# Hot spring 6 are the only spring/seep entries, but keeping them in a
# separate facet panel avoids visually implying they belong to the same
# "population" as the production/domestic wells.
cl_focus_timeline$site_category <- ifelse(cl_focus_timeline$site_type %in% c("spring","seep"),
                                            "Springs & Seeps", "Wells")
ggplot(cl_focus_timeline, aes(x = date, y = Cl, color = group2)) +
  geom_line(aes(group = feature), alpha = 0.4, linewidth = 0.5, show.legend = FALSE) +
  geom_point(size = 2.5) +
  ggrepel::geom_text_repel(aes(label = label), size = 3.2, max.overlaps = 20, seed = 4172) +
  scale_y_log10() +
  facet_wrap(~ site_category, scales = "free_x") +
  labs(x = NULL, y = "Chloride (mg/L, log scale)", color = "Sample group",
       title = "Chloride concentration by well/spring through time, 1950-2026",
       subtitle = "Faceted by well vs. spring/seep; wells with real repeat measurements over time (Steinhardt,\nBoyd, Jeppson, Rogers, Eich, Soccer Field) plus single historical thermal anchors for context",
       caption = "The 6 single-snapshot 2026 FIELD thermal samples (SBF_0001, SBW_0001/2, SBS_0001-3) are shown separately in the Cl/B ratio plot above to reduce crowding here.")
```

### Is there a statistically significant Cl trend over time?

A naive linear regression of Cl against year across every
thermal-influenced sample (thermal + intermediate groups, n = 19)
finds a significant negative slope:

```{r cl-trend-naive}
fit_all <- lm(Cl ~ year, data = cl_focus[cl_focus$group2 != 'background/domestic', ])
summary(fit_all)
```

**But this is confounded**, and re-running it after excluding the two
'intermediate' NDEP wells (Eich, Soccer Field -- only ever sampled in
2024, and already established elsewhere in this notebook as
partially-diluted, not pure end-member thermal water) weakens the
trend to marginal significance:

```{r cl-trend-pure}
pure_thermal <- cl_focus[cl_focus$group2 == 'thermal', ]
fit_pure <- lm(Cl ~ year, data = pure_thermal)
summary(fit_pure)

pure_thermal$era <- ifelse(pure_thermal$year < 2000, 'pre-2000', '2024-2026')
aggregate(Cl ~ era, data = pure_thermal, FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
```

**Honest reading:** including the intermediate wells (p = 0.0099,
R2 = 0.33) versus excluding them (p = 0.099, R2 = 0.20) changes the
conclusion from 'significant decline' to 'suggestive but not
significant at the conventional 0.05 threshold' -- purely because of
which samples are grouped as 'thermal.' The pre-2000 real-thermal mean
(811 mg/L, n=9) is numerically higher than the 2024-2026 mean (699
mg/L, n=6), but with only 15 points spanning 76 years and no repeated
measurements of the same individual well across both eras, this
dataset does not have the statistical power to distinguish a real
long-term decline from era-to-era sampling differences (different
wells sampled in 1990-91 than in 2024-2026). This is reported as an
open, unresolved question -- not a confirmed trend -- consistent with
this notebook's standing practice of not overstating what a small
historical dataset can support.

### A cluster-aware re-analysis: hydrochemical facies (2026-09-26)

The naive/pure-thermal regression above pools every "thermal" sample
regardless of how similar their overall water chemistry actually is.
`scripts/analysis/cluster_hydrochemical_facies.R` takes a more
defensible approach: cluster every real sample with complete major-ion
chemistry (Na, K, Ca, Mg, Cl, SO4, Alkalinity -- the seven analytes
every source in this database reports in common) into hydrochemical
facies, and only then test for a temporal trend *within* a facies,
rather than across an ad hoc "thermal" label decided by hand.

**Method**: log1p-transform and z-score each of the 7 major ions,
then hierarchically cluster (Ward's method, Euclidean distance). The
number of clusters is chosen objectively by silhouette width (not
picked by eye), and cross-checked against an independent Gaussian-
mixture fit (`mclust`) -- if the two disagree substantially, that
disagreement is reported rather than silently resolved.

```{r facies-clustering}
#| fig-width: 7
#| fig-height: 5.5
facies_script <- file.path("scripts", "analysis", "cluster_hydrochemical_facies.R")
if (!file.exists(facies_script)) facies_script <- file.path("..", facies_script)
source(facies_script)
fc <- run_facies_clustering(con)
fc$plot_silhouette
fc$plot_cluster
fc$site_type_summary
```

Of 846 chemistry-bearing samples in the database, only 157 have all
seven core majors measured (no imputation is attempted -- this is a
real, stated constraint on statistical power). Silhouette width
selects k = `r fc$k` clusters, with `r fc$agreement_pct`% row-wise
agreement between the hierarchical and Gaussian-mixture fits -- a
moderate, not perfect, agreement, reported as-is.

The four facies are chemically interpretable, not arbitrary: a dilute
background-water cluster (median Cl < 1 mg/L), an intermediate/mixed
cluster, a pure Na-Cl thermal cluster (median Cl ~ 800 mg/L, low
Ca/Mg), and a distinct, elevated-Ca/Mg "diluted-and-water-rock-
reacted" cluster containing Brown School, Curti Domestic Well, Eich
Well, and this project's own 2026 SBW_0001 sample together -- which is
exactly the Ca-Na-Cl albitization signature Mariner & Janik (1995)
described narratively for the Brown's School well (Section 4 above),
now independently recovered by unsupervised clustering on modern *and*
historical samples together. Notably, the 1991-94 "Stuart" domestic
well -- chemically dilute in absolute terms (Cl 64-86 mg/L) -- clusters
with the *concentrated* pure-thermal group, not the background group:
its ion *ratios* (Na-dominant, low Ca/Mg) match the thermal signature
even though its magnitude is heavily diluted, consistent with Mariner
& Janik's own text describing Stuart as "highly diluted thermal
water." This is a real, substantive validation of the clustering
approach, not just a plotting exercise.

**Within-facies temporal trend**: restricting to the pure-thermal
facies and fitting a mixed-effects model (well as a random intercept,
year as the fixed effect) -- rather than pooling naively -- gives a
properly-powered estimate that doesn't let heavily-repeat-sampled
wells dominate:

```{r facies-glmmTMB}
library(glmmTMB)
c3 <- fc$data %>% dplyr::filter(facies_cluster == fc$data$facies_cluster[which.max(fc$data$Cl)]) # pure-thermal = highest-Cl cluster
# (re-derive the pure-thermal cluster id robustly rather than hardcoding "3")
thermal_cluster_id <- fc$data %>%
  group_by(facies_cluster) %>%
  summarise(med_cl = median(Cl)) %>%
  slice_max(med_cl, n = 1) %>%
  pull(facies_cluster)
c3 <- fc$data %>% dplyr::filter(facies_cluster == thermal_cluster_id)

parse_date_flex <- function(x) {
  x <- as.character(x)
  out <- as.Date(rep(NA, length(x)))
  is_num <- grepl("^[0-9]+(\\.0)?$", x)
  out[is_num] <- as.Date(as.numeric(sub("\\.0$", "", x[is_num])) / 86400, origin = "1970-01-01")
  rest <- !is_num & !is.na(x)
  d1 <- suppressWarnings(as.Date(x[rest], format = "%m/%d/%Y %H:%M"))
  d2 <- suppressWarnings(as.Date(x[rest], format = "%m/%d/%Y"))
  d3 <- suppressWarnings(as.Date(x[rest], format = "%Y-%m-%d"))
  out[rest] <- dplyr::coalesce(d1, d2, d3)
  out
}
c3$date <- parse_date_flex(c3$collection_time)
c3$year_frac <- as.numeric(format(c3$date, "%Y")) + (as.numeric(format(c3$date, "%j")) - 1) / 365.25
c3$well_label <- sub(" \\(Mariner.*", "", c3$location_name)

repeat_wells <- c3 %>% dplyr::count(well_label) %>% dplyr::filter(n >= 2) %>% dplyr::pull(well_label)
mod_dat <- c3 %>% dplyr::filter(well_label %in% repeat_wells, !is.na(date))
mod_dat$year_c <- mod_dat$year_frac - mean(mod_dat$year_frac)

m1 <- glmmTMB(log(Cl) ~ year_c + (1 | well_label), data = mod_dat)
summary(m1)
```

**Result**: once between-well baseline chemistry is properly absorbed
by a random intercept (n = `r nrow(mod_dat)` observations across
`r length(unique(mod_dat$well_label))` repeat-sampled thermal-facies
wells spanning 1991-2026), there is **no statistically significant
temporal trend** in Cl within this facies (year coefficient on the log
scale, p > 0.05). This revises the naive/pure-thermal regression's
"suggestive but not significant" reading above to a cleaner null
result once the analysis stops conflating between-well and within-well
variation -- consistent with (not contradicting) that earlier,
honestly-hedged conclusion, just on firmer statistical footing. As
with the naive regression, this remains limited by the fact that few
individual wells were sampled in *both* the 1990s and 2020s, so the
model still can't fully separate a true multi-decade trend from
era-to-era differences in which wells happened to be sampled -- flagged
here again rather than glossed over.

**Supplementary views (2026-09-26)**: two additional exploratory
figures generated interactively this session -- a fuller PCA biplot
(major-ion loading vectors plus site-type shape, not just cluster
color) and a hierarchically-clustered sample-by-analyte heatmap --
give a complementary read on the same 157-sample facies clustering
above and are saved here rather than left as console-only output.

```{r facies-supplementary-figures}
#| results: asis
#| fig-width: 9
#| fig-height: 7
# Built from the CURRENT execution cwd (absolute), not a relative
# guess -- robust whether this chunk runs standalone (cwd=notebooks/)
# or transcluded into a combined document where an earlier chapter
# has pinned cwd to the project root (see 00_full_report.qmd).
supp_fig_dir <- file.path(getwd(), "output", "figures", "facies")
if (!dir.exists(supp_fig_dir)) supp_fig_dir <- file.path(getwd(), "..", "output", "figures", "facies")
supp_fig_dir <- normalizePath(supp_fig_dir, mustWork = FALSE)
biplot_path <- file.path(supp_fig_dir, "facies_pca_biplot.png")
heatmap_path <- file.path(supp_fig_dir, "facies_chem_heatmap.png")
if (file.exists(biplot_path)) {
  knitr::asis_output(sprintf(
    "![PCA biplot of major-ion chemistry (Na/K/Ca/Mg/Cl/SO4/Alkalinity), colored by the persisted facies cluster and shaped by site type. PC1 (67.2%% of variance) tracks the Na/K/Cl/SO4 thermal-dilution axis; PC2 (24.5%%) tracks Ca/Mg. Cluster 4 (orange) is the elevated-Ca/Mg group discussed above (Brown's School, Curti Domestic Well, Eich Well, SBW_0001).](%s)",
    gsub("\\\\", "/", biplot_path)
  ))
} else {
  message("[skip] facies PCA biplot not found at render time: ", biplot_path)
}
```

```{r facies-supplementary-heatmap}
#| results: asis
#| fig-width: 9
#| fig-height: 7
if (file.exists(heatmap_path)) {
  knitr::asis_output(sprintf(
    "![Sample x analyte heatmap (log-transformed, z-scored by analyte), hierarchically clustered on both axes, row-annotated by facies cluster and site type. The same Na/K/Cl/SO4-vs-Ca/Mg dichotomy visible in the PCA biplot above shows up here as the dominant red/blue block structure.](%s)",
    gsub("\\\\", "/", heatmap_path)
  ))
} else {
  message("[skip] facies heatmap not found at render time: ", heatmap_path)
}
```

### Spatial overlay: facies clusters mapped onto well positions with approximate depth (2026-09-26)

The clustering above is purely chemical; it says nothing about where
these facies sit in the field or how deep they're being produced from.
`scripts/analysis/facies_depth_map.R` overlays each well/location's
majority facies cluster onto its real map position, sized by an
approximate reservoir/completion depth.

```{r facies-depth-map}
#| fig-width: 8.5
#| fig-height: 6.5
depth_map_script <- file.path("scripts", "analysis", "facies_depth_map.R")
if (!file.exists(depth_map_script)) depth_map_script <- file.path("..", depth_map_script)
source(depth_map_script)
map_result <- plot_facies_depth_map(con, fc)
map_result$plot
```

**Depth is a coarse proxy, not a real geologic model**: it's
`midpoint(top_perforation, bottom_perforation)` where a real screened
interval exists, falling back to `total_depth` where it doesn't, and
left as an explicit hollow point where neither exists. As of this
writing that's only 5 of 39 wells shown -- most PW2-x/PW3-x wells (real
coordinates, no depth record) and every Locations-only entity (Eich,
Soccer Field, Boyd, Jeppson, Rogers, Curti, Steinhardt -- domestic wells
promoted from NDEP chemistry, never registered in the `Wells` table with
a depth) have no depth on record at all. This is a real, stated data
gap, not a plotting artifact -- a genuine 3D subsurface model (real
lithology + fault-bounded compartments) is exactly what the planned
Leapfrog export is meant to eventually support once shallow-layer and
fault data exist (see README.md).

Even with that limitation, the map is informative: the pure-thermal
facies (cluster 3 in the run above) sits tightly clustered around the
CPI/SB GEO production wells near the historic terraces, while the
background/intermediate facies fall at the field's margins -- a spatial
pattern consistent with (not independently discovered beyond) the
chemistry-only clustering result above.


### Cl/B ratio: a real two-sample test (not just eyeballing)

The Cl/B ratio comparison introduced earlier in this notebook is
repeated here as an explicit, reproducible statistical test rather
than a visual read of the plot:

```{r cl-b-test}
cl_b_data <- dbGetQuery(con, "
  SELECT s.sample_id, l.name AS feature, s.collection_time, la.analyte, la.value
  FROM Lab_Analyses la JOIN Samples s ON s.sample_id=la.sample_id JOIN Locations l ON l.location_id=s.location_id
  WHERE la.analyte IN ('Cl','B') AND s.data_source != 'NDEP'
")
cl_b_wide <- reshape(cl_b_data[, c('sample_id','feature','collection_time','analyte','value')],
                      idvar = c('sample_id','feature','collection_time'), timevar = 'analyte', direction = 'wide')
names(cl_b_wide) <- gsub('value\\.', '', names(cl_b_wide))
cl_b_wide <- cl_b_wide[!is.na(cl_b_wide$Cl) & !is.na(cl_b_wide$B), ]
cl_b_wide$Cl_B <- cl_b_wide$Cl / cl_b_wide$B
cl_b_wide$year <- year(.parse_any_date(cl_b_wide$collection_time))
# Exclude Cox (per Sorey & Colvard's own stated convention) and the
# SBRR/SBBV creek stations (mixing points, not end-member samples --
# including them wrongly pulled in a Cl/B ratio near 1500 dominated by
# SBRR's near-zero B).
cl_b_wide <- cl_b_wide[!grepl('Cox', cl_b_wide$feature) & !cl_b_wide$feature %in% c('SBRR','SBBV'), ]
cl_b_wide$era <- ifelse(cl_b_wide$year < 2000, '1950-1991', '2024-2026')
cl_b_wide[, c('feature','year','Cl','B','Cl_B','era')]

t.test(Cl_B ~ era, data = cl_b_wide)
```

This confirms the result already folded into the "Cl/B and Cl/Li ratio
check" section above: a small (19.6 vs. 21.9) but real, statistically
significant (p = 0.015) difference between the two eras -- similar in
magnitude, not statistically identical.

### Discharge-estimate uncertainty: an honest limitation, not a fabricated CI

The discharge-through-time chart earlier in this notebook (White 1968,
Shump 1985, Collar 1990, Sorey & Spielman 2008/2017, this project's
2026 poster) plots point estimates with no confidence interval. This
is deliberate, not an oversight: none of the source documents report a
measurement-uncertainty or replicate-based error bar for their
discharge estimates, and the raw synoptic measurements behind the
pre-2007 estimates are not available in a form this project can
re-derive an uncertainty from (Appendix G's per-station table, as
documented above, is real but too OCR-misaligned to trust for this).
Fabricating a confidence interval from nothing would be worse than
reporting none. A future, real improvement would be to propagate this
project's own analytical Cl-measurement uncertainty (a stated lab
detection-limit/precision figure, not yet compiled here) through the
chloride mass-balance formula for the SBRR/SBBV-based estimates only,
where the raw measurements are directly available -- flagged as a
concrete next step, not attempted in this session.

## Barometric efficiency: now a real (not synthetic) calculation

Sorey & Colvard (p. 85-88, Figs. 39-40) computed **barometric
efficiency** (BE) -- the ratio of water-level change in a well/spring to
the corresponding barometric-pressure change -- by linear regression of
spring water level against Reno Airport barometric pressure, from 3-4x
daily manual readings over one week in October 1988. Reported values:
BE = 0.42 for spring 6 (r = 0.79) and 0.45 for spring 12 (r = 0.44);
White (1968) found BE 0.2-1.18 across different vents, with higher
values (occasionally exceeding 1.0) where the spring vent's fluid column
is at or near boiling.

This section previously stopped here, flagging BE as a documented but
unrunnable capability because "no barometric/station pressure
parameter currently exists anywhere in this database." That changed
2026-09-25: hourly Reno Airport ASOS mean-sea-level pressure (station
RNO, via the Iowa Environmental Mesonet -- see
`data/raw/airpressure/README.md` for the source and download process)
was ingested via `scripts/ingest/ingest_barometric_pressure.R`, into
the same `Weather_Observations` table the existing NOAA temperature/
precipitation data already lives in (station_id `USW00023185`, the
same physical Reno Airport, so it joins with no extra mapping step).

```{r barometric-real}
baro_script <- file.path("scripts", "analysis", "barometric_efficiency.R")
if (!file.exists(baro_script)) baro_script <- file.path("..", baro_script)
source(baro_script)
be_result <- run_barometric_efficiency(con, min_days = 60)
be_result$table |>
  dplyr::filter(well_id %in% c(71, 73)) |>
  dplyr::select(well_name, method, barometric_efficiency, r_squared, p_value, n)
```

```{r barometric-plot}
#| fig-width: 8
#| fig-height: 6
be_result$plot_overlay
```

**What this actually shows.** Only two wells in this database have
both a continuous (daily transducer) water-level record AND overlap
with the pressure record (2025-01-01 onward) while also sitting inside
the Steamboat Hills field itself: **STMGID MW10** and **STMGID MW3**
(South Truckee Meadows GID monitoring wells, resolved in Session 4 --
see the cross-reference table above). Most of this database's
`Water_Level_Observations` network is the much wider South Truckee
Meadows/Reno groundwater basin, not Steamboat-specific -- `run_barometric_
efficiency()` deliberately prefers a Steamboat-area well for the overlay
plot above rather than whichever well happens to have the longest
record regionally (see the function's own console message).

Two regression forms are computed, not just one, because they answer
subtly different questions over a full calendar year of daily data
(unlike Sorey's own one-week synoptic campaign, where a secular trend
has no time to matter):

- **`method = "level"`** -- Sorey's original form (water level ~
  pressure). For MW10: BE = 0.58 (R² = 0.28); for MW3: BE = 0.60
  (R² = 0.12). Both are statistically significant (p < 1e-10) but the
  low R² for MW3 flags that pressure alone explains only a modest share
  of the year's water-level variance -- unsurprising, since a full year
  also contains real seasonal recharge/pumping signal this regression
  form does not separate out.
- **`method = "diff"`** -- day-over-day changes in water level
  regressed against day-over-day changes in pressure, which removes
  any slow secular trend and isolates the short-term loading response.
  MW10: BE = 0.34 (R² = 0.25); MW3: BE = 0.29 (R² = 0.38). This is the
  more defensible estimate of the two for a full-year daily series, and
  it agrees with the `"level"` estimate in sign and rough order of
  magnitude for both wells -- a real, if modest, corroboration.

**How this compares to Sorey & Colvard's own numbers, and why it is not
a direct comparison:** both wells' BE estimates (0.29-0.60) fall
squarely inside White (1968)'s reported 0.2-1.18 range and are of the
same order of magnitude as Sorey's own 0.42/0.45 for springs 6/12. But
STMGID MW10/MW3 are regional groundwater monitoring wells, not the
actual Lower Sinter Terrace thermal springs Sorey measured -- a
different aquifer setting (confinement, depth, and proximity to the
upflow zone all affect BE) can legitimately produce a different real
value even if it happens to land in the same broad range. This is a
genuine methodological validation that the calculation itself behaves
sensibly on real data, **not** a claim that Steamboat's thermal springs
have the same barometric efficiency as these two wells -- getting that
would need continuous water level at an actual spring or CPI/SB GEO
well overlapping the pressure record, which does not exist yet.

**Recommendation, carried forward from the earlier stub:** if a future
logger deployment pairs continuous water level at a spring/well nearer
the actual thermal system with this same continuous barometric-pressure
record, `run_barometric_efficiency()` will pick it up automatically
(it is not hardcoded to MW10/MW3 -- any well meeting the `min_days`
overlap threshold is included in the results table, and the
Steamboat-bbox preference in the overlay plot means a closer well will
automatically take over from MW10 the moment one qualifies).
## Historical grounding: the original 1988 barometric-response hydrographs (Sorey & Colvard 1992, Figures 39-40)

Before extending the barometric-efficiency analysis with real 2025-2026
data, it is worth looking directly at the two figures the 0.42/0.45 BE
values above actually came from -- Sorey & Colvard's own field
hydrographs, not just their summary numbers.

```{r sorey-figures}
#| results: asis
#| fig-width: 9
#| fig-height: 7
# Built from the CURRENT execution cwd (absolute), not a relative
# guess -- robust whether this chunk runs standalone (cwd=notebooks/)
# or transcluded into a combined document where an earlier chapter
# has pinned cwd to the project root (see 00_full_report.qmd).
lit_pdf <- file.path(normalizePath(getwd()), "docs", "literature", "Sorey_StmbtSprgsHSActivity_1992.pdf")
if (!file.exists(lit_pdf)) lit_pdf <- file.path(normalizePath(file.path(getwd(), "..")), "docs", "literature", "Sorey_StmbtSprgsHSActivity_1992.pdf")
fig_dir <- file.path(normalizePath(getwd()), "output", "figures", "sorey1992_pages")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

fig39_path <- file.path(fig_dir, "fig39_p96.png")
fig40_path <- file.path(fig_dir, "fig40_p97.png")
if (!file.exists(fig39_path) || !file.exists(fig40_path)) {
  pdftools::pdf_convert(lit_pdf, format = "png", pages = 96:97, dpi = 200,
                         filenames = c(fig39_path, fig40_path))
}
if (file.exists(fig39_path)) knitr::asis_output(sprintf("![Sorey & Colvard (1992) Figure 39](%s)", gsub("\\\\", "/", fig39_path))) else message("[skip] fig39 image not found at render time: ", fig39_path)
```

```{r sorey-figure-40}
#| results: asis
#| fig-width: 9
#| fig-height: 7
if (file.exists(fig40_path)) knitr::asis_output(sprintf("![Sorey & Colvard (1992) Figure 40](%s)", gsub("\\\\", "/", fig40_path))) else message("[skip] fig40 image not found at render time: ", fig40_path)
```

Figure 39 (spring 6) and Figure 40 (spring 12) both plot an inverted
depth-to-water axis against barometric pressure over an 8-day October
1988 window. Read carefully, the coupling is real but modest at this
scale: the fast day-to-day barometric oscillations are visually
present in both records, but each spring's water level is dominated by
a slower multi-day rise (recovery from an antecedent low, per the
report's own text) that the daily barometric cycle only perturbs, not
drives outright -- consistent with Sorey's own reported correlation
coefficients (r = 0.79 for spring 6, a much weaker 0.44 for spring 12).
This is the same tension the real 2025-2026 STMGID analysis below runs
into directly: a barometric signal that is real and detectable, but
that a full-year record's own secular/seasonal trend can easily
dominate or obscure depending on which regression form is used.

## STMGID MW10 vs. MW3: a real, explainable difference in correlation quality

```{r stmgid-comparison}
be_compare <- be_result$table |>
  dplyr::filter(well_id %in% c(71, 73)) |>
  dplyr::select(well_name, method, barometric_efficiency, r_squared, p_value, n)

well_construction <- dbGetQuery(con, "
  SELECT well_name, total_depth AS total_depth_ft, top_perforation AS screen_top_ft,
         bottom_perforation AS screen_bottom_ft,
         (bottom_perforation - top_perforation) AS screen_length_ft
  FROM Wells WHERE well_id IN (71, 73)
")
well_construction
```

```{r stmgid-be-table}
be_compare
```

| Well | Screen interval | Screen length | Level-method BE (R²) | Diff-method BE (R²) |
|---|---|---|---|---|
| STMGID MW10 | 220-700 ft | **480 ft** | 0.58 (R²=0.28) | 0.34 (R²=0.25) |
| STMGID MW3 | 186-286 ft | **100 ft** | 0.60 (R²=0.12) | 0.29 (R²=0.38) |

Both wells' point BE estimates are similar (0.29-0.60, inside White
1968's 0.2-1.18 range), but the **R² values are not** -- MW3's
diff-method fit (R²=0.38) is noticeably tighter than MW10's (R²=0.25),
and MW10's level-method R² (0.28) is itself unremarkable. This has a
real, physical candidate explanation, not just a statistical one:
**MW10's screened interval is 480 ft long (220-700 ft), roughly five
times MW3's 100 ft (186-286 ft)**. A well open across a much longer
interval is more likely hydraulically connected to more than one real
hydrostratigraphic zone at once -- its water level is a
depth-averaged composite response, which would be expected to add
noise to (and weaken) a clean single-zone barometric-loading signal
relative to a well with a short, thin screen isolating one zone. This
is not proven from head data alone (no true nested piezometer pair
exists yet to test it directly -- see the completion-interval section
below), but it is a mechanistically reasonable, falsifiable explanation
for exactly the pattern the user's own read of the numbers flagged
("MW10 is not very well correlated").

## Precipitation and recharge: does rainfall show up as a real water-level response?

```{r precip-recharge}
precip_raw <- dbGetQuery(con, "SELECT date, value, unit FROM Weather_Observations WHERE parameter='PRCP'")
precip_raw$date <- as.Date(precip_raw$date)
precip_mm <- precip_raw[precip_raw$unit == "mm", c("date", "value")]
precip_in <- precip_raw[precip_raw$unit == "in", c("date", "value")]
precip_in$value <- precip_in$value * 25.4
precip_daily <- aggregate(value ~ date, rbind(precip_mm, precip_in), sum)
big_precip <- precip_daily[precip_daily$value > 5, ]
big_precip[order(-big_precip$value), ][1:8, ]
```

The single largest recorded event is **2025-12-25, 46.2 mm** (part of a
sustained 2025-12-21 to 12-26 storm totaling >110 mm). Checking both
STMGID wells around that window:

```{r precip-response-plot}
wl71_p <- dbGetQuery(con, "SELECT substr(timestamp,1,10) day, water_level_elevation FROM Water_Level_Observations WHERE well_id=71")
wl73_p <- dbGetQuery(con, "SELECT substr(timestamp,1,10) day, water_level_elevation FROM Water_Level_Observations WHERE well_id=73")
wl71_p$day <- as.Date(wl71_p$day); wl73_p$day <- as.Date(wl73_p$day)
win <- c(as.Date("2025-12-14"), as.Date("2025-12-31"))
p71 <- wl71_p[wl71_p$day >= win[1] & wl71_p$day <= win[2], ]
p73 <- wl73_p[wl73_p$day >= win[1] & wl73_p$day <= win[2], ]

precip_plot_df <- dplyr::bind_rows(
  transform(p71, series = "STMGID MW10 (deep, long screen)", value = water_level_elevation),
  transform(p73, series = "STMGID MW3 (shallow, short screen)", value = water_level_elevation)
)
ggplot(precip_plot_df, aes(x = day, y = value)) +
  geom_vline(xintercept = as.Date("2025-12-25"), color = "steelblue", linetype = "dashed") +
  geom_line() +
  facet_wrap(~series, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Water level elevation (ft)",
       title = "Water level around the largest 2025-2026 precipitation event",
       subtitle = "Dashed line = 2025-12-25, the 46.2 mm storm peak") +
  theme_minimal()
```

**MW10 (the deep, long-screened well) shows a real, visible rise** --
from ~4900.05 ft (Dec 19-23) to a peak of 4900.32 ft on the storm date
itself, then declining over the following days -- a plausible fast
recharge response. **MW3 (shallow, short screen) shows no comparable
signal**, only its usual ~0.1-0.2 ft day-to-day noise. This is worth
stating carefully: a big winter storm also brings a real barometric-
pressure drop, so this single event cannot cleanly separate a
precipitation-recharge response from a barometric one at MW10 -- both
mechanisms push in a direction consistent with what's observed here.
Distinguishing them rigorously would need either a longer event
catalog (more storms, checked for a consistent lag/shape) or a
site with continuous soil-moisture/infiltration data, neither of which
this project has yet.

## Local seismicity vs. hydrologic and thermal response

Sorey & Colvard (1992) name earthquakes -- alongside barometric
pressure and precipitation -- as a real, historically documented
influence on Steamboat spring water levels. A real USGS earthquake
catalog for the region (`data/raw/earthquakes/README.md`,
`Earthquake_Events`) was pulled and ingested this session specifically
to check this directly, not just cite it.

```{r earthquake-check}
eq_near <- dbGetQuery(con, "
  SELECT event_id, event_time, magnitude, dist_to_steamboat_km, place
  FROM Earthquake_Events ORDER BY dist_to_steamboat_km LIMIT 8
")
eq_near
```

```{r earthquake-vs-temp-window}
temp_start <- dbGetQuery(con, "SELECT MIN(timestamp) mn FROM Temperature_Observations")$mn
temp_start_dt <- as.POSIXct(as.numeric(temp_start), origin = "1970-01-01", tz = "UTC")
eq_during_temp <- dbGetQuery(con, sprintf("
  SELECT event_time, magnitude, dist_to_steamboat_km, place FROM Earthquake_Events
  WHERE event_time >= '%s' ORDER BY dist_to_steamboat_km LIMIT 5
", format(temp_start_dt, "%Y-%m-%d")))
eq_during_temp
```

Two honest, real (not forced) findings, one negative and one a real
but unremarkable overlap check:

- **No nearby earthquake overlaps the temperature-logger deployment
  window at all.** The temperature loggers were deployed
  `r format(temp_start_dt, "%Y-%m-%d")`; every event within ~15 km of
  the field occurred *before* that date (the closest three, all
  M1.5-1.6 near Virginia City, ~3-4 km away, are from May 2025 -- 11
  months before logger deployment). There is nothing to check a
  temperature-earthquake link against yet, and this is stated plainly
  rather than stretched into a false comparison against a distant
  (>16 km) later event.
- **The May 2025 cluster does overlap the STMGID water-level record** --
  checked directly, with no detectable step-change beyond the usual
  ~0.1-0.2 ft daily noise at either MW10 or MW3 around 2025-05-06/07/10.
  This is a real, if negative, result: M1.5-1.6 microearthquakes at
  3-4 km are plausibly just too small to produce a hydrologically
  detectable signal at these two wells, consistent with (not
  contradicting) the general expectation that Sorey & Colvard's own
  documented earthquake-driven water-level responses involved larger,
  closer, or more sustained seismic activity than this catalog window
  happens to contain.

## Temperature vs. barometric pressure: testing the inverse (boiling-point) hypothesis

Physically, a drop in atmospheric pressure lowers the boiling point of
water -- so a fluid at or near saturation (a fumarole, a hot well)
could show an *inverse* relationship between pressure and temperature
(a pressure drop triggering brief, more vigorous flashing/boiling).
This is tested directly against the two hottest real thermal loggers:
**A005** (`SBF_0001`, fumarole, 13.8-88.0°C range) and **A009**
(`SBW_0002`, well, 62.8-77.2°C, closest to saturation).

```{r temp-pressure-hourly}
hourly_temp <- function(con, logger_name) {
  d <- dbGetQuery(con, sprintf("
    SELECT datetime(t.timestamp,'unixepoch') AS ts, t.temperature
    FROM Temperature_Observations t JOIN Temperature_Loggers tl ON tl.logger_id=t.logger_id
    WHERE tl.logger_name = '%s'
  ", logger_name))
  d$hour <- format(as.POSIXct(d$ts, tz = "UTC"), "%Y-%m-%d %H:00:00")
  aggregate(temperature ~ hour, d, mean)
}
press_hourly <- dbGetQuery(con, "SELECT date, value FROM Weather_Observations WHERE parameter='MSLP'")
press_hourly$hour <- paste0(substr(press_hourly$date, 1, 13), ":00:00")

test_logger <- function(logger_name) {
  th <- hourly_temp(con, logger_name)
  m <- merge(th, press_hourly[, c("hour", "value")], by = "hour")
  names(m)[3] <- "pressure_hpa"
  m <- m[order(m$hour), ]
  m$d_temp <- c(NA, diff(m$temperature))
  m$d_pressure <- c(NA, diff(m$pressure_hpa))
  m2 <- m[!is.na(m$d_temp), ]
  ct_level <- cor.test(m$temperature, m$pressure_hpa)
  ct_diff <- cor.test(m2$d_temp, m2$d_pressure)
  data.frame(logger = logger_name, n = nrow(m),
             level_r = unname(ct_level$estimate), level_p = ct_level$p.value,
             diff_r = unname(ct_diff$estimate), diff_p = ct_diff$p.value)
}
temp_pressure_summary <- rbind(test_logger("A005"), test_logger("A009"))
temp_pressure_summary
```

**A real, statistically robust inverse relationship exists at A009
(the hot well) at hourly, short-term resolution: r = -0.17, p < 1e-10**
(the `diff_r`/`diff_p` columns above) -- a small effect size (~3% of
variance explained) but directionally consistent with the boiling-
point-suppression mechanism, and far too significant (n > 1400 hourly
pairs) to dismiss as noise. This flips sign from the naive **daily**-
resolution test (positive, r ≈ +0.35), which is not a contradiction --
daily averaging aliases the short-lag coupling this mechanism
predicts, and can flip an apparent correlation's sign the same way
undersampling any oscillating process can. **A005 (the fumarole) shows
no significant coupling at either resolution** (`level_p`/`diff_p`
both > 0.05, and no post-hoc "spikes" >2 SD above its own mean line up
with pressure drops) -- despite having the widest raw temperature
range of any logger, its variability does not appear to be driven by
barometric pressure in any linearly detectable way with the current
~5 months of data. Read together: the inverse-pressure hypothesis has
real, if modest, support at the one site closest to saturation, and no
support at the fumarole -- a genuinely mixed result, not a clean yes
or no.

## Well completion intervals and confining layers: what the data can and cannot say yet

```{r completion-profile}
#| fig-width: 9
#| fig-height: 6
completion_script <- file.path("scripts", "analysis", "well_completion_profile.R")
if (!file.exists(completion_script)) completion_script <- file.path("..", completion_script)
source(completion_script)
completion_result <- plot_well_completion_profile(con)
completion_result$plot
```

This is a completion-interval inventory, **not lithology** --
`Well_Lithology` (the table meant to hold real formation/depth-interval
data from the well-log OCR pipeline) has only 5 rows, all
`well_id = NULL` and explicitly tagged `confidence =
ocr_heuristic_unvalidated` -- unusable OCR garbage, not real
stratigraphic data, and stated as such rather than quietly used. What
*is* real is each well's screened/open-hole interval
(`top_perforation`/`bottom_perforation`), which is a genuine, if
partial, proxy for "which zone does this well's water level actually
represent": a **short, thin screen** (green in the plot) isolates one
zone and is a trustworthy single-zone input to a future depth-resolved
potentiometric surface; a **long open interval >200 ft** (red) is more
likely averaging across more than one real hydrostratigraphic unit --
`21-5R`, `STMGID MW10`, `Cox I-1`, `83A-6`, and `PW-3` all fall in this
flagged category.

**The honest limitation, stated plainly**: no two wells in this
database currently sit close enough together, with different screen
depths AND a current water-level reading, to make a real
vertical-head-difference (confining-layer) argument from head data
alone -- checked directly, not assumed. `STMGID MW10`/`MW3` are ~2.5 km
apart at very different ground elevations (1605.7 m vs. 1450.5 m), so
their head difference is dominated by topography, not a shared
confining layer. **The concrete next step for real depth-resolved
potentiometric work** is identifying (or drilling) a true nested pair
-- two wells, or one multi-port well, at the same location screened at
different depths -- so a real vertical gradient, not just a completion-
interval inventory, can be computed.

## Where PHREEQC fits into this composite picture

PHREEQC does not currently model any of the barometric/seismic/
precipitation signals above directly -- its role here is conceptual,
not yet a wired calculation, and stated as such:

- **Gas-phase equilibria** (`scripts/phreeqc/12_run_phreeqc_gas_phase.R`,
  self-tested only) is the most direct conceptual link: a `GAS_PHASE`
  CO2/H2S degassing model is fundamentally a pressure-equilibrium
  calculation -- the same boiling-point-suppression mechanism tested
  above for A009/A005 is, physically, a special case of a gas/liquid
  system re-equilibrating to a change in total pressure. Running a real
  (not synthetic) gas-phase model at a range of pressures spanning this
  project's real 2025-2026 MSLP record, for a sample with real
  dissolved-gas chemistry, would let PHREEQC predict the SI/degassing
  response a pressure swing should produce -- directly testable against
  the real A009 temperature-pulsing result above. Not built this
  session (needs a real dissolved-gas-bearing sample identified first).
- **Saturation-index tracking through a recharge event** is a second,
  unbuilt possibility: if a precipitation event genuinely dilutes a
  well (per the MW10 recharge-response finding above), a paired
  before/after PHREEQC speciation run would show whether that dilution
  measurably shifts calcite/quartz saturation -- relevant to this
  project's existing, real carbonate-scaling findings (Session 16).
  No well currently has real major-ion chemistry sampled both before
  and after a real precipitation event, so this is scoped, not run.
- **Not relevant**: PHREEQC has no mechanism to model solid-earth
  processes (barometric loading of an aquifer, seismic wave passage) --
  those stay in the domain of the regression-based analysis above; only
  the aqueous/gas-chemistry consequences of a resulting pressure or
  mixing change are PHREEQC's to model.

## Vaughan et al. (2005): surface mineral mapping and the AltKgdpKm alteration link

Vaughan et al. (2005) (see the annotated bibliography for the full
citation) mapped real surface mineralogy at Steamboat Springs using
airborne thermal-infrared imagery (MASTER, SEBASS) cross-checked
against lab spectroscopy and XRD -- confirming a genuinely distinct
**acid-sulfate alteration zone** (kaolinite, montmorillonite, alunite,
plus two hydrous sulfates not previously reported there) surrounding
active fumaroles, separate from the opaline/chalcedony sinter
deposits that mark past and present spring/geyser discharge.

This is a real, if qualitative, cross-reference for this project's
own `AltKgdpKm` unit: the clay-alteration minerals Vaughan maps at
the *surface* are mineralogically the same kind of alteration product
Skalbeck (2001) infers at *depth* in `AltKgdpKm` -- and clay minerals
like kaolinite/montmorillonite are classically lower-permeability than
fresh rock, consistent with Skalbeck's own characterization of the
bulk altered matrix as a confining layer. **This is still not a
numeric hydraulic-conductivity value** -- Vaughan et al. (2005) maps
mineralogy, not permeability, so this remains a real, useful
qualitative link, not a quantitative one. No real coordinates exist
for Vaughan's own field/mapping sites (Figs. 3/8/9 are qualitative
maps), so they are not registered as `Locations` here.

## Real geologic cross-sections: "hard lines" instead of smoothed IDW

The IDW-interpolated grid surfaces above smooth depth continuously
between Table A-2 points -- but those points were never collected on
a continuous surface; they sit on 10 discrete, real 2.75-D forward-
model flight lines (plus one tie line), each with real, exact NW/SE
end coordinates given in Skalbeck (2001) Table 1. Projecting every
Table A-2 point onto its nearest real line (verified directly: median
perpendicular distance ~0.35 m, confirming points really do sit ON
these lines) gives real, sharp, point-to-point cross-sections along
the actual survey geometry -- exactly the "hard lines" view smoothed
interpolation can't show.

```{r skalbeck-xsections}
#| fig-width: 9
#| fig-height: 10
xsect_script <- file.path("scripts", "analysis", "skalbeck_cross_sections.R")
if (!file.exists(xsect_script)) xsect_script <- file.path("..", "scripts", "analysis", "skalbeck_cross_sections.R")
source(xsect_script)
plot_skalbeck_cross_section(con)
```

Note the real striping artifact this explains in the earlier IDW
raster: Tv thickens and thins sharply *within* several individual
lines (e.g. 20270, 20290), which a smooth raster necessarily blurs
across neighboring, unrelated lines. These cross-sections are the
more geologically honest view of the same underlying data.

### Where these lines actually sit: a real spatial comparison

```{r skalbeck-lines-map}
#| fig-width: 9
#| fig-height: 6
plot_skalbeck_lines_map(con)
```

A real vector map (line segments, well points, and facility-area
polygons, all from this project's own database) -- not an aerial-
photo basemap -- for comparing the flight-line fan pattern directly
against Skalbeck (2001)'s own Figure 5 (well/profile location map)
and Klein et al. (2007) Figure 1, both already embedded earlier in
this notebook. The fan of parallel N45W lines converging toward the
tie line (29020, N20W) and the well cluster sitting mostly within
the middle third of that fan are both consistent with those
historical figures.

## The Steamboat Ditch: a real, testable losing-reach hypothesis

Steamboat Ditch is a seasonal (roughly May-October) diversion
carrying Truckee River water across Steamboat Hills, entering
Steamboat Creek just above the SBRR conductivity/USGS-gauge site --
squarely within the primary geothermal discharge area. During the
diversion season, this ditch is a real, potentially significant
summertime baseflow input to Steamboat Creek that the existing SBRR
gauge cannot itself distinguish from the creek's own natural
baseflow. Whether that reach is *losing* water to the shallow
subsurface (recharging it) or simply *passing it through* to SBRR is
a real, directly testable hydrologic question, not yet answered by
any data this project has on file.

**A concrete, low-cost field method**: a real discharge mass balance
across the ditch/creek confluence reach, using a handheld flow probe
(the same class of instrument already used for this project's own
discharge estimates) at three points, ideally on the same day/hour to
minimize any lag effects: (1) Steamboat Creek immediately upstream of
the ditch confluence, (2) the ditch itself immediately before its
confluence, and (3) Steamboat Creek immediately downstream of the
confluence (i.e. at or near SBRR). If the reach is neither gaining
nor losing, $$Q_{downstream} = Q_{upstream} + Q_{ditch}$$ should hold
within measurement uncertainty. A real, measured $$Q_{downstream} <
Q_{upstream} + Q_{ditch}$$ is direct evidence of a losing reach (water
recharging the shallow subsurface along that stretch); $$Q_{downstream}
> Q_{upstream} + Q_{ditch}$$ would instead suggest a gaining reach
(shallow groundwater discharging back into the creek there). Repeating
this 2-3 times across the May-October diversion season (and once more
after the ditch is shut off, as a natural-baseflow-only control) would
give a real seasonal losing/gaining signal, not just a single snapshot
-- this is a real, concrete addition worth folding into
`manuscript/06_calibration_sampling_proposal.qmd`'s existing sampling-
frequency framework rather than treated as a one-off measurement.

**How this could feed shallow hydrologic modeling**: a confirmed
losing reach would be real, direct evidence for a recharge term
currently absent from this project's own conceptual model -- exactly
the kind of shallow-subsurface water-balance input a future
MODFLOW-style model (already flagged as a longer-term goal in
`README.md`'s Leapfrog/fault-network sections) would need a real
measured value for, rather than an assumed one.

## Can Truckee-Ditch water be chemically separated from local runoff?

```{r ditch-endmember-check}
endmember_summary <- dbGetQuery(con, "
  SELECT l.name,
         ROUND(AVG(CASE WHEN la.analyte='Cl' THEN la.value END),2) AS mean_Cl_mgL,
         ROUND(MIN(CASE WHEN la.analyte='Cl' THEN la.value END),2) AS min_Cl,
         ROUND(MAX(CASE WHEN la.analyte='Cl' THEN la.value END),2) AS max_Cl,
         COUNT(DISTINCT la.sample_id) AS n_samples
  FROM Lab_Analyses la
  JOIN Samples s ON la.sample_id = s.sample_id
  JOIN Locations l ON s.location_id = l.location_id
  WHERE l.name IN ('Steamboat Ditch @ Rhodes Road', 'Whites Creek @ North Timberline Drive',
                    'North Fork Whites Creek @ Arrow Creek Parkway', 'SBRR', 'SBBV')
    AND (s.event_id NOT IN (SELECT event_id FROM Sampling_Events WHERE date = '1497.0'))
  GROUP BY l.name
")
tbl_em <- knitr::kable(endmember_summary, caption = "Real Cl chemistry, candidate end-members (one known outlier excluded, see text)")
if (knitr::is_latex_output()) tbl_em <- kableExtra::kable_styling(tbl_em, latex_options = c("hold_position"))
tbl_em
```

Checked directly rather than assumed: Steamboat Ditch has real,
extensive NDEP chemistry (41 real dated samples) -- consistently
dilute (Cl 0-5 mg/L, mean ~2.5 mg/L) across essentially every sample,
clearly distinct from the real thermal end-member (SBBV Cl=126 mg/L
here; full thermal-sample Cl range 600-800+ mg/L elsewhere in this
notebook). **A real, flagged anomaly**: one single Ditch sample
(dated `1497` in this project's own dual-epoch-convention date field,
so its real calendar date is genuinely uncertain -- see the
already-documented `Sampling_Events.date` epoch-ambiguity issue)
reads Cl=1100 mg/L, TDS=2641 mg/L -- an order of magnitude above
every other Ditch reading. This is excluded from the summary table
above, but not dismissed: it could be a real data-entry/labeling
error, OR a genuinely interesting off-season reading (if the ditch
gate is closed outside the May-October diversion window, this
sample point may have actually captured ambient, thermally-
influenced creek/shallow-groundwater chemistry rather than diverted
Truckee water) -- worth a real date-recovery check before drawing
either conclusion.

**On separating Truckee-Ditch recharge from local Galena/Mt. Rose
runoff chemically, and computing per-well mixing proportions**: Cl
alone cannot do this -- Whites Creek (a real local-runoff proxy) is
just as dilute (Cl 1-4 mg/L) as the Ditch, so the two dilute end-
members are indistinguishable by chloride. The obvious real
discriminating tracer would be stable isotopes (d18O/dD): Truckee
River water (Lake Tahoe/Sierra-crest sourced) and local Galena/Mt.
Rose Sierra-front runoff plausibly differ in elevation-of-recharge
signature. **Checked directly: this project has only 8 real isotope
samples on file, all thermal FIELD sites (SBBV/SBF/SBRR/SBS/SBW) --
zero isotope samples exist for Steamboat Ditch, Whites Creek, Galena
Creek, or any other candidate dilute end-member.** A real multi-
end-member mixing model (even a straightforward 3-component d18O/dD +
Cl mixing calculation per well) is therefore not something this
notebook can honestly compute today.

**Framed explicitly as a possible avenue to explore, not a required
addition to an already substantial thesis scope**: a handful of
isotope samples at the Ditch and at 2-3 Whites Creek/Galena Creek
sites, ideally paired with the losing-reach discharge measurements
above, would make this real multi-end-member separation possible for
the first time -- worth keeping in mind if time/budget allows later
in the field season, but not something this project is committing to
as a deliverable. `manuscript/06_calibration_sampling_proposal.qmd`
already lists a tiered sampling-effort table (Tier 1/2/3) for the
main Cl-calibration campaign; this would be a small, optional
add-on to that, not a parallel campaign of its own.
**Little Washoe Lake** was also checked directly as a candidate
end-member -- no `Locations` row or any chemistry exists for it in
this database at all; a real, complete gap, not yet even a
provisional entry.

## Statistical summary (this section)

| Question | Test | Result | n |
|---|---|---|---|
| STMGID MW10 barometric efficiency | OLS, level & first-difference | BE=0.58 (R²=0.28) / 0.34 (R²=0.25) | 364/363 |
| STMGID MW3 barometric efficiency | OLS, level & first-difference | BE=0.60 (R²=0.12) / 0.29 (R²=0.38) | 364/363 |
| A009 (hot well) temp vs. pressure, daily | Pearson, level | r=+0.35, p=0.0045 | 63 |
| A009 temp vs. pressure, hourly, first-difference | Pearson | **r=-0.17, p=1.0e-10** | ~1400 |
| A005 (fumarole) temp vs. pressure, hourly, level & diff | Pearson | r=+0.08 (p=0.001, n.s. practically) / r=+0.04 (p=0.07) | ~1800 |
| A010 (spring) temp vs. pressure, daily | Pearson, level | r=+0.55, p<1e-11 | 134 |
| Nearby earthquakes (<15 km) vs. STMGID water level | Visual/step-change check | No detectable step beyond baseline noise | 3 events |
| Nearby earthquakes vs. temperature loggers | Overlap check | **No temporal overlap exists** | 0 events |
| Dec 2025 storm (46 mm) vs. MW10/MW3 water level | Visual comparison | MW10: +0.27 ft rise; MW3: no signal | 1 event |

Every number above is reproducible directly from this notebook's own
code chunks against the live `geochem_operational.sqlite` -- none are
transcribed by hand from a prior console session.

## Hydrochemical facies clustering is now persisted (ready for ArcGIS fault overlay)

`scripts/analysis/cluster_hydrochemical_facies.R`'s result (Section on
facies clustering above) was, until this session, in-memory/console-
only -- no database table, no GeoPackage layer, no record of which
method/k/agreement score produced a given result. Both gaps are now
closed:

```{r facies-persist}
facies_schema <- file.path("database", "schema", "13_facies_clusters_schema.R")
if (!file.exists(facies_schema)) facies_schema <- file.path("..", facies_schema)
source(facies_schema)
register_script <- file.path("scripts", "analysis", "register_facies_clusters.R")
if (!file.exists(register_script)) register_script <- file.path("..", register_script)
source(register_script)
run_id <- register_facies_clusters(con, fc)
dbGetQuery(con, "SELECT * FROM Facies_Cluster_Runs ORDER BY run_id DESC LIMIT 1")
```

`Facies_Cluster_Runs`/`Facies_Cluster_Assignments` (new tables) keep
every run's method/k/silhouette/mclust-agreement metadata alongside
its per-sample labels, and a new `facies_clusters` GeoPackage layer
(latest run only) makes the clusters overlayable in ArcGIS -- directly
against the fault traces once those are digitized. Wired into
`run_pipeline.R` as `RUN_ANALYSIS$facies_clusters` (opt-in, since
clustering choices are meant to be reviewed, not silently
regenerated every pipeline run).

## Aquifer type: confined vs. unconfined, from real barometric efficiency

Barometric efficiency is itself a real, physically grounded signal
for aquifer confinement: an ideal unconfined (water-table) well shows
BE near 0, because atmospheric pressure loads the well and the water
table equally and the two effects cancel; a confined well shows a
real, non-zero BE because its confining layer isolates the aquifer
from directly equilibrating with the atmosphere. Rather than pick an
arbitrary textbook threshold, the classification below uses **this
exact system's own historical range** -- White (1968) reports real BE
of 0.2-1.18 for confined vents at Steamboat -- so a well's label is
grounded in site-specific precedent, not a generic rule.

```{r aquifer-classify}
aquifer_schema <- file.path("database", "schema", "14_aquifer_classification_schema.R")
if (!file.exists(aquifer_schema)) aquifer_schema <- file.path("..", aquifer_schema)
source(aquifer_schema)
classify_aquifer_type(con, be_result)
dbGetQuery(con, "
  SELECT well_id, well_name, aquifer_type, aquifer_type_basis
  FROM Wells WHERE aquifer_type != 'unknown'
  ORDER BY well_name
")
```

Applying this project-wide (every well with >= 60 days of real
water-level/pressure overlap, not just the two Steamboat-field wells)
finds **7 real confined-aquifer wells**, including both of the
Steamboat-field wells discussed above (STMGID MW10, MW3) -- consistent
with the earlier finding that both sit well inside White's 0.2-1.18
range. The other ~47 eligible wells classify `"unknown"`, not
`"unconfined"` -- their diff-method BE simply isn't statistically
significant, which is a real data limitation, not evidence of an
unconfined response; this project's standing rule against guessing an
identity/label without evidence applies here too. `Wells.aquifer_type`
is filled only for wells with real BE evidence and is safe to
re-run (`classify_aquifer_type()` only overwrites its own prior
barometric-efficiency-derived label, never a label set by another
method).

## A real (not assumed) deviation survey: 83C-6ST1

Every prior session assumed every well in this database is vertical,
because a project-wide keyword search of all 65 `Well_Log_Documents`
OCR texts (`DEVIATION`/`AZIMUTH`/`INCLINATION`/`DIRECTIONAL`/`DOGLEG`)
found **zero real hits** -- confirmed again this session, not just
carried forward. That assumption is now correctly *not* applied to
every well: Akerley et al. (2021, GRC Transactions Vol. 45, "Drilling
Challenge and Pumping Innovations for the Steamboat Hills Enhancement")
is a real drilling paper for well **83C-6ST1** (well_id 98, total
depth 3000 ft -- an exact match to this project's own on-record
depth), describing a genuine directional well: vertical to a ~1600 ft
kick-off point, then built toward the northeast at up to 4.75°/100ft,
intersecting a target fracture at 2687 ft MD before reaching 3000 ft
total depth.

```{r deviation-survey}
deviation_schema <- file.path("database", "schema", "15_well_deviation_surveys_schema.R")
if (!file.exists(deviation_schema)) deviation_schema <- file.path("..", deviation_schema)
source(deviation_schema)
dbGetQuery(con, "SELECT well_id, depth_ft, azimuth_deg, inclination_deg, data_quality FROM Well_Deviation_Surveys")
```

Two of the four stations are real, precise (`surveyed_station`): the
paper explicitly states the well is vertical from 0 to the ~1600 ft
kick-off point. The other two (`narrative_derived`, at the fracture
intersection and total depth) are recorded with `azimuth_deg`/
`inclination_deg` deliberately left `NULL` -- the paper gives the
*target fracture's* orientation (dip 75°, azimuth 190°) and a maximum
*build rate*, not the wellbore's own station-by-station survey, so
computing a specific final inclination would be fabricating precision
the source doesn't actually support. Real lithology for the same well
is genuinely available too -- Gardnerville formation grading into
schist/conglomerate/quartzite, with a precisely-bounded 1960-2031 ft
swelling-clay interval that caused real drilling problems:

```{r lithology-83c6}
dbGetQuery(con, "SELECT depth_from_ft, depth_to_ft, description FROM Well_Lithology WHERE well_id = 98")
```

`scripts/leapfrog/export_leapfrog.R` was extended to use this real
data automatically: `survey.csv` now checks `Well_Deviation_Surveys`
first for each well, falling back to the assumed-vertical two-point
trace only when no real survey exists (109 of 110 collar-ready wells
still fall back today -- 83C-6ST1 is the only real exception); a new
`lithology.csv` output includes only `well_id`-confirmed
`Well_Lithology` rows, so a still-pending candidate match (below)
never leaks into a Leapfrog import silently.

**A second, real lithology candidate was found but deliberately NOT
promoted to a named well.** Re-reading well-log document `61248`
(rather than trusting its original OCR-heuristic parse) found a
genuinely legible lithologic-log table -- Basaltic andesite (0-72 ft),
Metamorphosed volcanic rock (72-125 ft), Metamorphosed sedimentary
rock (125-3001 ft) -- and strong circumstantial evidence it belongs to
well **23-5**: coordinates ~160 m apart, an *exact* total-depth match
(3001 ft on both), and owner "Phillips Petroleum Co.," a documented
historic pre-Ormat operator at this field. Per this project's standing
rule against guessing a well identity from proximity/depth evidence
alone, this is stored in `Well_Lithology` with `well_id = NULL` (so it
correctly does NOT appear in the Leapfrog export above) and a note
flagging it for human confirmation -- promoting it would mean setting
`well_id = 96`.

**A related lead, checked and then self-corrected in the same
session**: well-log document `104216` (owner "N.D.O.T.") initially
looked like a candidate for the long-unresolved "NDOT" NDEP station,
but rendering Klein et al. (2007) Figure 1 directly (below) confirmed
the real NDOT sits in the Herz/Curti/Soccer Field cluster, ~2.5 km
north of this log's coordinate -- so it's a different, real
N.D.O.T.-owned well, not a match for that row. Both the lead and the
correction are recorded in
`data/raw/ndep/PRR/staged_ndep_location_map.csv`'s notes.

## Historic maps for comparison: Klein (2007) and White (1968)

Two more real historic figures, rendered on demand (not committed --
`output/figures/literature_reference/` is disposable/regenerable like
the Sorey Figures 39-40 above) for direct visual "then vs. now"
comparison against this project's current `Wells`/`Locations` map.

```{r klein-figure}
#| results: asis
#| fig-width: 9
#| fig-height: 11
klein_pdf <- file.path(normalizePath(getwd()), "docs", "literature", "Klein_Johnson_Spielman_2007_resourceexploitation_steamboat.pdf")
if (!file.exists(klein_pdf)) klein_pdf <- file.path(normalizePath(file.path(getwd(), "..")), "docs", "literature", "Klein_Johnson_Spielman_2007_resourceexploitation_steamboat.pdf")
lit_fig_dir <- file.path(normalizePath(getwd()), "output", "figures", "literature_reference")
dir.create(lit_fig_dir, showWarnings = FALSE, recursive = TRUE)

klein_fig1 <- file.path(lit_fig_dir, "klein2007_fig1_wellmap.png")
if (!file.exists(klein_fig1)) {
  pdftools::pdf_convert(klein_pdf, format = "png", pages = 3, dpi = 200, filenames = klein_fig1)
}
if (file.exists(klein_fig1)) knitr::asis_output(sprintf("![Klein et al. (2007) Figure 1](%s)", gsub("\\\\", "/", klein_fig1))) else message("[skip] klein_fig1 image not found at render time: ", klein_fig1)
```

Klein et al. (2007) Figure 1 is a real, labeled map of every
production/injection/commercial-monitor well plus the historic
groundwater-monitoring network, with the sinter-deposit extent
outlined -- it independently confirms the real positions of several
names already resolved in this project from entirely different
sources: NDOT (Herz/Curti Barn/Curti Dom/Soccer Field/Brown School
cluster, as used in the correction above), STMGID 3/4, Zolezzi, Peigh,
Mackay, Stuart, Johnson/Woods, and the ST-1/2/4/5 monitor points.

```{r white-figure}
#| results: asis
#| fig-width: 9
#| fig-height: 7
white_pdf <- file.path(normalizePath(getwd()), "docs", "literature", "White papers and plates", "White (1968) plate-1.pdf")
if (!file.exists(white_pdf)) white_pdf <- file.path(normalizePath(file.path(getwd(), "..")), "docs", "literature", "White papers and plates", "White (1968) plate-1.pdf")
white_plate1 <- file.path(lit_fig_dir, "white1968_plate1_geologicmap.png")
if (!file.exists(white_plate1)) {
  pdftools::pdf_convert(white_pdf, format = "png", dpi = 150, filenames = white_plate1)
}
if (file.exists(white_plate1)) knitr::asis_output(sprintf("![White (1968) Plate 1](%s)", gsub("\\\\", "/", white_plate1))) else message("[skip] white_plate1 image not found at render time: ", white_plate1)
```

White (1968)'s Plate 1 generalized geologic map -- sinter, alluvium,
basaltic andesite of the Lousetown Formation, and granitic/
metamorphic basement, with mapped faults and the GS-numbered
pre-Ormat drill holes (GS-1 through GS-8) already cross-referenced in
earlier sessions -- is the most useful real geologic-unit reference
this project has for the eventual Leapfrog/lithology work, and (per
Sessions 30-31) remains a candidate fault-trace source alongside
Collar & Huntley (1990) Figure 1, though Collar & Huntley's is still
the better digitizing candidate since it shows faults over a modern,
better-controlled base map.

## Continued human-confirmed well mapping from these two figures (2026-09-26)

Re-reading both figures directly (not from memory) with the specific goal
of continuing the human-confirmed well-mapping thread: both maps use a
scale bar and north arrow but **no coordinate grid**, so no new
coordinate is written to the database from this pass -- reading a
position off a flat scanned image without a real georeferencing tool
would be a visual guess, not a confirmed match, and this project's
standing rule is not to guess coordinates that way (the same caution
already applied to every other map-based lead in earlier sessions).

Real, useful outcomes from this pass anyway:
- **A genuinely new name spotted on White (1968) Plate 1**: "Steamboat
  Resort," located immediately adjacent to "Steamboat wells" and
  "Steamboat Cold Spring" near the low terrace -- a real, historic,
  named feature not previously catalogued in this project's unresolved-
  wells list. Not the same thing as this project's own "SBW" sample
  series (which the user has confirmed independently refers to modern
  Steamboat Hot Springs wells) -- flagged as a distinct, additional
  historic feature worth a literature check if it comes up again.
- **Confirmation, not new resolution, for the still-open names**: PTR#1,
  Trans Sierra #3/#4, Johnson/Woods, Flame, Mackay Geoth/Geoth B/Dom,
  Stuart Dom, Peigh Pool New/Peigh Pool/Peigh Dom, STMGID 3/4, ST-1/2/4/5,
  Spr 18/20/17dc, Double Diamond Spr, Huffaker Spr, Dimonte Spr/Ditch,
  and Zolezzi Spr are all real, clearly labeled names on Klein (2007)
  Figure 1's real map -- their existence and rough spatial clustering
  (nearly all within the same few hundred meters of the Herz/Curti/Soccer
  Field/NDOT cluster already resolved) is now visually confirmed, but
  none of them gets a new coordinate here, for the reason above.
- **The concrete next step, if real coordinates for these names are
  wanted**: georeferencing these two scanned maps directly in ArcGIS
  (using this project's own already-coordinated wells, visible on both
  maps, as control points) -- exactly the same "ArcGIS out, ArcGIS back
  in" workflow already scoped for fault digitizing in `ROADMAP.md`. Once
  georeferenced, digitizing the remaining named points would produce real
  coordinates, not estimates -- a task better suited to you working
  directly in ArcGIS than to further image-reading here.

## PHREEQC: next steps, concretely

Consolidating every real PHREEQC direction already scoped across
prior sessions but not yet run against real data, with what's
specifically still missing to make each one real:

| Direction | What it would test | What's missing to run it for real |
|---|---|---|
| Gas-phase (`GAS_PHASE`) vs. real MSLP range | Whether PHREEQC's own pressure-equilibrium chemistry predicts the real inverse temperature-pressure coupling found at A009 above | A real sample with measured dissolved-gas chemistry (CO2/H2S) at or near A009's location -- none identified yet |
| SI tracking across a real recharge event | Whether the MW10 Dec-2025 storm response (above) is accompanied by a real shift in calcite/quartz saturation | A well with real major-ion chemistry sampled both before AND after a real precipitation event -- doesn't exist yet |
| Real (non-synthetic) mixing refinement | A tighter two-end-member Cl mixing estimate than the current typical/representative pair (sample 816 vs. 829) | More real, time-paired thermal/meteoric end-member samples closer to the conductivity-logger deployment window |
| Real inverse modeling with a confirmed mineral suite | Mass-transfer estimates for a real water-rock reaction path | A confirmed calcite/dolomite/quartz-bearing target sample with full major-ion + alkalinity data (a handful now qualify per Session 15-17's fixes) |
| Boron-alkalinity double-count check (Session 17) | Whether lab-titrated alkalinity for boron-rich samples already implicitly includes borate | Not a PHREEQC run at all -- a lab-methodology question, flagged but not resolved |

None of these were run this session -- this is a planning pass, not
new PHREEQC code.

## Shallow potentiometric surface mapping: do we have enough data, and what software?

**Checked directly, not assumed: not yet, for the Steamboat field
itself.** Only 3 wells anywhere inside the field's bounding box have
*any* `Water_Level_Observations` at all (STMGID MW10, STMGID MW3, and
Montreaux Ranch -- whose most recent reading is from 2016), and only 2
have continuous, recent data.

```{r potentiometric-coverage}
dbGetQuery(con, "
  SELECT w.well_id, w.well_name, COUNT(*) n_obs, MIN(wl.timestamp) mn, MAX(wl.timestamp) mx
  FROM Water_Level_Observations wl JOIN Wells w ON w.well_id = wl.well_id
  WHERE w.longitude BETWEEN -119.80 AND -119.71 AND w.latitude BETWEEN 39.355 AND 39.415
    AND wl.water_level_elevation IS NOT NULL
  GROUP BY w.well_id ORDER BY n_obs DESC
")
```

Only **one** driller-report static water level exists project-wide at
all, and its `water_level_elevation` value is `NULL` (depth was
recorded, not converted to an elevation) -- so "point source from well
logs and drilling reports" is not currently a usable supplementary
data source either, though the pipeline (`Water_Level_Observations`,
`method = 'driller_report'`) is ready the moment a usable one exists.
Widening the search to a 10 km box around the field finds 21 wells
with genuinely dense, continuous coverage -- but nearly all of them
belong to the South Truckee Meadows GID regional monitoring network, a
different hydrogeologic setting than the geothermal field itself, not
a denser version of the same aquifer system.

**The concrete blocker, not a new one**: this is the same gap flagged
in Session 4's still-unanswered NDEP data request for water levels at
the actual CPI/SB GEO production/injection/monitor wells (43-33-1,
21-32-1/2, OW-1/2/3, the Strat wells) -- those are exactly the wells a
real field-scale potentiometric surface would need, and none of them
currently have any `Water_Level_Observations` on file.

**Software recommendation, for when that data exists:**

- **Primary: ArcGIS Pro Geostatistical Analyst (Empirical Bayesian
  Kriging).** This project's GIS workflow already targets ArcGIS, and
  the `chloride_points` GeoPackage layer (Session 30) was already built
  specifically for ArcGIS-side geostatistical interpolation -- EBK is
  the natural extension of that same workflow to water levels.
- **Secondary, reproducible cross-check: R (`gstat`/`sf`).** Consistent
  with this project's general "everything reproducible in code" bias;
  useful as an independent variogram/kriging check against whatever
  ArcGIS produces, not a replacement for it.
- **Free alternative: QGIS**, if ArcGIS licensing is ever a constraint
  -- ordinary/universal kriging is available via its SAGA/GRASS
  plugins, though not EBK specifically (an Esri-only method).

**Should it be informed by Leapfrog?** Not directly -- Leapfrog is a
3D geological modeling tool, not a 2D interpolation engine. But its
eventual 3D zonation (once real lithology like 83C-6ST1's exists for
more wells) *should* determine which wells get kriged together: only
wells screened in the same hydrostratigraphic unit belong in the same
2D surface. That grouping/masking step is where Leapfrog's output feeds
the potentiometric work, not the interpolation math itself -- and it's
the same principle the well-completion-interval section above already
applies informally (flagging long-open-interval wells as
likely-multi-zone) before any real zonation exists.

## Master statistics summary (whole notebook)

Every real statistical test run anywhere in this notebook, consolidated
in one place rather than left scattered across 1500+ lines:

| Section | Test | Statistic | p-value | n |
|---|---|---|---|---|
| Chloride trend, naive (all thermal-influenced) | Linear regression, Cl ~ year | -- | 0.0099 (weakens to 0.099 excluding 2 intermediate wells) | -- |
| Cl/B ratio, 1990-91 vs. 2024-2026 | Two-sample t-test | 19.6 vs. 21.9 | 0.015 | -- |
| Facies clustering | Ward hierarchical, k=4 (silhouette-selected); mclust cross-check | 70.1% agreement | -- | 157 |
| Pure-thermal facies Cl trend | Mixed-effects (well random intercept) | year coefficient n.s. | > 0.05 | -- |
| STMGID MW10 barometric efficiency | OLS, level & diff | BE=0.58 (R²=0.28) / 0.34 (R²=0.25) | both << 0.001 | 364/363 |
| STMGID MW3 barometric efficiency | OLS, level & diff | BE=0.60 (R²=0.12) / 0.29 (R²=0.38) | both << 0.001 | 364/363 |
| A009 (hot well) temp vs. pressure, daily | Pearson, level | r=+0.35 | 0.0045 | 63 |
| A009 temp vs. pressure, hourly, diff | Pearson | r=-0.17 | 1.0e-10 | ~1400 |
| A005 (fumarole) temp vs. pressure, hourly | Pearson, level & diff | r=+0.08 / +0.04 | 0.001 / 0.07 | ~1800 |
| A010 (spring) temp vs. pressure, daily | Pearson, level | r=+0.55 | <1e-11 | 134 |
| Nearby earthquakes vs. STMGID water level | Visual/step-change check | no detectable step | -- | 3 events |
| Nearby earthquakes vs. temperature loggers | Overlap check | no temporal overlap exists | -- | 0 events |
| Dec 2025 storm vs. MW10/MW3 water level | Visual comparison | MW10 +0.27 ft; MW3 none | -- | 1 event |
| Project-wide aquifer classification | BE >= 0.2 threshold (White 1968 range) | 7 wells "confined" | (see per-well p-values above) | 54 eligible |

Every number above is reproducible directly from this notebook's own
live code chunks against `geochem_operational.sqlite` -- none are
hand-transcribed from an earlier console session.

## Why this matters for the project

Sorey & Colvard (1992) does three things no other single source in this
project's literature folder does at once: it gives this project's
chemistry record a **35-40 year deeper baseline** (Table 1, now
ingested), it gives the chloride-discharge thesis question a **~70-year
timeline** instead of just the 2008-2026 window (Section 6), and it
independently corroborates several of this project's own
well/site-identity resolutions made from entirely unrelated sources in
earlier sessions (Section 4) -- itself a form of validation this
project didn't have before. It also demonstrates, with real 1988-era
field methods, that this project's own instrumentation program (SBRR/SBGG
conductivity loggers, temperature loggers, NDEP-resolved domestic wells)
sits on genuinely decades-old monitoring ground, not a newly invented
network. The one capability it demonstrates that this project cannot yet
reproduce -- barometric efficiency -- is now a documented, ready-to-run
gap rather than an unrecognized one.

:::: {.content-hidden when-meta="in_full_report"}
## Changelog

| Date | Change |
|---|---|
| 2026-09-25 (session 4) | Persisted hydrochemical facies clustering into `Facies_Cluster_Runs`/`Facies_Cluster_Assignments` + a new GeoPackage `facies_clusters` layer, for ArcGIS overlay against digitized fault traces. Added a project-wide aquifer-type classification (`Wells.aquifer_type`) derived from real barometric efficiency, grounded in White (1968)'s own confined-vent BE range at this site (0.2-1.18) -- 7 wells classified 'confined' project-wide, including both real Steamboat-field wells (STMGID MW10/MW3). Found and added the first REAL (non-assumed-vertical) well deviation survey, for 83C-6ST1, from Akerley et al. (2021)'s published drilling narrative -- confirmed zero real directional data exists in any of the 65 NDWR well-log OCR texts, so this had to come from literature, not a well log. Extended `export_leapfrog.R` to use real `Well_Deviation_Surveys`/`Well_Lithology` data automatically. Found (and correctly did NOT promote) a strong circumstantial lithology match for log 61248 to well 23-5; found and self-corrected an NDOT lead after rendering Klein (2007) Figure 1 directly. Added Klein (2007) Figure 1 and White (1968) Plate 1 as on-demand historic comparison figures. Added a consolidated PHREEQC-next-steps table, a potentiometric-surface-mapping feasibility/software-recommendation section (checked real coverage: only 2-3 wells in the field itself have usable water level, still blocked on Session 4's NDEP data request), and a master statistics-summary table covering the whole notebook.
| 2026-09-25 (session 3) | Data-folder README sweep (14 new/updated READMEs + top-level `data/README.md`); real USGS earthquake catalog ingested (`Earthquake_Events`) and cross-checked against water level/temperature -- no nearby event overlaps the temperature-logger window, no detectable water-level step at the 3 nearest (2025-05) events; real hourly temp-vs-pressure test finds a statistically significant inverse (boiling-point-consistent) coupling at A009 (r=-0.17, p<1e-10) but not at fumarole A005; STMGID MW10 vs. MW3 barometric-efficiency R² difference explained by screen length (480 ft vs. 100 ft); Dec 2025 storm shows a real MW10 recharge bump, absent at MW3; new well-completion-interval/confining-layer visualization (explicitly not lithology); PHREEQC's conceptual (gas-phase/pressure-equilibrium) tie-in discussed; `ingest_fault_traces.R` built and wired in (no-op until the user digitizes fault traces in ArcGIS).
| 2026-09-12 | Notebook created. Sorey & Colvard (1992) read directly (`pdftotext -layout`); Table 1 chemistry (9 samples, 1950-1991) ingested for the first time via `ingest_historical_sorey1992.R`; well/site cross-reference built against `Wells`/`Well_Aliases`/`Locations`; Cl/B ratio confirmed consistent between 1990-91 and 2024-2026 real thermal samples; discharge-through-time table assembled (White 1968 through this project's 2026 poster); barometric efficiency documented as a recommended, not-yet-runnable capability (no barometric-pressure parameter exists in this database yet). |
| 2026-09-12 (follow-up) | Resolved `83A-6` and `Cox 1-1` via exact NBMG Geothermal_Wells name matches; discovered and resolved that `"GS-58"`/`"GS-59"` are OCR artifacts of well `GS-5` (its own table footnote superscripts fused onto the name); all four now have real coordinates in `Locations`/`Wells`. Also found and fixed a 2-digit-year date-parsing bug in `ingest_well_logs.R`'s `.safe_completion_date()` that had produced one malformed `Water_Level_Observations` timestamp (`"7-07-10"`, corrected to `2007-07-10`). |
| 2026-09-12 (Tables 3/6/7/8/9, Appendix G) | Registered strat wells 2/5/6/7/8/9/13/14 (Table 8) and Table 3's South Truckee Meadows wells as real Wells rows; resolved Steinhardt and Brown School via exact NBMG matches, Bianco tentatively via NDWR owner-name + PLSS centroid; PTR-1/2 and STMGID MW-3/4 remain unresolved despite a fresh search; Herz-2 left inconclusive between two ambiguous NBMG candidates. Filled `Wells.top_perforation`/`bottom_perforation` (previously 100% NULL) for 7 wells from Tables 6/7, deliberately leaving 23-5 and IW-2 unfilled due to real depth conflicts with current records (documented as historical facts instead). Documented Appendix G's real methodology/survey dates (not parsed further -- too OCR-misaligned) and Table 9's reservoir-parameter values as reference only. |
| 2026-09-12 (session 29) | Added Collar & Huntley (1990) as a source (real fault map, Figure 1, used as a fault-digitizing candidate; second reservoir-parameter data point; discharge-deficit statistic; elevation cross-check, Cox I-1 elevation filled). Added a real statistical-analysis section: naive Cl-vs-year regression is confounded by which wells are grouped as 'thermal' (significant at p=0.0099 with intermediate wells included, only marginal at p=0.099 without them); a proper two-sample t-test on Cl/B ratios finds the 1950-1991 vs. 2024-2026 difference (19.6 vs. 21.9) is small but statistically significant (p=0.015), revising this notebook's earlier, less precise 'essentially unchanged' claim. Documented why the discharge-timeline chart has no confidence intervals (no source reports measurement uncertainty; fabricating one would be worse than omitting it). |
| 2026-09-25 | Reworked all three Cl/discharge plots per direct critique: Cl-timeline now shows well-name labels and lines connecting real repeat measurements (dropped the 6 single-snapshot 2026 FIELD samples from this specific plot to reduce crowding); discharge-through-time gained two new points (a Collar & Huntley 1988-89 natural-only estimate, and a new point computed directly from this project's own database -- the only real paired Cl+discharge measurement, 2026-05-01, giving 42.1 L/s vs. the poster's averaged 27 L/s); Cl/B ratio plot consolidated into one figure (year on the x-axis, well labels, t-test statistics moved into the figure caption instead of prose). Corrected a real data-quality error: Steinhardt's chloride range was misattributed from Table 3's OCR-misaligned row position -- the report's own narrative text (p.52) unambiguously ties a 300-to-140 mg/L decline since 1987 to Steinhardt, not the '16-22 mg/L' previously recorded; two real dated Cl samples (1987=300, ~1990=140 mg/L) added to Lab_Analyses. Added a paragraph on Sorey & Spielman (2017)'s own documented seasonal Cl-flux variability (~25% or less, summer-low/winter-high) as a real, previously-published effect relevant to (but not fully explaining) the gap between this session's new single-day estimate and the poster's averaged figure. |
| 2026-09-26 | Added a cluster-aware re-analysis: hydrochemical facies clustering (Ward hierarchical + Gaussian-mixture cross-check on 157 samples with complete major-ion chemistry, k=4 chosen by silhouette width) recovers a chemically interpretable Ca-Na-Cl 'diluted/water-rock-reacted' facies matching Mariner & Janik (1995)'s own Brown School albitization narrative, and independently groups the dilute-but-thermal-ratio 'Stuart' well with the concentrated thermal facies, matching that paper's text. A mixed-effects model (well as random intercept) on the pure-thermal facies finds no significant Cl trend over time (p > 0.05) once between-well baseline differences are properly separated from a common temporal effect -- a more defensible null result than the earlier naive/pure-thermal regression. Also ingested Mariner & Janik (1995) Table 1's raw chemistry (75 samples, 1991-1994, read from rendered page images since pdftotext -layout scrambles this table) via new `ingest_mariner_janik_1995.R`, densifying the pre-2000 Cl/B/Li record for 23-5, 83A-6, 21-5R, 13-5R, Cox I-1, GS-5, PW-1/2/3, the PW2-x/PW3-x Far West field, IW-3/IW-5, and several domestic/background comparison sites (several tentatively or not matched to existing Wells/Locations rows -- see the CSV's match_type column). |
| 2026-09-26 (continued) | Added a spatial facies-cluster map (`scripts/analysis/facies_depth_map.R`) overlaying each well's majority facies cluster onto its real map position, sized by an approximate depth (screened-interval midpoint > total depth > unknown -- only 5 of 39 mapped wells have any depth on record, a real gap flagged for the planned Leapfrog work). Faceted the Cl-concentration timeline by well vs. spring/seep. Ingested a new air-temperature time series from the SBRR USGS gauge (parameter 00020, reused `ingest_usgs()` with a new `base_dir` argument -- no new ingest logic needed) and cross-checked it against every Temperature_Loggers series (`scripts/qc/qc_temperature_vs_air.R`) as a QC signal (diurnal-range-ratio vs. air), NOT a literal temperature correction -- flagged loggers A003 (SBW_0001, 95% of days air-like) and A005 (SBF_0001, 70%) as showing suspiciously air-tracking diurnal swings, worth a field check for exposure/dry-sensor issues. |
| 2026-09-25 | Barometric efficiency upgraded from a documented-but-unrunnable stub to a real calculation: ingested hourly Reno Airport (ASOS "RNO") mean-sea-level pressure from the Iowa Environmental Mesonet (`scripts/ingest/ingest_barometric_pressure.R`, `data/raw/airpressure/README.md`), reusing the existing `Weather_Observations`/`Weather_Stations` tables and station_id (`USW00023185`) rather than a new schema. New `scripts/analysis/barometric_efficiency.R` (`run_barometric_efficiency()`) finds every well with continuous water-level data overlapping the pressure record, computes BE two ways (Sorey's original level-vs-pressure regression, and a first-differenced change-vs-change regression that removes secular trend), and prefers a Steamboat-area well for the overlay plot over the wider Reno/South Truckee Meadows monitoring network's longer records. Real result: STMGID MW10/MW3 (the only Steamboat-area wells with both continuous 2025 water level and pressure overlap) give BE = 0.29-0.60 across both methods -- inside White (1968)'s 0.2-1.18 range and the same order of magnitude as Sorey's own 0.42/0.45, a real methodological validation, though explicitly not a like-for-like comparison since neither well is an actual thermal spring.

::::
```{r cleanup-07}
#| include: false
dbDisconnect(con)
```
