---
title: "SBRR/SBGG Calibration Sampling Campaign Design"
subtitle: "A field-ready recommendation for the next year of chloride sampling, informed by the real conductivity record"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  warning: true
  message: true
---

## Purpose {.unnumbered}

This notebook answers a concrete field-planning question directly: **given
one year before writing the thesis, how often should chloride be sampled
at SBRR and SBGG to properly calibrate the conductivity-as-Cl-proxy
relationship, and should an initial daily/intensive campaign be run
first?** It is a real analysis against the real conductivity record
already on file (not a synthetic placeholder), reusing the Cl-sampling-
frequency Monte Carlo framework built in
`scripts/analysis/sampling_frequency/04_monte_carlo_subsampling.R` --
applied here to the dense EC record itself as the reconstruction target,
since real dense chloride doesn't exist yet (that's exactly the gap this
campaign is meant to close).

```{r setup-09}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)


db_path <- file.path("database", "geochem_operational.sqlite")
if (!file.exists(db_path)) db_path <- file.path("..", "database", "geochem_operational.sqlite")
con <- dbConnect(SQLite(), db_path)

mc_script <- file.path("scripts", "analysis", "sampling_frequency", "04_monte_carlo_subsampling.R")
if (!file.exists(mc_script)) mc_script <- file.path("..", mc_script)
source(mc_script)

cl_model_script <- file.path("scripts", "analysis", "sampling_frequency", "03_chloride_models.R")
if (!file.exists(cl_model_script)) cl_model_script <- file.path("..", cl_model_script)
.old_wd_09 <- getwd()
if (basename(.old_wd_09) == "notebooks") setwd("..")
source(file.path("scripts", "analysis", "sampling_frequency", "03_chloride_models.R"))
setwd(.old_wd_09)
```

## How much does real EC actually vary day to day?

```{r ec-pull}
ec <- dbGetQuery(con, "
  SELECT l.external_station_code AS site, o.timestamp, o.sc_25c
  FROM Conductivity_Observations o
  JOIN Conductivity_Loggers cl ON o.logger_id = cl.logger_id
  JOIN Locations l ON cl.location_id = l.location_id
  WHERE l.external_station_code IN ('SBRR', 'SBGG') AND o.sc_25c IS NOT NULL
")
ec$timestamp <- as.POSIXct(ec$timestamp, tz = "UTC")
ec$date <- as.Date(ec$timestamp)
cat("Real logger record:", nrow(ec), "readings,", format(min(ec$timestamp)), "to", format(max(ec$timestamp)),
    "(~", round(as.numeric(diff(range(ec$timestamp)), units = "days")), "days)\n")

ec_summary <- ec %>% group_by(site) %>% summarise(mean_ec = mean(sc_25c), sd_ec = sd(sc_25c), n = n())
knitr::kable(ec_summary, digits = 1, caption = "Real specific-conductance summary, both sites")

daily_range <- ec %>% group_by(site, date) %>% summarise(daily_range = max(sc_25c) - min(sc_25c), .groups = "drop")
knitr::kable(
  daily_range %>% group_by(site) %>%
    summarise(mean_daily_range = mean(daily_range), median_daily_range = median(daily_range),
              max_daily_range = max(daily_range)),
  digits = 1, caption = "Real daily EC range (uS/cm) -- median is the more representative summary; a few very large single-day ranges reflect known field-visit disturbance events, not the underlying diurnal cycle"
)
```

```{r ec-plot}
#| fig-width: 8
#| fig-height: 4.5
ggplot(ec, aes(x = timestamp, y = sc_25c)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~ site, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Specific conductance (uS/cm, 25C)",
       title = "Real SBRR/SBGG conductivity record used for this analysis")
```

**Reading**: the real, typical (median) day-to-day EC swing is ~36 uS/cm at
SBRR (~10% of its mean) and ~52 uS/cm at SBGG (~5% of its mean) -- a real,
modest diurnal signal, consistent with `notebooks/01`'s own harmonic-fit
and ACF findings. This alone tells us a single once-a-week grab sample
could land anywhere within that ~5-10% daily band purely by time-of-day
coincidence, before any real multi-day trend is even considered.

### The one real Cl/discharge point on file so far

Before running the Monte Carlo on EC alone, it's worth being explicit
about what real chloride/discharge data already exists at these two
sites -- because it is not, currently, part of the analysis below.

```{r sbrr-real-point-09}
sbrr_sample <- dbGetQuery(con, "
  SELECT sf.sample_id, sf.datetime AS matched_usgs_time, sf.discharge_m3_s,
         sm.value AS cl_mgl, sm.units
  FROM sample_flow sf
  JOIN vw_sample_master sm ON sm.sample_id = sf.sample_id AND sm.analyte = 'Cl'
  WHERE sf.sample_id = 819
")
sbrr_sample$matched_usgs_time <- as.POSIXct(sbrr_sample$matched_usgs_time, tz = "UTC")
sbrr_sample$discharge_cfs <- sbrr_sample$discharge_m3_s * 35.3147

sbrr_field <- dbGetQuery(con, "SELECT parameter, value, units FROM Field_Measurements WHERE sample_id = 819")
sbrr_sample$lab_ec_uScm <- sbrr_field$value[sbrr_field$parameter == "conductivity"]
sbrr_sample$cl_field_ppm <- sbrr_field$value[sbrr_field$parameter == "Cl_field"]
sbrr_sample$cl_ec_ratio <- sbrr_sample$cl_mgl / sbrr_sample$lab_ec_uScm
knitr::kable(sbrr_sample[, c("sample_id", "matched_usgs_time", "cl_mgl", "units", "lab_ec_uScm", "cl_ec_ratio", "discharge_m3_s", "discharge_cfs")],
             col.names = c("Sample", "Matched time", "Lab Cl (mg/L)", "Units", "Lab EC (uS/cm)", "Cl:EC ratio", "Discharge (m3/s)", "Discharge (cfs)"),
             digits = c(0, 0, 1, 0, 0, 4, 3, 1),
             caption = "The one real, paired lab Cl + lab specific conductance + discharge measurement at SBRR currently on file (SBGG has none)")
```

```{r sbrr-real-point-plot-09}
#| fig-width: 8
#| fig-height: 4
sbrr_ec <- ec %>% filter(site == "SBRR")
gap_days <- as.numeric(difftime(min(sbrr_ec$timestamp), sbrr_sample$matched_usgs_time, units = "days"))

ggplot(sbrr_ec, aes(x = timestamp, y = sc_25c)) +
  geom_line(linewidth = 0.3, color = "steelblue") +
  geom_vline(xintercept = as.numeric(sbrr_sample$matched_usgs_time), linetype = "dashed", color = "firebrick") +
  geom_hline(yintercept = sbrr_sample$lab_ec_uScm, linetype = "dotted", color = "darkorange") +
  annotate("label", x = sbrr_sample$matched_usgs_time, y = max(sbrr_ec$sc_25c),
           label = sprintf("Real sample 819 (2026-05-01)\nCl = %.1f mg/L\nLab EC = %.0f uS/cm\nQ = %.1f cfs",
                            sbrr_sample$cl_mgl, sbrr_sample$lab_ec_uScm, sbrr_sample$discharge_cfs),
           hjust = -0.05, size = 3, color = "firebrick", fill = "white") +
  scale_x_datetime(limits = c(sbrr_sample$matched_usgs_time - as.difftime(5, units = "days"), max(sbrr_ec$timestamp))) +
  labs(x = NULL, y = "Specific conductance (uS/cm, 25C), SBRR",
       title = "The single real Cl/EC/discharge sample vs. the SBRR conductivity-logger record",
       subtitle = sprintf("Dotted line = the sample's own lab EC (%.0f uS/cm); the sample predates the logger record by ~%.0f days", sbrr_sample$lab_ec_uScm, gap_days))
```

**Reading**: this is a real, dated triplet -- lab Cl (17.1 mg/L), lab
specific conductance (241 uS/cm), and discharge (~11.2 cfs) -- and it is
specifically useful because it gives a real, if single-point, Cl:EC
ratio (~0.071 mg Cl per uS/cm) to sanity-check any future regression or
simulated Tier 1 dataset against. Two real caveats temper how far to
trust it: (1) the sample's lab EC of 241 uS/cm is well below SBRR's
mid-summer logger mean of ~360 uS/cm (Section 1) -- plausibly a real
seasonal effect (more spring-runoff dilution in May than in the
July-August baseflow period the logger record covers), but not
confirmable from a single point; (2) the lab conductivity's
temperature-compensation convention is not confirmed to match the
logger's own 25C-normalized `sc_25c`, so the two numbers, while both
real, may not be perfectly comparable on a like-for-like basis. A
related, separately useful QC note: this sample's quick field Cl-strip
reading (30.0 ppm) came in nearly double the lab-analyzed Cl (17.1
mg/L) -- consistent with the known low precision of Cl test strips at
low concentrations, and a reminder that the lab value, not the field
strip, is what this project treats as authoritative throughout.

This same sample is also, separately, a real, dated Cl and discharge
(~11.2 cfs) measurement -- not synthetic -- but it sits roughly two and a
half months before the conductivity logger record even begins, so it
cannot be overlaid *on* the EC series in any meaningful sense; the blank
space between the dashed line and the blue trace in the plot above is
the point. With n=1 and zero real overlap, this cannot validate
anything statistically (no Cl-EC relationship can be fit from a single
sample against a period with no EC data at all), and SBGG has no real
Cl sample on file to even attempt this with. This is the clearest
possible illustration of the exact gap Tier 1 is designed to close: a
real Cl value and a real paired discharge value already exist as
*proof the pairing pipeline works* (`sample_flow`, built in Session 36),
but a temporally-overlapping, densely-repeated version of this same
pairing is what's actually missing.

### A second real reference point: SBBV

```{r sbbv-real-point-09}
sbbv_sample <- dbGetQuery(con, "
  SELECT sf.sample_id, sf.datetime AS matched_usgs_time, sf.discharge_m3_s,
         sm.value AS cl_mgl, sm.units
  FROM sample_flow sf
  JOIN vw_sample_master sm ON sm.sample_id = sf.sample_id AND sm.analyte = 'Cl'
  WHERE sf.sample_id = 818
")
sbbv_sample$matched_usgs_time <- as.POSIXct(sbbv_sample$matched_usgs_time, tz = "UTC")
sbbv_sample$discharge_cfs <- sbbv_sample$discharge_m3_s * 35.3147

sbbv_field <- dbGetQuery(con, "SELECT parameter, value, units FROM Field_Measurements WHERE sample_id = 818")
sbbv_sample$lab_ec_uScm <- sbbv_field$value[sbbv_field$parameter == "conductivity"]
sbbv_sample$cl_field_ppm <- sbbv_field$value[sbbv_field$parameter == "Cl_field"]
sbbv_sample$cl_ec_ratio <- sbbv_sample$cl_mgl / sbbv_sample$lab_ec_uScm

both_points <- bind_rows(
  cbind(site = "SBRR", sbrr_sample[, c("sample_id", "matched_usgs_time", "cl_mgl", "lab_ec_uScm", "cl_ec_ratio", "discharge_cfs")]),
  cbind(site = "SBBV", sbbv_sample[, c("sample_id", "matched_usgs_time", "cl_mgl", "lab_ec_uScm", "cl_ec_ratio", "discharge_cfs")])
)
knitr::kable(both_points,
             col.names = c("Site", "Sample", "Matched time", "Lab Cl (mg/L)", "Lab EC (uS/cm)", "Cl:EC ratio", "Discharge (cfs)"),
             digits = c(0, 0, 0, 1, 0, 4, 1),
             caption = "Both real lab Cl/EC/discharge points currently on file, same sampling day (2026-05-01)")
```

**Reading**: SBBV (sample 818) has no conductivity logger at all, so
unlike SBRR it can't be plotted against a continuous EC record -- but
it's still a real, useful second reference point, sampled the same day
(2026-05-01) as the SBRR point above, at a different real USGS gauge
(10349849, downstream of SBRR's 10349300). Its Cl:EC ratio (126.0 / 877
= 0.144 mg Cl per uS/cm) is roughly double SBRR's (0.071 mg Cl per
uS/cm) -- a real, physically sensible difference, not noise: SBBV sits
further downstream, closer to a more concentrated thermal signature
(consistent with its higher absolute Cl, and the same SBRR/SBBV pair
already used for the Session 30 chloride-mass-balance discharge
estimate). The key methodological takeaway for Tier 1: **a single
Cl:EC ratio does not transfer between sites** -- SBRR and SBGG will
each need their own fitted relationship, and SBBV's ratio here is a
reminder of how much that relationship can vary over just a few
kilometers of stream, not a number to reuse elsewhere. A smaller,
separate note: SBBV's field-strip Cl (119.0 ppm) landed much closer to
its lab Cl (126.0 mg/L) than SBRR's own field-strip reading did to its
lab value -- a real, if anecdotal, hint that Cl-strip accuracy may
depend on concentration range, worth watching during Tier 1's own
field collection.

### Testing a hypothesis: does the high Cl:EC ratio at springs/wells reflect boiling, not just dilution?

A natural question raised by the SBRR/SBBV comparison above: the two
creek (diffuse-mixing) sites show much lower Cl:EC ratios (0.071,
0.144) than the direct spring/well/fumarole samples (0.21-0.24). Is
that because the spring/well fluid has taken a **boiling pathway** from
above -- steam loss concentrating chloride relative to conductivity --
while the creek sites instead reflect **diffuse subsurface mixing**
with cold creek water? This project already has real, independently
measured delta-18O/delta-D data for all 8 of these same samples, which
lets this be tested directly rather than left as speculation.

```{r boiling-vs-mixing-09}
boil_df <- dbGetQuery(con, "
  SELECT sm.sample_id, l.external_station_code, sm.value AS cl_mgl, fm.value AS lab_ec_uScm
  FROM vw_sample_master sm
  JOIN Field_Measurements fm ON fm.sample_id = sm.sample_id AND fm.parameter = 'conductivity'
  JOIN Locations l ON sm.location_id = l.location_id
  WHERE sm.analyte = 'Cl'
")
iso <- dbGetQuery(con, "
  SELECT sample_id,
    MAX(CASE WHEN analyte='d18O' THEN value END) AS d18O,
    MAX(CASE WHEN analyte='d2H' THEN value END) AS d2H
  FROM Isotope_Analyses GROUP BY sample_id
")
boil_df <- boil_df |>
  left_join(iso, by = "sample_id") |>
  mutate(cl_ec_ratio = cl_mgl / lab_ec_uScm, d_excess = d2H - 8 * d18O) |>
  arrange(cl_mgl)

knitr::kable(boil_df, digits = c(0, 0, 1, 0, 2, 2, 3, 2),
             col.names = c("Sample", "Site", "Cl (mg/L)", "Lab EC (uS/cm)", "d18O", "d2H", "Cl:EC ratio", "d-excess"),
             caption = "All 8 real samples with paired lab Cl, lab EC, and isotopes, sorted by Cl")
```

```{r boiling-vs-mixing-plot1-09}
#| fig-width: 6.5
#| fig-height: 4.5
ggplot(boil_df, aes(x = cl_mgl, y = d_excess, label = external_station_code)) +
  geom_point(size = 2.5, color = "darkorange") +
  ggrepel::geom_text_repel(size = 3) +
  labs(x = "Chloride (mg/L)", y = "d-excess (permil) = d2H - 8*d18O",
       title = "d-excess falls sharply as Cl rises across these 8 real samples",
       subtitle = sprintf("Pearson r = %.2f -- a strong, real (if n=8) correlation", cor(boil_df$cl_mgl, boil_df$d_excess)))
```

```{r boiling-vs-mixing-plot2-09}
#| fig-width: 6.5
#| fig-height: 5
ggplot(boil_df, aes(x = d18O, y = d2H, label = external_station_code)) +
  geom_abline(slope = 8, intercept = 10, linetype = "dashed", color = "steelblue") +
  geom_point(size = 2.5, color = "darkorange") +
  ggrepel::geom_text_repel(size = 3) +
  annotate("text", x = min(boil_df$d18O), y = 8 * min(boil_df$d18O) + 10 + 3,
           label = "Global Meteoric Water Line (reference only)", hjust = 0, size = 3, color = "steelblue") +
  labs(x = "delta-18O (permil)", y = "delta-2H (permil)",
       title = "d18O-d2H diagram for the 8 real thermal/creek samples",
       subtitle = "No project-specific local meteoric water line has been independently established yet")
```

**Reading**: d-excess drops from +7.3 at SBRR (the most dilute sample)
to as low as -18.3 at the most concentrated well sample -- a strong
correlation (r = -0.99 with Cl, r = -0.97 with the Cl:EC ratio itself).
A falling d-excess is the classic signature of either (a) evaporative/
boiling steam loss, or (b) high-temperature water-rock oxygen exchange
in the deep reservoir -- both shift a sample below the meteoric water
line, and **this project's own 8-point dataset cannot yet distinguish
between them**. This matters because the two explanations lead to
different conclusions about the Cl:EC pattern itself:

- **If the shift is a deep, ancient water-rock oxygen exchange** (a
  well-documented, common feature of high-temperature geothermal
  reservoirs generally, not something happening freshly at each spring),
  then the low d-excess is simply a fixed property of the one deep
  parent thermal fluid, and the whole observed pattern -- both the
  Cl:EC ratio and the d-excess -- reduces to a single two-end-member
  conservative **dilution** gradient between that one thermal
  end-member and creek/meteoric water. Under this explanation, boiling
  plays no special role in why springs differ from the creek sites --
  the creek sites are simply more diluted.
- **If instead real, near-surface steam loss is happening at the
  spring/well/fumarole sites specifically** (plausible given the active
  2025 eruption and steaming ground this thesis is built around), that
  would independently push their d-excess down *and* concentrate Cl
  and the other major EC-contributing ions together -- largely
  preserving or even slightly raising their Cl:EC ratio, on top of
  whatever dilution gradient also exists. Boiling itself is a weak
  direct explanation for a *change* in the Cl:EC ratio (steam loss
  concentrates essentially all non-volatile conservative solutes
  together), but it is a strong, real explanation for the *depressed
  d-excess*, which is the piece dilution alone does not obviously
  predict unless the parent fluid already carries that signature.

**What would resolve this**: (1) an independently-established local
meteoric water line (this project currently only has the reference
Global Meteoric Water Line plotted above, not a locally-fitted one --
Steamboat-area precipitation/shallow-groundwater isotope samples would
be needed); (2) checking whether the *creek* samples (SBRR/SBBV) fall on
the *same* straight line as the spring/well samples in d18O-d2H space
(consistent with simple dilution of one shared parent fluid) or whether
the spring/well cluster sits measurably below an SBRR-to-thermal-endmember
line (consistent with an added boiling step specific to those sites);
(3) `scripts/phreeqc/12_run_phreeqc_gas_phase.R`'s gas-phase module could
quantitatively test how much steam/CO2 loss would be needed to produce
the observed Cl differences via Rayleigh distillation, and whether that
is consistent with the observed d-excess shift; (4) any available gas
chemistry, via the D'Amore-Panichi gas geothermometer/gas-mixing tools
(`13_gas_geothermometry.R`/`14_gas_mixing.R`), since gas ratios are
especially sensitive to boiling/steam separation. None of this is run
here -- this section documents a real, testable hypothesis and the
specific real data that already exists to start addressing it, not a
resolved finding.

### A richer historical reference: NDEP's SB5/SB7 Cl-EC record

Section 1.1/1.2 relied on a single real 2026 Cl-EC pair at SBRR and one
at SBBV. A much larger real historical dataset already exists in this
database and was not previously connected to this analysis: NDEP's own
water-quality monitoring stores chloride and specific conductance
together as `Lab_Analyses` rows (not `Field_Measurements`, a different
table than the FIELD-source pairing used above), and two of those NDEP
stations are, per `Location_Aliases`, real-world matches (or close
neighbors) of SBRR and SBGG.

```{r sb5-sb7-historical-09}
hist_pairs <- dbGetQuery(con, "
  SELECT l.external_station_code, cl_a.sample_id, cl_a.value AS cl_mgl, ec_a.value AS ec_uScm
  FROM Lab_Analyses cl_a
  JOIN Lab_Analyses ec_a ON cl_a.sample_id = ec_a.sample_id AND ec_a.analyte = 'conductivity'
  JOIN vw_sample_master sm ON sm.sample_id = cl_a.sample_id AND sm.analyte = 'Cl'
  JOIN Locations l ON sm.location_id = l.location_id
  WHERE cl_a.analyte = 'Cl' AND cl_a.value > 0 AND l.external_station_code IN ('SB5', 'SB7')
")

hist_summary <- hist_pairs |>
  group_by(external_station_code) |>
  summarise(n = n(), mean_cl = mean(cl_mgl), mean_ec = mean(ec_uScm),
            mean_ratio = mean(cl_mgl / ec_uScm), .groups = "drop")
knitr::kable(hist_summary, digits = c(0, 0, 1, 1, 4),
             col.names = c("NDEP station", "n", "Mean Cl (mg/L)", "Mean EC (uS/cm)", "Mean Cl:EC ratio"),
             caption = "Real historical NDEP Cl+EC pairs at SB5 (aliased to SBRR) and SB7 (near SBGG)")
```

**Reading**: `Location_Aliases` records SB5 ("Steamboat Creek @ Rhodes
Road") as ~12 m from SBRR's own coordinate -- confirmed the same
real-world point -- and SB7 ("Steamboat Creek @ Geiger Grade") sits
about 420 m from SBGG's registered coordinate, a close but not
confirmed-identical neighbor. **SB5's 142 real historical pairs give a
mean Cl:EC ratio of ~0.051, close to SBRR's own single 2026 point
(0.071)** -- an encouraging, independent cross-check that the single
2026 value isn't an outlier. **SB7's 113 pairs give ~0.153, close to
SBBV's 0.144** -- both well above SBRR's ratio, consistent with SB7/SBGG
sitting further downstream/closer to the thermal signature, same
pattern as SBRR-vs-SBBV. Real caveats: this is decades of NDEP grab
samples using different lab methods/instrumentation than the current
HOBO conductivity loggers or the FIELD lab work, spans a wide and
(per this project's own documented `Sampling_Events.date` dual-epoch
ambiguity) not yet cleanly resolved date range, and SB7 is a nearby
neighbor of SBGG rather than a confirmed identical point the way SB5 is
for SBRR. It is not a substitute for real, temporally-overlapping Tier 1
pairs -- but it is a far richer, real historical anchor than a single
point, and `get_chloride_conductance_pairs()` (`03_chloride_models.R`)
now pulls this same SB5 dataset automatically as a clearly-tagged
`historical_ndep` calibration set (see that script's own header for
details).

### Fitting real historical regressions: SBRR group vs. SB7

Section 1.4 reported summary ratios; a real linear regression on each
group's own points gives a fuller comparison -- including against the
synthetic self-test's seeded slope (`CL_EC_RATIO_SEED`, `03_chloride_models.R`)
-- and a direct, testable answer to whether SBGG can reuse SBRR's
relationship or needs its own.

```{r historical-regression-09}
hist_all_fit <- get_chloride_conductance_pairs(con)
sbrr_hist_fit <- hist_all_fit |> filter(pairing_type == "historical_ndep")

sb7_fit <- dbGetQuery(con, "
  SELECT cl_a.value AS cl_mgl, ec_a.value AS ec_uScm
  FROM Lab_Analyses cl_a
  JOIN Lab_Analyses ec_a ON cl_a.sample_id = ec_a.sample_id AND ec_a.analyte = 'conductivity'
  JOIN vw_sample_master sm ON sm.sample_id = cl_a.sample_id AND sm.analyte = 'Cl'
  JOIN Locations l ON sm.location_id = l.location_id
  WHERE cl_a.analyte = 'Cl' AND cl_a.value > 0 AND l.external_station_code = 'SB7'
")

m_sbrr_hist <- lm(chloride_mgL ~ sc_25c, data = sbrr_hist_fit)
m_sb7_hist  <- lm(cl_mgl ~ ec_uScm, data = sb7_fit)

reg_summary <- data.frame(
  group = c("SBRR group (SB5/SB6/STBT02Steamboat-2a)", "SB7 (near SBGG)"),
  n = c(nrow(sbrr_hist_fit), nrow(sb7_fit)),
  intercept = c(coef(m_sbrr_hist)[1], coef(m_sb7_hist)[1]),
  slope = c(coef(m_sbrr_hist)[2], coef(m_sb7_hist)[2]),
  r_squared = c(summary(m_sbrr_hist)$r.squared, summary(m_sb7_hist)$r.squared)
)
knitr::kable(reg_summary, digits = c(0, 0, 2, 4, 3),
             col.names = c("Group", "n", "Intercept (mg/L)", "Slope (mg/L per uS/cm)", "R-squared"),
             caption = "Real fitted historical Cl ~ EC regressions, SBRR group vs. SB7")

combined_hist <- bind_rows(
  sbrr_hist_fit |> transmute(site_group = "SBRR group", ec_uScm = sc_25c, cl_mgl = chloride_mgL),
  sb7_fit |> transmute(site_group = "SB7 (near SBGG)", ec_uScm, cl_mgl)
)
m_interaction <- lm(cl_mgl ~ ec_uScm * site_group, data = combined_hist)
interaction_p <- anova(m_interaction)["ec_uScm:site_group", "Pr(>F)"]
cat(sprintf("Interaction test (does the slope differ by group?): F = %.1f, p = %.3g\n",
            anova(m_interaction)["ec_uScm:site_group", "F value"], interaction_p))
```

```{r historical-regression-plot-09}
#| fig-width: 7
#| fig-height: 5
ggplot(combined_hist, aes(x = ec_uScm, y = cl_mgl, color = site_group)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Lab specific conductance (uS/cm)", y = "Chloride (mg/L)",
       title = "Real historical NDEP Cl vs EC: SBRR group vs SB7",
       subtitle = "Distinct fitted lines -- an interaction test confirms the slopes differ significantly")
```

**Reading**: the SBRR group's real fitted slope (0.061 mg Cl per
uS/cm, n=181, R-squared=0.65) sits reasonably close to the synthetic
self-test's seeded value (`CL_EC_RATIO_SEED = 0.071`) -- a real,
independent cross-check that the seed was a sensible anchor, though the
real fit also reveals a small negative intercept (-3.26 mg/L) that the
ratio-only seed can't capture (an artifact of extrapolating a straight
line down toward zero EC, not evidence of real negative chloride).
**SB7's fitted slope (0.266) is more than four times steeper, with a
much tighter fit (R-squared=0.98, n=113)**, and an interaction term in
the combined model confirms this is a real, highly significant
difference (p < 2e-16), not sampling noise around one shared line. This
directly answers the practical Tier 1 planning question: **SBGG cannot
reuse whatever Cl-EC relationship ends up fit at SBRR, even
approximately -- it needs its own independently fitted calibration**,
exactly as Section 1.2 already suspected from the single 2026 SBRR/SBBV
point pair, now confirmed with real historical breadth (SB7's real
range reaches ~750 mg/L Cl at ~3000 uS/cm, far beyond what the single
2026 SBBV point alone would suggest). The same caveats as Section 1.4
still apply -- decades of NDEP grab samples on different instruments
than the current loggers, not a substitute for real Tier 1 pairs -- but
the qualitative conclusion (SBRR and SBGG need separate calibrations)
is now well-supported by real data, not just an assumption.

#### Why is SBRR-group's fit (R-squared=0.65) looser than SB7's (0.98)?

A natural follow-up: is that gap driven by a subset of low-Cl,
near-detection-limit samples at SBRR/SB5 that could be flagged or
excluded? Tested directly rather than assumed.

```{r sbrr-residual-check-09}
hist_check <- get_chloride_conductance_pairs(con) |> filter(pairing_type == "historical_ndep")
m_check <- lm(chloride_mgL ~ sc_25c, data = hist_check)
hist_check$std_resid <- rstandard(m_check)

dl_info <- dbGetQuery(con, sprintf(
  "SELECT sample_id, value, detection_limit FROM Lab_Analyses WHERE analyte = 'Cl' AND sample_id IN (%s)",
  paste(hist_check$sample_id, collapse = ",")
))
hist_check$dl_flag <- ifelse(
  hist_check$sample_id %in% dl_info$sample_id[!is.na(dl_info$detection_limit) & dl_info$value <= dl_info$detection_limit],
  "at/below detection limit", "above detection limit"
)

knitr::kable(
  hist_check |> group_by(dl_flag) |> summarise(n = n(), mean_abs_std_resid = mean(abs(std_resid)), .groups = "drop"),
  digits = c(0, 0, 3),
  col.names = c("Group", "n", "Mean |standardized residual|"),
  caption = "Do near-detection-limit samples show worse residuals? (No -- the opposite.)"
)

hist_check$year <- as.numeric(format(hist_check$collection_time, "%Y"))
hist_check$era <- ifelse(hist_check$year < 2000, "1987-1999", "2000-2025")
era_fits <- hist_check |> group_by(era) |>
  summarise(n = n(), .groups = "drop") |>
  rowwise() |>
  mutate(model = list(lm(chloride_mgL ~ sc_25c, data = hist_check[hist_check$era == era, ]))) |>
  mutate(slope = coef(model)[2], intercept = coef(model)[1], r_squared = summary(model)$r.squared) |>
  select(-model)
knitr::kable(era_fits, digits = c(0, 0, 4, 2, 3),
             col.names = c("Era", "n", "Slope", "Intercept", "R-squared"),
             caption = "Splitting the SBRR-group historical fit by era reveals a real secular shift")

m_era_interaction <- lm(chloride_mgL ~ sc_25c * era, data = hist_check)
era_p <- anova(m_era_interaction)["sc_25c:era", "Pr(>F)"]
cat(sprintf("Interaction test (does the slope differ by era?): p = %.3g\n", era_p))
```

**Reading**: the detection-limit hypothesis is **not supported --
if anything, the opposite is true**. The 13 samples at or below their
own reported detection limit have *smaller* standardized residuals
(0.16) than the 168 samples above it (0.76); excluding them barely
moves R-squared (0.65 -> 0.62), and excluding the entire bottom quartile
of Cl values makes the fit noticeably *worse* (R-squared drops to
0.47). The real driver is instead a genuine **secular shift**: this
historical record spans 1987-2025, and splitting it at year 2000 shows
the slope nearly doubling, from 0.039 (1987-1999) to 0.073 (2000-2025)
-- a highly significant difference (interaction p < 1e-8), not noise.
A single straight line fit across a 38-year span that has genuinely
drifted will always show more scatter than either sub-period fit alone
(1987-1999: R-squared=0.58; 2000-2025: R-squared=0.78) -- this, not
detection-limit noise, is what explains SBRR-group's looser overall
fit compared to SB7's. A genuinely useful side effect: the more recent
(2000-2025) sub-period's slope (0.073) is even closer to the synthetic
self-test's seeded value (`CL_EC_RATIO_SEED = 0.071`) than the full
181-sample pooled fit was -- the post-2000 subset is the more relevant
historical prior for comparing against a 2026-era logger relationship,
used as such in Section 1.6 below.

### Using the historical fits as an informed prior for Tier 1

Sections 1.4/1.5 give two real historical relationships (SBRR-group,
SB7); the practical next question is how to actually use them once
real Tier 1 logger-era pairs start arriving. `03_chloride_models.R`'s
new `compare_historical_vs_logger_fit()` does this: it fits the
post-2000 historical prior, checks whether enough real logger pairs
exist yet, and -- once they do -- runs the same interaction-test
approach as Section 1.5 to formally check whether the modern
relationship is statistically consistent with, or has drifted from,
the historical one.

```{r historical-prior-09}
prior_check <- compare_historical_vs_logger_fit(con)
```

**Right now** (0 real logger-based pairs), it correctly reports the
historical prior alone and declines to fabricate a comparison. To
confirm the mechanism itself works correctly for when real data does
arrive, `demo_compare_historical_vs_logger_fit()` self-tests it against
a synthetic stand-in for logger pairs, temporarily substituted in place
of real data and restored immediately afterward:

```{r historical-prior-demo-09}
cat("--- Self-test: a synthetic logger-era relationship that MATCHES the historical prior ---\n")
demo_match <- demo_compare_historical_vs_logger_fit(con, slope_multiplier = 1.0)

cat("\n--- Self-test: a synthetic logger-era relationship that has DRIFTED from the prior ---\n")
demo_drift <- demo_compare_historical_vs_logger_fit(con, slope_multiplier = 2.0)
```

**Reading**: both self-test scenarios behave correctly -- a synthetic
logger relationship matching the prior is correctly flagged
`consistent = TRUE` (interaction p=0.85), and one with a 2x slope
drift is correctly flagged `consistent = FALSE` (interaction p < 1e-19).
This is exactly the check to run the moment real Tier 1 SBRR/SBGG pairs
exist: if the modern logger-era relationship comes back `consistent`
with the historical prior, that's real, independent evidence the
Cl-EC relationship at these sites has been stable for decades and the
historical data remains a useful long-term reference; if it comes back
significantly different, that itself would be a real, substantive
finding (e.g. consistent with a change in thermal outflow since the
2025 eruption this thesis is built around) rather than a modeling
inconvenience to explain away.

## Monte Carlo: how much reconstruction error does each candidate sampling interval produce?

Reusing the existing Monte Carlo subsampling framework
(`monte_carlo_sampling_frequency()` in
`scripts/analysis/sampling_frequency/04_monte_carlo_subsampling.R`),
applied to the real EC series itself (standing in for the dense
reference a real Cl-EC regression would eventually predict from). Concretely,
for each candidate chemistry-sampling interval (1, 2, 3, 7, and 14 days):

1. Pick a random start offset within the first interval (e.g. for a
   7-day interval, the first "sample" might fall on day 2, day 5, or
   any other day of that first week).
2. Subsample the real, dense EC record at that fixed interval starting
   from the chosen offset -- this mimics "what if we had only grabbed a
   sample every N days, starting on this particular day."
3. Linearly interpolate between just those sparse points to reconstruct
   an approximation of the full continuous series, exactly as if the
   sparse samples were all the information available.
4. Compare the reconstruction to the real, fully-resolved record
   point-by-point (RMSE, MAE) and, when a paired discharge series is
   available, as a mass-flux percent error.
5. Repeat steps 1-4 200 times per interval, each time with a new random
   offset, and summarize the resulting 200 RMSE values.

Step 5 is why this is a *Monte Carlo* exercise rather than a single
calculation: which specific days a real field campaign happens to land
on is arbitrary, and a sampling schedule that is unlucky enough to
always catch the daily peak (or always catch the trough) will look
artificially better or worse than one that samples more evenly through
the cycle. Repeating with many random offsets characterizes that
phase-dependent uncertainty directly, rather than reporting a single
number that might just reflect a lucky or unlucky start day.

```{r monte-carlo-real}
run_mc_for_site <- function(site_code) {
  df <- ec %>% filter(site == site_code) %>% arrange(timestamp) %>%
    transmute(time = timestamp, chloride_mgL = sc_25c)  # EC stands in as the dense-proxy target
  mc <- monte_carlo_sampling_frequency(df, frequencies_days = c(1, 2, 3, 7, 14), n_draws = 200, seed = 7391)
  out <- summarize_sampling_frequency(mc)
  out$pct_of_mean <- 100 * out$rmse_mean / ec_summary$mean_ec[ec_summary$site == site_code]
  out
}

sbrr_mc <- run_mc_for_site("SBRR")
sbgg_mc <- run_mc_for_site("SBGG")

knitr::kable(sbrr_mc[, c("frequency_days", "n_draws", "rmse_mean", "rmse_p95", "pct_of_mean")],
             digits = 2, caption = "SBRR: EC reconstruction error by candidate sampling interval")
knitr::kable(sbgg_mc[, c("frequency_days", "n_draws", "rmse_mean", "rmse_p95", "pct_of_mean")],
             digits = 2, caption = "SBGG: EC reconstruction error by candidate sampling interval")
```

```{r monte-carlo-plot}
#| fig-width: 7
#| fig-height: 4.5
mc_combined <- bind_rows(
  cbind(sbrr_mc, site = "SBRR"),
  cbind(sbgg_mc, site = "SBGG")
)
ggplot(mc_combined, aes(x = frequency_days, y = pct_of_mean, color = site)) +
  geom_line() + geom_point(size = 2) +
  geom_vline(xintercept = c(1, 2), linetype = "dotted", color = "grey40") +
  annotate("text", x = 1, y = max(mc_combined$pct_of_mean), label = "Option A\n(daily)", size = 3, hjust = -0.1, vjust = 1, color = "grey30") +
  annotate("text", x = 2, y = max(mc_combined$pct_of_mean), label = "Option B\n(every-other-day)", size = 3, hjust = -0.1, vjust = 1, color = "grey30") +
  labs(x = "Candidate sampling interval (days)", y = "Reconstruction RMSE (% of mean EC)",
       title = "Real error-vs-sampling-frequency curve, SBRR and SBGG",
       subtitle = "Dotted lines mark Tier 1 Option A (daily) and Option B (every-other-day)")
```

**Reading**: daily sampling keeps reconstruction error under ~5% of the
mean at both sites; weekly sampling roughly doubles that error (~6-8%);
by biweekly, error reaches ~8-13%. There's no sharp "cliff" -- error grows
smoothly -- which means the choice of interval is a real cost/precision
trade-off, not a hard threshold.

**Precision loss, Option A (daily) vs. Option B (every-other-day), explicit real numbers**: at SBRR, mean reconstruction error rises from 4.74% (daily) to 5.38% (every-other-day) -- a modest ~0.6 percentage-point increase in the typical case, but the worst-case (95th percentile across the 200 draws) rises more, from 5.05% to 6.46% (+1.4 points, a ~28% relative jump). At SBGG the pattern is the same shape: mean error rises from 3.07% to 3.46%, while the worst case rises from 3.21% to 4.34% (+1.1 points, a ~35% relative jump). In plain terms: switching from daily to every-other-day costs relatively little in the typical case, but meaningfully widens how bad an unlucky-start-day campaign could be -- exactly the trade-off Section 2.1's boxplot below visualizes directly.

### How much does the random start-offset itself matter?

The line plot above only shows the *mean* RMSE across the 200 draws per
interval -- it hides how much a single real field campaign's result
could vary just from which day it happened to start on. Plotting the
full distribution of draws (not just their mean) makes that spread
visible directly:

```{r mc-spread-09}
#| fig-width: 7
#| fig-height: 4.5
run_mc_raw <- function(site_code) {
  df <- ec %>% filter(site == site_code) %>% arrange(timestamp) %>%
    transmute(time = timestamp, chloride_mgL = sc_25c)
  mc <- monte_carlo_sampling_frequency(df, frequencies_days = c(1, 2, 3, 7, 14), n_draws = 200, seed = 7391)
  mc$site <- site_code
  mc$pct_of_mean <- 100 * mc$rmse / ec_summary$mean_ec[ec_summary$site == site_code]
  mc
}
mc_raw_combined <- bind_rows(run_mc_raw("SBRR"), run_mc_raw("SBGG"))

ggplot(mc_raw_combined, aes(x = factor(frequency_days), y = pct_of_mean, fill = site)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.8) +
  annotate("text", x = 1, y = max(mc_raw_combined$pct_of_mean) * 1.02, label = "Option A", size = 3, color = "grey30") +
  annotate("text", x = 2, y = max(mc_raw_combined$pct_of_mean) * 1.02, label = "Option B", size = 3, color = "grey30") +
  labs(x = "Candidate sampling interval (days)", y = "Reconstruction RMSE (% of mean EC, one draw)",
       title = "Spread across the 200 Monte Carlo draws (random start-offset), by interval and site",
       subtitle = "Each box summarizes 200 different 'what if the campaign started on a different day' outcomes -- Option A/B marked")
```

**Reading**: the boxes widen noticeably at longer intervals -- at 14
days, a real SBGG campaign could plausibly land anywhere from roughly
6% to 18% RMSE depending purely on which day it started, whereas at 1-2
days nearly every draw lands in a tight band near the mean. This is the
concrete, quantified reason Tier 1's daily cadence (Section 4 below) is
recommended over a coarser interval: not only is the *average* error
lower, the *worst-case* outcome from an unlucky start day is far less
severe. Comparing just the two marked boxes (Option A vs. Option B): both stay in a tight, low band relative to the 7- and 14-day boxes' much wider spread -- the real cost of every-other-day sampling is a modestly wider box, not a jump into a different regime the way weekly or biweekly sampling is.

## Time of day: does it matter when during the day you sample?

Section 1 already showed a real, modest diurnal EC cycle at both sites.
That matters for a grab-sample campaign for a different reason than the
day-to-day interval question above: if the sampling time of day drifts
from visit to visit, part of the apparent day-to-day change in Cl/EC
will really just be a change in what point of the daily cycle was
caught, not a real trend. The fix is cheap: sample at a consistent time
of day, chosen to sit close to each site's own 24-hour mean so it is
also reasonably representative of "typical" conditions.

```{r diurnal-timing-09}
ec$hour_bin <- floor(as.numeric(format(ec$timestamp, "%H")) + as.numeric(format(ec$timestamp, "%M")) / 60)
hourly <- ec %>% group_by(site, hour_bin) %>% summarise(mean_ec = mean(sc_25c), .groups = "drop")

hourly_summary <- hourly %>% group_by(site) %>%
  summarise(peak_hour = hour_bin[which.max(mean_ec)], peak_ec = max(mean_ec),
            trough_hour = hour_bin[which.min(mean_ec)], trough_ec = min(mean_ec))
knitr::kable(hourly_summary, digits = 1,
             caption = "Real hourly mean specific conductance: peak and trough clock hour, both sites")
```

```{r diurnal-timing-plot-09}
#| fig-width: 7
#| fig-height: 4
ggplot(hourly, aes(x = hour_bin, y = mean_ec, color = site)) +
  geom_line() + geom_point(size = 1.5) +
  geom_vline(xintercept = c(10, 13), linetype = "dashed", color = "grey40") +
  labs(x = "Hour of day (local, 24-hr)", y = "Mean specific conductance (uS/cm, 25C)",
       title = "Real diurnal cycle by clock hour, both sites",
       subtitle = "Dashed lines mark the recommended 10:00-13:00 sampling window")
```

**Reading**: the two sites are close to *anti-phased* -- SBRR peaks
around mid-morning to early afternoon (roughly hours 8-14) and troughs
overnight, while SBGG peaks overnight and troughs during that same
mid-morning-to-afternoon window. There is no single clock hour where
both sites sit exactly at their own daily mean simultaneously, but a
**late-morning-to-early-afternoon window (about 10:00-13:00 local
time)** comes closest for both -- SBRR is within its own plateau there,
and SBGG's midday values sit close to (within a few percent of) its
overall mean, well inside the day-to-day noise already quantified in
Section 1. This window is also the most practical one for a single
field day covering both sites (daylight, travel time between Rhodes
Road and Geiger Grade, and typical safety/access considerations).

**Recommendation for Tiers 1 and 2 alike**: visit both sites within the
same ~10:00-13:00 window on every sampling day, and keep the order and
approximate visit time consistent trip to trip (e.g. always SBRR first,
then SBGG, or vice versa) so that any residual within-window drift is at
least systematic rather than random. This is a small logistical
constraint that removes a real, quantifiable source of noise from every
downstream Cl-EC comparison.

## Recommendation: a two-tier campaign

**Tier 1 -- short, intensive calibration window (2-4 weeks, early field
season):** Daily (or, resources permitting, sub-daily/every-other-day)
chloride grab samples at both SBRR and SBGG, run concurrently with the
already-active conductivity loggers. Purpose: this is the *only* way to
finally get enough real, temporally-overlapping Cl-EC pairs to fit a real
calibration relationship (Section 3 of `01_conductivity_temporal_structure.qmd`,
`03_chloride_models.R`, is already built and waiting for exactly this
data) and to directly test whether short-term Cl variability tracks EC's
own real diurnal signal (Section 1/2 above) or behaves differently.

**Isotopes (delta-18O/delta-D) at SBRR and SBGG:** yes, worth including
for a subset of Tier 1 visits at *both* sites, not necessarily every
single day. SBRR and SBGG are surface-water stream sites, so any real
Cl/EC increase observed between them could in principle reflect either
real thermal loading along that reach or simple evaporative
concentration during transit -- Cl and EC alone cannot tell those apart,
since both rise under either mechanism. Isotopes can: evaporative
concentration shifts delta-18O/delta-D together along a local
evaporation line, while conservative thermal mixing does not. Pairing
isotope samples with the every-3rd-day cadence in the Tier 1 calendar
below (Section 4.1) is enough to test this without isotope costs
dominating the whole campaign; see
`manuscript/06_calibration_sampling_proposal.qmd` for the broader,
project-wide isotope/major-ion rationale.

**Discharge measurement, every visit, both tiers:** every Cl grab sample
must be paired with an instantaneous discharge value taken at the same
time. Concentration alone cannot support the chloride mass-flux
calculation (`scripts/analysis/chloride_mass_balance.R`, the two-station
flux formula introduced in Session 39) that this project's actual
thermal-discharge estimates depend on -- a Cl-EC calibration without
paired Q only ever answers "what is the concentration," never "how much
thermal water is discharging." **SBRR** is co-located with USGS gauge
10349300, so pairing is just reading the live gauge at sample time
(exactly how the real Section 1.1 point was obtained). **SBGG has no
permanent gauge**, so a manual discharge measurement (wading-rod
current-meter transect, using the `stream_discharge.xlsx` transect
template already built into the schema) is a hard requirement at every
SBGG visit, not an optional add-on -- without it, SBGG can only ever
contribute to the Cl-EC concentration calibration, never to a real flux
estimate.

**Tier 2 -- lower-frequency campaign for the rest of the year (biweekly to
monthly):** Once Tier 1 establishes the calibration relationship, routine
monitoring can drop to biweekly-or-monthly grab samples with a
quantified (not guessed) reconstruction-error cost (Section 2 above) --
**plus targeted extra samples bracketing any large storm or spring-runoff
event**, since `notebooks/07`/`notebooks/08` already found a real,
detectable water-level response to large storms (the Dec-2025 event at
STMGID MW10) -- the same logic likely applies to Cl/EC at SBRR/SBGG and is
worth testing directly with real paired data once Tier 1 samples exist.
The same discharge-pairing requirement above continues at Tier 2's lower
frequency -- a Cl value without a paired Q still can't be converted into
a real Q_TW estimate, no matter how infrequently it's collected -- and
the same 10:00-13:00 sampling window from Section 3 still applies.

**Why not just do one long daily campaign for the whole year?** The
Section 2 error curve shows diminishing returns below about a week's
interval once the initial calibration is established -- the real payoff of
daily sampling is in *establishing the relationship itself* (Tier 1), not
in sustaining it indefinitely. A full year of daily sampling would be the
most statistically ideal choice but is very unlikely to be logistically
realistic for a one-person thesis field campaign; the two-tier design is
the pragmatic compromise, explicitly justified rather than assumed.

### Timing: should Tier 1 start now, in the fall wet-season transition?

This proposal is being finalized 2026-09-29. A real practical
constraint -- funding/reporting deadlines favor real Tier 1 data
existing before 2027 -- pushes toward starting in October or November
2026 rather than waiting for a more hydrologically "ideal" window. That
means Tier 1 will likely run during the fall wet-season transition, not
the dry summer baseflow period the conductivity record (Section 1) and
Monte Carlo results (Section 2) are built from. Worth stating plainly,
with real benefits and drawbacks, rather than leaving it implicit.

```{r timing-precip-09}
precip_by_month <- dbGetQuery(con, "
  SELECT strftime('%Y-%m', date) AS year_month,
         SUM(CASE WHEN value > 0 THEN 1 ELSE 0 END) AS wet_days,
         ROUND(SUM(value), 1) AS total_mm
  FROM Weather_Observations
  WHERE parameter = 'PRCP'
  GROUP BY year_month ORDER BY year_month
")
knitr::kable(precip_by_month, col.names = c("Year-month", "Wet days", "Total precip (mm)"),
             caption = "Real NOAA precipitation record at Reno Airport -- the only fall/winter season currently on file")
```

**Reading**: the only real fall/winter record on file (2025-2026) shows
October was notably dry (2 wet days, 0.2 mm total in Oct 2025), while
November and December each had ~8 wet days and several mm of real
precipitation, with January-February 2026 the wettest months on
record. This is a single real year, not a multi-year climatology -- a
real limitation, stated honestly -- but it is consistent with this
region's broadly known dry-fall/wet-winter pattern. Starting Tier 1 in
the first half of October has some real historical basis for still
being in a relatively dry window, with wet-season risk increasing
noticeably by November.

**Benefits of running Tier 1 now (fall 2026), even with real storm
risk**:

- Meets the real funding/reporting deadline -- real Tier 1 data before
  2027 is more useful than a scientifically "ideal" campaign that
  arrives too late to matter for that deadline.
- If a real storm does occur during Tier 1, that is not purely a cost:
  `notebooks/07`/`notebooks/08` already found a real water-level
  response to the Dec-2025 storm at STMGID MW10 -- a storm captured
  *during* Tier 1 would be the first real chance to test whether Cl/EC
  at SBRR/SBGG show an analogous response, a genuinely valuable dataset
  this project doesn't have yet.
- Broadens the conductivity record's seasonal coverage beyond the
  current July-August-only window (Section 1's own stated limitation),
  which matters directly for the diurnal-timing caveat below.

**Drawbacks of running Tier 1 during a precipitation event**:

- A storm mid-campaign introduces a real third end-member (fresh
  meteoric precipitation) into what Section 1.3's mixing discussion
  already treats as a two-end-member (thermal/creek) system --
  storm-affected Tier 1 samples should be analyzed and flagged
  separately, not pooled uncritically into the same simple linear
  Cl-EC fit the rest of Tier 1 is meant to establish.
- Section 3's 10:00-13:00 sampling-window recommendation was derived
  from a July-August, solar-heating-driven diurnal cycle -- that
  physical driver is real but weaker in fall (lower sun angle, shorter
  days) and could be masked entirely on a storm day. The specific
  clock-hour recommendation should be treated as provisional for a
  fall campaign, not assumed to transfer directly from the summer
  record.
- SBGG's manual current-meter discharge measurement (the paired-discharge
  requirement above) becomes both more hazardous and less accurate
  during a real high-turbidity, high-turbulence storm flow -- a real
  field-safety and data-quality consideration, not just an
  inconvenience.

**Recommendation**: proceed with Tier 1 in fall 2026 given the real
funding constraint, but explicitly flag any storm-affected sampling
days (directly joinable against this same real `Weather_Observations`
PRCP record) as a distinct analysis subset -- both a genuine bonus
dataset for storm-response testing and a necessary exclusion from the
"clean baseflow" calibration fit Tier 1 is primarily designed to
produce.

### Tier 1 Option A: 14 consecutive days

A concrete, complete 14-consecutive-day version of Tier 1, generated
directly (not hand-typed) so it stays consistent with the recommendations
above -- isotopes on roughly every 3rd visit, discharge paired every
visit, both sites within the same time window each day:

**Why 14 samples specifically?** Two real constraints, not one, set this number. First, Section 2's error-vs-frequency graph (marked "Option A" above) shows daily sampling already sits in a low, flat part of the error curve (4.7-4.9% of mean EC) -- sampling *more often* than daily buys essentially nothing further, since there is no sub-daily structure left to resolve once inside that band. Second, a practical statistical floor matters more than the interval curve does here: `manuscript/06_calibration_sampling_proposal.qmd` cites n=15-20 real paired samples as the conventional minimum for a stable simple linear calibration regression. 14 consecutive days is a deliberate, honest compromise against that guidance -- it is the longest single intensive block a one-person thesis field campaign can realistically sustain, and it sits just *under* the ideal n=15-20 floor rather than comfortably above it. In practice, real-world attrition (a missed day, an equipment issue, a field-access problem) will likely pull the realized n toward the low end of 14 anyway, so extending to a full 3 weeks (n=21), resources permitting, would be a real improvement worth considering rather than treating 14 as untouchable.

```{r tier1-calendar-09}
tier1_days <- 1:14
tier1_calendar <- data.frame(
  day = tier1_days,
  time_window = "10:00-13:00",
  sites = "SBRR & SBGG",
  cl_major_ions = "Yes (both sites)",
  isotopes_d18O_dD = ifelse(tier1_days %% 3 == 1, "Yes (both sites)", ""),
  discharge_measurement = "SBRR: USGS gauge reading | SBGG: current-meter transect"
)
knitr::kable(tier1_calendar,
             col.names = c("Day", "Time window", "Sites", "Cl + major ions", "Isotopes (d18O/dD)", "Discharge measurement"),
             caption = "Proposed Tier 1 sampling calendar: 14 consecutive days, example start (isotopes on days 1, 4, 7, 10, 13)")
```

This 14-consecutive-day layout is the tightest, highest-information
version of Tier 1. An equally valid variant, if daily field visits
aren't logistically realistic, is the same structure stretched to
**every-other-day over 4 weeks** (14 visits total, same isotope/discharge
cadence relative to visit number rather than calendar day) -- Section
2.1's spread analysis shows this costs some precision at longer gaps
but is far better than jumping straight to Tier 2's biweekly/monthly
cadence without ever establishing the calibration.

### Tier 1 Option B: every-other-day over 4 weeks (lower logistical cost)

If committing to 14 *consecutive* days isn't realistic (competing
obligations, no coverage for other fieldwork, etc.), the same 14 visits
can instead be spread across a 4-week window, one visit every other
day. This is generated the same way as Option A, indexed by visit
number rather than calendar day so the isotope/discharge cadence stays
identical:

```{r tier1-option-b-09}
visit_number <- 1:14
tier1b_calendar <- data.frame(
  visit = visit_number,
  calendar_day = 2 * visit_number - 1,
  time_window = "10:00-13:00",
  sites = "SBRR & SBGG",
  cl_major_ions = "Yes (both sites)",
  isotopes_d18O_dD = ifelse(visit_number %% 3 == 1, "Yes (both sites)", ""),
  discharge_measurement = "SBRR: USGS gauge reading | SBGG: current-meter transect"
)
knitr::kable(tier1b_calendar,
             col.names = c("Visit #", "Calendar day (of 28)", "Time window", "Sites",
                           "Cl + major ions", "Isotopes (d18O/dD)", "Discharge measurement"),
             caption = "Tier 1 Option B: every-other-day over 4 weeks (14 visits across 28 calendar days; isotopes on visits 1, 4, 7, 10, 13)")
```

**What this option costs and buys, honestly**: the total real lab
cost is *identical* to Option A -- same 14 Cl+major-ion analyses, same
5 isotope pairs, same 14 paired discharge measurements -- this option
only trades calendar time, not lab budget, for reduced field-logistics
pressure (no 14-day unbroken commitment). The real statistical cost,
stated explicitly rather than left to the graph alone: mean
reconstruction error rises from 4.74% to 5.38% at SBRR and from 3.07%
to 3.46% at SBGG (Section 2's marked graph); the worst-case (95th
percentile) cost is larger -- 5.05% to 6.46% at SBRR, 3.21% to 4.34% at
SBGG (Section 2.1's marked boxplot). Both intervals still sit
in the
tight, low-error band, well short of weekly's roughly-doubled error --
so an every-other-day cadence is a genuinely reasonable compromise, not
a steep step down from daily. If the real limiting factor is a hard
budget cap on lab *analyses* rather than field days, neither option
reduces cost -- that would mean cutting the total visit count itself
(e.g. daily for 7 days instead of 14), which would need to be weighed
directly against Section 2.1's real precision-loss curve before
recommending it.

### Tier 2 sampling timeline (the rest of the year)

Tier 2 was described qualitatively above (biweekly-to-monthly, plus
storm-triggered extras); here is a concrete, generated timeline picking
up right after either Tier 1 option ends (week 4), transitioning from
biweekly to monthly cadence over the rest of the year, with periodic
isotope re-checks:

```{r tier2-timeline-09}
transition_weeks <- seq(5, 11, by = 2)     # biweekly for ~2 months after Tier 1
steady_weeks <- seq(16, 52, by = 4)         # monthly for the remaining ~10 months
tier2_weeks <- c(transition_weeks, steady_weeks)
tier2_phase <- c(rep("Transition (biweekly)", length(transition_weeks)),
                  rep("Steady-state (monthly)", length(steady_weeks)))

tier2_calendar <- data.frame(
  visit = seq_along(tier2_weeks),
  week = tier2_weeks,
  approx_month = ceiling(tier2_weeks / 4.345),
  phase = tier2_phase,
  cl_major_ions = "Yes (both sites)",
  isotopes_d18O_dD = ifelse(seq_along(tier2_weeks) %% 4 == 1, "Yes (quarterly check)", ""),
  discharge_measurement = "SBRR: USGS gauge reading | SBGG: current-meter transect"
)
knitr::kable(tier2_calendar,
             col.names = c("Visit #", "Week (post Tier-1 start)", "Approx. month", "Phase",
                           "Cl + major ions", "Isotopes (d18O/dD)", "Discharge measurement"),
             caption = "Tier 2 timeline: biweekly transition (months 2-3), then monthly (months 4-12), quarterly isotope re-checks")
```

**Reading**: this gives 14 routine Tier 2 visits across the year (4
biweekly transition visits, 10 monthly steady-state visits) -- combined
with Tier 1's own 14 visits, that's 28 core Cl+major-ion visits/year,
broadly in line with the ~40-total-analyses "Tier 2" target already
recommended in `manuscript/06_calibration_sampling_proposal.qmd` once
the isotope subset and any storm-triggered extras are added. **This
fixed calendar is deliberately a floor, not a ceiling**: per Section
4's own recommendation, any large storm or spring-runoff event
identified in real time should add extra, unscheduled visits bracketing
it (before/during/after), the same way `notebooks/07`/`notebooks/08`
found a real water-level response to the Dec-2025 storm at STMGID MW10
-- those extra visits are not pre-datable and so are intentionally not
in the fixed table above.

## What this does and doesn't tell us yet

- This analysis uses **EC as a stand-in for chloride** because no dense
  real Cl record exists yet -- exactly the gap Tier 1 above is designed to
  close. Once Tier 1 samples exist, this same Monte Carlo framework should
  be re-run on the *real* Cl series (via the fitted `03_chloride_models.R`
  relationship), not just EC, for a final, fully-real recommendation.
- **Neither real ALS lab chemistry nor a paired discharge series currently
  feeds the Monte Carlo above** -- Section 1.1's single real Cl/discharge
  pair at SBRR (and SBGG's complete absence of one) is shown for context
  only; it predates the conductivity record by ~2.5 months and is nowhere
  near enough (n=1, zero overlap) to validate or calibrate anything.
  Discharge pairing at every Tier 1/2 visit (Section 4) is what will
  finally make a real flux-based version of this Monte Carlo possible.
- The ~5-week record used here (mid-July to late-August 2026) does not yet
  span a full season -- the daily-range/diurnal-amplitude numbers above may
  not hold in winter/spring; this is itself a reason to keep sampling
  through Tier 2 rather than assuming summer behavior generalizes.
- This document is deliberately reproducible (a live notebook, not a
  static writeup) so it can be re-run and refined as real Tier 1 data
  starts arriving.

```{r cleanup-09}
#| echo: false
dbDisconnect(con)
```
