---
title: "Pipeline QC & Status Report"
output: html_document
params:
  db_path: "database/geochem_demo.sqlite"
  run_time: "unknown"
  mode: "DEMO"
---

# Overview

A fast, mechanical status check generated after every pipeline run
(`scripts/run_pipeline.R`) -- table counts, ingest recap, data
availability, and a `QC_Issues`/`PHREEQC_Run_Failures` summary. This is
deliberately **not** a science/interpretation document: geochemical
interpretation, geothermometry, mixing models, and well-network detail
live in the versioned Quarto notebooks under `notebooks/` (in
particular `05_data_inventory_and_well_network.qmd` for well/facility
network detail and `06_phreeqc_geochemical_modeling.qmd` for PHREEQC),
which are rendered on demand rather than on every pipeline run, and on
the project website. See those for the "why" behind any number here.

- **Mode:** `r params$mode`
- **Run time:** `r params$run_time`

---

# Database Connection

```{r}
library(DBI)
library(RSQLite)
library(here)
library(dplyr)

db_path <- if (!is.null(params$db_path)) params$db_path else file.path("database", "geochem_demo.sqlite")
con <- dbConnect(SQLite(), here::here(db_path))
```

# Table Row Counts

```{r}
tables <- c(
  "Locations",
  "Wells",
  "Water_Level_Observations",
  "Temperature_Observations",
  "Conductivity_Observations",
  "Sampling_Events",
  "Lab_Analyses",
  "Isotope_Analyses",
  "Well_Log_Documents",
  "Sampling_Ports",
  "Production_Port_Links",
  "Port_Injection_Links",
  "Facility_Areas",
  "Hydraulic_Gradients",
  "PHREEQC_Solutions",
  "PHREEQC_Results",
  "QC_Issues"
)

counts <- do.call(rbind, lapply(tables, function(t) {
  n <- tryCatch(dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM ", t))$n, error = function(e) NA)
  data.frame(table = t, rows = n)
}))

counts
```

# Ingest Recap

Most recent ingest run per source (`Ingest_Run_Log`) -- a quick way to
confirm every expected source actually ran this pipeline execution, and
roughly how much it inserted.

```{r}
tryCatch({
  dbGetQuery(con, "
    SELECT data_source, MAX(timestamp) AS last_run,
           SUM(locations_inserted) AS locations,
           SUM(events_inserted) AS events,
           SUM(samples_inserted) AS samples,
           SUM(measurements_inserted) AS measurements
    FROM Ingest_Run_Log
    GROUP BY data_source
    ORDER BY last_run DESC
  ")
}, error = function(e) cat("Ingest_Run_Log not available:", conditionMessage(e)))
```

# QC Issues Summary

Counts from `QC_Issues` (populated by `run_qc_checks()` /
`scripts/qc/qc_data_integrity_checks.R`), broken down by type and
severity -- a genuine pass/fail signal, not exploratory. See
`qc_reports/*.csv` and `docs/data/qc_summary.csv` for the full,
record-level detail behind each count.

```{r}
tryCatch({
  qc_summary <- dbGetQuery(con, "
    SELECT issue_type, severity, COUNT(*) AS n
    FROM QC_Issues
    GROUP BY issue_type, severity
    ORDER BY n DESC
  ")
  if (nrow(qc_summary) == 0) {
    cat("No QC_Issues rows -- either QC has not run yet, or nothing was flagged.")
  } else {
    qc_summary
  }
}, error = function(e) cat("QC_Issues not available:", conditionMessage(e)))
```

```{r}
tryCatch({
  worst <- dbGetQuery(con, "
    SELECT table_name, issue_type, severity, COUNT(*) AS n
    FROM QC_Issues
    WHERE severity = 'ERROR'
    GROUP BY table_name, issue_type, severity
    ORDER BY n DESC
    LIMIT 10
  ")
  if (nrow(worst) > 0) {
    cat("Top ERROR-severity issue types:\n")
    worst
  } else {
    cat("No ERROR-severity QC issues.")
  }
}, error = function(e) NULL)
```

# PHREEQC Run Failures

Samples that failed to converge with every thermodynamic database tried
(`PHREEQC_Run_Failures`) -- should normally be empty or very small; a
growing count here is worth investigating in
`notebooks/06_phreeqc_geochemical_modeling.qmd`.

```{r}
tryCatch({
  failures <- dbGetQuery(con, "SELECT sample_id, location_name, site_type, reason, logged_at FROM PHREEQC_Run_Failures ORDER BY logged_at DESC")
  if (nrow(failures) == 0) {
    cat("No PHREEQC_Run_Failures rows.")
  } else {
    failures
  }
}, error = function(e) cat("PHREEQC_Run_Failures not available:", conditionMessage(e)))
```

# Data Availability

Which sources have data, and over what time period -- see
`scripts/analysis/data_availability.R`. Fuller well/facility-network
detail (production -> port -> injection routing, wells by role) lives in
`notebooks/05_data_inventory_and_well_network.qmd`, not here, to avoid
maintaining the same summary in two places.

```{r}
avail_path <- here::here("data", "derived", "data_availability", "data_availability.csv")
if (file.exists(avail_path)) {
  read.csv(avail_path)
} else {
  cat("Data availability output not yet generated -- run the pipeline's export stage.")
}
```

```{r, out.width="100%"}
avail_plot <- here::here("output", "figures", "data_availability_timeline.png")
if (file.exists(avail_plot)) knitr::include_graphics(avail_plot)
```

```{r}
dbDisconnect(con)
```
