---
title: "Pipeline Run Report"
output: html_document
params:
  db_path: "database/geochem_demo.sqlite"
  run_time: "unknown"
  mode: "DEMO"
---

# Overview

This report summarizes the results of the data pipeline run.

It includes:

- database status
- ingest summary
- hydrologic analysis outputs
- QA/QC checks

---

# Database Connection

```{r}
library(DBI)
library(RSQLite)
library(here)

db_path <- if (!is.null(params$db_path)) params$db_path else file.path("database", "geochem_demo.sqlite")
con <- dbConnect(SQLite(), here::here(db_path))
```


```{r}
tables <- c(
  "Locations",
  "Wells",
  "Water_Level_Observations",
  "Temperature_Observations",
  "Conductivity_Observations",
  "Sampling_Events",
  "Lab_Analyses",
  "Isotope_Analyses",
  "Well_Log_Documents",
  "Sampling_Ports",
  "Production_Port_Links",
  "Port_Injection_Links",
  "Facility_Areas",
  "Hydraulic_Gradients"
)

counts <- do.call(rbind, lapply(tables, function(t) {
  n <- tryCatch(dbGetQuery(con, paste0("SELECT COUNT(*) AS n FROM ", t))$n, error = function(e) NA)
  data.frame(table = t, rows = n)
}))

counts
```

# Data Availability

Which sources have data, and over what time period -- see
`scripts/analysis/data_availability.R`.

```{r}
avail_path <- here::here("data", "derived", "data_availability", "data_availability.csv")
if (file.exists(avail_path)) {
  read.csv(avail_path)
} else {
  cat("Data availability output not yet generated -- run the pipeline's export stage.")
}
```

```{r, out.width="100%"}
avail_plot <- here::here("output", "figures", "data_availability_timeline.png")
if (file.exists(avail_plot)) knitr::include_graphics(avail_plot)
```

# Well & Facility Flow Network Summary

Production well -> port -> injection well routing (see README's "Well &
Facility Flow Network" section for the schema/provenance behind this).

```{r}
tryCatch({
  wells_by_role <- dbGetQuery(con, "SELECT well_role, COUNT(*) n FROM Wells GROUP BY well_role ORDER BY n DESC")
  wells_by_role
}, error = function(e) cat("Wells.well_role not available:", conditionMessage(e)))
```

```{r}
tryCatch({
  dbGetQuery(con, "
    SELECT sp.port_name, COUNT(*) n_production_wells
    FROM Production_Port_Links ppl JOIN Sampling_Ports sp ON ppl.port_id = sp.port_id
    GROUP BY sp.port_name ORDER BY n_production_wells DESC
  ")
}, error = function(e) cat("Production_Port_Links not available:", conditionMessage(e)))
```


```{r}
grad <- dbGetQuery(con, "SELECT * FROM Hydraulic_Gradients")

head(grad)
```

```{r}
summary(grad$gradient)
```

```{r}
library(ggplot2)

ggplot(grad, aes(x = gradient)) +
  geom_histogram(bins = 50) +
  theme_minimal()
```

```{r}
table(grad$gradient_class)
```

```{r}
sum(is.na(grad$gradient))
```

```{r}
subset(grad, abs(grad$gradient) > 0.05)[1:10, ]
```

```{r}
summary(grad$distance_m)
```

```{r}
temp <- dbGetQuery(con, "
  SELECT temperature FROM Temperature_Observations
")

summary(temp$temperature)
```

```{r}
library(sf)

grad_sf <- st_read(
  here::here("output", "geopackage", "hydro_data.gpkg"),
  layer = "hydraulic_gradients"
)

plot(st_geometry(grad_sf))
```

```{r}
dbDisconnect(con)
```
