---
title: "Data and Schema"
output: html_document
---

# Data & Structure

This system integrates multiple environmental datasets into a unified, structured database designed for reproducible scientific analysis.

---

## 📥 Data Sources

The database integrates heterogeneous inputs collected across multiple workflows:

- **Regulatory data (NDEP)** — long-term chemistry records  
- **Field measurements** — site-based sampling campaigns  
- **Laboratory analyses** — high-quality analytical chemistry (in progress)  
- **Temperature loggers** — continuous thermal monitoring  

Each source has different structure, frequency, and quality constraints, requiring controlled ingestion and validation.

---

## 🔄 Data Flow

All data entering the system pass through a controlled sequence of transformations:

**Raw Input → Validation → Standardization → Database → Derived Outputs**

Key features of this process:

- Raw inputs are never modified  
- All transformations are scripted and version-controlled  
- External identifiers (e.g., station codes) are translated into internal database keys  
- Invalid or inconsistent data are rejected during ingest  

### Archival and Provenance

To preserve reproducibility:

- All raw field templates are **archived before ingestion**  
- Templates are **cleared only after successful ingest**  
- This prevents duplication while maintaining a complete history of inputs  

---

## 🧱 Database Structure

The system is built on a normalized relational schema with the following core tables:

- **Locations** — spatial reference points  
- **Sampling_Events** — time-based groupings of observations  
- **Samples** — discrete physical observations  
- **Field_Measurements** — in-situ measurements  
- **Lab_Analyses** — analytical chemistry results  
- **Temperature_Observations** — high-frequency time-series data  

### Design Principles

- normalized relational structure  
- strict foreign key relationships  
- append-only time series (no overwriting)  
- explicit separation of raw vs. derived data  

---

## 🌡️ Continuous vs Discrete Data

A key feature of the system is the integration of:

### Discrete observations
- field measurements  
- laboratory chemistry  

### Continuous observations
- temperature logger data  

Continuous data enable:

- detection of short-term variability  
- temporal pattern recognition  
- higher-resolution system monitoring  

---

## 📊 Data Behavior

### Source Integration

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(DBI)
library(RSQLite)
library(ggplot2)

con <- DBI::dbConnect(RSQLite::SQLite(), "database/geochem_demo.sqlite")

if (dbGetQuery(con, "SELECT COUNT(*) n FROM Lab_Analyses")$n == 0) {

  cat("Laboratory data not yet available.")

} else {

  source_df <- dbGetQuery(con, "
SELECT
  CASE
    WHEN source_id = 1 THEN 'NDEP'
    WHEN source_id = 2 THEN 'Field'
    ELSE 'Other'
  END AS data_source,
  COUNT(*) AS n
FROM Lab_Analyses
GROUP BY data_source
")

# Ensure both categories exist
all_sources <- data.frame(
  data_source = c("NDEP", "Field")
)

source_df <- merge(all_sources, source_df, all.x = TRUE)
source_df$n[is.na(source_df$n)] <- 0

# Plot
print(
ggplot(source_df, aes(x = data_source, y = n, fill = data_source)) +
  geom_col() +
  labs(
    title = "Integration of Chemistry Data Sources",
    subtitle = "Field chemistry not yet ingested",
    x = "Data Source",
    y = "Number of Records"
  ) +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme_minimal()
)
}
```

---

### Ingestion Through Time

cat("**Note:** In DEMO mode, all ingestion events are replayed from scratch. In OPERATIONAL mode, this plot reflects real-time database growth.")


```{r echo=FALSE, message=FALSE, warning=FALSE}
ingest_log <- dbGetQuery(con, "
SELECT
  timestamp,
  data_source,
  measurements_inserted
FROM Ingest_Run_Log
ORDER BY timestamp;
")

library(lubridate)
library(dplyr)

ingest_log$timestamp <- ymd_hms(ingest_log$timestamp)

ingest_log <- ingest_log |>
  arrange(timestamp) |>
  mutate(cumulative = cumsum(measurements_inserted))

print(
ggplot(ingest_log,
       aes(x = timestamp,
           y = cumulative,
           color = data_source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_vline(aes(xintercept = timestamp),
           linetype = "dashed",
           alpha = 0.2) +
  labs(
    title = "Cumulative Database Growth",
    subtitle = "Each step represents an ingestion event",
    x = "Time",
    y = "Total Records",
    color = "Data Source"
  ) +
  theme_minimal()
)
```

---

These patterns reflect:

- incremental database growth  
- integration across multiple data sources  
- reproducibility of ingestion workflows  

---

## 🔬 Geochemical Modeling

Validated chemistry is transformed into **PHREEQC-ready inputs**, enabling:

- forward speciation modeling  
- inverse modeling of hydrothermal processes  

Model outputs are written back into the system, preserving the relationship between:

```
raw observations → derived chemistry → interpreted processes
```

---

## 🔁 Reproducibility

All steps in the system are:

- scripted  
- version-controlled  
- reproducible  

No manual edits are made to the database.

Every dataset can be:

- traced to a specific input  
- reprocessed from raw data  
- validated through ingestion logs  

---

## 📸 Image Data

Photographic observations are stored as structured metadata and linked to:

- locations  
- sampling events  
- field measurements  

Images are stored externally, while metadata remain relational.

---

## ✅ Current State

The system is actively evolving.

- Temperature logger data are fully integrated  
- Field and sampling data workflows are operational  
- Laboratory chemistry ingestion is ongoing  

&gt; In operational mode, ingestion is explicitly controlled, and the database evolves incrementally over time.