---
title: "Data and Schema"
output: html_document
---

# Data & Structure

This system integrates heterogeneous environmental observations into a unified relational database designed to support reproducible scientific analysis.  

Rather than storing datasets independently, all observations are structured within a shared schema that enforces spatial, temporal, and relational consistency. This enables direct comparison across data types, monitoring scales, and time.

The following sections explore how the database behaves as an integrated system, highlighting patterns in data sources, temporal growth, structural

---

## 📥 Data Sources

The database integrates multiple observational domains, each representing a different aspect of the hydrothermal system:

- **Regulatory data (NDEP)** — long-term geochemical records  
- **Field measurements** — site-specific sampling campaigns  
- **Laboratory analyses** — high-quality analytical chemistry  
- **Temperature loggers** — continuous thermal monitoring
- **Continuous conductivity loggers, USGS discharge & historic chemistry, NOAA weather, and field-photo EXIF GPS** (added 2026-09) — see `AGENTS.md`/`README.md` for the current per-source drop-folder and ingest-script mapping

Each source differs in temporal resolution, structure, and analytical purpose.  

Bringing these datasets into a common framework allows relationships to be established between measurements that were previously isolated.

> This integration is essential for interpreting the hydrothermal system as a coupled hydraulic, thermal, and geochemical process.


---

## 🔄 Data Flow

All data entering the system pass through a controlled sequence of transformations:

**Raw Input → Validation → Standardization → Database → Derived Outputs**

Key features of this process:

- Raw inputs are never modified  
- All transformations are scripted and version-controlled  
- External identifiers (e.g., station codes) are translated into internal database keys  
- Invalid or inconsistent data are rejected during ingest  

### Archival and Provenance

To preserve reproducibility:

- All raw field templates are **archived before ingestion**  
- Templates are **cleared only after successful ingest**  
- This prevents duplication while maintaining a complete history of inputs  

---

## 🧱 Database Structure

The system is built on a normalized relational schema with the following core tables:

- **Locations** — spatial reference points  
- **Sampling_Events** — time-based groupings of observations  
- **Samples** — discrete physical observations  
- **Field_Measurements** — in-situ measurements  
- **Lab_Analyses** — analytical chemistry results  
- **Temperature_Observations** — high-frequency time-series data  

### Design Principles

- normalized relational structure  
- strict foreign key relationships  
- append-only time series (no overwriting)  
- explicit separation of raw vs. derived data  

---

## 🌡️ Continuous vs Discrete Data

A key feature of the system is the integration of:

### Discrete observations
- field measurements  
- laboratory chemistry  

### Continuous observations
- temperature logger data  

Continuous data enable:

- detection of short-term variability  
- temporal pattern recognition  
- higher-resolution system monitoring  

---

## 🔗 Integration of Data Sources

The degree to which different datasets contribute to the system provides insight into both data availability and monitoring strategy.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(ggplot2)
library(dplyr)

source_path <- "../docs/data/source_summary.csv"

if (!file.exists(source_path)) {
  
  cat("Source breakdown available after pipeline export.")
  
} else {
  
  sources <- read_csv(source_path, show_col_types = FALSE)
  
  p <- ggplot(sources,
              aes(x = reorder(source, n), y = n, fill = source)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Integration of Data Sources",
      x = "",
      y = "Records"
    ) +
    theme_minimal()
  
  print(p)
}

```
Interpretation:
This distribution reflects the heterogeneous nature of environmental monitoring. Regulatory datasets typically dominate long-term records, while field and sensor data provide targeted and high-resolution observations.
Key Insight:
The strength of the system lies not in any single dataset, but in the ability to integrate multiple observational scales into a unified structure.

---

## 📈 System Growth Through Time

The database evolves through discrete ingestion events rather than continuous accumulation, reflecting how environmental data are collected and integrated.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)

path <- "../docs/data/ingest_log.csv"

if (!file.exists(path)) {
  cat("Ingestion history will appear after pipeline runs.")
  
} else {
  
  ingest <- read_csv(path, show_col_types = FALSE) %>%
    mutate(timestamp = ymd_hms(timestamp)) %>%
    arrange(timestamp) %>%
    mutate(cumulative = cumsum(measurements_inserted))
  
p <- ggplot(ingest,
              aes(timestamp, cumulative, color = data_source)) +
    geom_line(linewidth = 1.2) +
    labs(
      title = "Cumulative Database Growth",
      x = "Time",
      y = "Total Records"
    ) +
    theme_minimal()
  
  ggplotly(p)
}
```

Interpretation:
Each increase corresponds to a specific ingestion event, such as a field campaign, regulatory update, or logger deployment.
Key Insight:
The stepwise growth pattern provides explicit provenance, allowing users to trace how and when data entered the system.

---

## ⚖️ Data Granularity: Continuous vs Discrete Observations

A defining characteristic of the system is the coexistence of continuous and discrete data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("docs/data/table_counts.csv")) {
  
  table_counts <- read_csv("../docs/data/table_counts.csv", show_col_types = FALSE)
  
  summary_df <- table_counts %>%
    mutate(type = case_when(
      grepl("Temperature", table) ~ "Continuous",
      TRUE ~ "Discrete"
    )) %>%
    group_by(type) %>%
    summarise(n = sum(n_rows))
  
  p <- ggplot(summary_df,
              aes(x = type, y = n, fill = type)) +
    geom_col() +
    labs(
      title = "Continuous vs Discrete Observations",
      y = "Number of Records",
      x = ""
    ) +
    theme_minimal()
  
  print(p)
}
```

Interpretation:
Continuous temperature observations dominate the total number of records, while discrete datasets (chemistry and field measurements) provide analytical depth.
Key Insight:
This multi-scale structure enables the system to capture both high-frequency variability and chemically diagnostic signals.

---

## 🕒 Temporal Distribution of Observations

Understanding when data are collected is critical for interpreting system dynamics.

```{r echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("docs/data/temp_sample.csv")) {
  
  temp <- read_csv("../docs/data/temp_sample.csv", show_col_types = FALSE) %>%
    mutate(timestamp = as.POSIXct(timestamp, origin = "1970-01-01")) %>%
    filter(!is.na(timestamp))
  
  p <- ggplot(temp,
              aes(timestamp)) +
    geom_histogram(bins = 50, fill = "steelblue") +
    labs(
      title = "Temporal Distribution of Observations",
      x = "Time",
      y = "Frequency"
    ) +
    theme_minimal()
  
  print(p)
}

```

Interpretation:
Clustering reflects episodic data acquisition, including sensor deployments, field campaigns, and regulatory sampling.
Key Insight:
Temporal structure reveals not only system behavior, but also sampling strategy and data availability constraints.

## 🗺️ Spatial Coverage (Sample Points)

Spatial distribution determines how well the system is observed across the hydrothermal field.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(leaflet)

path <- "../docs/data/well_sample.csv"

if (!file.exists(path)) {
  
  cat("Spatial data will appear after export.")
  
} else {
  
  wells <- read_csv(path, show_col_types = FALSE)
  
  leaflet(wells) |>
    addTiles() |>
    addCircleMarkers(
      lng = ~longitude,
      lat = ~latitude,
      radius = 4,
      color = "blue",
      popup = ~paste("Well:", well_id)
    )
}
```
Interpretation:
Observed clusters represent areas of active monitoring, while gaps indicate potential regions of uncertainty or under-sampling.
Key Insight:
Spatial structure directly influences the ability to interpret hydrothermal flow paths and system heterogeneity.

---

## 📊 System-Level Interpretation

Taken together, these patterns describe a system that is:

- incrementally constructed through controlled ingestion workflows  
- composed of heterogeneous but integrated data sources  
- dominated by high-frequency continuous observations  
- structured across irregular temporal sampling intervals  
- spatially distributed according to monitoring access and priorities  

> The database is not simply a collection of datasets—it is a structured representation of system observation itself.

---

## 🔬 Geochemical Modeling Integration

Validated chemistry is transformed into **PHREEQC-ready inputs**, enabling process-based interpretation of hydrothermal conditions.

These transformations support:

- forward speciation modeling  
- inverse geochemical modeling  
- evaluation of mineral saturation states  

> This step links observational data to thermodynamic interpretation, extending the system beyond monitoring into predictive modeling.

---

## 🔁 Reproducibility and Provenance

All transformations within the system are:

- scripted  
- version-controlled  
- reproducible  

No manual edits are made to database contents.

Every dataset can be:

- traced to its original input  
- reconstructed from raw data  
- validated through ingestion logs  

> This ensures that all analyses performed using the system are transparent, auditable, and reproducible.

---

## 📸 Image Data

Photographic observations are stored as structured metadata and linked to:

- locations  
- sampling events  
- field measurements  

Images are stored externally, while metadata remain relational.

---

## 📄 NDEP Public Records Request (PRR) Staging & Promotion

Chemistry extracted from NDEP PRR PDF lab reports never becomes "real"
data automatically -- it always passes through two deliberately separate
steps:

1. **`ingest_ndep_prr.R`** parses a supported PDF layout (currently the
   SGS lab-report format) and writes every result to `Staging_NDEP_WQ`
   only. Nothing here touches `Locations`, `Samples`, or `Lab_Analyses`.
2. **`promote_staged_ndep.R`** (a separate, manually-invoked step) reads
   a human-maintained mapping,
   `data/raw/ndep/PRR/staged_ndep_location_map.csv`
   (`station_name -> external_station_code, site_type, latitude,
   longitude, coordinate_source, coordinate_uncertainty_m, notes`), and
   only promotes a station's staged rows into `Samples`/`Lab_Analyses`
   once that mapping gives it real coordinates.

**Where coordinates come from:** NBMG's statewide "Geothermal_Wells"
dataset (ArcGIS Open Data,
`data-nbmg.opendata.arcgis.com/datasets/72341ba987e34c12a575c83f1d7c5367_0`),
matched by **exact well name** -- preferred over PLSS-derived
(quarter-section-precision) or digitized-from-topo-map estimates, both of
which carry too much uncertainty to register with confidence. Matched
stations get `coordinate_source = 'nbmg_geothermal_wells'` and a
documented `coordinate_uncertainty_m` (100 m by default, since NBMG
doesn't publish its own uncertainty for these records) rather than being
treated as precise.

As of this writing, 4 of 13 staged station names (Boyd Dom, Jeppson Dom,
Rogers Well, Soccer Field) have an unambiguous exact-name match and have
been promoted; the rest remain intentionally unresolved -- either because
multiple, mutually-inconsistent NBMG records exist for the same name
(e.g. "Herz Deep": two different NBMG sources disagree on its coordinate
by over a kilometer), or because no match exists at all (e.g. the
"Outlet" stations, likely Ormat's own creek monitoring points rather than
registered water wells). Their chemistry stays staged, not promoted,
until a confident coordinate is found.

---


## ✅ Current State

The system is actively evolving:

- continuous temperature monitoring is fully integrated  
- field and sampling workflows are operational  
- laboratory chemistry ingestion is expanding  

> In operational mode, the database evolves incrementally, providing a live representation of the hydrothermal system through time.