---
title: "Results"
output: html_document
---

# Results

This section presents observational insights derived from the integrated hydrothermal monitoring system.

By combining continuous temperature measurements with spatial and temporal structure, the system reveals patterns that are not accessible through traditional discrete sampling alone.

Rather than focusing on database structure, these results emphasize **system behavior**, **variability**, and **emerging hydrothermal signals**.

---

## 🌡️ Temperature Dynamics

Continuous temperature monitoring provides direct insight into subsurface processes and system variability.

```{r plotly, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(plotly)

temp_path <- "../docs/data/temp_sample.csv"

if (!file.exists(temp_path)) {
  
  cat("Temperature data will appear after pipeline export.")
  
} else {
  
  temp <- read_csv(temp_path, show_col_types = FALSE) %>%
    mutate(
      timestamp = as.POSIXct(timestamp, origin = "1970-01-01", tz = "UTC")
    ) %>%
    filter(!is.na(timestamp), !is.na(temperature)) %>%
    mutate(logger_id = factor(logger_id))
  
  p <- ggplot(temp,
              aes(timestamp, temperature, color = logger_id)) +
    
    geom_line(alpha = 0.6, linewidth = 0.6) +
    
    geom_smooth(se = FALSE, linewidth = 1.0, alpha = 0.3) +
    
    labs(
      title = "Temperature Time Series",
      subtitle = "Continuous monitoring reveals system variability",
      x = "Time",
      y = "Temperature (°C)",
      color = "Logger"
    ) +
    
    theme_minimal() +
    theme(legend.position = "bottom")
  
  ggplotly(p)
}
```
**Interpretation:**
Temperature exhibits continuous variability across multiple monitoring locations, reflecting dynamic and evolving subsurface conditions.

**Key Insight:**  
These signals capture both short-term fluctuations and longer-term trends, providing a level of temporal resolution that is fundamentally inaccessible through discrete sampling approaches.

---

## 📊 Temporal Patterns in Monitoring

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)

path <- "../docs/data/temp_sample.csv"

if (!file.exists(path)) {
  
  cat("Temporal distribution will appear after export.")
  
} else {
  
  temp <- read_csv(path, show_col_types = FALSE) %>%
    mutate(timestamp = as.POSIXct(timestamp, origin = "1970-01-01")) %>%
    filter(!is.na(timestamp))
  
  ggplot(temp, aes(timestamp)) +
    geom_histogram(bins = 60, fill = "#2c7fb8", alpha = 0.8) +
    labs(
      title = "Temporal Distribution of Observations",
      subtitle = "Clustering reflects deployments and monitoring periods",
      x = "Time",
      y = "Observation Frequency"
    ) +
    theme_minimal()
}
```
**Interpretation:**
Observations are temporally clustered, reflecting discrete monitoring deployments, field campaigns, and sensor operation periods.

**Key Insight:**  
Temporal clustering highlights how data collection strategies shape the dataset and influence the ability to resolve system dynamics.

---

## 📊 Thermal Variability Across Monitoring Locations

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)

path <- "../docs/data/temp_sample.csv"

if (!file.exists(path)) {
  
  cat("Logger variability will appear after export.")
  
} else {
  
  temp <- read_csv(path, show_col_types = FALSE) %>%
    filter(!is.na(temperature))
  
  ggplot(temp,
         aes(x = factor(logger_id), y = temperature)) +
    geom_boxplot(fill = "#fc8d59", alpha = 0.8) +
    labs(
      title = "Temperature Variability by Logger",
      subtitle = "Different locations exhibit distinct thermal behavior",
      x = "Logger",
      y = "Temperature (°C)"
    ) +
    theme_minimal()
}
```
**Interpretation:**
Different monitoring locations exhibit distinct temperature ranges and variability patterns.

**Key Insight:**  
These differences reflect spatial heterogeneity in the hydrothermal system, potentially indicating variations in flow paths, heat sources, or fluid mixing.

---

## 🗺️ Spatial Structure of Monitoring Network

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(leaflet)
library(readr)

path <- "../docs/data/well_sample.csv"

if (!file.exists(path)) {
  
  cat("Spatial data will appear after export.")
  
} else {
  
  wells <- read_csv(path, show_col_types = FALSE)
  
  leaflet(wells) |>
    addTiles() |>
    addCircleMarkers(
      lng = ~longitude,
      lat = ~latitude,
      radius = 5,
      color = "#2b8cbe",
      fillOpacity = 0.7,
      popup = ~paste("Well:", well_id)
    )
}
```
**Interpretation:**
Monitoring locations are spatially clustered in accessible or actively studied areas of the hydrothermal field.

**Key Insight:**  
Spatial distribution constrains where system behavior can be observed, influencing interpretations of subsurface flow and thermal structure.

---

## 🔬 Emerging System Behavior

Combining temporal, spatial, and thermal observations reveals several first-order characteristics of the hydrothermal system:

- temperature signals are highly dynamic and continuous  
- monitoring is spatially structured and unevenly distributed  
- observations are temporally clustered around data collection events - different locations exhibit distinct thermal regimes  

> Together, these patterns indicate a system that is both spatially heterogeneous and temporally variable, requiring integrated multi-scale observation to fully characterize.

---

## 🚀 Future Analysis

As additional datasets are integrated, the system will support more advanced interpretations, including:

- linking temperature anomalies with hydraulic gradients  
- identifying hydrothermal upflow and recharge zones  
- integrating geochemical signatures with thermal signals  
- developing process-based hydrothermal models (PHREEQC)  

> These analyses will extend the system from observational monitoring to predictive understanding of hydrothermal processes.


