---
title: "Results"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
```

## What the Data Show So Far

This page is deliberately narrower in scope than the [Data System](data.html) page: rather than describing the database's structure, it asks what the observations collected so far actually reveal about how the Steamboat Hills system behaves. The honest answer, at this stage, is that the continuous temperature record is the richest signal currently available, and it already shows more structure than any discrete sampling program could have captured on its own.

```{r temp-timeseries, fig.cap="Continuous temperature record across all active temperature loggers, colored by temperature logger. Interactive: hover to inspect individual readings, or drag to zoom into a shorter window."}
temp_path <- "../docs/data/temp_sample.csv"
if (file.exists(temp_path)) {
  temp <- read_csv(temp_path, show_col_types = FALSE) %>%
    mutate(timestamp = as.POSIXct(timestamp, origin = "1970-01-01", tz = "UTC")) %>%
    filter(!is.na(timestamp), !is.na(temperature)) %>%
    mutate(logger_id = factor(logger_id)) %>%
    group_by(logger_id) %>%
    slice(seq(1, n(), by = 15)) %>%
    ungroup()

  p <- ggplot(temp, aes(timestamp, temperature, color = logger_id)) +
    geom_line(alpha = 0.6, linewidth = 0.6) +
    labs(title = "Temperature Time Series by Temperature Logger", x = "Time", y = "Temperature (\u00b0C)", color = "Temperature Logger") +
    theme_minimal(base_size = 13) + theme(legend.position = "bottom")
  ggplotly(p)
} else {
  cat("Temperature data will appear after export.")
}
```

Two things stand out in that record. First, individual temperature loggers hold distinct, persistent temperature ranges rather than converging on one system-wide value, a strong indication that the field is spatially heterogeneous, with different monitoring points sitting on different flow paths, at different depths, or at different distances from the main upflow zones. Second, within any single logger's record there is real short-term variability that a quarterly or even monthly chemistry sample would average away entirely; that variability is exactly the kind of signal this project's monitoring-frequency question (below) is trying to characterize rigorously rather than just observe qualitatively.

```{r temp-boxplot, fig.cap="Distribution of temperature readings by temperature logger, showing how much thermal ranges differ across monitoring locations."}
if (file.exists(temp_path)) {
  temp2 <- read_csv(temp_path, show_col_types = FALSE) %>% filter(!is.na(temperature))
  ggplot(temp2, aes(x = factor(logger_id), y = temperature)) +
    geom_boxplot(fill = "#c65d2e", alpha = 0.85) +
    labs(title = "Temperature Variability by Temperature Logger", x = "Temperature Logger", y = "Temperature (\u00b0C)") +
    theme_minimal(base_size = 13)
}
```

## Where the Network Actually Observes the System

Any pattern in the charts above has to be read alongside where the underlying instruments actually sit, since a monitoring network that is spatially uneven will always look more (or less) variable than the system truly is, depending on where its gaps fall. The map below repeats the well/spring layer from the [Data System](data.html) page and adds the temperature and conductivity loggers themselves, so the instrument that produced each pattern above can be located directly: it is the map to have open while interpreting anything else on this page.
```{r spatial-map-results}
library(leaflet)
well_path <- "../docs/data/well_sample.csv"
loc_path <- "../docs/data/locations_sample.csv"
temp_logger_path <- "../docs/data/temp_logger_locations.csv"
cond_logger_path <- "../docs/data/cond_logger_locations.csv"
if (file.exists(well_path)) {
  wells <- read_csv(well_path, show_col_types = FALSE) |>
    mutate(label = ifelse(is.na(well_name) | well_name == "", paste("Well", well_id), well_name))
  m <- leaflet() |> addTiles() |>
    addCircleMarkers(data = wells, lng = ~longitude, lat = ~latitude,
      radius = 5, color = "#2b8cbe", fillOpacity = 0.7,
      popup = ~paste0("<b>", label, "</b><br>Role: ", well_role), group = "Wells")
  if (file.exists(loc_path)) {
    locs <- read_csv(loc_path, show_col_types = FALSE) |> filter(site_type != "well")
    m <- m |> addCircleMarkers(data = locs, lng = ~longitude, lat = ~latitude,
      radius = 5, color = "#e6550d", fillOpacity = 0.7,
      popup = ~paste0("<b>", coalesce(name, external_station_code), "</b><br>Type: ", site_type),
      group = "Springs / Seeps / Other")
  }
  if (file.exists(temp_logger_path)) {
    tloggers <- read_csv(temp_logger_path, show_col_types = FALSE)
    m <- m |> addCircleMarkers(data = tloggers, lng = ~longitude, lat = ~latitude,
      radius = 6, color = "#e9a13b", stroke = FALSE, fillOpacity = 0.95,
      popup = ~paste0("<b>Temperature logger ", logger_name, "</b><br>Status: ", status, "<br>Site: ", location_name),
      group = "Temperature Loggers")
  }
  if (file.exists(cond_logger_path)) {
    cloggers <- read_csv(cond_logger_path, show_col_types = FALSE)
    m <- m |> addCircleMarkers(data = cloggers, lng = ~longitude, lat = ~latitude,
      radius = 10, color = "#c65d2e", weight = 3, fill = FALSE, opacity = 0.95,
      popup = ~paste0("<b>Conductivity logger ", logger_name, "</b><br>Role: ", role, "<br>Status: ", status, "<br>Site: ", location_name),
      group = "Conductivity Loggers")
  }
  m |> addLayersControl(
    overlayGroups = c("Wells", "Springs / Seeps / Other", "Temperature Loggers", "Conductivity Loggers"),
    options = layersControlOptions(collapsed = FALSE))
} else {
  cat("Spatial data will appear after export.")
}
```

Taken together, these first-order patterns &mdash; persistent, location-specific thermal regimes; short-term variability layered on top of them; and a monitoring network that is spatially uneven by necessity, concentrated where access and prior study made it possible &mdash; describe a system that genuinely requires integrated, multi-scale observation to characterize well. No single well, and no single data type, tells the whole story on its own.

## A First Answer: Reactivation, Not a New Source

A poster presented at the 2026 Geothermal Rising Conference (Irvin and Lindsey, 2026, listed in full on the [References](references.html) page) used this database directly to ask the question the eruption raised in the first place: is the 2025 Lower Sinter Terrace event a chemically distinct new geothermal source, or a reactivation and redistribution of flow within the hydrothermal system that was already there? A chloride mass balance across the USGS Rhodes Road (SBRR) and Bella Vista (SBBV) stations on Steamboat Creek,

$$Q_{TW} = \frac{C_{Cl,SBBV} - C_{Cl,SBRR}}{820\ \text{mg/L}}$$

(where 820 mg/L is the typical Steamboat brine chloride concentration used as the deep end-member), puts current thermal water inflow to the creek at roughly **27 L/s**. That is well below the 60 L/s average (45-90 L/s range) Sorey and Spielman (2017) reported for 2008-2016, suggesting overall discharge has declined even as flow paths reorganize at the surface. Stream-water stable isotopes shift toward thermal-water compositions moving downstream between SBRR and SBBV, and paired temperature loggers show a moderate inverse relationship (r = -0.68) consistent with discharge redistributing within a connected shallow, fracture-controlled flow system rather than a separate source turning on. Together, these point toward conduit reactivation and flow-path reorganization as the more likely explanation.

This is a poster-stage finding, not yet a fully documented analysis in this codebase: the chloride mass-balance calculation above has not yet been re-run as a reproducible script against the live database the way the rest of this site's figures are, so treat the specific numbers as a snapshot from that presentation rather than a live, auto-updating result. Folding that calculation into the pipeline itself, so it updates the same way every other number on this page does, is a natural near-term follow-up.


## What Comes Next

Three pieces of modeling sit directly downstream of the data assembled here, and each is at a different stage of readiness.

<div class="callout-box future">
**PHREEQC geochemical modeling.** Validated major-ion and isotope chemistry is the direct input to forward speciation, mineral-saturation, and inverse mixing models in PHREEQC: the step that turns a table of concentrations into a statement about what water types are mixing, and where. A starter script for building PHREEQC-ready input tables already exists in this codebase, but it is not yet wired into the automated pipeline and has not been re-verified against the current database; getting it fully integrated is the next major modeling milestone.
</div>

<div class="callout-box future">
**Calibrating conductivity against chloride.** The conductivity logger deployed in Steamboat Creek is only useful as a chloride-discharge proxy once its relationship to real chloride measurements is calibrated against paired samples collected during its deployment window. That calibration, and the Monte Carlo analysis built around it to identify the coarsest chemistry sampling interval that still keeps flux estimates accurate, are both implemented and self-tested against synthetic data already &mdash; they are simply waiting on chloride samples that temporally overlap the logger's live deployment period to produce a real, rather than simulated, recommendation.
</div>

<div class="callout-box future">
**A potentiometric surface and contaminant-transport picture.** Water levels and chemistry are both already tied to precise well coordinates in the database, which is exactly the input ArcGIS needs to interpolate a potentiometric surface across the field and trace how a chemical signature &mdash; a chloride plume, say &mdash; might be migrating through the subsurface between wells. That interpolation and transport-mapping step hasn't been built yet, but the data pipeline was designed from the start with this specific export in mind, and it is the natural next spatial-analysis milestone once well coordinates and roles are more fully resolved.
</div>

None of these three are blocked on new instrumentation or new fieldwork so much as on continuing to run the pipeline that already exists: more chloride samples collected during the conductivity logger's deployment, more wells with confirmed roles and coordinates, and a dedicated pass to wire the existing PHREEQC starter script into the automated build. The infrastructure to take each of these the rest of the way is already in place.
