---
title: "Results"
output: html_document
---

This system produces continuously growing environmental datasets that can be analyzed across temporal and (eventually) spatial dimensions.

## Leaflet visualization

```{r leaflet, echo=FALSE, message=FALSE, warning=FALSE}
library(leaflet)
library(sf)
library(readxl)
library(readr)
library(ggplot2)
library(dplyr)


# read exported geopackage
gpkg <- st_read(
  "output/geopackage/hydro_data.gpkg",
  layer = "temperature_timeseries",
  quiet = TRUE
)

print(names(gpkg))

leaflet(gpkg) |>
  addTiles() |>
  addCircleMarkers(
    radius = 4,
    color = "red",
    popup = ~paste("Temp:", temperature)
  )
```

---

## 📈 Temporal Data Growth

```{r echo=FALSE, message=FALSE, warning=FALSE}
ingest_path <- "website/data/ingest_log.csv"

if (file.exists(ingest_path)) {

  ingest_log <- read_csv(ingest_path, show_col_types = FALSE)

  library(lubridate)

ingest_log <- ingest_log %>%
  mutate(
    timestamp = suppressWarnings(ymd_hms(timestamp, tz = "UTC"))
  )

# fallback for numeric timestamps
if (all(is.na(ingest_log$timestamp))) {
  ingest_log$timestamp <- as.POSIXct(
    as.numeric(ingest_log$timestamp),
    origin = "1970-01-01",
    tz = "UTC"
  )
}

  ingest_log$cumulative <- cumsum(ingest_log$measurements_inserted)

  print(
    ggplot(ingest_log,
           aes(timestamp, cumulative, color = data_source)) +
      geom_line(linewidth = 1.2) +
      labs(
        title = "System Growth Over Time",
        x = "Time",
        y = "Total Records"
      ) +
      theme_minimal()
  )

}
```

**Insight:**  
The database grows incrementally as new field campaigns, regulatory data, and logger deployments are ingested.

---

## 🧱 Database Composition

```{r echo=FALSE, message=FALSE, warning=FALSE}
table_path <- "website/data/table_counts.csv"

if (file.exists(table_path)) {

  table_counts <- read_csv(table_path, show_col_types = FALSE)

  print(
    ggplot(table_counts,
           aes(x = reorder(table, n_rows), y = n_rows)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      labs(
        title = "Database Composition",
        x = "",
        y = "Records"
      ) +
      theme_minimal()
  )

}
```

---

##🌡Temperature logger data over time

```{r plotly, echo = FALSE, warning = FALSE, message = FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(plotly)

temp_path <- "website/data/temp_sample.csv"

if (!file.exists(temp_path)) {
  
  cat("Temperature data will appear after pipeline export.")
  
} else {
  
  temp <- read_csv(temp_path, show_col_types = FALSE)
  
  # --- Clean + convert ---
  temp <- temp |>
  mutate(
    timestamp = as.POSIXct(timestamp, origin = "1970-01-01", tz = "UTC"),
    location_name = paste("Logger", logger_id)
  ) |>
  filter(
    !is.na(timestamp),
    !is.na(temperature)
  )

  
  # --- Determine grouping ---
  if ("location_name" %in% colnames(temp)) {
    temp <- temp |> mutate(group_var = location_name)
    color_label <- "Location"
    title_text <- "Temperature Time Series by Monitoring Location"
    
  } else if ("logger_id" %in% colnames(temp)) {
    temp <- temp |> mutate(group_var = factor(logger_id))
    color_label <- "Logger"
    title_text <- "Temperature Time Series by Logger"
    
  } else {
    temp <- temp |> mutate(group_var = "Temperature")
    color_label <- NULL
    title_text <- "Temperature Time Series"
  }
  
  # --- Plot ---
  p <- ggplot(temp,
              aes(x = timestamp,
                  y = temperature,
                  color = group_var,
                  group = group_var)) +
    
    geom_line(alpha = 0.8, linewidth = 0.6) +
    
    geom_smooth(se = FALSE, linewidth = 1.0, alpha = 0.3) +
    
    geom_hline(
      yintercept = c(40, 60, 80),
      linetype = "dashed",
      color = "gray50",
      alpha = 0.4
    ) +
    
    labs(
      title = title_text,
      subtitle = "High-frequency geothermal signals",
      x = "Time",
      y = "Temperature (°C)",
      color = color_label
    ) +
    
    theme_minimal() +
    theme(legend.position = "bottom")
  
  ggplotly(p)

}
```

**Insight:**  
Continuous temperature monitoring captures variability that is completely missed by discrete sampling campaigns, revealing short-term fluctuations and longer-term system behavior.

---

## 🗺️ Spatial Outputs (Upcoming)

The system exports GIS-ready layers:

```
output/geopackage/temperature_data.gpkg
```

These datasets enable:

- spatial visualization of temperature distributions  
- integration with ArcGIS time-aware layers  
- comparison across hydrothermal features  

*A full spatial analysis interface will be incorporated as the dataset matures.*

---

## 🔬 Interpretation

Even at this stage, several key observations emerge:

- Temperature variability is highly dynamic and continuous  
- Data volume grows predictably with ingestion runs  
- The database structure supports integration across multiple data types  

Future work will integrate:

- geochemical analysis (laboratory results)  
- spatial clustering and mapping  
- process-based modeling (PHREEQC)  

 